\ea\label{ex:nrexamaike}
Am ai kep Kupi, Kupi Am karimpo. Am kakarimpopno Am Ilöp karimpo.\\
\gll Am	ai	kep	Kupi	Kupi	Am	karim-po	Am	k-karim-pi-p-no	Am	Ilöp	karim-po\\
     Am	father	3\textsc{sg}.\textsc{poss}	Kupi	Kupi	Am	beget(\textsc{tp})-\textsc{lv}.\textsc{pp}	Am	\textsc{sub}-beget(\textsc{tp})-\textsc{lv}-\textsc{pc}-\textsc{co}	Am	Ilöp	beget(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `Am, his father is Kupi, Kupi begot Am. When he had begotten Am, Am begot Ilöp.'
\z

\ea\label{ex:nrexilopkk}
Ilöp kakarimpopno Ilöp karim Kaso, karim Dowo. Kaso Dowoyo karimpowe.\\
\gll Ilöp	k-karim-pi-p-no	Ilöp	karim	Kaso	karim	Dowo	Kaso	Dowo-yo	karim-po-we\\
     Ilöp	\textsc{sub}-beget(\textsc{tp})-\textsc{lv}-\textsc{pc}-\textsc{co}	Ilöp	beget(\textsc{tp})	Kaso	beget(\textsc{tp})	Dowo	Kaso	Dowo-\textsc{loc}	beget(\textsc{tp})-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `When he had begotten Ilöp, then Ilöp begets Kaso and begets Dowo. Kaso and Dowo he begot.'
\z

\ea\label{ex:nrexkasoap}
Kaso Apai karimpo. Apai kamapim Dowo seken, Dowo Baes Waei.\\
\gll Kaso	Apai	karim-po	Apai	kamapim	Dowo	seken	Dowo	Baes	Waei\\
     Kaso	Apai	beget(\textsc{tp})-\textsc{lv}.\textsc{pp}	Apai	originate(\textsc{tp})	Dowo	second(\textsc{tp})	Dowo	Baes	Waei\\
\glt `Kaso begot Apai, Apai is the father of the second Dowo. Dowo, Baes and Waei (are his sons).'
\z

\ea\label{ex:nrexnaketr}
Uke oki nake tripela solo. Tripela brata kamap, uke tripela kamappo.\\
\gll uke	o-ki	nake	tripela	solo	tripela	brata	kamap	uke	tripela	kamap-po\\
     we.\textsc{excl}	\textsc{prox}-\textsc{aph}	live	three(\textsc{tp})	only	three(\textsc{tp})	brother(\textsc{tp})	come.up(\textsc{tp})	we.\textsc{excl}	three(\textsc{tp})	come.up(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `We live here, we three only. Three brothers came up, the three of us are the offspring.'
\z

\ea\label{ex:nrex}
Stori Am kep roki.\\
\gll stori	Am	kep	ro-ki\\
     story(\textsc{tp})	Am	3\textsc{sg}.\textsc{poss}	\textsc{prox}-\textsc{aph}\\
\glt `This is Am's story.'
\z

\ea\label{ex:nrexnuiyae}
Yaeau nui. Yaeau kunuipno moni buar, yaeau kepro moni buaryo.\\
\gll yaeau	nui	yaeau	k-nui-p-no	moni	buar	yaeau	kep=ro	moni	buar-yo\\
     traditional.ceremony	do.intentionally	traditional.ceremony	\textsc{sub}-do.intentionally-\textsc{pc}-\textsc{co}	money(\textsc{tp})	stone.axe	traditional.ceremony	3\textsc{sg}.\textsc{poss}=\textsc{emph}	money(\textsc{tp})	stone.axe-\textsc{loc}\\
\glt `They celebrate a traditional ceremony. After having celebrated the ceremony there was money and stone axes. Their ceremony (brought) money and stone axes.'\footnote{It may be that this sequence refers to the appointment of Am as leader of the clan after the death of his father Kupi. In Nimboran society such an installation was done with a big feast that lasted over days (\cite[29-33]{Kouwenhoven:1956mg}).}
\z

\ea\label{ex:nrexguitar}
Guitar paitimpo moni kipiyapopno bekno.\\
\gll guitar	paitim-po	moni	k-piye-pi-p-no	bek-no\\
     guitar	beat(\textsc{tp})-\textsc{lv}.\textsc{pp}	money(\textsc{tp})	\textsc{sub}-take-\textsc{lv}-\textsc{pc}-\textsc{co}	bag(\textsc{tp})-\textsc{ins}\\
\glt `(Am) played the guitar. Having taken the money, (Am put it) in the bag.'
\z

\ea\label{ex:nrexlubruk}
Yena dawa kipiyapno dawa lu brukwepu.\\
\gll yena	dawa	k-piye-p-no	dawa	lu	bruk-wepu\\
     people	axe	\textsc{sub}-take-\textsc{pc}-\textsc{co}	axe	blade	break(\textsc{tp})-\textsc{quant}.\textsc{s}.\textsc{pp}\\
\glt `(Just) as the people were grabbing the axes, the blades broke in pieces.'
\z

\ea\label{ex:nrexrimpiy}
Serimpi yena royeini, Bu ki royeini ronpua rokiniro, tripela royeini.\\
\gll serim-pi	yena	roye-ini	Bu	ki	roye-ini	ronpua\_rokini=ro	tripela	roye-ini\\
     share(\textsc{tp})-\textsc{lv}	people	give-\textsc{nsg}.\textsc{or}	Bu	\textsc{aph}	give-\textsc{nsg}.\textsc{or}	three=\textsc{emph}	three(\textsc{tp})	give-\textsc{nsg}.\textsc{or}\\
\glt `They share them. He gives them to the people. Bu gives them three, he gives to them three.'\footnote{It is not clear what kind of items the numeral ``three'' refers to; maybe the narrator wants to say that Bu gives three new axes to the people.}
\z

\ea\label{ex:nrexrokpip}
Ro kipopno Am yip poro; lopos klokni pewo, yena riyo nake.\\
\gll ro	k-pi-p-no	Am	yip	po=ro	lopos	klokni	pewo	yena	ri-yo	nake\\
     \textsc{prox}	\textsc{sub}-do-\textsc{pc}-\textsc{co}	Am	house	make.\textsc{pp}=\textsc{emph}	post	one	erect.\textsc{pp}	people	\textsc{dist}-\textsc{loc}	live\\
\glt `After doing this Am built a house. He erected one (tall) post. The people live there.'
\z

\ea\label{ex:nrexpaeniy}
Paeni, Yob, Baepiye, Sin. Yena kukpo. Kaikai si, rapue si, ruri isaeaupi.\\
\gll Paeni	Yob	Baepiye	Sin	yena	kuk-po	kaikai	si	rapue	si	ruri	isaeaupi\\
     Paeni	Yob	Baepiye	Sin	\textsc{aph}.\textsc{pl}	cook(\textsc{tp})-\textsc{lv}.\textsc{pp}	food(\textsc{tp})	cook	food	cook	child	take.care.\textsc{pl}.\textsc{o}\\
\glt `(The women) Paeni, Yob, Baepiye, and Sin cooked. They cook the food and take care of the children.'
\z

\ea\label{ex:nrexpeknak}
Epe kanakeno ruri roipi emi lupi seku. Rumkari ke ar lakiyo epeyo. Roipi solo ukeli yip riyo, yipyo woko haus palimen Siopo.\\
\gll epe	k-nake-no	ruri	roipi	emi\_lupi	seku	rumkari	ke	ar	lakiyo	epe-yo	roipi	solo	ukeli	yip	ri-yo	yip-yo	woko	haus	palimen	Siopo\\
     mother	\textsc{sub}-give.birth-\textsc{co}	child	boy	umbilical.cord	fall.\textsc{pp}	girl	\textsc{aph}	\textsc{neg}	fetch.people.\textsc{pp}	mother-\textsc{loc}	boy	only	take.with.oneself	house	\textsc{dist}-\textsc{loc}	house-\textsc{loc}	accompany.\textsc{pp}	house	parliament	Siopo\\
\glt `When the mothers were giving birth, the umbilical cords of the children, of the boys, fell down. They did not fetch the girls from their mothers. The boys only they fetched to that house. They accompanied them to the house, the House Parliament (or) Siopo.'\footnote{Traditionally, the women preserved the umbilical cords of the newborn babies. The dried umbilical cords were believed to have an effect on the baby's sex. If they had children of one sex only, they used the cords in the following way: If a mother had only girls and wished for a boy, she put the last umbilical cord into the rib of a \textit{duam}-palm, a type of sago palm with thorns. If a mother had only boys and wished for a girl, she put the last umbilical cord into the rib of a \textit{numomo}-palm, a type of sago palm without thorns. (Margaret Osi, p. c.). The House Siopo is most probably identical with the \textit{house tambaran} which was abandoned after the Mission had settled in the village of Ossima. The \textit{haus tambaran} was the place where the men used to debate any kind of topic regarding village and life.}
\z

\ea\label{ex:nrexssiopo}
Yip, nem kep roki, haus palimen, haus Siopo.\\
\gll yip	nem	kep	ro-ki	haus	palimen	haus	Siopo\\
     house	name	3\textsc{sg}.\textsc{poss}	\textsc{prox}.\textsc{emph}-\textsc{aph}	house(\textsc{tp})	parliament	house(\textsc{tp})	Siopo\\
\glt `The name of the house is this: Parliament House (or) Siopo House.'
\z

\newpage
\ea\label{ex:nrexyipsep}
Yip sepolo, mi Mosbika lo. Yipro Mosbiyo poli.\\
\gll yip	sepolo	mi	Mosbi-ka	lo	yip=ro	Mosbi-yo	poli\\
     house	disappear.\textsc{pp}	then	Port.Moresby-\textsc{path}	go.\textsc{pp}	house=\textsc{emph}	Port.Moresby-\textsc{loc}	be.there\\
\glt `The house disappeared. Then it went to Port Moresby. This house is now in Port Moresby.'\footnote{Note the metaphorical use of the verb \textit{le} `go'. The transfer was institutionally, but the narrator says that the house itself moved.}
\z

\ea\label{ex:nrextelefo}
Telefon kepro, Bu Ipor Dawap telefon ikep, bi wiyo, du mauo.\\
\gll telefon	kep=ro	Bu	Ipor	Dawap	telefon	ikep	bi	wiyo	du	mauo\\
     phone(\textsc{tp})	3\textsc{sg}.\textsc{poss}=\textsc{emph}	Bu	Ipor	Dawap	phone(\textsc{tp})	3\textsc{sg}.\textsc{poss}.\textsc{emph}	animal	hold.\textsc{pp}	bush	roam-\textsc{pp}\\
\glt `His phone. Bu (had) Ipor and Dawap as his messengers. He caught animals. He roamed the bush.'\footnote{Ipor and Dawap are trees that act like personal messengers conveying news.}
\z

\ea\label{ex:nrexbukimu}
Bu ki muli: ``Bi ereyo nui.'' Yenaro solo sokuro, bi riyo lu.\\
\gll Bu	ki	muli	bi	ere-yo	nui	yena=ro	solo	sokuro	bi	ri-yo	lu\\
     Bu	\textsc{aph}	say	pig	\textsc{prox}-\textsc{loc}	sleep	people=\textsc{emph}	only	build.fences.for.hunting.\textsc{pp}	pig	\textsc{dist}-\textsc{loc}	shoot.\textsc{pp}\\
\glt `Bu says: ``Pigs live here.'' People simply built hunting fences and shot the pigs from there.'
\z

\ea\label{ex:nrexembuit}
``Biper ereyo nake,'' em Bu i tok olsem.\\
\gll biper	ere-yo	nake	em	Bu	i	tok	olsem\\
     possum	\textsc{prox}-\textsc{loc}	sit	\textsc{aph}(\textsc{tp})	Bu	\textsc{pred}(\textsc{tp})	speak(\textsc{tp})	like.this(\textsc{tp})\\
\glt `{``}Possums stay here,'' that's what Bu was saying.'
\z

\ea\label{ex:nrexkimosa}
Bu ki mosaupi telefon ikep roki: Ipor Dawapyo.\\
\gll Bu	ki	mosaupi	telefon	ikep	ro-ki	Ipor	Dawap-yo\\
     Bu	\textsc{aph}	teach	phone(\textsc{tp})	3\textsc{sg}.\textsc{poss}.\textsc{emph}	\textsc{prox}-\textsc{aph}	Ipor	Dawap-\textsc{loc}\\
\glt `Bu teaches and his messengers are these: Ipor and Dawap.'
\z

\newpage
\ea\label{ex:nrexbraser}
``Bras ereyo nake.'' Ki kiwiyapno du mauo. ``Biep ereyo nui.''\\
\gll bras	ere-yo	nake	ki	k-wiye-no	du	mauo	biep	ere-yo	nui\\
     bandicoot	\textsc{prox}-\textsc{loc}	sit	\textsc{aph}	\textsc{sub}-hold-\textsc{co}	bush	roam.\textsc{pp}	boar	\textsc{prox}-\textsc{loc}	sleep\\
\glt `{``}Bandicoots stay here.'' Catching them he roamed the bush. ``Boars live here.'''
\z

\ea\label{ex:nrexemipor}
Em Ipor ki mui bo Dawapyo: ``Ako depi e luwali.'' Em Bu ki muli mosaupi.\\
\gll em	Ipor	ki	mui	bo	Dawap-yo	ako	de-pi	e	luwali	em	Bu	ki	muli	mosaupi\\
     \textsc{aph}(\textsc{tp})	Ipor	\textsc{aph}	say	word	Dawap-\textsc{loc}	wife	2\textsc{sg}-\textsc{poss}	lover	have.sexual.intercourse	\textsc{aph}(\textsc{tp})	Bu	\textsc{aph}	speak	teach\\
\glt `Ipor tells news, with Dawap: ``Your wife has sex with a lover.'' Bu (also) provides information (of this kind).'
\z

\ea\label{ex:nrexankini}
Ono kilimpo yenaro an kini. Bu ki muelo: ``Tripela wiyo, ere ere ere wiyo.''\\
\gll ono	kilim-po	yena=ro	an\_kini	Bu	ki	muelo	tripela	bia	wiyo	ere	ere	ere	wiyo\\
     man	kill(\textsc{tp})-\textsc{lv}.\textsc{pp}	people=\textsc{emph}	five	Bu	\textsc{aph}	talk.to.\textsc{pp}	three(\textsc{tp})	corpse	hold.\textsc{pp}	\textsc{prox}	\textsc{prox}	\textsc{prox}	hold.\textsc{pp}\\
\glt `They killed the lover, five people altogether. Bu talked (to them). Three held  the corpse, this one, this one, and this one took (it) (together).'
\z

\ea\label{ex:nrexpalime}
Kupulupipno yena rais ileulipop, yena rais nop. Due komoipno riyo sonipop, yip riyo palimen yip.\\
\gll k-pulupi-p-no	yena	rais	ile-uli-pi-p	yena	rais	ni-p	due	k-moi-p-no	ri-yo	soni-pi-p	yip	ri-yo	palimen	yip\\
     \textsc{sub}-come.\textsc{pl}-\textsc{pc}-\textsc{co}	people	rice(\textsc{tp})	eat.\textsc{pl}.\textsc{a}-\textsc{prog}-\textsc{lv}-\textsc{pc}	people	rice(\textsc{tp})	eat-\textsc{pc}	sago.palm	\textsc{sub}-fell-\textsc{pc}-\textsc{co}	\textsc{dist}-\textsc{loc}	pulverise.sago.pith-\textsc{lv}-\textsc{pc}	house	\textsc{dist}-\textsc{loc}	parliament	house\\
\glt `When they had come, the people used to eat rice. They were eating rice. Having cut a sago palm, they processed the sago pith there, in that house, the Parliament House.'
\z

\newpage
\ea\label{ex:nrexpaekpu}
Pu paekyo lulipop, ep epi rka rka.\\
\gll pu\_paek-yo	luli-pi-p	ep	epi	r(i)-ka	r(i)-ka\\
     water.hole-\textsc{loc}	mix.with.water-\textsc{lv}-\textsc{pc}	rinse.water	side	\textsc{dist}-\textsc{path}	\textsc{dist}-\textsc{path}\\
\glt `They washed (the sago pith) near the water. The rinse water (streamed) to all sides.'
\z

\ea\label{ex:nrexnoahki}
Paua. Noah ki wokimpo, Noah Noah ponamo. Yena ke God ruri, yena ke God plas. Yelo roise po, wokimpo Am.\\
\gll paua	Noah	ki	wokim-po	Noah	Noah	ponamo	yena	ke	God	ruri	yena	ke	God	plas	yelo	roise	po	wokim-po	Am\\
     power(\textsc{tp}) Noah	\textsc{aph}	work(\textsc{tp})-\textsc{lv}.\textsc{pp}	Noah	Noah	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	people	\textsc{aph}	God	child	people	\textsc{aph}	God	semen	ground	together	make.\textsc{pp}	work(\textsc{tp})-\textsc{lv}.\textsc{pp}	Am\\
\glt `Noah created the power. Noah, Noah gave it (to Bu). (All) the people are children of God, they are the sperm of God. He created them and the earth. He created Am.'
\z

\ea\label{ex:nrexwapidi}
Wapi diri kepro Boyau, Bu Ilop diri kep, roki kamapimporo uke oki nake.\\
\gll Wapi	diri	kep=ro	Boyau	Bu	Ilöp	diri	kep	ro-ki	kamapim-po=ro	uke	o-ki	nake\\
     Wapi	younger.brother	3\textsc{sg}.\textsc{poss}=\textsc{emph}	Boyau	Bu	Ilöp	younger.brother	3\textsc{sg}.\textsc{poss}	\textsc{prox}-\textsc{aph}	originate(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	we.\textsc{excl}	\textsc{prox}-\textsc{aph}	live\\
\glt `Wapi and his younger brother Boyau, Bu and his younger brother Ilöp come from  here, (where) we live (now).'
\z

\ea\label{ex:nrexmonipa}
Moni pastaim uke maloro, moni ukeyo polip moni dere.\\
\gll moni	pastaim	uke	malo=ro	moni	uke-yo	poli-p	moni	dere\\
     money(\textsc{tp})	first(\textsc{tp})	we.\textsc{excl}	hear.\textsc{pp}=\textsc{emph}	money(\textsc{tp})	we.\textsc{excl}-\textsc{loc}	be.there-\textsc{pc}	money(\textsc{tp})	ancestor\\
\glt `When we heard of money first, we (already) had money. Money (and richness) from the ancestors.'
\z

\newpage
\ea\label{ex:nrexoudere}
Ukeyo solo polip, Öu dere mi ukero muli Öu klen.\\
\gll uke-yo	solo	poli-p	Öu	dere	mi	uke=ro	muli	Öu	klen\\
     we.\textsc{excl}-\textsc{loc}	only	be.there-\textsc{pc}	Öu	ancestor	again	we.\textsc{excl}=\textsc{emph}	speak	Öu	clan(\textsc{tp})\\
\glt `(The money) was simply with us, from the Öu ancestor, as we call him, the Öu clan.'
\z

\ea\label{ex:nrexpulbol}
Kuppuapno, Wiba Yaebu kinapno Omoi siyui, Pupulboli.\\
\gll k-ppue-p-no	Wiba	Yaebu	ki-na-p-no	Omoi	siyu-i	Pupulboli\\
     \textsc{sub}-go.up-\textsc{pc}-\textsc{co}	Wiba	Yaebu	\textsc{sub}-come.inside-\textsc{pc}-\textsc{co}	Omoi	throw.\textsc{pp}-\textsc{du}.\textsc{a}	Pupulboli\\
\glt `Wiba and Yaebu came to the land here. Having come in, they (founded) Omoi and dug (the well) Pupulboli.'\footnote{The name Pupulboli literally means `origin of shining water'.}
\z

\ea\label{ex:nrexyaebub}
Yaebu bilong Waes dere, Wiba bilong Öu dere.\\
\gll Yaebu	bilong	Waes	dere	Wiba	bilong	Öu	dere\\
     Yaebu	\textsc{poss}(\textsc{tp})	Waes	ancestor	Wiba	\textsc{poss}(\textsc{tp})	Öu	ancestor\\
\glt `Yaebu is a Waes ancestor, and Wiba is an Öu ancestor.'
\z

\ea\label{ex:nrexthirdk}
Yaebu third komanoko, ruwaesi nambawan ipei kopi. Mi Yaebu third karimpo.\\
\gll Yaebu	third	komanoko	ruwaesi	nambawan	ipei	ko-pi	mi	Yaebu	third	karim-po\\
     Yaebu	third(\textsc{tp})	? 	small.child	first(\textsc{tp})	firstborn	1\textsc{sg}-\textsc{poss}	again	Yaebu	third(\textsc{tp})	beget(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `The third, Yaebu -- as I named him -- is my firstborn child. I begot the third,  Yaebu.'
\z

\ea\label{ex:nrexupulbo}
Pupulboli pu ulno ipiyopi. Pu kipiyapno wepuloi, Bu ponamoi.\\
\gll Pupulboli	pu	ul-no	ipiyo-i	pu	k-ipiye-i-p-no	wepulo-i	Bu	ponamo-i\\
     Pupulboli	water	bamboo-\textsc{ins}	fill.in.a.container.\textsc{pp}-\textsc{du}.\textsc{a}	water	\textsc{sub}-fill.in.a.container-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	bring.\textsc{pp}-\textsc{du}.\textsc{a}	Bu	give.3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `They filled water from Pupulboli in bamboo containers. After filling them they brought them (to a certain place) and gave them to Bu.'\footnote{The two men might have given some water to Bu as a kind of offering.}
\z

\ea\label{ex:nrexrokriy}
Yena ro kiriyapno yilau layepanepo. Onume layepanepo, Omoiyo molo.\\
\gll yena	ro	k-riye-p-no	yilau	layepane-po	Onume	layepane-po	Omoi-yo	molo\\
     people	\textsc{prox}	\textsc{sub}-see.\textsc{o}.\textsc{inanim}-\textsc{pc}-\textsc{co}	place	leave.behind-\textsc{lv}.\textsc{pp}	Onume	leave.behind-\textsc{lv}.\textsc{pp}	Omoi-\textsc{loc}	go.\textsc{pl}.\textsc{pp}\\
\glt `When the people saw this, they left the place. They left Onume and went to Omoi.'
\z

\ea\label{ex:nrexnauoli}
Nau ol i stap long Omoi, Omoiyo mape.\\
\gll nau	ol	i	stap	long	Omoi	Omoi-yo	mape\\
     now(\textsc{tp})	\textsc{aph}.\textsc{pl}(\textsc{tp})	\textsc{pred}(\textsc{tp})	live(\textsc{tp})	\textsc{prep}(\textsc{tp})	Omoi	Omoi-\textsc{loc}	live.\textsc{pl}\\
\glt `Nowadays they live in Omoi, they live in Omoi.'
\z

\ea\label{ex:nrexmikoon}
Ko midorilo. Mi ko Onumeyo nake, yelo riyoro iki nakapno bonporo, yilau uke bonporo.\\
\gll ko	mi-dorilo	mi	ko	Onume-yo	nake	yelo	ri-yo=ro	iki	nake-p-no	bon-po=ro	yilau	uke	bon-po=ro\\
     I	\textsc{iter}-go.back.\textsc{pp}	again	I	Onume-\textsc{loc}	live	ground	\textsc{dist}-\textsc{loc}=\textsc{emph}	\textsc{aph}.\textsc{pl}	live-\textsc{pc}-\textsc{co}	born(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	place	we.\textsc{excl}	born(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}\\
\glt `I went back there. I live in Onume again, on that ground (where) they used to live. They had already been born there, (in) the place we were born.'
\z

\ea\label{ex:nrexaewanu}
Ko o lo. Oki yelo kopi yilau kopi. Onume Onume Aewanup Aewanup, Omoi Omoi.\\
\gll ko	o	lo	o-ki	yelo	ko-pi	yilau	ko-pi	Onume	Onume	Aewanup	Aewanup	Omoi	Omoi\\
     I	\textsc{prox}	go.\textsc{pp}	\textsc{prox}-\textsc{aph}	ground	1\textsc{sg}-\textsc{poss}	place	1\textsc{sg}-\textsc{poss}	Onume	Onume	Aewanup	Aewanup	Omoi	Omoi\\
\glt `I went (back) here. This is my ground, my place: Onume, Onume, Aewanup, Aewanup, Omoi, Omoi.'
\z

\ea\label{ex:nrexrokama}
Ukero kamapporo yilau riyo.\\
\gll uke=ro	kamap-po=ro	yilau	ri-yo\\
     we.\textsc{excl}=\textsc{emph}	arrive(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	place	\textsc{dist}-\textsc{loc}\\
\glt `We finally arrived at that place.'
\z

\ea\label{ex:nrexnoahno}
Noah Noah em karim dispela graun, Noah yelo piyo. Mi riyo nakap, yelo karimporo.\\
\gll Noah	Noah	em	karim	dispela	graun	Noah	yelo	piyo	mi	ri-yo	nake-p	yelo	karim-po=ro\\
     Noah	Noah	\textsc{aph}(\textsc{tp})	beget(\textsc{tp})	this(\textsc{tp})	ground(\textsc{tp})	Noah	ground	take.\textsc{pp}	then	\textsc{dist}-\textsc{loc}	live-\textsc{pc}	ground	beget(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}\\
\glt `Noah, Noah, he begot this ground. Noah took this ground, then he was living there. He  begot this ground.'
\z

\ea\label{ex:nrexyaesid}
Yau Yaesi Duweriyebili Duwarui Wekanei Onume, Onume dupua.\\
\gll Yau	Yaesi	Duweriyebili	Duwarui	Wekanei	Onume	Onume	dupua\\
     Yau	Yaesi	Duweriyebili	Duwarui	Wekanei	Onume	Onume	two\\
\glt `(The places are) Yau, Yaesi, Duweriyebili, Duwarui, Wekanei, Onume, two (places called) Onume.'
\z

\ea\label{ex:nrexantapu}
Onume antap uke klokni muli, dupua Onume dere roki. Onume ewe roki yip riyo neki. Bu riyo nake, yip Am, yip riyo po.\\
\gll Onume	antap	uke	klokni	muli	dupua	Onume	dere	ro-ki	Onume	ewe	ro-ki	yip	ri-yo	neki	Bu	ri-yo	nake	yip	Am	yip	ri-yo	po\\
     Onume	above(\textsc{tp})	we.\textsc{excl}	one	say	two	Onume	ancestor	\textsc{prox}-\textsc{aph}	Onume	older.brother	\textsc{prox}-\textsc{aph}	house	\textsc{dist}-\textsc{loc}	erect	Bu	\textsc{dist}-\textsc{loc}	live	house	Am	house	\textsc{dist}-\textsc{loc}	make.\textsc{pp}\\
\glt `The Onume Antap as we call one (of them), (because there are) two. The ancestor's Onume is this. Onume, the older brother builds a house there, Bu lives there. Am's house, he built that house.'
\z

\ea\label{ex:nrexorikop}
Stori kopi ba roki. Tenkyu em tasol.\\
\gll stori	ko-pi	ba	ro-ki	tenkyu	em	tasol\\
     story	1\textsc{sg}-\textsc{poss}	other	\textsc{prox}-\textsc{aph}	thank.you(\textsc{tp})	\textsc{aph}(\textsc{tp})	only(\textsc{tp})\\
\glt `This is one of my stories. Thank you, that's it.'
\z
