\ea\label{ex:nrexomosau}
Ko mosaupi. Ono nem kep Sakou. Dari werino puyo panai, nem kiyopi Ipp Salyo.\\
\gll ko	mosaupi	ono	nem	kep	Sakou	dari	weri-no	pu-yo	pana-i	nem	kiyo-pi	Ipp	Sal-yo\\
     I	teach	man	name	3\textsc{sg}.\textsc{poss}	Sakou	older.sister	younger.sister-\textsc{ins}	river-\textsc{loc}	put.thither.\textsc{pp}-\textsc{du}.\textsc{a}	name	\textsc{aph}.\textsc{du}-\textsc{poss}	Ipp	Sal-\textsc{loc}\\
\glt `I will tell (a story). (Once upon a time there was) a man, his name was Sakou. His sisters threw him into the river, their names were Ipp and Sal.'
\z
\largerpage[2]

\ea\label{ex:nrexkoupun}
Sakou pu wenap. Wo kup waliyo, koliyeno.\\
\gll Sakou	pu	wene-p	wo\_kup	waliyo	koliye-no\\
     Sakou	river	carry.thither-\textsc{pc}	kind.of.creeper	twine.around.\textsc{pp}	hang-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `The river carried Sakou away, (but) a \textit{kup}-liana clung to him and he held  on it.'
\z

\ea\label{ex:nrexkkoliy}
Kokoliyenapno ki ipiyo ppuo olyo. Dob seppue, ``Woppuo aeppu ule.''\\
\gll k-koliye-ne-p-no	ki	ipi-yo	ppuo	ol-yo	dob	seppue	woppuo	aeppu	ule\\
     \textsc{sub}-hang-3\textsc{sg}.\textsc{or}-\textsc{pc}-\textsc{co}	\textsc{aph}	clay-\textsc{loc}	go.up.\textsc{pp}	mountain-\textsc{loc}	eye	go.up	kind.of.fruit	ripe	be.there.\textsc{pl}\\
\glt `Holding on to the liana, he climbed up from the mud onto the river bank. He looks up: ``There are (a lot of) ripe \textit{woppuo}-fruits.'''
\z

\ea\label{ex:nrexkppuep}
Kuppuapno ini kepyo nakap. Kuppuapno kilîpno ki kinepno. Dob seku.\\
\gll k-ppue-p-no	ini	kep-yo	nake-p	k-ppue-p-no	k-lî-pi-no	ki	k-ni-p-no	dob	seku\\
     \textsc{sub}-go.up-\textsc{pc}-\textsc{co}	branch	3\textsc{sg}.\textsc{poss}-\textsc{loc}	sit-\textsc{pc}	\textsc{sub}-go.up-\textsc{pc}-\textsc{co}	\textsc{sub}-pluck-\textsc{pc}-\textsc{co}	\textsc{aph}	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	eye	fall.\textsc{pp}\\
\glt `Having climbed (the tree), he was sitting on one of its branches. Having climbed up and plucked (some fruits) and eaten (them), he looked down.'
\z

\ea\label{ex:nrexkirowo}
Bike kupulapno kiro woppuo aeppu niulipop. Mi kupulapno miro nep.\\
\gll bike	k-pule-p-no	ki=ro	woppuo	aeppu	ni-uli-pi-p	mi	k-pule-p-no	mi=ro	ni-p\\
     cassowary	\textsc{sub}-come-\textsc{pc}-\textsc{co}	\textsc{aph}=\textsc{emph}	kind.of.fruit	ripe	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	then	\textsc{sub}-come-\textsc{pc}-\textsc{co}	so=\textsc{emph}	eat-\textsc{pc}\\
\glt `A cassowary had come, and it was eating ripe \textit{woppuo}-fruits. After having come (again) it was eating (more fruits).'
\z

\ea\label{ex:nrexsakouu}
Sakou umul nek, ``Yala ko asa pi?'' An seke kep sayo; bike kûm sekeyo lilip.\\
\gll Sakou	umul\_nek	yala	ko	asa	pi	an	seke	kep	sayo	bike	kûm	seke-yo	lili-p\\
     Sakou	think.\textsc{pp}	now	I	how	do	hand	hair	3\textsc{sg}.\textsc{poss}	touch.\textsc{pp}	cassowary	sharp.bone	hair-\textsc{loc}	be.there-\textsc{pc}\\
\glt `Sakou thought, ``What do I do now?'' (His) hand touched his hair; there is always a sharp cassowary bone in his hair.'
\z

\ea\label{ex:nrexiyobik}
Riyo bike kûm paliyepako. Woppuo aeppu kilîpno bike kûmyo disanewepu.\\
\gll ri-yo	bike	kûm	paliyepako	woppuo	aeppu	k-lî-p-no	bike	kûm-yo	disane-wepu\\
     \textsc{dist}-\textsc{loc}	cassowary	sharp.bone	pull.out.\textsc{pp}	kind.of.fruit	ripe	\textsc{sub}-pluck-\textsc{pc}-\textsc{co}	cassowary	sharp.bone-\textsc{loc}	skewer-\textsc{quant}.\textsc{o}.\textsc{pp}\\
\glt `He pulled the cassowary bone out (of his hair). (Then) he plucked ripe	\textit{woppuo}-fruits and speared them on the sharp cassowary bone.'
\z

\ea\label{ex:nrexkdisan}
Kidisanewepupno riyopuno pakeno.\\
\gll k-disane-wepi-p-no	riyopuno	pake-no\\
     \textsc{sub}-skewer-\textsc{quant}.\textsc{o}-\textsc{pc}-\textsc{co}	then	throw-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Having speared (them), he threw (them) (to the cassowary).'
\z

\ea\label{ex:nrexnipnom}
Kinepno mi ba koniyo. Kûm moni paspo. Bike mi dusui.\\
\gll k-ni-p-no	mi	ba	koniyo	kûm	moni	pas-po	bike	mi	d-sui\\
     \textsc{sub}-eat-\textsc{pc}-\textsc{co}	so	\textsc{neg}.\textsc{emph}	swallow.\textsc{pp}	sharp.bone	throat	block-\textsc{lv}.\textsc{pp}	cassowary	so	\textsc{lkh}-die\\
\glt `As (the cassowary) ate, it (could) not swallow (the pierced fruits). The sharp bone got stuck in its throat. So the cassowary would surely die.'
\z

\ea\label{ex:nrexkesuip}
Bike sui, paliya. ``Yala ko asa pi?'' Dari werino wonien yilauyo, wonienro, ari.\\
\gll bike	sui	paliya	yala	ko	asa	pi	dari	weri-no	woni-en	yilau-yo	woni-en=ro	ari\\
     cassowary	die	be.dead	now	I	how	do	older.sister	younger.sister-\textsc{ins}	call-\textsc{nsg}.\textsc{or}.\textsc{pp}	village-\textsc{loc}	call-\textsc{nsg}.\textsc{or}.\textsc{pp}=\textsc{emph}	no\\
\glt `The cassowary is dying. It is dead. ``What do I do now?'' He called out for his sisters in the village, he called and called for them. Nothing.'
\z


\newpage
\ea\label{ex:nrexariwer}
Dari werino ari. Oke mi masalai bekimpo, nem kep sukupu.\\
\gll dari	weri-no	ari	o-ke	mi	masalai	bekim-po	nem	kep	sukupu\\
     older.sister	younger.sister-\textsc{ins}	no	\textsc{prox}-\textsc{aph}	again	bush.spirit(\textsc{tp})	answer(\textsc{tp})-\textsc{lv}.\textsc{pp}	name	3\textsc{sg}.\textsc{poss}	bush.spirit\\
\glt `(His) older and younger sister, no. This one again, the bush spirit, answered. Its name is Sukupu.'
\z

\ea\label{ex:nrexlaimir}
Masalai miro pulo. Sukupu kupulapno Sakou ke miro wonip. Robekimpono.\\
\gll masalai	mi=ro	pulo	sukupu	k-pule-p-no	Sakou	ke	mi=ro	woni-p	ro=bekim-pi-no\\
     bush.spirit(\textsc{tp})	so=\textsc{emph}	come.\textsc{pp}	bush.spirit	\textsc{sub}-come-\textsc{pc}-\textsc{co}	Sakou	\textsc{aph}	again=\textsc{emph}	call-\textsc{pc}	\textsc{emph}=answer(\textsc{tp})-\textsc{lv}-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `So the bush spirit came. When Sukupu had come, Sakou was calling out again, (and) the bush spirit answered him.'
\z

\ea\label{ex:nrexakouum}
Sakou umul senek, ``Dari werino ipulero yena ba roisero.'' Ari.\\
\gll Sakou	umul\_senek	dari	weri-no	i-pule=ro	yena	ba	roise=ro	ari\\
     Sakou	think.\textsc{pp}	older.sister	younger.sister-\textsc{ins}	\textsc{du}.\textsc{s}-come=\textsc{emph}	people	other	together=\textsc{emph}	no\\
\glt `Sakou thought, ``The two sisters are coming together with other people,'' (but) no.'
\z

\ea\label{ex:nrexseneka}
Sakou umul senek, ``Ari, oke mi sukupu ki pulo.'' Min dopyo, woppuo boliyo nekip.\\
\gll Sakou	umul\_senek	ari	o-ke	mi	sukupu	ki	pulo	min	dopyo	woppuo	boliyo	neki-p\\
     Sakou	think.\textsc{pp}	no	\textsc{prox}-\textsc{aph}	again	bush.spirit	\textsc{aph}	come.\textsc{pp}	come.hither.\textsc{pp}	next.to	kind.of.fruit.tree	at.the.foot.of	stand-\textsc{pc}\\
\glt `Sakou realised, ``No, this one is the bush spirit who came.'' (The bush spirit) came closer and stood at the base of the \textit{woppuo}-tree.'
\z

\ea\label{ex:nrexdobsep}
Kupulapno dob seppuo, ``De ke woppuo aeppu niro. Ba ko powaip, ko ni!''\\
\gll k-pule-p-no	dob	seppuo	de	ke	woppuo	aeppu	ni=ro	ba	ko	powai-p	ko	ni\\
     \textsc{sub}-come-\textsc{pc}-\textsc{co}	eye	go.skywards.\textsc{pp}	you	\textsc{top}	kind.of.fruit	ripe	eat=\textsc{emph}	some	I	give.1\textsc{sg}.\textsc{or}-\textsc{imp}	I	eat\\
\glt `Having come (closer), it looked up: ``What about you? You are eating ripe \textit{woppuo}-fruits. Give me some, (so) I eat (them)!'''
\z

\ea\label{ex:nrexinapwo}
Sakou inap woppuo aeppu ppueno, pakeno.\\
\gll Sakou	ina-p	woppuo	aeppu	ppue-no	pake-no\\
     Sakou	hurry-\textsc{pc}	kind.of.fruit	ripe	go.up-3\textsc{sg}.\textsc{or}.\textsc{pp}	throw-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Sakou was rushing. He climbed up and threw some ripe \textit{woppuo}-fruits to the spirit.'
\z

\ea\label{ex:nrexnepnor}
Kuppuenapno, riyopuno pakeno. Somapaip, dû wayap, koniyo. Eku kepka seku.\\
\gll k-ppue-ne-p-no	riyopuno	pake-no	somapai-p	dû	waye-p	koniyo	eku	kep-ka	seku\\
     \textsc{sub}-go.up-3\textsc{sg}.\textsc{or}-\textsc{pc}-\textsc{co}	then	throw-3\textsc{sg}.\textsc{or}.\textsc{pp}	taste-\textsc{pc}	flesh	try-\textsc{pc}	swallow.\textsc{pp}	anus	3\textsc{sg}.\textsc{poss}-\textsc{path}	fall.\textsc{pp}\\
\glt `Having climbed up, (Sakou) threw (some fruits) (down) to it. (The bush spirit) was tasting, it was trying the pulp. It swallowed (them), and (the fruits) fell right out of its arse.'\footnote{When food comes out of a person's anus undigested, this is a sign that the person in fact is a bush spirit.}
\z

\ea\label{ex:nrexkipkoy}
Sakou umul nekip, ``Ko yala asa pi?''\\
\gll Sakou	umul\_neki-p	ko	yala	asa	pi\\
     Sakou	think-\textsc{pc}	I	now	how	do\\
\glt `Sakou was thinking (back and forth), ``What should I do now?'''
\z

\ea\label{ex:nrexokekoy}
``Masalai oke, ko yala koniyam. Sukupu oke, oke ono pari. Oke masalai, ou.''\\
\gll masalai	o-ke	ko	yala	koniye-m	sukupu	o-ke	o-ke	ono	pari	o-ke	masalai	ou\\
     bush.spirit(\textsc{tp})	\textsc{prox}-\textsc{aph}	I	\textsc{mod}	swallow-\textsc{pos}	bush.spirit	\textsc{prox}-\textsc{aph}	\textsc{prox}-\textsc{aph}	man	\textsc{neg}.\textsc{cop}	\textsc{prox}-\textsc{aph}	bush.spirit(\textsc{tp})	yes\\
\glt `{``}A bush spirit. It could devour me. A bush spirit is this one, it is not a human. Yes, it is a bush spirit.'''
\z

\ea\label{ex:nrexkeroli}
Bikero lilip. Muelno: ``Bike puenpep, ko dob powe, de kûnap!''\\
\gll bike-ro	lili-p	mueli-no	bike	puenpi-p	ko	dob	pi-we	de	kûne-p\\
     cassowary-\textsc{emph}	be.there-\textsc{pc}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	cassowary	cut.meat-\textsc{imp}	I	eye	\textsc{lv}-\textsc{ter}	you	go.down-\textsc{imp}\\
\glt `The (dead) cassowary was lying there. (The bush spirit) said to (Sakou), ``Cut up the flesh of the cassowary, I will watch you (do it), come down!'''
\z

\ea\label{ex:nrexookoke}
``O o, ko ke asa kûnam? Oki de puenpep!''\\
\gll o	o	ko	ke	asa	kûne-m	o-ki	de	puenpi-p\\
     oh	oh	I	\textsc{top}	how	go.down-\textsc{pos}	\textsc{prox}-\textsc{aph}	you	cut.meat-\textsc{imp}\\
\glt `{``}Oh no. How could I go down? Here you go, cut (it) up (yourself)!'''
\z

\ea\label{ex:nrexpuorok}
Woppuo roke nowap, amainu. Am app sayap.\\
\gll woppuo	ro-ke	nowe-p	a-mainu	am	app	saye-p\\
     kind.of.fruit.tree	\textsc{prox}-\textsc{aph}	grow-\textsc{pc}	\textsc{imp}3-high	almost	sky	touch-\textsc{pc}\\
\glt `This \textit{woppuo}-tree was growing very tall. It would almost touch the sky.'
\z

\ea\label{ex:nrexpopuen}
Sakou dob pop, puenpep. Pili kep yana royo, dû kep yana royo, kili kep yana royo.\\
\gll Sakou	dob	po-p	puenpi-p	pili	kep	yana	royo	dû	kep	yana	royo	kili	kep	yana	royo\\
     Sakou	eye	\textsc{lv}-\textsc{pc}	cut.meat-\textsc{pc}	skin	3\textsc{sg}.\textsc{poss}	other	lay.\textsc{pp}	meat	3\textsc{sg}.\textsc{poss}	other	lay.\textsc{pp}	bone	3\textsc{sg}.\textsc{poss}	other	lay.\textsc{pp}\\
\glt `Sakou was watching as (the bush spirit) was cutting up (the cassowary's flesh). Its skin (the spirit) put in one (pile), its meat it put in another (pile), and its bones it put in a third (pile).'
\z

\ea\label{ex:nrexelipuk}
Eli pûke. ``Eli, ko asa pi?,'' sukupu muelne.\\
\gll eli	pûke	eli	ko	asa	pi	sukupu	mueli-ne\\
     intestines	lose	intestines	I	how	do	bush.spirit	talk.to-3\textsc{sg}.\textsc{or}\\
\glt `The intestines are hanging out. ``The intestines, what do I do (with them)?'' the bush spirit asks him.'
\z

\ea\label{ex:nrexlineel}
Sakou muelne: ``Eli de sewili numuelyo, yorer. Pu kippuyo lili, de sewili bike eli.''\\
\gll Sakou	mueli-ne	eli	de	sewili	numuelyo	yorer	pu	kippuyo	lili	de	sewili	bike	eli\\
     Sakou	talk.to-3\textsc{sg}.\textsc{or}	intestines	you	carry.away	far.away	very.far.away	river	far.away	be.there	you	carry.away	cassowary	intestines\\
\glt `Sakou says to it, ``Carry the intestines far away, very far away. The river is very far away. You carry the cassowary's intestines (there).'''
\z

\ea\label{ex:nrexwelros}
Welro sepue. Riyopuno eli pulolu, epo popiyo, puyo pusiyo.\\
\gll wel=ro	sepue	riyopuno	eli	pulolu	epo	popiyo	pu-yo	pusiyo\\
     carry.\textsc{pp}=\textsc{emph}	run.away	eventually	intestines	take.out.\textsc{pl}.\textsc{o}.\textsc{pp}	faeces	take.out.\textsc{pp}	river-\textsc{loc}	wash.\textsc{pp}\\
\glt `(The bush spirit) ran off carrying (the intestines). Finally it emptied the intestines, rinsed out the faeces and washed (the intestines) in the river.'
\z

\ea\label{ex:nrexakouku}
Sakou kukûnapno yeloyo, bike dû pele pili piyo.\\
\gll Sakou	k-kûne-p-no	yelo-yo	bike	dû	pele	pili	piyo\\
     Sakou	\textsc{sub}-go.down-\textsc{pc}-\textsc{co}	ground-\textsc{loc}	cassowary	meat	rib.cage	skin	take.\textsc{pp}\\
\glt `Having climbed onto the ground, Sakou took the meat, the rib bones, and the skin of the cassowary.'\footnote{In the next episode, Sakou shares the cassowary with five friendly spirits, namely anthropomorphised grasses, ferns, and ants that bear names.}
\z

\ea\label{ex:nrexnamodu}
Puop namo. ``Dû ko ar muli, oke ko pili powaip, dû ke ko delaine.''\\
\gll Puop	namo	dû	ko	ar	muli	o-ke	ko	pili	powai-p	dû	ke	ko	d-laine\\
     Puop	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	meat	I	\textsc{neg}	want	\textsc{prox}-\textsc{aph}	I	skin	give.1\textsc{sg}.\textsc{or}-\textsc{imp}	meat	\textsc{top}	I	\textsc{lkh}-lose.balance\\
\glt `He gave Puop (some of it) (who replied): ``I don't want the meat, give me the skin instead! The meat is too heavy for me.'''
\z

\ea\label{ex:nrexnepmue}
Nep muelne: ``Ko misoru, ari, dû ke ko delaine, ko pili powaip!'' Pili namo, mappeyo.\\
\gll Nep	mueli-ne	ko	misoru	ari	dû	ke	ko	d-laine	ko	pili	powai-p	pili	namo	mappeyo\\
     Nep	talk.to-3\textsc{sg}.\textsc{or}	I	too	no	meat	\textsc{aph}	I	\textsc{lkh}-lose.balance	I	skin	give.1\textsc{sg}.\textsc{or}-\textsc{imp}	skin	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	cook.over.fire.\textsc{pp}\\
\glt `Nep says to him: ``Me neither, the meat is too heavy for me, give me the skin!'' He gave him some skin and (Nep) cooked it.'
\z

\ea\label{ex:nrexlooyos}
Milo, oyo Sakumyo. Sakum mi bo so muelne: ``Pili ko powaip, ko mappeye.''\\
\gll mi-lo	o-yo	Sakum-yo	Sakum	mi	bo	so	mueli-ne	pili	ko	powai-p	ko	mappeye\\
     \textsc{iter}-go.\textsc{pp}	\textsc{prox}-\textsc{loc}	Sakum-\textsc{loc}	Sakum	again	word	like	talk.to-3\textsc{sg}.\textsc{or}	skin	I	give.1\textsc{sg}.\textsc{or}-\textsc{imp}	I	cook.over.fire\\
\glt `He went on here, to Sakum, and Sakum talks to him like this: ``Give me some skin, I'll cook (it).'''
\z

\ea\label{ex:nrexloeppu}
Lo Eppulyo, ki muelne: ``Ko misoru, ko delaine, ko pili powaip!''\\
\gll lo	Eppul-yo	ki	mueli-ne	ko	misoru	ko	d-laine	ko	pili	powai-p\\
     go.\textsc{pp}	Eppul-\textsc{loc}	\textsc{aph}	talk.to-3\textsc{sg}.\textsc{or}	I	too	I	\textsc{lkh}-lose.balance	I	skin	give.1\textsc{sg}.\textsc{or}-\textsc{imp}\\
\glt `He went to Eppul who says to him: ``Me neither, (the meat) is too heavy for me, give me some skin!'''
\z

\ea\label{ex:nrexyomuem}
Lo. Klapno oyo, Muemyo: ``Ko dû moniseso powaikap! De ke miweppuap olyo!''\\
\gll lo	k-le-p-no	o-yo	Muem-yo	ko	dû	moniseso	powai-ke-p	de	ke	mi-weppue-p	ol-yo\\
     go.\textsc{pp}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	\textsc{prox}-\textsc{loc}	Muem-\textsc{loc}	I	meat	very.little	give.1\textsc{sg}.\textsc{or}-\textsc{ingr}-\textsc{imp}	you	\textsc{top}	\textsc{iter}-carry.up-\textsc{imp}	mountain-\textsc{loc}\\
\glt `(Sakou) went on. When he came to Muem (the termite said): ``Give me a small piece of meat! Then you carry (the cassowary) up the mountain!'''
\z

\ea\label{ex:nrexpuosep}
Sakou riyo ppuo, sepue, mainu. Bikero melko, royenen kepno dari werino, Bo Bapulyo.\\
\gll Sakou	ri-yo	ppuo	sepue	mainu	bike=ro	meli-ko	roye-nen	kep-no	dari	weri-no	Bo	Bapul-yo\\
     Sakou	\textsc{dist}-\textsc{loc}	go.up.\textsc{pp}	run.away	high	cassowary=\textsc{emph}	carry.\textsc{pl}.\textsc{o}-\textsc{rts}	give-\textsc{nsg}.\textsc{or}.\textsc{pp}	3\textsc{sg}.\textsc{poss}-\textsc{ins}	older.sister	younger.sister-\textsc{ins}	Bo	Bapul-\textsc{loc}\\
\glt `Sakou went up there, far and high. He carried the cassowary and gave it with his (own things) to his sisters Bo and Bapul.'\footnote{The two women in the tree, Bo and Bapul, behave like sisters towards Sakou. This reminds of the four women living in a tree house, who take care of male children, as told in the story of Am (Text \ref{sc0201}).}
\z

\ea\label{ex:nrexbapuly}
Ukenakap Bo Bapulyo. Bo Bapulyo Sakou yaikoi. Riyo kuppuapno Bo Bapulyo lakii.\\
\gll uke-nake-p	Bo	Bapul-yo	Bo	Bapul-yo	Sakou	yai-ko-i	ri-yo	k-ppue-p-no	Bo	Bapul-yo	laki-i\\
     jointly-live-\textsc{pc}	Bo	Bapul-\textsc{loc}	Bo	Bapul-\textsc{loc}	Sakou	take.care.of-\textsc{rts}-\textsc{du}.\textsc{a}	tree-\textsc{loc}	\textsc{sub}-go.up-\textsc{pc}-\textsc{co}	Bo	Bapul-\textsc{loc}	fetch.people-\textsc{du}.\textsc{a}\\
\glt `He lived together with Bo and Bapul, and they took care of Sakou. When he had climbed the tall tree, Bo and Bapul accomodated him.'
\z

\ea\label{ex:nrexuriyup}
Sukupu riyo puloro, ``Eh, arka lo?'' Sakou dorka ki yek.\\
\gll sukupu	ri-yo	pulo=ro	eh	arka	lo	Sakou	dor-ka	ki	yek\\
     bush.spirit	\textsc{dist}-\textsc{loc}	come.\textsc{pp}=\textsc{emph}	eh	where	go.\textsc{pp}	Sakou	foot-\textsc{path}	\textsc{aph}	follow.one's.traces.\textsc{pp}\\
\glt `The bush spirit came (back) there, ``Ah, where did he go?'' It followed Sakou's foot steps.'\footnote{In the next episode, the bush spirit meets all the friendly spirits whom Sakou has met before, and it also shares food with them.}
\z

\ea\label{ex:nrexopsano}
Puop sano: ``Oh, ko pili powaikap! Lap, oki le.''\\
\gll Puop	sa-no	oh	ko	pili	powai-ke-p	le-p	o-ki	le\\
     Puop	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	oh	I	skin	give.1\textsc{sg}.\textsc{or}-\textsc{ingr}-\textsc{imp}	go-\textsc{imp}	\textsc{prox}-\textsc{aph}	go\\
\glt `Puop asked (the spirit): ``Oh, give me some skin! (Then) go, go this way!'''
\z

\ea\label{ex:nrexigonep}
Igo Nepyo. Nep muelne: ``Ko misoru, pili ko powaikap!''\\
\gll igo	Nep-yo	Nep	mueli-ne	ko	misoru	pili	ko	powai-ke-p\\
     and.so.forth(\textsc{tp})	Nep-\textsc{loc}	Nep	talk.to-3\textsc{sg}.\textsc{or}	I	too	skin	I	give.1\textsc{sg}.\textsc{or}-\textsc{ingr}-\textsc{imp}\\
\glt `And so forth to Nep. Nep says: ``Me too, give me also some skin!'''
\z

\ea\label{ex:nrexeppulm}
Eppul muli: ``Ko misoru, bike pili ko powaikap!''\\
\gll Eppul	muli	ko	misoru	bike	pili	ko	powai-ke-p\\
     Eppul	say	I	too	cassowary	skin	I	give.1\textsc{sg}.\textsc{or}-\textsc{ingr}-\textsc{imp}\\
\glt `Eppul says: ``Me too, give me some skin of the cassowary, too!'''
\z

\ea\label{ex:nrexlosaku}
Lo Sakumyo. Sakum muli: ``Ko misoru, ko pili powaikap!''\\
\gll lo	Sakum-yo	Sakum	muli	ko	misoru	ko	pili	powai-ke-p\\
     go.\textsc{pp}	Sakum-\textsc{loc}	Sakum	say	I	too	I	skin	give.1\textsc{sg}.\textsc{or}-\textsc{ingr}-\textsc{imp}\\
\glt `(The bush spirit) went on to Sakum who says: ``Me too, give me some skin, too!'''
\z

\ea\label{ex:nrexropeka}
Rope kana kamappi Muemyo: ``Ko dû moni powaikap, ko pari!\\
\gll rope	kana	kamap-pi	Muem-yo	ko	dû	moni	powai-ke-p	ko	pari\\
     why	fast	arrive(\textsc{tp})-\textsc{lv}	Muem-\textsc{loc}	I	meat	small	give.1\textsc{sg}.\textsc{or}-\textsc{ingr}-\textsc{imp}	I	\textsc{neg}.\textsc{cop}\\
\glt `And now? (The spirit) quickly arrives at Muem (who says): ``Give me a small (piece of) meat, I have none!'''
\z

\ea\label{ex:nrexdeolyo}
De olyo ppuap!''  ``De lakiyo?'' ``Oh, ko pari, kero ko pari lakiye.''\\
\gll de	ol-yo	ppue-p	de	lakiyo	oh	ko	pari	ke=ro	ko	pari	lakiye\\
     you	mountain-\textsc{loc}	go.up-\textsc{imp}	you	fetch.people.\textsc{pp}	oh	I	\textsc{neg}.\textsc{cop}	\textsc{top}=\textsc{emph}	I	\textsc{neg}.\textsc{cop}	fetch.people\\
\glt `(Muem continues), ``Go up the mountain!'' -- ``Did you catch (him)?'' -- ``Oh no, I didn't. Such a thing? I never catch people!'''
\z

\ea\label{ex:nrexolrono}
Olro nowap, igo. Sukupu ppue.\\
\gll ol=ro	nowe-p	igo	sukupu	ppue\\
     mountain=\textsc{emph}	grow-\textsc{pc}	and.so.forth(\textsc{tp})	bush.spirit	go.up\\
\glt `The mountain was growing taller and taller as the bush spirit climbed up.'
\z

\ea\label{ex:nrexwosuel}
Wo suelpakoi. Yeloyo basekiko. Lelpapporo, riyopuno puana.\\
\gll wo	suelpako-i	yelo-yo	ba-seki-ko	lelpappo=ro	riyopuno	puana\\
     rope	cut.through.\textsc{pp}-\textsc{du}.\textsc{a}	ground-\textsc{loc}	\textsc{fac}-fall-\textsc{fac}	lie.motionless.\textsc{pp}=\textsc{emph}	eventually	stand.up.\textsc{pp}\\
\glt `(The sisters) cut the rope. (The bush spirit) has fallen onto the ground. It lay there like dead. (But) eventually it got up (again).'
\z

\ea\label{ex:nrexehkonu}
``Eh, ko nuipapporo!'' Boyo mippue, so solo igo.\\
\gll eh	ko	nuipappo=ro	boyo	mi-ppue	so	solo	igo\\
     eh	I	sleep.motionless.\textsc{pp}=\textsc{emph}	later	\textsc{iter}-go.up	like	only	and.so.forth(\textsc{tp})\\
\glt `{``}Ah, I was fast asleep!'' (thinks the bush spirit). Then it climbs up again, like before, further and further.'
\z

\ea\label{ex:nrexyoyaup}
Bo Bapulyo yaup si, ``De mek kwewap!''\\
\gll Bo	Bapul-yo	yaup	si	de	mek	kwe-we-p\\
     Bo	Bapul-\textsc{loc}	hot.water	cook	you	mouth	open.mouth-\textsc{ter}-\textsc{imp}\\
\glt `Bo and Bopul boil hot water (and say to the bush spirit): ``Open your mouth and leave it open!'''
\z

\newpage
\ea\label{ex:nrexppuesu}
Ppue, sukupu mek kweulip. Yaup sipakenoi mekyo.\\
\gll ppue	sukupu	mek	kwe-uli-p	yaup	sipake-no-i	mek-yo\\
     go.up	bush.spirit	mouth	open.mouth-\textsc{prog}-\textsc{pc}	hot.water	pour.down-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	mouth-\textsc{loc}\\
\glt `It climbs up. The bush spirit held its mouth open, and they poured hot water into its mouth.'
\z

\ea\label{ex:nrexriseku}
Doriseku yeloyo, paliya, lelpapporo. Puana, miro ppue.\\
\gll doriseku	yelo-yo	paliya	lelpappo=ro	puana	mi=ro	ppuo\\
     fall.back.\textsc{pp}	ground-\textsc{loc}	be.dead	lie.motionless.\textsc{pp}=\textsc{emph}	stand.up.\textsc{pp}	again=\textsc{emph}	go.up.\textsc{pp}\\
\glt `It fell backwards on the ground and (seems to be) dead. (The spirit) lay there like dead. (But) it got up and climbed up again.'
\z

\ea\label{ex:nrexyoasap}
``Yala koyo asa pi? Luo sûno mappeye.'' Luo mappeyoi.\\
\gll yala	koyo	asa	pi	luo	sû-no	mappeye	luo	mappeyo-i\\
     now	we.\textsc{du}.\textsc{excl}	how	do	stone	fire-\textsc{ins}	roast.over.fire	stone	roast.over.fire.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `{``}What do we do now? We heat stones in the fire,'' (and) they heated stones.'
\z

\ea\label{ex:nrexmekkwe}
Mek kweuliwe ppue, luo pakoi. Em i pundaun. Doriseku yeloyo, lelro, puana, ppue.\\
\gll mek	kwe-uli-we	ppue	luo	pako-i	em	i	pundaun	doriseku	yelo-yo	lel=ro	puana	ppue\\
     mouth	open.mouth-\textsc{prog}-\textsc{ter}	go.up	stone	throw.\textsc{pp}-\textsc{du}.\textsc{a}	\textsc{aph}(\textsc{tp})	\textsc{pred}(\textsc{tp})	fall(\textsc{tp})	fall.back.\textsc{pp}	ground-\textsc{loc}	be.there.\textsc{pp}=\textsc{emph}	stand.up.\textsc{pp}	go.up\\
\glt `(The bush spirit) opens the mouth wide as it climbs, and they threw stones (into its mouth). It fell. It fell backwards onto the ground, lay (there), got up, and climbs (again).'
\z

\ea\label{ex:nrexpiumul}
``Ko asa pi?'' umul nek Sakou roise dari werino.\\
\gll ko	asa	pi	umul\_nek	Sakou	roise	dari	weri-no\\
     I	how	do	think.\textsc{pp}	Sakou	together	older.sister	younger.sister-\textsc{ins}\\
\glt `{``}What am I going to do?'' Sakou thought together with his sisters.'
\z

\ea\label{ex:nrexoyoyen}
Oyo yena yilau kiniyo leki. Yena yilau kiniyo riyo pulup.\\
\gll o-yo	yena	yilau	kiniyo	leki	yena	yilau	kiniyo	ri-yo	pulup\\
     \textsc{prox}-\textsc{loc}	people	village	all	fetch.people.\textsc{pl}.\textsc{o}	people	village	all	\textsc{dist}-\textsc{loc}	come.\textsc{pl}.\textsc{pp}\\
\glt `This is where he brought all the people from the village. All the people from the village came there.'
\z

\ea\label{ex:nrexenamon}
Yena moni, yena moni poyap. Ono bekulu bulika nekip.\\
\gll yena	moni	yena	moni	poye-p	ono	bekulu	bulika	neki-p\\
     people	small	people	small	stand.\textsc{pl}-\textsc{pc}	man	huge	side.by.side	stand-\textsc{pc}\\
\glt `(There were) little people. The little people were standing (in line). Next to (them) a huge man was standing.'
\z

\ea\label{ex:nrexmonimi}
Moni mike bakonupko. Yena moni kukonupipno, ekuka masakaikûno.\\
\gll moni	mike	ba-konupi-ko	yena	moni	k-konupi-p-no	eku-ka	masakaikûno\\
     small	first	\textsc{fac}-swallow.\textsc{pl}.\textsc{o}-\textsc{fac}	people	small	\textsc{sub}-swallow.\textsc{pl}.\textsc{o}-\textsc{pc}-\textsc{co}	anus-\textsc{path}	fall.down.\textsc{pl}.\textsc{pp}\\
\glt `The little (people) were devoured first (by the bush spirit). After it had swallowed the little humans, they fell out of its arse in large numbers.'
\z

\ea\label{ex:nrexkoipun}
Ikoipuno ono ikoi bulika nekip. Ki koniyouro, paliya, basuiko.\\
\gll ikoipuno	ono	ikoi	bulika	neki-p	ki	koniye-ou=ro	paliya	ba-sui-ko\\
     at.last	man	big	side.by.side	stand-\textsc{pc}	\textsc{aph}	swallow-\textsc{frus}=\textsc{emph}	be.dead	\textsc{fac}-die-\textsc{fac}\\
\glt `At last the big man was standing next to (the bush spirit). It tried to devour him without success. The spirit is dead. It has died.'
\z

\ea\label{ex:nrexpiyeiy}
Riyopuno piyeip, yipyo panapo, sû piyamu. Sû paloyo layepanepo. Sû beriro. Ûli kep pulwolo.\\
\gll riyopuno	piye-p	yip-yo	panapo	sû	piyamu	sû	palo-yo	layepane-po	sû	beri=ro	ûli	kep	pulwolo\\
     eventually	take-\textsc{pc}	house-\textsc{loc}	put.inside.\textsc{pp}	fire	collect.firewood.\textsc{pp}	fire	sago.thatches-\textsc{loc}	leave.behind-\textsc{lv}.\textsc{pp}	fire	burn=\textsc{emph}	gall.bladder	3\textsc{sg}.\textsc{poss}	burst.\textsc{pp}\\
\glt `Finally, they took (the dead bush spirit) and brought him into its house. They collected firewood and set fire on the sago palm leaves. The fire blazed, and bush spirit's gall bladder burst.'
\z

\ea\label{ex:nrexkouroi}
Riyopuno Sakou roise Bo Bapulyo ukeinakap.\\
\gll riyopuno	Sakou	roise	Bo	Bapul-yo	uke-nake-p\\
     then	Sakou	together	Bo	Bapul-\textsc{loc}	jointly-live-\textsc{pc}\\
\glt `Then Bo and Bopul were living together with Sakou.'
\z

\ea\label{ex:nrexlinode}
Sakou muelno: ``De rapiyekap!'' Sakou loro, riyo piye. ``De amulip o?''\\
\gll Sakou	mueli-no	de	rapiye-ke-p	Sakou	lo=ro	ri-yo	piye	de	a-muli-p	o\\
     Sakou	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	fetch-\textsc{ingr}-\textsc{imp}	Sakou	go.\textsc{pp}=\textsc{emph}	\textsc{dist}-\textsc{loc}	take	you	\textsc{imp}3-like-\textsc{pc}	\textsc{prox}\\
\glt `They said to Sakou: ``Go get (us something)!'' Sakou left. He brings (something and asks): ``Is this what you wanted?'''\footnote{It is not clear what the sisters want. Perhaps they want some special bush food like fruits or eggs.}
\z

\ea\label{ex:nrexohropa}
``O? Ro pari, ko ke ba mulip. Ba piyekap!'' Loro.\\
\gll o	ro	pari	ko	ke	bayana	muli-p	bayana	piye-ke-p	lo=ro\\
     o	\textsc{prox}.\textsc{emph}	\textsc{neg}.\textsc{cop}	I	\textsc{top}	different	want-\textsc{pc}	different	take-\textsc{ingr}-\textsc{imp}	go.\textsc{pp}=\textsc{emph}\\
\glt `{``}Oh, this is not it. I wanted something different, go get something different!'' So off he went.'
\z

\ea\label{ex:nrexoukoro}
``Ou, ko ro mulip, upuna. Koyo ro mulip, upuna.''\\
\gll ou	ko	ro	muli-p	upuna	koyo	ro	muli-p	upuna\\
     yes	I	\textsc{prox}.\textsc{emph}	want-\textsc{pc}	alright	we.\textsc{du}.\textsc{excl}	\textsc{prox}	want-\textsc{pc}	alright\\
\glt `{``}Yes, that's what I wanted. Alright, that's what we wanted, alright.'''
\z

\ea\label{ex:nrexlenokr}
Leno kredimponoipno, le rapiyenoi, melnoi. Wokekoi yilau kepyo, due dû roise le kiniyo roise wokoi.\\
\gll le-no	k-redim(\textsc{tp})-pi-no-i-p-no	le	rapiyo-no-i	meli-no-i	woke-ko-i	yilau	kep-yo	due\_dû	roise	le	kiniyo	roise	woko-i\\
     things-\textsc{ins}	\textsc{sub}-make.ready-\textsc{lv}-3\textsc{sg}.\textsc{or}-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	things	fetch-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	carry.\textsc{pl}.\textsc{o}-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	accompany-\textsc{rts}-\textsc{du}.\textsc{a}	village	3\textsc{sg}.\textsc{poss}-\textsc{loc}	sago.flour	together	belongings	all	together	accompany.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `When the two had finished with his things, they fetched (everything) and carried everything for him. They went with him to his village, with sago flour and with all his belongings they accompanied him (there).'
\z

\ea\label{ex:nrexpleske}
Asples kepyo, yilau duki kepyo wokoi. Yena riyo bal lilip.\\
\gll asples	kep-yo	yilau	duki	kep-yo	woko-i	yena	ri-yo	bal\_lili-p\\
     village(\textsc{tp})	3\textsc{sg}.\textsc{poss}-\textsc{loc}	village	true	3\textsc{sg}.\textsc{poss}-\textsc{loc}	accompany.\textsc{pp}-\textsc{du}.\textsc{a}	people	\textsc{dist}-\textsc{loc}	play.ball-\textsc{pc}\\
\glt `To his village, his home village, they went with him. People were playing ball there.'
\z

\ea\label{ex:nrexikilil}
Iki liliko. Bal, bal somon, Sakou paeau. Sakou pewo boliyo nekip, bal dorno piyelayowe.\\
\gll iki	lili-ko	bal	bal	somon	Sakou	paeau	Sakou	pewo	boliyo	neki-p	bal	dor-no	piyelayo-we\\
     \textsc{aph}.\textsc{pl}	play-\textsc{rts}	ball	ball	come.hither.\textsc{pp}	Sakou	reach	Sakou	banana	at.the.foot.of	stand-\textsc{pc}	ball	foot-\textsc{ins}	hold.with.one's.foot.\textsc{pp}-\textsc{ter}\\
\glt `They played. The ball, the ball came close and reached Sakou, who was standing at the base of a banana tree and firmly held the ball with his foot.'
\z

\ea\label{ex:nrexukarur}
Buka ruri ikep Sakoupi, Sakou ki nuni, ki lipelamuko.\\
\gll buka	ruri	ikep	Sakou-pi	Sakou	ki	nuni	ki	lipelami-ko\\
     child.of.a.man's.sister	child	3\textsc{sg}.\textsc{poss}.\textsc{emph}	Sakou-\textsc{poss}	Sakou	\textsc{aph}	maternal.uncle	\textsc{aph}	seek.hither-\textsc{rts}\\
\glt `His nephew, Sakou's own child, Sakou is the uncle. (His nephew) had already looked for (the ball).'
\z

\largerpage
\ea\label{ex:nrexnuniko}
``De nuni! Ko ke soro de ke basuiko!'' ``Ari, ko ar sui. Ko unake, ou!''\\
\gll de	nuni	ko	ke	so=ro	de	ke	ba-sui-ko	ari	ko	ar	sui	ko	u-nake	ou\\
     you	maternal.uncle	I	\textsc{top}	believe=\textsc{emph}	you	\textsc{top}	\textsc{fac}-die-\textsc{fac}	no	I	\textsc{neg}	die	I	\textsc{dfac}-live	yes\\
\glt `{``}You are (my) uncle! I really thought you had died (long ago)!'' -- ``No, I didn't die. Yes, I am still alive!'''
\z

\ea\label{ex:nrexloroep}
Ki loro. Epe aino muelien: ``Nuni kopi unakap!''\\
\gll ki	lo=ro	epe	ai-no	mueli-en	nuni	ko-pi	u-nake-p\\
     \textsc{aph}	go.\textsc{pp}=\textsc{emph}	mother	father-\textsc{ins}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	maternal.uncle	1\textsc{sg}-\textsc{poss}	\textsc{dfac}-live-\textsc{pc}\\
\glt `(The nephew) went and told his father and mother: ``My uncle was alive the whole time.'''
\z

\ea\label{ex:nrexyopana}
``Koyo puyo panai.'' ``Ari, bapulok. Pewo boliyo neki, nuni kopi bapulok.''\\
\gll koyo	pu-yo	pana-i	ari	ba-pulo-ko	pewo	boliyo	neki	nuni	ko-pi	ba-pule-ko\\
     we.\textsc{du}.\textsc{excl}	river-\textsc{loc}	put.thither.\textsc{pp}-\textsc{du}.\textsc{a}	no	\textsc{fac}-come-\textsc{fac}	banana	at.the.foot.of	stand	maternal.uncle	1\textsc{sg}-\textsc{poss}	\textsc{fac}-come-\textsc{fac}\\
\glt `(The sisters said), ``We threw (him) into the river. (But) no (he didn't drown).'' -- ``He has come (back) and is standing at the foot of the banana tree. My uncle has returned.'''
\z

\ea\label{ex:nrexwiyoiw}
Anno wiyoi, wo mopui. Riyopuno yipyo wokoi.\\
\gll an-no	wiyo-i	wo\_mopu-i	riyopuno	yip-yo	woko-i\\
     arm-\textsc{ins}	hold.\textsc{pp}-\textsc{du}.\textsc{a}	weep.\textsc{pp}-\textsc{du}.\textsc{a}	then	house-\textsc{loc}	accompany.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `They hugged (him) and wept. Then they accompanied him to the house.'
\z

\ea\label{ex:nrexkwokei}
Kowokeipno muelien: ``Inepe kaikai aro ni?'' ``Uke kaikai o ni.'' ``O ari, eroke yeloro.''\\
\gll k-woke-i-p-no	mueli-en	ine-pe	kaikai	a=ro	ni	uke	kaikai	o	ni	o	ari	ero-ke	yelo=ro\\
     \textsc{sub}-accompany-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{pl}-\textsc{q}	food(\textsc{tp})	which=\textsc{emph}	eat	we.\textsc{excl}	food(\textsc{tp})	\textsc{prox}	eat	\textsc{prox}	no	\textsc{prox}-\textsc{aph}	soil=\textsc{emph}\\
\glt `They had left with him, and he said to them: ``Which kind of food do you eat?'' -- ``We eat this food, this (kind of food).'' -- ``This? No, this is just earth.'''
\z

\ea\label{ex:nrexmekmon}
Mek moniseso wal puyo mapenoso.\\
\gll mek	moniseso	wal	pu-yo	mape-no-so\\
     mouth	very.small	fish	water-\textsc{loc}	stay.\textsc{pl}-\textsc{co}-\textsc{sim}\\
\glt `(Their) mouths were very small, like those of fish living in water.'
\z

\ea\label{ex:nrexkaioki}
``Ko ke kaikai oki ni. Ine mek pusapopup!'' Ro ponien bi, mike yena nosopuo.\\
\gll ko	ke	kaikai	o-ki	ni	ine	mek	pusapopi-p	ro	ponien	bi	mike	yena	nosopuo\\
     I	\textsc{top}	food	\textsc{prox}-\textsc{aph}	eat	you.\textsc{pl}	mouth	wash.\textsc{pl}.\textsc{o}-\textsc{imp}	\textsc{prox}.\textsc{emph}	give.\textsc{nsg}.\textsc{or}.\textsc{pp}	meat	first	people	vomit.\textsc{pp}\\
\glt `{``}I eat this food here. Wash (your) mouths!'' He gave them this (food), meat (of the 	cassowary). First the people vomited.'
\z

\ea\label{ex:nrexsekenn}
Seken nosopuo. Third ponienro, upuna, ar nosopuo. Mek ikepro ikoi bapok.\\
\gll seken	nosopuo	third	ponien=ro	upuna	ar	nosopuo	mek	ikep=ro	ikoi	ba-pi-ko\\
     second(\textsc{tp})	vomit.\textsc{pp}	third(\textsc{tp})	give.\textsc{nsg}.\textsc{or}.\textsc{pp}=\textsc{emph}	alright	\textsc{neg}	vomit.\textsc{pp}	mouth	3\textsc{pl}.\textsc{poss}=\textsc{emph}	big	\textsc{fac}-\textsc{lv}-\textsc{fac}\\
\glt `The second time they (also) vomited. The third time he gave them (meat), alright, they did not vomit. Their mouths have become larger.'
\z

\ea\label{ex:nrexmbiro}
Mi bi royeen, kloknino dupuano rokinino. Upuna, mek mepien. ``Ine yala upuna ni.''\\
\gll mi	bi	roye-en	klokni-no	dupua-no	rokini-no	upuna	mek	mepi-en	ine	yala	upuna	ni\\
     again	meat	give-\textsc{nsg}.\textsc{or}.\textsc{pp}	one-\textsc{ins}	two-\textsc{ins}	three-\textsc{ins}	alright	mouth	taste.nice-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{pl}	now	alright	eat\\
\glt `He gave them meat again, once, twice, three times. All right, (their) mouths taste nice  to them. ``Now you are eating properly.'''
\z

\ea\label{ex:nrexloinaw}
Iki ke sukoipu nip. Yena duyo molo, moloro. Löina wepulupip ba.\\
\gll iki	ke	sukoipu	ni-p	yena	du-yo	molo	molo=ro	löi-na	wepulupi-p	ba\\
     \textsc{aph}.\textsc{pl}	\textsc{aph}	millipede	eat-\textsc{pc}	people	bush-\textsc{loc}	go.\textsc{pl}.\textsc{pp}	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	löi-\textsc{aff}	bring.\textsc{pl}.\textsc{a}-\textsc{pc}	other\\
\glt `(But) they still ate millipedes. People went into the forest, went (there) and \textit{löi löi löi} (calling) they brought something (home).'
\z

\ea\label{ex:nrexehbiop}
``Eh, bi ope ine asa ile? Ppulae pi, maki ari. Oke ko ke ba niwe, ppulae pi!''\\
\gll eh	bi	o-pe	ine	asa	ile	ppulae	pi	maki	ari	o-ke	ko	ke	ba	ni-we	ppulae	pi\\
     eh	meat	\textsc{prox}-\textsc{q}	you.\textsc{pl}	how	eat.\textsc{pl}.\textsc{a}	bad	\textsc{lv}	good	no	\textsc{prox}-\textsc{aph}	I	\textsc{top}	\textsc{neg}.\textsc{emph}	eat-\textsc{ter}	bad	\textsc{lv}\\
\glt `{``}Ah, how do you eat this meat? It's bad, it's not good! This (stuff), I never eat it, it's bad!'''
\z

\ea\label{ex:nrexkaikep}
Kaikai kepro wepuloro. Sakou yena ukenakap.\\
\gll kaikai	kep=ro	wepulo=ro	Sakou	yena	uke-nake-p\\
     food(\textsc{tp})	3\textsc{sg}.\textsc{poss}=\textsc{emph}	bring.\textsc{pp}=\textsc{emph}	Sakou	people	jointly-sit-\textsc{pc}\\
\glt `He brought his food, (and) Sakou continued to live with the people.'
\z

\ea\label{ex:nrexnaropu}
Yenaro puloro romiso, yena muelien: ``Nuko due klokni ukenui.''\\
\gll yena=ro	pulo=ro	ro=miso	yena	mueli-en	nuko	due	klokni	uke-nui\\
     people=\textsc{emph}	come.\textsc{pp}=\textsc{emph}	\textsc{emph}=too	people	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	we.\textsc{incl}	night	one	jointly-sleep\\
\glt `The people (from other clans) also came there, (and Sakou) said to them: ``For one night we will all sleep together.'''
\z

\ea\label{ex:nrexkaikin}
Yena kaikai kiniyo rapiyo, puno rino due. ``Klokni nui, ukenui.''\\
\gll yena	kaikai	kiniyo	rapiyo	pu-no	ri-no	due	klokni	nui	uke-nui\\
     people	food(\textsc{tp})	all	fetch.\textsc{pp}	water-\textsc{ins}	wood-\textsc{ins}	night	one	do.intentionally	jointly-sleep\\
\glt `The people fetched all kinds of food and water and firewood, (and Sakou repeated): ``For one night we will all sleep together.'''
\z

\ea\label{ex:nrexnaepul}
Yena epul malemayo. Riyopuno due nui.\\
\gll yena	epul	malemayo	riyopuno	due\_nu\\
     people	ear	obey.\textsc{pp}	then	sleep.\textsc{pp}\\
\glt `The people obeyed, and finally they slept.'
\z

\ea\label{ex:nrexuampul}
Duam pul bi kili yûr kili bras kili bike kili royepiyi, kiniyo.\\
\gll duam	pul	bi	kili	yûr	kili	bras	kili	bike	kili	royepiyi	kiniyo\\
     kind.of.sago	seed	pig	bone	bird	bone	bandicoot	bone	cassowary	bone	throw.around	all\\
\glt `Sago seeds, pig bones, bird bones, bandicoot bones, cassowary bones, all sorts (of bones) (Sakou) throws around.'
\z

\ea\label{ex:nrexkriwon}
Kiniyo royepiyi, roseliyo. Bike kri woni, biep woni, yûr bo mui. Duam kwe seki.\\
\gll kiniyo	royepiyu	ro=seliyo	bike	kri	woni	biep	woni	yûr	bo	mui	duam	kwe	seki\\
     all	throw.around.\textsc{pp}	\textsc{emph}=sow.\textsc{pp}	cassowary	kri	call	boar	call	bird	sound	speak	kind.of.\textit{}sago	dried.rib.of.sago.palm	fall\\
\glt `All (these bones) he throws around. He sowed (and created the animals): the cassowaries call \textit{kri}, the boars call, the birds call, dried sago palm ribs fall (down).'
\z

\ea\label{ex:nrexduamso}
Riyopuno duruwa, yena dob soriyap: duam sokiromaliyenake.\\
\gll riyopuno	duruwa	yena	dob	soriye-p	duam	sokiromaliye-nake\\
     eventually	early.morning	people	eye	realise.\textsc{o}.\textsc{inanim}-\textsc{pc}	kind.of.sago	be.there.in.abundance-\textsc{dur}\\
\glt `Finally it is early morning (and) the people were looking with amazement: there are \textit{duam}-palms in abundance.'
\z

\ea\label{ex:nrexdukudu}
Duku duamno nekiyenake, dikana numomo, kiniyo nekiyenake.\\
\gll duku	duam-no	nekiye-nake	dikana	numomo	kiniyo	nekiye-nake\\
     kind.of.sago	kind.of.sago-\textsc{ins}	stand-\textsc{dur}	kind.of.sago	kind.of.sago	all	stand-\textsc{dur}\\
\glt `\textit{Duku}- and \textit{duam}-palms, \textit{dikana}- and \textit{numumo}-palms, all kinds of palmtrees abound.'
\z

\ea\label{ex:nrexkebiep}
Bike biepno kri woninake, yûr bo mui. Riyopuno duruwei, du baruweko.\\
\gll bike	biep-no	kri	woni-nake	yûr	bo	mui	riyopuno	duruwei	du	ba-ruwe-ko\\
     cassowary	boar-\textsc{ins}	kri	call-\textsc{dur}	bird	sound	speak	then	early.morning	darkness	\textsc{fac}-break-\textsc{fac}\\
\glt `Cassowaries and boars call all the time, birds sing. Then it is early morning, the darkness is over.'
\z

\ea\label{ex:nrexbosako}
Bo Sakoupi kuru, bo maki.\\
\gll bo	Sakou-pi	kuru	bo	maki\\
     story	Sakou-\textsc{poss}	be.finished	story	good\\
\glt `Sakou's story is over. It's a good story.'
\z
