\ea\label{ex:nrexhausta}
Ko bo mosaupi.\\
\gll ko	bo	mosaupi\\
     I	story	teach\\
\glt `I will tell a story.'
\z

\ea\label{ex:nrextambar}
Haus tambaran ipei uke ba riyo.\\
\gll haus\_tambaran	ipei	uke	ba	riyo\\
     traditional.male.cult.house(\textsc{tp})	first.ranking	we.\textsc{excl}	\textsc{neg}.\textsc{emph}	see.\textsc{o}.\textsc{inanim}.\textsc{pp}\\
\glt `We did not see the first ranking male cult house.'\footnote{All the women including the story teller were forbidden to see this secret male cult house, because it was taboo for the sight of women.}
\z

\ea\label{ex:nrexsaulbo}
Haus tambaran ppusi uke riyo, Saul Bopule Iwan uke riyo.\\
\gll haus\_tambaran	ppusi	uke	riyo	Saul	Bopule	Iwan	uke	riyo\\
     traditional.male.cult.house(\textsc{tp})	second.ranking	we.\textsc{excl}	see.\textsc{o}.\textsc{inanim}.\textsc{pp}	Saul	Bopule	Iwan	we.\textsc{excl}	see.\textsc{o}.\textsc{inanim}.\textsc{pp}\\
\glt `(But) we saw the second ranking male cult house; the Saul Bopule Iwan house we saw.'
\z

\ea\label{ex:nrexosokin}
Uke kiriyapno kaikai kero oso kiniyo ukeno.\\
\gll uke	k-riye-p-no	kaikai	kep=ro	oso	kiniyo	uke-no\\
     we.\textsc{excl}	\textsc{sub}-see.\textsc{o}.\textsc{inanim}-\textsc{pc}-\textsc{co}	food(\textsc{tp})	3\textsc{sg}.\textsc{poss}=\textsc{emph}	more	all	jointly-eat.\textsc{pp}\\
\glt `In the presence (of this cult house) food is abundant. We ate together all (sorts of food).'\footnote{That means, the cult(house) with its rituals was the source of food abundance.}
\z

\ea\label{ex:nrexkaiuke}
Kaikai ukeno nakap, kaikai kep kiniyo.\\
\gll kaikai	uke-no	nake-p	kaikai	kep	kiniyo\\
     food(\textsc{tp})	we.\textsc{excl}-\textsc{ins}	stay-\textsc{pc}	food(\textsc{tp})	3\textsc{sg}.\textsc{poss}	many\\
\glt `Food was staying with us. (The culthouse's food) variety was plentiful.'
\z

\newpage
\ea\label{ex:nrexnakoka}
Uke nako, kaikai sosolo polinakap.\\
\gll uke	nako	kaikai	sosolo	poli-nake-p\\
     we.\textsc{excl}	be.born	food(\textsc{tp})	like.this	be.there-\textsc{dur}-\textsc{pc}\\
\glt `We were born, and food like this was always around.'
\z

\ea\label{ex:nrexkusono}
Kaikai kusono nakap, kusono nakap.\\
\gll kaikai	kuso-no	nake-p	kuso-no	nake-p\\
     food(\textsc{tp})	always-\textsc{ins}	stay-\textsc{pc}	always-\textsc{ins}	stay-\textsc{pc}\\
\glt `There used to be always something to eat, always.'\footnote{The verb \textit{nake} `stay, sit, live' which is used for living entities hints at animals for hunting.}
\z

\ea\label{ex:nrexbuaiye}
Buai yena pulo yilau ukepiyo.\\
\gll buai	yena	pulo	yilau	uke-pi-yo\\
     white.people	people	come.\textsc{pp}	place	we.\textsc{excl}-\textsc{poss}-\textsc{loc}\\
\glt `(Then) the white people came to our place.'
\z

\ea\label{ex:nrexmisinp}
Riyopuno kupulapno ukenakap, misin pulap.\\
\gll riyopuno	k-pule-p-no	uke-nake-p	misin	pulap\\
     then	\textsc{sub}-come-\textsc{pc}-\textsc{co}	jointly-live-\textsc{pc}	mission	be.filled(\textsc{tp})\\
\glt `Then ... Having come they lived with us and the mission became crowded.'
\z

\ea\label{ex:nrexlaukpi}
Yilau kipiyapno bo aipiro uke muelien, ukero malo.\\
\gll yilau	k-piye-p-no	bo	ai-pi=ro	uke	mueli-en	uke=ro	malo\\
     place	\textsc{sub}-take-\textsc{pc}-\textsc{co}	word	father-\textsc{poss}=\textsc{emph}	we.\textsc{excl}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	we.\textsc{excl}=\textsc{emph}	hear.\textsc{pp}\\
\glt `Having settled down, they proclaimed the word of God to us, and we heard it ourselves.'\footnote{Note that the coming of the whites is referred to by the same verb \textit{piye} `take' as the appropriation of land by the people's ancestor Si (cf. \textref{sc0202} above).}
\z

\ea\label{ex:nrexmikeuk}
Kimike uke ar malo ai bo. Uke so nakap, uke sosolo nakap bipo.\\
\gll kimike	uke	ar	malo	ai	bo	uke	so	nake-p	uke	sosolo	nake-p	bipo\\
     before	we.\textsc{excl}	\textsc{neg}	hear.\textsc{pp}	father	word	we.\textsc{excl}	like	live-\textsc{pc}	we.\textsc{excl}	like.this	live-\textsc{pc}	before(\textsc{tp})\\
\glt `Before, we didn't know the word of God. We were living, we simply lived.'
\z

\ea\label{ex:nrexaibouk}
Ai bo uke kipiyapno, ou, bo maki savepoli.\\
\gll ai	bo	uke	k-piye-p-no	ou	bo	maki	save\_poli\\
     father	word	we.\textsc{excl}	\textsc{sub}-take-\textsc{pc}-\textsc{co}	yes	word	good	have.as.knowledge\\
\glt `Having accepted the word of God, yes, we (then) had knowledge of a good word.'
\z

\ea\label{ex:nrexbipobo}
Bipo bo uke ba piyap, uke sosolo nakap.\\
\gll bipo	bo	uke	ba	piye-p	uke	sosolo	nake-p\\
     before(\textsc{tp})	word	we.\textsc{excl}	\textsc{neg}.\textsc{emph}	take-\textsc{pc}	we.\textsc{excl}	like.this	live-\textsc{pc}\\
\glt `Before we weren't aware of the word (of God), we were just living.'
\z

\ea\label{ex:nrexarkalo}
Emkapuno kaikai mono arka loro, arka loro?\\
\gll emkapuno	kaikai	mono	arka	lo=ro	arka	lo=ro\\
     some.time.before	food(\textsc{tp})	road	where	go.\textsc{pp}=\textsc{emph}	where	go.\textsc{pp}=\textsc{emph}\\
\glt `Where did the food sources go some time ago, where did (the food) go?'
\z

\ea\label{ex:nrexmakyas}
Ai kopi, nuni kopi, memi kopi muelipi: ``Bese makina yasiye, ral makina yasiye, yesi makina yasiye, yala de monanem.''\\
\gll ai	ko-pi	nuni	ko-pi	memi	ko-pi	muel-ipi	bese	maki-na	yasiye	ral	makina	yasiye	yesi	maki-na	yasiye	yala	de	mona-ni-m\\
     father	1\textsc{sg}-\textsc{poss}	maternal.uncle	1\textsc{sg}-\textsc{poss}	grandmother	1\textsc{sg}-\textsc{poss}	talk.to-1\textsc{sg}.\textsc{or}	tulip	good-\textsc{adv}	plant	kind.of.tree	good-\textsc{adv}	plant	aibika	good-\textsc{adv}	plant	\textsc{mod}	you	\textsc{irr}-eat-\textsc{irr}\\
\glt `My father, my uncle, my grandmother, they tell me: ``Plant \textit{tulip}-trees, plant \textit{ral}-trees, plant \textit{aibika}-shrubs, (do that) properly, otherwise you will not eat (vegetables).”'
\z

\ea\label{ex:nrexkekowi}
Kaikai oke ko wili, yala kûnam, yala sepolam.\\
\gll kaikai	o-ke	ko	wili	yala	kûne-m	yala	sepole-m\\
     food(\textsc{tp})	\textsc{prox}-\textsc{aph}	I	carry	now	go.down-\textsc{pos}	now	vanish-\textsc{pos}\\
\glt `This (kind of) food I carry (home from the garden), will soon become less and soon disappear.'
\z

\ea\label{ex:nrexbiopma}
``Biopo makina yai, kakaruk makina yai, de mek ino ire.''\\
\gll biopo	maki-na	yai	kakaruk	maki-na	yai	de	mek	i-no	ire\\
     domesticated.pig	good-\textsc{adv}	take.care.of	chicken(\textsc{tp})	good-\textsc{adv}	take.care.of	you	mouth	\textsc{dist}-\textsc{ins}	make.feel.good\\
\glt `{``}Take good care of the pigs, take good care of the chicken, then you will have a  sweet mouth.”'
\z

\ea\label{ex:nrexarikai}
Kaikai ke yala basepolopisiko. Ari. Kaikai arka loro?\\
\gll kaikai	ke	yala	ba-sepole-pisi-ko	ari	kaikai	arka	lo=ro\\
     food(\textsc{tp})	\textsc{top}	now	\textsc{fac}-vanish-\textsc{cpl}-\textsc{fac}	no	food(\textsc{tp})	where	go.\textsc{pp}=\textsc{emph}\\
\glt `Now the food has vanished completely, oh no, where did the food go?'
\z

\ea\label{ex:nrexbepuno}
Kaikai bepu noro, bi noro, ari. Bipo ke epeyo polip.\\
\gll kaikai	bepu	no=ro	bi	no=ro	ari	bipo	ke	epeyo	poli-p\\
     food(\textsc{tp})	sago.grub	eat.\textsc{pp}=\textsc{emph}	pig	eat.\textsc{pp}=\textsc{emph}	no	before(\textsc{tp})	\textsc{aph}	openly	be.there-\textsc{pc}\\
\glt `The food we ate was sago grubs, pigs. Not anymore. Back then this food was abundant.'
\z

\ea\label{ex:nrexkanwal}
Kaikai kana walpuap, wal kana walpuap, bermepu due bepu kana walpuap. Ukero nep.\\
\gll kaikai	kana	walpue-p	wal	kana	walpue-p	bermepu	due	bepu	kana	walpue-p	uke=ro	ni-p\\
     food(\textsc{tp})	quickly	spread.about-\textsc{pc}	fish	quickly	spread.about-\textsc{pc}	kind.of.sago.grub	sago	kind.of.sago.grub	quickly	spread.about-\textsc{pc}	we.\textsc{excl}=\textsc{emph}	eat-\textsc{pc}\\
\glt `Food was spreading quickly, the fish were spreading quickly, the sago grubs, cultivated sago grubs were spreading quickly. We were eating (more than enough).'
\z

\ea\label{ex:nrexarrkai}
Yalaka ari. Kaikai mi klokni, mi ukero klokni ni.\\
\gll yalaka	ari	kaikai	mi	klokni	mi	uke=ro	klokni	ni\\
     now	no	food(\textsc{tp})	again	one	again	we.\textsc{excl}=\textsc{emph}	one	eat\\
\glt `Nowadays, instead of abundance, there is one piece of food (left). (Being hungry) we eat just one piece.'
\z

\ea\label{ex:nrexokninn}
Kloknina sepiyowolopisiko.\\
\gll klokni-na	sepiye-wole-pisi-ko\\
     one-\textsc{adv}	take.away-move.further-\textsc{cpl}-\textsc{rts}\\
\glt `Occasionally, there was a complete lack (of food) in all places.'\footnote{There is indeed a severe shortage of faunal food resources. A piece of game is found only occasionally. This is caused by the intrusion of the logging companies and the destruction of the primary forest.}
\z

\ea\label{ex:nrexbuaikk}
Buai kupulapno moni mipulo.\\
\gll buai	k-pule-p-no	moni	mi-pulo\\
     white.people	\textsc{sub}-come-\textsc{pc}-\textsc{co}	money(\textsc{tp})	\textsc{iter}-come.\textsc{pp}\\
\glt `After the white people came, more and more money came in.'
\z

\ea\label{ex:nrexmileik}
Ere mile, ikoina melippue moni.\\
\gll ere	mi-le	ikoina	melippue	moni\\
     here	\textsc{iter}-go	much	carry.up.\textsc{pl}.\textsc{o}	money(\textsc{tp})\\
\glt `Here it goes and goes, the price (for food) is increasing considerably.'
\z

\ea\label{ex:nrexarniar}
Yala ko ikoina ar ni, ar ni. Kaikai buaipi moni ikoiele.\\
\gll yala	ko	ikoina	ar	ni	ar	ni	kaikai	buai-pi	moni	ikoiele\\
     now	I	much	\textsc{neg}	eat	\textsc{neg}	eat	food(\textsc{tp})	white.people-\textsc{poss}	money(\textsc{tp}) very.big\\
\glt `Now I don't eat. White people's food is very expensive.'
\z

\ea\label{ex:nrexkesoso}
Uke sosolo mape, moni aska.\\
\gll uke	sosolo	mape	moni	aska\\
     we.\textsc{excl}	like.this	live.\textsc{pl}	money(\textsc{tp})	none\\
\glt `We live like this. We have no money.'
\z

\ea\label{ex:nrexukedli}
Uke dilipeli, uke ikoina nip.\\
\gll uke	d-lipeli	uke	ikoina	ni-p\\
     we.\textsc{excl}	\textsc{lkh}-find	we.\textsc{excl}	much	eat-\textsc{pc}\\
\glt `If we earned (money), we would eat a lot.'
\z
