\ea\label{ex:nrexnimayo}
Ba kep nomayo, ki niulipoko. Epe kepro dop kemiye. Ba pul aska, pul baslaupoko.\\
\gll ba	kep	ni-mayo	ki	ni-uli-pi-ko	epe	kep=ro	dop\_kemiye	ba	pul	aska	pul	ba-slau-pi-ko\\
     breast	3\textsc{sg}.\textsc{poss}	eat-\textsc{mal}.\textsc{pp}	\textsc{aph}	eat-\textsc{prog}-\textsc{lv}-\textsc{rts}	mother	3\textsc{sg}.\textsc{poss}=\textsc{emph}	feel.tired	breast	liquid	none	liquid	\textsc{fac}-dry-\textsc{lv}-\textsc{fac}\\
\glt `She sucked her (mother's) breasts. She indulged in sucking them. Her mother is tired (of this). There is no milk (any more), the breasts have dried up.'
\z

\newpage
\ea\label{ex:nrexopkemi}
Epe aino dop kemiye. Epe aino yilau bayo yilau bayo ilo, riyo inakap.\\
\gll epe	ai-no	dop\_kemiye	epe	ai-no	yilau	ba-yo	yilau	ba-yo	i-lo	ri-yo	i-nake-p\\
     mother	father-\textsc{ins}	feel.tired	mother	father-\textsc{ins}	place	other-\textsc{loc}	place	other-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	\textsc{dist}-\textsc{loc}	\textsc{du}.\textsc{s}-stay-\textsc{pc}\\
\glt `Mother and father are tired of this. The parents went to another place, to another place. There they were staying.'
\z
%
\ea\label{ex:nrexamauli}
Ruri kama uliyopanai. Ruri ki kama nakap. Wor bi sû wapenoi.\\
\gll ruri	kama	uliyepana-i	ruri	ki	kama	nake-p	wor	bi	sû	wape-no-i\\
     child	alone	abandon.\textsc{pp}-\textsc{du}.\textsc{a}	child	\textsc{aph}	alone	stay-\textsc{pc}	dog	meat	fire	put.together-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `They abandoned the child. The child was staying alone. They left a dog, meat, and fire for her.'
\z

\ea\label{ex:nrexnosepi}
Epe aino sepuei. Ruri roipiro wokoi, rowonakeip.\\
\gll epe	ai-no	sepue-i	ruri	roipi=ro	woko-i	ro=wo-nake-i-p\\
     mother	father-\textsc{ins}	run.away-\textsc{du}.\textsc{s}	child	boy=\textsc{emph}	go.together.with.\textsc{pp}-\textsc{du}.\textsc{a}	\textsc{emph}=\textsc{accom}-stay-\textsc{du}.\textsc{s}-\textsc{pc}\\
\glt `Mother and father walk away. They took the (other) child, the boy, with them. They were staying together with him.'
\z

\ea\label{ex:nrexrumkar}
Ruri riyo puloro, rumkari riyo puloro. Wonip yipyo paeau.\\
\gll ruri	ri-yo	pulo=ro	rumkari=ro	ri-yo	pulo=ro	woni-p	yip-yo	paeau\\
     child	\textsc{dist}-\textsc{loc}	come.\textsc{pp}=\textsc{emph}	girl=\textsc{emph}	\textsc{dist}-\textsc{loc}	come.\textsc{pp}=\textsc{emph}	call-\textsc{pc}	house-\textsc{loc}	arrive\\
\glt `The child came there, the girl came there. She was calling out and arrived at the house.'
\z

\ea\label{ex:nrexdiskop}
``Ari, epe kopi ar nake, ai kopi ar nake, disei kopi ar nake. Balok, bamolko.''\\
\gll ari	epe	ko-pi	ar	nake	ai	ko-pi	ar	nake	disei	ko-pi	ar	nake	ba-le-ko	ba-mole-ko\\
     no	mother	1\textsc{sg}-\textsc{poss}	\textsc{neg}	sit	father	1\textsc{sg}-\textsc{poss}	\textsc{neg}	sit	brother	1\textsc{sg}-\textsc{poss}	\textsc{neg}	sit	\textsc{fac}-go-\textsc{fac}	\textsc{fac}-go.\textsc{pl}-\textsc{fac}\\
\glt `No, my mother isn't there, my father isn't there, my brother isn't there. He has gone. They have left (for good).'
\z

\ea\label{ex:nrexumaeau}
Riyopuno smep paliyepana. Bi umaeau nouli, bi puaku bou ulap, sû worno roise. Wor nakap sû mopoyo. Uliyeno.\\
\gll riyopuno	smep	paliyepana	bi	umaeau	ne-uli	bi	puaku	bou	ule-p	sû	wor-no	roise	wor	nake-p	sû	mopo-yo	uliye-no\\
     then	door	open.wide.\textsc{pp}	meat	high.up	go.thither-\textsc{prog}	pig	head	back.limbs	be.there.\textsc{pl}-\textsc{pc}	fire	dog-\textsc{ins}	together	dog	sit-\textsc{pc}	fire	ashes-\textsc{loc}	leave-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Then she opened the door wide. There was meat (hanging) high up, a pig's head and back limbs were there. Embers and a dog were there. The dog was sitting on the ashes. (Those were the things) they (had) left for her.'
\z

\ea\label{ex:nrexpupnop}
Riyopuno kaikai kopopno kinipno. Riyopuno wonpappo pupno, pupno wonpappo.\\
\gll riyopuno	kaikai	k-pi-p-no	k-ni-p-no	riyopuno	wonpappo	pup-no	pup-no	wonpappo\\
     then	food(\textsc{tp})	\textsc{sub}-make-\textsc{pc}-\textsc{co}	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	then	call.blowing.\textsc{pp}	shell-\textsc{ins}	shell-\textsc{ins}	call.blowing.\textsc{pp}\\
\glt `Then, having prepared and eaten the food, she called out by blowing a horn. She blew a (conch) shell.'
\z

\ea\label{ex:nrexweiafi}
Onoro pupno bekimporo, tumbuna Weiafi wuino. Nem kep Siyu. Tumbuna bekimpono.\\
\gll ono=ro	pup-no	bekim-po=ro	tumbuna	Weiafi	wui-no	nem	kep	Siyu	tumbuna	bekim-pi-no\\
     man=\textsc{emph}	shell-\textsc{ins}	answer(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	ancestor(\textsc{tp})	Weiafi	answer-3\textsc{sg}.\textsc{or}.\textsc{pp}	name	3\textsc{sg}.\textsc{poss}	Siyu	ancestor(\textsc{tp})	answer(\textsc{tp})-\textsc{lv}-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `A man replied with a shell, the ancestor Weiafi replied to her. His name is Siyu. The ancestor answered her.'
\z

\ea\label{ex:nrexpuesro}
Wapuesro woni. Epe aino kep woninip, epe aino wonwonnakeinep.\\
\gll Wapues=ro	woni	epe	ai-no	kep	woni-ini-p	epe	ai-no	woni-woni-nake-ini-p\\
     Wapues=\textsc{emph}	call	mother	father-\textsc{ins}	3\textsc{sg}.\textsc{poss}	call-\textsc{nsg}.\textsc{or}-\textsc{pc}	mother	father-\textsc{ins}	call-call-\textsc{dur}-\textsc{nsg}.\textsc{or}-\textsc{pc}\\
\glt `Wapues calls out. She was calling for her mother and father; she kept on calling her parents.'
\z

\ea\label{ex:nrexiyuros}
Siyuro, Siyu bo Wapues wuinap pupno. Yipyo sepa.\\
\gll Siyu=ro	Siyu	bo	Wapues	wui-ne-p	pup-no	yip-yo	sepa\\
     Siyu=\textsc{emph}	Siyu	sound	Wapues	answer-3\textsc{sg}.\textsc{or}-\textsc{pc}	shell-\textsc{ins}	house-\textsc{loc}	arrive.\textsc{pp}\\
\glt `As for Siyu, he answered the calls of Wapues with a horn shell. He arrived at the house.'
\z

\ea\label{ex:nrexahheko}
``Ah, de, ko de wonip? Ko kero epe aino woninip.\\
\gll ah	de	ko	de	woni-p	ko	ke=ro	epe	ai-no	woni-ini-p\\
     ah	you	I	you	call-\textsc{pc}	I	\textsc{top}=\textsc{emph}	mother	father-\textsc{ins}	call-\textsc{nsg}.\textsc{or}-\textsc{pc}\\
\glt `(Wapues says) ``Ah, you, I was calling you? I was calling out for mother and father.”'
\z

\ea\label{ex:nrexkemias}
De kemi asa pulo? Ko de ar wonmap. De ke milap! Ko ke epe aino woninip.''\\
\gll de	ke-mi	asa	pulo	ko	de	ar	woni-me-p	de	ke	mi-le-p	ko	ke	epe	ai-no	woni-ini-p\\
     you	\textsc{top}-too	how	come.\textsc{pp}	I	you	\textsc{neg}	call-2\textsc{sg}.\textsc{or}-\textsc{pc}	you	\textsc{top}	\textsc{iter}-go-\textsc{imp}	I	\textsc{top}	mother	father-\textsc{ins}	call-\textsc{nsg}.\textsc{or}-\textsc{pc}\\
\glt `(Wapues continues), ``You, how did you come? I was not calling you. You go (away)! I was calling mother and father.”'
\z


\ea\label{ex:nrexdobbor}
Dob soreyap: eli bekulu, dop kep wisi, wapoyo nakap. Dob bîka ponap, smep musimayo.\\
\gll dob	soreye-p	eli	bekulu	dop	kep	wisi	wapo-yo	nake-p	dob	bî-ka	pi-ne-p	smep	musi-mayo\\
     eye	realise.\textsc{o}.\textsc{anim}.\textsc{sg}-\textsc{pc}	fat	huge	skin	3\textsc{sg}.\textsc{poss}	black	porch-\textsc{loc}	sit-\textsc{pc}	eye	hole-\textsc{path}	\textsc{lv}-3\textsc{sg}.\textsc{or}-\textsc{pc}	door	shut-\textsc{mal}.\textsc{pp}\\
\glt `She realised (that he was) a fat, huge (man) with black skin. He was sitting on the porch. She was looking at him through a hole (in the wall). She locked the door.'
\z

\ea\label{ex:nrexaisiya}
Riyopuno kaikai si. Yaup sûyo kiyowapno ya mappo. Bi si bese ap roise.\\
\gll riyopuno	kaikai	si	yaup	sû-yo	k-yowe-p-no	ya\_mappo	bi	si	bese	ap	roise\\
     then	food(\textsc{tp})	cook	water	fire-\textsc{loc}	\textsc{sub}-boil-\textsc{pc}-\textsc{co}	stir.sago.\textsc{pp}	meat	cook.\textsc{pp}	tulip	spinach	together\\
\glt `Then she cooked food. After boiling water on the fire, she stirred in sago and cooked meat with \textit{tulip} vegetables and spinach.'
\z

\ea\label{ex:nrexikapon}
Bîka ponamo, sukupuro. Koponamapno kinepno ekuka ar seku.\\
\gll bî-ka	ponamo	sukupu=ro	k-poname-p-no	k-ni-p-no	eku-ka	ar	seku\\
     hole-\textsc{path}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	bush.spirit=\textsc{emph}	\textsc{sub}-give.3\textsc{sg}.\textsc{or}-\textsc{pc}-\textsc{co}	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	anus-\textsc{path}	\textsc{neg}	fall.\textsc{pp}\\
\glt `She gave it to the bush spirit through a hole. When she had given it to him (and) he had eaten it, it did not fall out of his arse.'\footnote{The fact that Siyu digested the food in the normal way shows Wapues that he is a human and not a bush spirit.}
\z

\ea\label{ex:nrexohonou}
Muelne: ``Oh, ono, upuna. Ko so sukupuro, yala ko koniyam.''\\
\gll mueli-ne	oh	ono	upuna	ko	so	sukupu=ro	yala	ko	koniye-m\\
     talk.to-3\textsc{sg}.\textsc{or}	oh	man	alright	I	believe	bush.spirit=\textsc{emph}	\textsc{mod}	I	swallow-\textsc{pos}\\
\glt `She speaks to herself, ``Oh he's human. All right, I thought he is a bush spirit (and could therefore) devour me.”'
\z


\ea\label{ex:nrexyonuid}
Wapoyo nui dupikau punipino duruwa.\\
\gll wapo-yo	nui	dupikau	punipino	duruwa\\
     porch-\textsc{loc}	sleep	deep.darkness	morning	early.morning\\
\glt `He sleeps on the porch, (from) deep darkness (to) the morning, (until) early morning.'
\z

\ea\label{ex:nrexdeilan}
Muelne: ``De îlanap! De biliyo kanakewam, de îlanewe epika!\\
\gll mueli-ne	de	îlane-p	de	bili-yo	k-nake-we-m	de	îlane-we	epi-ka\\
     talk.to-3\textsc{sg}.\textsc{or}	you	give.passage-\textsc{imp}	you	opening-\textsc{loc}	\textsc{proh}-sit-\textsc{ter}-\textsc{proh}	you	give.passage-\textsc{ter}	side-\textsc{path}\\
\glt `She says to him: ``Make room! You can't sit (in front of) the door, move aside!'''
\z

\ea\label{ex:nrexeyoled}
Ko ike seleyo le, dueyo le.'' Smep paliyo, lo.\\
\gll ko\_ike	sele-yo	le	due-yo	le	smep	paliyo	lo\\
     I.myself	garden-\textsc{loc}	go	sago.swamp-\textsc{loc}	go	door	open.\textsc{pp}	go.\textsc{pp}\\
\glt `(Wapues continues), ``I'm going to the garden and the sago swamp myself.'' She opened the door and left.'
\z

\newpage
\ea\label{ex:nrexsiyumu}
Ukiro Siyu muelno: ``Pu de arka ipiyewilipi?''\\
\gll uki=ro	Siyu	mueli-no	pu	de	arka	ipiye\_wili-pi\\
     husband=\textsc{emph}	Siyu	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	water	you	where	bucket\_carry-\textsc{lv}\\
\glt `The (prospective) husband Siyu said to her: ``From where do you get buckets of water?”'
\z

\ea\label{ex:nrexkaropu}
``Ni puro lilikaro.'' ``Pul pu arka?'' ``Pul pu luikaro. Ko rika pulmopi.''\\
\gll ni\_pu=ro	lilika=ro	pul\_pu	arka	pul\_pu	luika=ro	ko	ri-ka	pul\_mopi\\
     drinking.water=\textsc{emph}	upriver=\textsc{emph}	bathing.water	where	bathing.water	downriver=\textsc{emph}	I	\textsc{dist}-\textsc{path}	bathe\\
\glt `(She replied), ``Drinking water is upriver.” -- (He asked) ``And where is the bathing place?” --  (She said), ``Bathing water is downriver. I bathe over there.”'
\z

\ea\label{ex:nrexunolow}
Riyopuno lo. Wapues seleyo lo.\\
\gll riyopuno	lo	Wapues	sele-yo	lo\\
     then	go.\textsc{pp}	Wapues	garden-\textsc{loc}	go.\textsc{pp}\\
\glt `Then she left. Wapues went to the garden.'
\z

\ea\label{ex:nrexakukpi}
Siyu ke kopopno, bi puaku kopopno kiyapopno wîlyo royowe. Yaup sûyo yowo, ya mappo, layowe.\\
\gll Siyu	ke	k-pi-p-no	bi	puaku	k-pi-p-no	k-yapi-p-no	wîl-yo	royo-we	yaup	sû-yo	yowo	ya\_mappo	layo-we\\
     Siyu	\textsc{aph}	\textsc{sub}-make-\textsc{pc}-\textsc{co}	pig	head	\textsc{sub}-make-\textsc{pc}-\textsc{co}	\textsc{sub}-clean-\textsc{pc}-\textsc{co}	dish-\textsc{loc}	lay.\textsc{pp}-\textsc{ter}	water.for.cooking	fire-\textsc{loc}	boil.\textsc{pp}	stir.sago.\textsc{pp}	put.\textsc{pp}-\textsc{ter}\\
\glt `When Siyu had prepared ... Having prepared the pig's head, having cleaned it, he put it in a dish. He boiled water on the fire, stirred in sago and put it (next to the other food).'
\z

\ea\label{ex:nrexweyaks}
Layowe. Ya kusupopopno: ``Bi, yala ko am ar saupo.''\\
\gll layo-we	ya	k-supopi-p-no	bi	yala	ko	am	ar	saupo\\
     put.\textsc{pp}-\textsc{ter}	sago	\textsc{sub}-cut.into.pieces-\textsc{pc}-\textsc{co}	meat	now	I	yet	\textsc{neg}	know\\
\glt `He put it (there). Having cut the sago into portions (he thought), ``Now the meat, I don't know yet (how to cook it).'''
\z

\newpage
\ea\label{ex:nrexpnobue}
Klapno puyo lo. Dop kepro kusupipno bue riyo royepako puyo.\\
\gll k-le-p-no	pu-yo	lo	dop	kep=ro	k-supupi-p-no	bue	ri-yo	royepako	pu-yo\\
     \textsc{sub}-go-\textsc{pc}-\textsc{co}	river-\textsc{loc}	go.\textsc{pp}	skin	3\textsc{sg}.\textsc{poss}=\textsc{emph}	\textsc{sub}-open.for.emptying.\textsc{pl}.\textsc{o}-\textsc{pc}-\textsc{co}	salt	\textsc{dist}-\textsc{loc}	lay.down.\textsc{pp}	river-\textsc{loc}\\
\glt `Having gone ... He went to the river. Having emptied his skin, he put the salt there; (he put it) into the river.'\footnote{Siyu's body contained salt under the skin. The man's supranatural connection with salt may have contributed to Wapues's assent that he is the right husband for her.}
\z

\ea\label{ex:nrexomapai}
Bue kep kroyepakapno somapaip: ``Mepi, upuna, yala ko ino si.''\\
\gll bue	kep	k-royepake-p-no	somapai-p	mepi	upuna	yala	rapue	ko	i-no	si\\
     salt	3\textsc{sg}.\textsc{poss}	\textsc{sub}-lay.down-\textsc{pc}-\textsc{co}	taste.nicely-\textsc{pc}	taste.nice	alright	now	food	I	\textsc{dist}-\textsc{ins}	cook\\
\glt `Having put his salt (into the river), he was tasting (the water), ``It tastes good, alright. Now I cook the food with it.”'
\z

\ea\label{ex:nrexipunin}
Pu bayana rari, pu nina.\\
\gll pu	bayana	rari	pu	ni-na\\
     water	different	dig	water	eat-\textsc{purp}\\
\glt `(Then) he digs for different water, for drinking water.'
\z

\ea\label{ex:nrexaprois}
Bue moniro piyo. Ipi sûyo kiyowapno bi sikûno bese roise ap roise.\\
\gll bue	moni=ro	piyo	ipi	sû-yo	k-yowe-p-no	bi	sikûno	bese	roise	ap	roise\\
     salt	little=\textsc{emph}	take.\textsc{pp}	pot	fire-\textsc{loc}	\textsc{sub}-boil-\textsc{pc}-\textsc{co}	meat	fill.in.\textsc{pp}	tulip	together	spinach	together\\
\glt `He took some salt. When the pot (of water) was boiling on the fire, he put the meat in together with \textit{tulip}-vegetables and spinach.'
\z

\ea\label{ex:nrexpnosom}
Kisipno somapaip, mepi. Puliye ako ripeinoro.\\
\gll k-si-p-no	somapai-p	mepi	puliye	ako	ripei-ne=ro\\
     \textsc{sub}-cook-\textsc{pc}-\textsc{co}	taste-\textsc{pc}	taste.nicely	take.out	wife	share-3\textsc{sg}.\textsc{or}=\textsc{emph}\\
\glt `After cooking, he was tasting it and it tastes good. He takes (the food) out and shares it with his wife.'
\z

\newpage
\ea\label{ex:nrexaesika}
Ki dop kepro pûke. As, mi aesi kamappo. Ono miro dop senispo.\\
\gll ki	dop	kep=ro	pûke	as	mi	aesi	kamap-po	ono	mi=ro	dop	senis-po\\
     \textsc{aph}	skin	3\textsc{sg}.\textsc{poss}=\textsc{emph}	lose	none	again	young	become(\textsc{tp})-\textsc{lv}.\textsc{pp}	man	so=\textsc{emph}	skin	change(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `He loses his skin, all gone. He became young again. The man changed his skin so much.'
\z

\ea\label{ex:nrexapuepu}
Riyopuno rapue puakyo panawe, kinipno nakap, rile nakenap.\\
\gll riyopuno	rapue	puak-yo	pana-we	k-ni-p-no	nake-p	rile	nake-ne-p\\
     then	food	board-\textsc{loc}	put.thither.\textsc{pp}-\textsc{ter}	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	sit-\textsc{pc}	above	sit-3\textsc{sg}.\textsc{or}-\textsc{pc}\\
\glt `Then he put the food on a board. After having eaten he stayed on; he was sitting above (in the house and waiting) for her.'
\z

\ea\label{ex:nrexesyipy}
Wapues yipyo paeau: ``De balok?''\\
\gll Wapues	yip-yo	paeau	de	ba-le-ko\\
     Wapues	house-\textsc{loc}	arrive	you	\textsc{fac}-go-\textsc{fac}\\
\glt `Wapues arrived at the house, ``Have you gone?”'
\z

\ea\label{ex:nrexemnobe}
Em no bekim long em. Bo ba puleno, monemno nakap.\\
\gll em	no	bekim	long	em	bo	ba	pule-no	monemno	nake-p\\
     \textsc{aph}(\textsc{tp})	\textsc{neg}(\textsc{tp})	answer(\textsc{tp})	\textsc{prep}(\textsc{tp})	\textsc{aph}(\textsc{tp})	word	\textsc{neg}.\textsc{emph}	come-3\textsc{sg}.\textsc{or}.\textsc{pp}	silent	sit-\textsc{pc}\\
\glt `He didn't answer her. Not a word came to her, he was sitting silently.'
\z

\ea\label{ex:nrexmeppal}
Smep paliyepanap, anno wiyo: ``Eh, de balok, de ke ono bayana!''\\
\gll smep	paliyepane-p	an-no	wiyo	eh	de	ba-le-ko	de	ke	ono	bayana\\
     door	open.wide-\textsc{pc}	arm-\textsc{ins}	hold.\textsc{pp}	eh	you	\textsc{fac}-go-\textsc{fac}	you	\textsc{top}	man	different\\
\glt `She was opening the door wide, and he embraced her: ``Eh, you've gone! You've gone! You're a different person!'''
\z

\newpage
\ea\label{ex:nrexkookir}
``Ono bayana pari. Oke ko emka ko puloro, ko okiro.''\\
\gll ono	bayana	pari	o-ke	ko	emka	ko	pulo=ro	ko	o-ki=ro\\
     man	different	\textsc{neg}.\textsc{cop}	\textsc{prox}-\textsc{aph}	I	yesterday	I	come.\textsc{pp}=\textsc{emph}	I	\textsc{prox}-\textsc{aph}=\textsc{emph}\\
\glt `(Siyu replied), ``I am not a different person. It is me. I came yesterday. I am the same person.”'
\z

\ea\label{ex:nrexiyaded}
``De makiro. Dedukoyo piyeiya, dedukoyo inake.''\\
\gll de	maki=ro	dedukoyo	piye-i=ya	dedukoyo	i-nake\\
     you	good=\textsc{emph}	we.\textsc{du}.\textsc{incl}	take-\textsc{du}.\textsc{a}=\textsc{emph}	we.\textsc{du}.\textsc{incl}	\textsc{du}.\textsc{s}-live\\
\glt `(She said), ``You're really good. We take each other. Let's live (together).'''
\z

\ea\label{ex:nrexrimboi}
Riyopuno inakero. Ruri faifpela karim, boi rumkari dupua, roipi tripela. So karimpo, faifpela.\\
\gll riyopuno	i-nake=ro	ruri	faifpela	i	karim	boi	rumkari	dupua	roipi	tripela	so	karim-po	faifpela\\
     then	\textsc{du}.\textsc{s}-live=\textsc{emph}	child	five(\textsc{tp})	\textsc{pred}(\textsc{tp})	give.birth(\textsc{tp})	boy(\textsc{tp})	girl	two	boy	three(\textsc{tp})	five	give.birth(\textsc{tp})-\textsc{lv}.\textsc{pp}	five(\textsc{tp})\\
\glt `Then they live together. (They have) five children. She gave birth to boys ... to two girls and to three boys. Five, she gave birth to five (children).'
\z

\ea\label{ex:nrexkurufi}
Kuru, bo kuru.\\
\gll kuru	bo	kuru\\
     be.finished	story	be.finished\\
\glt `(That's) the end. The story is finished.'
\z
