\ea\label{ex:nrexripaek}
Ripi Ripaekyo bûri\_seino. Epe aino basuiweko; ki kama inakap rupperie.\\
\gll Ripi	Ripaek-yo	bûri\_sei-no	epe	ai-no	ba-sui-we-ko	ki	kama	i-nake-p	rupperie\\
     Ripi	Ripaek-\textsc{loc}	sister\_brother-\textsc{ins}	mother	father-\textsc{ins}	\textsc{fac}-die-\textsc{du}.\textsc{s}-\textsc{fac}	\textsc{aph}	alone	\textsc{du}.\textsc{s}-live-\textsc{pc}	orphans\\
\glt `Ripi and Ripaek are brother and sister. (Their) parents had died, and they live  alone as orphans.'
\z

\ea\label{ex:nrexarponi}
Yena rapue ar poninipop. Kuso inakap: rapue an kepno riyepomapoip royeniulipop.\\
\gll yena	rapue	ar	ponini-pi-p	kuso	i-nake-p	rapue	an	kep-no	riye\_pomapi-i-p	roye\_ni-uli-pi-i-p\\
     people	food	\textsc{neg}	give.\textsc{nsg}.\textsc{or}-\textsc{lv}-\textsc{pc}	always	\textsc{du}.\textsc{s}-live-\textsc{pc}	food	hand	3\textsc{sg}.\textsc{poss}-\textsc{ins}	see.\textsc{o}.\textsc{inanim}\_stroll-\textsc{du}.\textsc{a}|\textsc{s}-\textsc{pc}	lay\_eat-\textsc{prog}-\textsc{lv}-\textsc{du}.\textsc{a}-\textsc{pc}\\
\glt `People didn't give them any food. So they lived (like this): Looking for food on their own they strolled around, and they ate what was lying around.'
\z

\ea\label{ex:nrexepopua}
Epo puakuyo ppaliyeinipop, ya pose puakuyo ppaliyeinipop, ya pose ya ise.\\
\gll epo	puaku-yo	ppaliye-ini-pi-p	ya	pose	puaku-yo	ppaliye-ini-pi-p	ya	pose	ya	ise\\
     faeces	head-\textsc{loc}	rub-\textsc{nsg}.\textsc{or}-\textsc{lv}-\textsc{pc}	sago	be.spoiled	head-\textsc{loc}	rub-\textsc{nsg}.\textsc{or}-\textsc{lv}-\textsc{pc}	sago	be.spoiled	sago	char\\
\glt `(The people) used to rub faeces on their heads, they used to rub bad sago on their heads,  foul sago, burned sago.'
\z

\ea\label{ex:nrexamuinu}
Ruri dupua umul nekpamui, ``Nuko ile kaikai painimpi. Yena nuko ar ponien.''\\
\gll ruri	dupua	umul\_nekpamu-i	nuko	i-le	kaikai	painim-pi-i	yena	nuko	ar	ponien\\
     child	two	ponder.\textsc{pp}-\textsc{du}.\textsc{s}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	food(\textsc{tp})	search(\textsc{tp})-\textsc{lv}-\textsc{du}.\textsc{a}	people	we.\textsc{incl}	\textsc{neg}	give.\textsc{nsg}.\textsc{or}.\textsc{pp}\\
\glt `The two children pondered: ``We go and search for food, because the people didn't give us any.'''
\z

\ea\label{ex:nrexipueli}
Pu riyepoi. Pu dob sekui, pu eli dob sekui.\\
\gll pu	riye-po-i	pu	dob	seku-i	pu\_eli	dob	seku-i\\
     water	see.\textsc{o}.\textsc{inanim}-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}	water	eye	fall.\textsc{pp}-\textsc{du}.\textsc{a}	pond.with.brackish.water	eye	fall.down.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `They saw water. Their eyes caught sight of water. Their eyes caught sight of a pond with brackish water.'
\z

\ea\label{ex:nrexikoehp}
Pu baîko. ``Eh pu ulili.'' Ri epeyo lilip. ``Puyo nuko yopi saulno.''\\
\gll pu	ba-î-ko	eh	pu	u-lili	ri	epeyo	lili-p	pu-yo	nuko	yopi-i	saul-no\\
     water	\textsc{fac}-dry.up-\textsc{fac}	eh	water	\textsc{dfac}-be.there	wood	open	be.there-\textsc{pc}	water-\textsc{loc}	we.\textsc{incl}	scoop-\textsc{du}.\textsc{a}	shovel-\textsc{ins}\\
\glt `The water has dried up. ``Eh, here is still (some) water,'' pieces of wood were poking out of the water. ``We will scoop (the remaining water) with a shovel.'''
\z

\ea\label{ex:nrexwaerip}
Bayopiko, baîko. Riyopuno wal luwapoi, waeripi wamo waeaup.\\
\gll ba-yopi-i-ko	ba-î-ko	riyopuno	wal	luwapi-i	waeripi	wamo	waeaup\\
     \textsc{fac}-scoop-\textsc{du}.\textsc{a}-\textsc{fac}	\textsc{fac}-dry.up-\textsc{fac}	then	fish	catch-\textsc{du}.\textsc{a}	kind.of.fish	kind.of.fish	kind.of.fish\\
\glt `They have scooped it, and (the pond) is dry. Then they catch the fish (with their  hands), \textit{waeripi}-fish, \textit{wamo}-fish, \textit{waeup}-fish.'
\z

\ea\label{ex:nrexaupsep}
Waeaup sepolo. Ri epeyo piapoiro.\\
\gll waeaup	sepolo	ri	epeyo	piapo-i=ro\\
     kind.of.fish	vanish.\textsc{pp}	wood	open	lift.up.\textsc{pp}-\textsc{du}.\textsc{a}=\textsc{emph}\\
\glt `The \textit{waeaup}-fish vanished. They uncovered the wood (at the bottom of the pond).'
\z

\ea\label{ex:nrexeherep}
``Eh erepe? Anayo?'' ``Koyo? Koyo memi depi.''\\
\gll eh	ere=pe	ana-yo	koyo	koyo	memi	de-pi\\
     eh	\textsc{prox}=\textsc{q}	who-\textsc{loc}	we.\textsc{du}.\textsc{excl}	we.\textsc{du}.\textsc{excl}	grandchildren	2\textsc{sg}-\textsc{poss}\\
\glt `{``}Eh, what is this? Who are you two?'' ``We two? We two are your grandchildren.'''
\z

\ea\label{ex:nrexkeriep}
``Oke ri epeyo ari, yip kopi. Awe! Deyo iminep!'' Dob soreyewap, ``Erepe? Deyo ba powe?''\\
\gll o-ke	ri	epeyo	ari	yip	ko-pi	awe	deyo	i-mini-p	dob	soreye-we-p	ere=pe	deyo	ba	pi-we\\
     \textsc{prox}-\textsc{aph}	wood	open	no	house	1\textsc{sg}-\textsc{poss}	come.\textsc{imp}	you.\textsc{du}	\textsc{du}.\textsc{s}-come.hither-\textsc{imp}	eye	look.intently.\textsc{o}.\textsc{anim}.\textsc{sg}-\textsc{du}.\textsc{o}-\textsc{pc}	\textsc{prox}=\textsc{q}	you.\textsc{du}	what	do-\textsc{ter}\\
\glt `{``}This is not uncovered wood, (this is) my house. Come! Come here you two!'' She  was scrutinising them: ``What is this? What did you two do?'''
\z

\largerpage
\ea\label{ex:nrexeponoy}
``Yena koyo epono ya poseno ya iseno ppaliyeen. Dop koyopi ppulae. Koyo makina ar inakap.''\\
\gll yena	koyo	epo-no	ya	pose-no	ya	ise-no	ppaliye-en	dop	koyo-pi	ppulae	koyo	maki-na	ar	i-nake-p\\
     people	we.\textsc{du}.\textsc{excl}	faeces-\textsc{ins}	sago	be.spoiled-\textsc{ins}	sago	char-\textsc{ins}	rub-\textsc{nsg}.\textsc{or}.\textsc{pp}	body	we.\textsc{du}.\textsc{excl}-\textsc{poss}	bad	we.\textsc{du}.\textsc{excl}	good-\textsc{adv}	\textsc{neg}	\textsc{du}.\textsc{s}-live-\textsc{pc}\\
\glt `{``}The people rubbed us with faeces, with bad sago, with burned sago. Our bodies are bad, we were not living well.'''
\z

\ea\label{ex:nrexiereru}
Pul mopien. Epo ya puno pusiyeen, seke sayeen. Dop maki, ereru.\\
\gll pul\_mopi-en	epo	ya	pu-no	pusiye-en	seke	saye-en	dop	maki	ereru\\
     bathe-\textsc{nsg}.\textsc{or}.\textsc{pp}	faeces	sago	water-\textsc{ins}	wash-\textsc{nsg}.\textsc{or}.\textsc{pp}	hair	shave-\textsc{nsg}.\textsc{or}.\textsc{pp}	body	good	strong\\
\glt `She bathed them. She washed the faeces and sago off their bodies with water and cut their hair. (Now, their) bodies are good and strong.'
\z

\ea\label{ex:nrexipsusi}
Dipsu sien, wîlno ripien. Riyopuno royenen.\\
\gll dipsu	si-en	wîl-no	ripi-en	riyopuno	roye-en\\
     rice	cook-\textsc{nsg}.\textsc{or}.\textsc{pp}	dish-\textsc{ins}	distribute-\textsc{nsg}.\textsc{or}.\textsc{pp}	then	lay-\textsc{nsg}.\textsc{or}.\textsc{pp}\\
\glt `She cooked rice for them and distributed it on plates for them. Then she gave it to them.'
\z

\ea\label{ex:nrexmakopi}
Kinoipno dipsu neno ponien. Ropyo uleien, muelien: ``Deyo ilap! Ko yala pu makopi.''\\
\gll k-ni-i-p-no	dipsu	neno	ponien	rop-yo	ulei-en	mueli-en	deyo	i-le-p	ko	yala	pu	makopi\\
     \textsc{sub}-eat-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	rice	raw	give.\textsc{nsg}.\textsc{or}.\textsc{pp}	basket-\textsc{loc}	put.inside-\textsc{nsg}.\textsc{or}.\textsc{pp}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{du}	\textsc{du}.\textsc{s}-go-\textsc{imp}	I	now	rain	perform.magic\\
\glt `When they had eaten, she gave them raw rice. She put it in a basket and said to them:  ``You two go! I will perform rain magic now.'''
\z

\ea\label{ex:nrexsilpok}
Riyopuno sowo sepiana. Pu silepokûne. Pu poro dupikau.\\
\gll riyopuno	sowo	sepiana	pu	silepokûne	pu	po=ro	dupikau\\
     then	ginger	fragment.\textsc{pp}	rain	drip.down.steadily	rain	\textsc{lv}.\textsc{pp}=\textsc{emph}	pervasive.darkness\\
\glt `Then she ground and scattered ginger. The rain falls steadily. It rained  heavily until darkness.'
\z

\ea\label{ex:nrexpepuol}
Pepuol peia kûno, ani duruwa.\\
\gll pepuol	peia	kûno	ani	duruwa\\
     kind.of.frog	kind.of.frog	go.down.\textsc{pp}	day	early.morning\\
\glt `\textit{Pepuol}-frogs and \textit{peia}-frogs went down (to waterpools), in daylight at dawn.'
\z

\ea\label{ex:nrexpiubak}
Punipino ilo, pu riyeip. Piu bakûnko, luwapoi. Piu baluwaoiko, loloi, klokni weloi yipyo.\\
\gll punipino	i-lo	pu	riye-i-p	piu	ba-kûne-ko	luwapo-i	piu	ba-luwapi-i-ko	lolo-i	klokni	welo-i	yip-yo\\
     morning	\textsc{du}.\textsc{s}-go.\textsc{pp}	water	see.\textsc{o}.\textsc{inanim}-\textsc{du}.\textsc{a}-\textsc{pc}	frog	\textsc{fac}-go.down-\textsc{fac}	catch.\textsc{pp}-\textsc{du}.\textsc{a}	frog	\textsc{fac}-catch-\textsc{du}.\textsc{a}-\textsc{fac}	wrap.\textsc{pp}-\textsc{du}.\textsc{a}	one	carry.away.\textsc{pp}-\textsc{du}.\textsc{a}	house-\textsc{loc}\\
\glt `In the morning (the children) went and saw the water. Frogs have jumped down (to waterpools), and they caught them with their hands. (Soon) they have caught enough frogs. They wrapped them (into leaves). One (package) they carried to the house.'\footnote{Probably they walk back to their old house in the village where they used to live before they encountered the spirit woman at the bottom of the pond. Then they would have persuaded the villagers to go catching fish.}
\z

\ea\label{ex:nrexripiri}
Ripi Ripaekyo yena muelien, ``Ine molap, piu kauna oso mape.''\\
\gll Ripi	Ripaek-yo	yena	mueli-en	ine	mole-p	piu	kauna	oso	mape\\
     Ripi	Ripaek-\textsc{loc}	people	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{pl}	go.\textsc{pl}-\textsc{imp}	frog	in.great.numbers	more	sit.\textsc{pl}\\
\glt `Ripi and Ripaek said to the people: ``Go (down there), (there are) frogs in great numbers, there are more (there).'''
\z

\ea\label{ex:nrexpiuluw}
Yena molo kiniyo. Epe aino ruri roise piu luwapo.\\
\gll yena	molo	kiniyo	epe	ai-no	ruri	roise	piu	luwapo\\
     people	go.\textsc{pl}.\textsc{pp}	all	mother	father-\textsc{ins}	child	with	frog	catch.\textsc{pp}\\
\glt `So the people went, all of them. Mothers and fathers together with the children, and they caught the frogs.'
\z

\ea\label{ex:nrexbusukn}
Pu busukna, kimike nomoina, boyopuno losna die poniyena. Waeripi pule. Ere mini wiyeyap! Waeripi sepolo.\\
\gll pu	busuk-na	kimike	nomoi-na	boyopuno	los-na	die\_poniye-na	waeripi	pule	ere	mini	wiye-yep	waeripi	sepolo\\
     water	shin-\textsc{aff}	before	ankle-\textsc{aff}	later	thigh-\textsc{aff}	waist-\textsc{aff}	kind.of.fish	come	here	come.hither	catch-\textsc{imp}.\textsc{pl}	kind.of.fish	vanish.\textsc{pp}\\
\glt `The water reached their shins, first their ankles, later their thigh and waist. The \textit{waeripi}-fish come, here they come, catch them! The \textit{waeripi}-fish vanished.'
\z

\ea\label{ex:nrexlowaer}
Mi rika baka lo. Waeripi ere mini, ere mini oke rkaro.\\
\gll mi	ri-ka	ba-ka	lo	waeripi	ere	mini	ere	mini	o-ke	r-ka=ro\\
     again	\textsc{dist}.\textsc{emph}-\textsc{path}	other-\textsc{path}	go.\textsc{pp}	kind.of.fish	here	come.hither	here	come.hither	\textsc{prox}-\textsc{aph}	\textsc{dist}.\textsc{emph}-\textsc{path}=\textsc{emph}\\
\glt `(The fish) went there, to the other side. The \textit{waeripi}-fish come here, come here; here and over there (they are moving).'
\z

\ea\label{ex:nrexinerna}
Pu inerna walina, yena puyo silekûnwepu.\\
\gll pu	iner-na	wali-na	yena	pu-yo	silekûne-wepu\\
     water	armpit-\textsc{aff}	neck-\textsc{aff}	people	water-\textsc{loc}	sink-\textsc{quant}.\textsc{s}.\textsc{pp}\\
\glt `The water reaches up to the armpit and to the neck. The people sank into the water.'
\z

\ea\label{ex:nrexinuges}
Buoko buar wepulo. Yena kiniyo leliewepu. Yena supuli basupuliko. Yena ulewolelayep, Inuges ol Ir Inuges pur Ir.\\
\gll Buoko	buar	wepulo	yena	kiniyo	lelie-wepu	yena	supuli	ba-supuli-ko	yena	ulewolelaye-p	Inuges	ol	Ir	Inuges	pur	Ir\\
     Buoko	stone.axe	bring.\textsc{pp}	people	all	kill.people-\textsc{quant}.\textsc{o}.\textsc{pp}	people	die.\textsc{pl}.\textsc{s}	\textsc{fac}-die.\textsc{pl}.\textsc{s}-\textsc{fac}	people	be.there.\textsc{pl}\_move.further\_lay-\textsc{pc}	Inuges	mountain	Ir	Inuges	plain	Ir\\
\glt `Buoko brought an axe and killed all people. The people die. They are dead.  The people('s bodies) were lying everywhere, the Inuges from the mountain Ir and the Inuges from the plain Ir.'
\z

\ea\label{ex:nrexiskolo}
Isko lopapien dopyo. Kuru.\\
\gll isko	lopapi-en	dop-yo	kuru\\
     black.colour	mark.with.paint.\textsc{pl}.\textsc{o}-\textsc{nsg}.\textsc{or}.\textsc{pp}	body-\textsc{loc}	be.finished\\
\glt `He marked their bodies with black paint. It is finished.'
\z

\ea\label{ex:nrexyaraak}
Amou ai kep sano, ``Yara ako kopi ari? Dupu arka lo?''\\
\gll Amou	ai	kep	sa-no	Yara	ako	ko-pi	ari	Dupu	arka	lo\\
     Amou	father	3\textsc{sg}.\textsc{poss}	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	Yara	wife	1\textsc{sg}-\textsc{poss}	no	Dupu	where	go.\textsc{pp}\\
\glt `Amou asked his father Yara: ``My wife is not (there)? Where did Dupu go?'''
\z

\ea\label{ex:nrexdupyen}
``Balok. Dupu yena wulien; yena piu lu.'' Monomno molo, dob sopop: yena kiniyo supuli.\\
\gll ba-le-ko	Dupu	yena	wuli-en	yena	piu	lu	monomno	molo	dob\_sopi-p	yena	kiniyo	supuli\\
     \textsc{fac}-go-\textsc{fac}	Dupu	people	follow-\textsc{nsg}.\textsc{or}.\textsc{pp}	people	frog	shoot.\textsc{pp}	along.the.path	go.\textsc{pl}.\textsc{pp}	look.intently-\textsc{pc}	people	all	die.\textsc{pl}\\
\glt `{``}She has left. Dupu followed the people who caught frogs.'' (Amou and his father Yara) walked along the path and realised: All the people died.'
\z

\ea\label{ex:nrexamouye}
Amou yena kiniyo lakwepu. Ako kepyo nekip. Ako elno; epul kepyo isau re polip.\\
\gll Amou	yena	kiniyo	lakwepu	ako	kep-yo	neki-p	ako	el\_no	epul	kep-yo	isau	re	poli-p\\
     Amou	people	all	count.names.\textsc{quant}.\textsc{o}.\textsc{pp}	wife	3\textsc{sg}.\textsc{poss}-\textsc{loc}	stand-\textsc{pc}	wife	pregnant	ear	3\textsc{sg}.\textsc{poss}-\textsc{loc}	kind.of.bird	feather	be.there-\textsc{pc}\\
\glt `Amou counted all the people by their names. He was standing next to his (dead) wife. The woman had been pregnant; in her ear was an \textit{isau}-feather.'
\z

\ea\label{ex:nrexlumiso}
Buoko dor epe Amoupi suelo. Amou suelno. Dor lumî solo upuna wopuem.\\
\gll Buoko	dor\_epe	Amou-pi	suelo	Amou	suel-no	dor	lumî	solo	upuna	wopue-m\\
     Buoko	big.toe	Amou-\textsc{poss}	cut.\textsc{pp}	Amou	cut-\textsc{co}	foot	scar	only	alright	be.marked-\textsc{pos}\\
\glt `Buoko cut Amou's big toe. (Now) Amou has a gash. The foot has a big scar. Alright, he  will be marked now.'\footnote{Amou is one of the survivors of the killing. He must have encountered Buoko who cut his foot. Amou fled and planned to take revenge.}
\z

\ea\label{ex:nrexolasau}
Amou ol epi baka sesiyo, Buoko nisenap. Ol epi baka nisenap, baka lenap, ol Asaul.\\
\gll Amou	ol	epi	baka	sesiyo	Buoko	nise-ne-p	ol	epi	baka	nise-ne-p	baka	le-ne-p	ol	Asaul\\
     Amou	mountain	side	other	get.out.of.the.way	Buoko	lie.in.wait-3\textsc{sg}.\textsc{or}-\textsc{pc}	mountain	side	other	lie.in.wait-3\textsc{sg}.\textsc{or}-\textsc{pc}	other	wait.for-3\textsc{sg}.\textsc{or}-\textsc{pc}	mountain	Asaul\\
\glt `Amou fled to the other side of the mountain and was lying in wait for Buoko. On the other side of the mountain he was lying in wait. He was waiting for him on the other (side), (at) Mount Asaul.'
\z

\ea\label{ex:nrexsepuep}
Sepue pele suko, bopap sepalo. Sepue pele wauna. Riyopuno wami bî sepeipana, dob pop, ``Buoko upule.'' Amou lu paliya.\\
\gll sepue\_pele	suko	bopap	sepalo	sepue\_pele	wauna	riyopuno	wami	bî	sepeipana	dob	pi-p	Buoko	u-pule	Amou	lu	paliya\\
     trough.leaf	cut.\textsc{pp}	pig.trap	fence.\textsc{pp}	trough.leaf	put.side.by.side.\textsc{pp}	then	window	hole	drill.into.\textsc{pp}	eye	\textsc{lv}-\textsc{pc}	Buoko	\textsc{dfac}-come	Amou	shoot.\textsc{pp}	be.dead\\
\glt `He cut \textit{limbum}-ribs and fenced a pig trap. He put the \textit{limbum}-ribs side by side. Then he drilled a spying hole and was looking (through): ``Here comes Buoko.''  Amou shot, (Buoko) is dead.'
\z

\ea\label{ex:nrexkikike}
Amou kike, kikero kikero. Ber pûke, ber yelo mono powolap. Yeloyo ye, boyo puana.\\
\gll Amou	kike	kike=ro	kike=ro	ber	pûke	ber	yelo\_mono	powole-p	yelo-yo	ye	boyo	puana\\
     Amou	run	run=\textsc{emph}	run=\textsc{emph}	tongue	lose	tongue	along.the.ground	burst-\textsc{pc}	ground-\textsc{loc}	fall.over.\textsc{pp}	later	stand.up.\textsc{pp}\\
\glt `Amou runs and runs. He loses his tongue as the tongue was dragging on the ground (behind him). He fell over onto the ground, but he got up (again).'
\z

\ea\label{ex:nrexuokoba}
Yena muelien, ``Ko Buoko baluik.''\\
\gll yena	mueli-en	ko	Buoko	ba-lui-ko\\
     people	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	I	Buoko	\textsc{fac}-shoot-\textsc{fac}\\
\glt `He said to the people: ``I have shot Buoko.'''
\z

\ea\label{ex:nrexpubues}
Pu bueso lili. Pu ikoiele. Pu nem kep roki: Pu ppulae. Puro yena lil lili, Inuges lil lili. Yena sukei riyo mape Pu\-ppulaeyo.\\
\gll pu	bue-so	lili	pu	ikoiele	pu	nem	kep	ro-ki	pu	ppulae	pu=ro	yena	lil	lili	Inuges	lil	lili	yena	sukei	ri-yo	mape	Puppulae-yo\\
     water	sea-\textsc{sim}	be.there	water	very.big	water	name	3\textsc{sg}.\textsc{poss}	\textsc{prox}.\textsc{emph}-\textsc{aph}	water	Ppulae	water=\textsc{emph}	people	blood	be.there	Inuges	blood	be.there	people	spirit	\textsc{dist}-\textsc{loc}	live.\textsc{pl}	Puppulae-\textsc{loc}\\
\glt `The water is (salty) like the sea. The lake is very big. The name of the water is this: Pu Ppulae. The water is the blood of the people, the blood of the Inuges. The spirit of the people lives there, in the Pu Ppulae lake.'\footnote{The name of the lake is ``Bad-water-lake''.}
\z
