\ea\label{ex:nrexomeduy}
Ome duyo lo. Duyo klapno bike lu.\\
\gll Ome	du-yo	lo	du-yo	k-le-p-no	bike	lu\\
     Ome	bush-\textsc{loc}	go.\textsc{pp}	bush-\textsc{loc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	cassowary	shoot.\textsc{pp}\\
\glt `Ome went to the bush. Having gone to the bush he shot a cassowary.'
\z

\ea\label{ex:nrexsanoak}
Bike wepulo yipyo, epe kep sano, ``Ako kopi ari?''\\
\gll bike	wepulo	yip-yo	epe	kep	sa-no	ako	ko-pi	ari\\
     cassowary	bring.\textsc{pp}	house-\textsc{loc}	mother	3\textsc{sg}.\textsc{poss}	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	wife	1\textsc{sg}-\textsc{poss}	no\\
\glt `He brought the cassowary home and asked his mother: ``Isn't my wife (here)?'''
\z

\ea\label{ex:nrexroburi}
``Ako depi kero bûri depi woko.''\\
\gll ako	de-pi	ke=ro	bûri	de-pi	woko\\
     wife	2\textsc{sg}-\textsc{poss}	\textsc{top}=\textsc{emph}	sister	2\textsc{sg}-\textsc{poss}	accompany.\textsc{pp}\\
\glt `{``}Your wife? She accompanied your sister.'''
\z

\ea\label{ex:nrexsukuip}
Yena sûkui pi. Yena pupno woni, yena garamut rino kowe. Yelo niyerip.\\
\gll yena	sûkui	pi	yena	pup-no	woni	yena	garamut	ri-no	kowe	yelo	niyeri-p\\
     people	dance	\textsc{lv}	people	conch-\textsc{ins}	call	people	garamut(\textsc{tp})	wood-\textsc{ins}	beat	ground	shake-\textsc{pc}\\
\glt `The people dance. They blow the conches and they beat the garamut. The ground was shaking.'\footnote{The \textit{garamut} is the native Papuan signal drum, the slit gong. It is also used for celebrations and entertainment as here.}
\z

\ea\label{ex:nrexukuiro}
Yuku yakono sûkui roki pi, Yiyo; sûkui kipiwe.\\
\gll yuku	yako-no	sûkui	roki	pi	Yi-yo	sûkui	kipiwe\\
     man	woman-\textsc{ins}	dance	sway	\textsc{lv}	Yi-\textsc{loc}	dance	close.together\\
\glt `Men and women sway, at the place Yi; they dance close together.'\footnote{The translation `close together' is not confirmed. It is due to paraphrasing the Kilmeri verb in Tok Pisin by Margaret Osi. The verb only occurred in this context and this text.}
\z

\ea\label{ex:nrexpakokp}
Ki pako kipiyapno pe roise wel.\\
\gll ki	pako	k-piye-p-no	pe	roise	wel\\
     \textsc{aph}	bow	\textsc{sub}-take-\textsc{pc}-\textsc{co}	arrow	together	carry.\textsc{pp}\\
\glt `After taking the bow he carried it together with arrows.'
\z

\ea\label{ex:nrexkuipip}
Yena sûkui popno ako kep reyap.\\
\gll yena	sûkui	pi-p-no	ako	kep	reye-p\\
     people	dance	\textsc{lv}-\textsc{pc}-\textsc{co}	wife	3\textsc{sg}.\textsc{poss}	see.\textsc{o}.\textsc{anim}.\textsc{sg}-\textsc{pc}\\
\glt `While people danced, he was looking at his wife.'
\z

\ea\label{ex:nrexomekid}
Ome ki dob sana, ``Eh ako kopi ke roki ono baya.'' Sûkui roki ipoi.\\
\gll Ome	ki	dob	sana	eh	ako	ko-pi	ke	roki	ono	baya	sûkui	roki	i-po-i\\
     Ome	\textsc{aph}	eye	look.thither.\textsc{pp}	eh	wife	1\textsc{sg}-\textsc{poss}	\textsc{top}	sway	man	different	dance	sway	\textsc{du}.\textsc{s}-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{s}\\
\glt `Ome looked there: ``Eh, my wife is dancing with another man”. They swayed in the dance.'
\z

\ea\label{ex:nrexyoulei}
``Ko pe pakoyo ulei, ko pulapi.''\\
\gll ko	pe	pako-yo	ulei	ko	pulapi\\
     I	arrow	bow-\textsc{loc}	put.inside	I	strain\\
\glt `{``}I will put an arrow on the bow and draw the bow”, (says Ome to himself).'
\z

\newpage
\ea\label{ex:nrexyokupo}
Biliyo nekipno ki pulapiro.\\
\gll bili-yo	neki-p-no	ki	pulapi=ro\\
     opening-\textsc{loc}	stand-\textsc{pc}-\textsc{co}	\textsc{aph}	strain=\textsc{emph}\\
\glt `While	standing at	the	entrance, he	draws	the	bow.'
\z

\ea\label{ex:nrexiyonek}
Yena kiniyo supuli, yelo bî kana powolo. Yena riyo kûno, yena kiniyo riyo kûpo.\\
\gll yena	kiniyo	supuli	yelo	bî	kana	powolo	yena	riyo	kûno	yena	kiniyo	riyo	kûpo\\
     people	all	die.\textsc{pl}	ground	hole	quickly	burst.\textsc{pp}	people	\textsc{dist}-\textsc{loc}	go.down.\textsc{pp}	people	all	\textsc{dist}-\textsc{loc}	go.down.\textsc{pl}.\textsc{pp}\\
\glt `All the people die, (because) a crack suddenly opened up in the ground. The people fell in there, all the people fell in there.'\footnote{It is not clear whether the people are hit by Ome's arrows and then fall down into the hole, or whether the fissure opened as a (supernatural) punishment for the people's immoral conduct some time ago. People about the age of 80-90, which belong to the oldest generation still alive, seem to remember an earthquake in the area. Their children told me this.}
\z

\ea\label{ex:nrexpaupiy}
Yena riyo kûkûnapno bili sowe. Yelo mi bî dori, paupiyo, dori solo.\\
\gll yena	ri-yo	k-kûne-p-no	bili	sowe	yelo	mi	bî	dori	paupiyo	dori	solo\\
     people	\textsc{dist}-\textsc{loc}	\textsc{sub}-go.down-\textsc{pc}-\textsc{co}	opening	cover	ground	again	hole	turn.back	shut.completely.\textsc{pp}	turn.back	only\\
\glt `After the people fell in there, the opening (of the ground) closes. The ground turns back again (to close the cave). It closed completely, it just turned back.'
\z

\ea\label{ex:nrexseiwel}
Pako ukepi yena dop sei wel.\\
\gll pako	uke-pi	yena	dop	sei	wel\\
     bow	we.\textsc{excl}-\textsc{poss}	people	skin	white	carry.\textsc{pp}\\
\glt `The people with white skin colour carried (away) our bows.'
\z

\ea\label{ex:nrexopwisi}
Pako dop seipi uke dop wisi piyo.\\
\gll pako	dop	sei-pi	uke	dop	wisi	piyo\\
     bow	skin	white-\textsc{poss}	we.\textsc{excl}	skin	black	take.\textsc{pp}\\
\glt `We black-skinned people took the bows of the white-skinned (people).'
\z

\ea\label{ex:nrexwilkob}
Bawelko, basepoloko. Aska.\\
\gll ba-wili-ko	ba-sepole-ko	aska\\
     \textsc{fac}-carry-\textsc{fac}	\textsc{fac}-lose-\textsc{fac}	nothing\\
\glt `They have been carried (away), they are lost. Nothing (is left).'
\z

\ea\label{ex:nrexbolapi}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `The story is finished.'
\z
