\ea\label{ex:nrexdetemu}
Kimike kimike dukiro dete mumunano kaikai puno solo siulipop. Bue aska, puno solo.\\
\gll kimike	kimike	duki=ro	dete	mumuna-no	kaikai	pu-no	solo	si-uli-pi-p	bue	aska	pu-no	solo\\
     in.the.past	in.the.past	true=\textsc{emph}	ancestor	ancestor-\textsc{ins}	food(\textsc{tp})	water-\textsc{ins}	only	cook-\textsc{prog}-\textsc{lv}-\textsc{pc}	salt	none	water-\textsc{ins}	only\\
\glt `In really past times, (our) ancestors used to cook their food with water only. There was no salt, only water.'
\z

\ea\label{ex:nrexmunako}
Riyopuno mumuna kopi nem kep ke Wumeye woppuo pul ppuo. Welro du mono. Oyo nu Wupepp yilauyo.\\
\gll riyopuno	mumuna	ko-pi	nem	kep	ke	Wumeye	woppuo	pul	ppuo	wel=ro	du	mono	o-yo	nu	Wupepp	yilau-yo\\
     then	ancestor	1\textsc{sg}-\textsc{poss}	name	3\textsc{sg}.\textsc{poss}	\textsc{aph}	Wumeye	kind.of.tree	fruit	climb.\textsc{pp}	carry.\textsc{pp}=\textsc{emph}	bush	path	\textsc{prox}-\textsc{loc}	sleep.\textsc{pp}	Wupepp	place-\textsc{loc}\\
\glt `Then my ancestor -- his name is Wumeye -- climbed for fruits of a \textit{woppuo}-tree  and carried them (along) the bush track. He spent the night here, in the place (called) Wupepp.'
\z

\ea\label{ex:nrexbuemap}
Nuknoko epul so malap, bue welipi. Woppuo pul kupurapopno bueno mapa.\\
\gll nuknoko	epul	so	male-p	bue	weli-pi	woppuo	pul	k-purapi-p-no	bue-no	mapa\\
     sleep.for.some.nights.\textsc{pp}	ear	like	hear-\textsc{pc}	sea	approach-\textsc{lv}	kind.of.tree	fruit	\textsc{sub}-break-\textsc{pc}-\textsc{co}	salt-\textsc{ins}	taste.\textsc{pp}\\
\glt `He slept for some days (in the bush) and heard the (sound of the) sea approaching.  Having opened the \textit{woppuo}-fruits, he tasted them with salt.'\footnote{Wumeye must have walked all the way down to the coast following the noise of the surf.}
\z

\ea\label{ex:nrexbueriy}
So nep oke bue. Riyopuno doripulo yilauyo.\\
\gll so	ni-p	o-ke	bue	riyopuno	doripulo	yilau-yo\\
     like	eat-\textsc{pc}	\textsc{prox}-\textsc{aph}	salt	then	come.back.\textsc{pp}	village-\textsc{loc}\\
\glt `So he was eating: this is salt. Then he returned to the village.'
\z

\newpage
\ea\label{ex:nrexabomos}
Yena bo mosaupoen: ``Bue ko ika riyeko, bue paeau. Bue ko bariyeko.''\\
\gll yena	bo	mosaupi-en	bue	ko	i-ka	riye-ko	bue	paeau	bue	ko	ba-riye-ko\\
     people	word	teach-\textsc{nsg}.\textsc{or}.\textsc{pp}	sea	I	\textsc{dist}-\textsc{path}	see.\textsc{o}.\textsc{inanim}-\textsc{rts}	sea	arrive	sea	I	\textsc{fac}-see.\textsc{o}.\textsc{inanim}-\textsc{fac}\\
\glt `He informed the people: ``The sea, I saw it there. I reached the sea, I have seen the sea.”'
\z

\ea\label{ex:nrexkirewl}
Riyopuno yena muelien. Mono kirewoloro, bue paeau.\\
\gll riyopuno	yena	muel-en	mono	kirewolo=ro	bue	paeau\\
     then	people	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	path	cut.foot.track.\textsc{pp}=\textsc{emph}	sea	arrive\\
\glt `Then he talked to the people, (and afterwards) they cut a bush track to reach the sea.'
\z

\ea\label{ex:nrexsompai}
Somapaip bue, ou. Riyopuno ul kusukelipno bue isaeaupo, yilauyo melpulup.\\
\gll somapai-p	bue	ou	riyopuno	ul	k-sukeli-p-no	bue	isaeaupo	yilau-yo	meli\_pulup\\
     taste-\textsc{pc}	sea	yes	then	bamboo	\textsc{sub}-cut.\textsc{pl}.\textsc{o}-\textsc{pc}-\textsc{co}	sea	fill.in.\textsc{pp}	village-\textsc{loc}	carry.\textsc{pl}.\textsc{o}\_come.\textsc{pl}.\textsc{pp}\\
\glt `They were tasting the salty water, yes. Then, having cut bamboo containers, they  filled them (with) salt water and carried them to the village.'
\z

\ea\label{ex:nrexwumbue}
Bue nem kep bue Wumeye.\\
\gll bue	nem	kep	bue	Wumeye\\
     salt	name	3\textsc{sg}.\textsc{poss}	salt	Wumeye\\
\glt `The salt's name is Wumeye salt.'
\z

\ea\label{ex:nrexkuskur}
Kuru bo kopi.\\
\gll kuru	bo	ko-pi\\
     be.finished	story	1\textsc{sg}-\textsc{poss}\\
\glt `My story is finished.'
\z
