\ea\label{ex:nrexsukmas}
Sukupu masalai, Kopukei ono.\\
\gll sukupu	masalai	Kopukei	ono\\
     bush.spirit	bush.spirit(\textsc{tp}) Kopukei man\\
\glt `Sukupu (is) a bush spirit, Kopukei (is) a human.'
\z

\ea\label{ex:nrexaipnop}
Kopukei aipo nopuanako, sukupu aipo nopuanako.\\
\gll Kopukei	aipo	nopuane-ko	sukupu	aipo	nopuane-ko\\
     Kopukei	taboo.sign	set.a.taboo.thither-\textsc{rts}	bush.spirit	taboo.sign	set.a.taboo.thither-\textsc{rts}\\
\glt `Kopukei had set a taboo sign and the bush spirit had set a taboo sign.'
\z

\ea\label{ex:nrexkeipul}
Kopukei pulo, bermepu mo. Komoipno yeloyo layo. Bermepu we yeloyo.\\
\gll Kopukei	pulo	bermepu	mo	k-moi-p-no	yelo-yo	layo	bermepu	we	yelo-yo\\
     Kopukei	come.\textsc{pp}	sago.grub	cut.\textsc{pp}	\textsc{sub}-cut-\textsc{pc}-\textsc{co}	ground-\textsc{loc}	put.\textsc{pp}	sago.grub	fall.down	ground-\textsc{loc}\\
\glt `Kopukei came and cut sago grubs. Having cut them, he put them on the ground.  (Some) sago grubs fell on the ground.'
\z

\ea\label{ex:nrexukeibe}
Kopukei bermepu puraporo. Rapiyo, luwapo ropyo, niskûno.\\
\gll Kopukei	bermepu	purapo=ro	rapiyo	luwapo	rop-yo	niskûno\\
     Kopukei	sago.grub	break.\textsc{pp}=\textsc{emph}	fetch.\textsc{pp}	catch.with.hands.\textsc{pp}	basket-\textsc{loc}	fill.in.\textsc{pp}\\
\glt `Kopukei broke the sago grubs, took them, grabbed them, and filled them in a basket.'
\z

\ea\label{ex:nrexwelmal}
Boyopuno sukupu pulo; sukupu welmalalpo.\\
\gll boyopuno	sukupu	pulo	sukupu	welmalalpo\\
     later	bush.spirit	come.\textsc{pp}	bush.spirit	come.near.like.noise.of.breaking.wood.\textsc{pp}\\
\glt `Later the bush spirit came. The bush spirit (heard) the noise of breaking wood.'
\z

\ea\label{ex:nrexupudob}
Sukupu dob po, ``Upunaro, bepu de bamoiko. Nuko purapii.''\\
\gll sukupu	dob	po	upuna=ro	bepu	de	ba-moi-ko	nuko	purapi-i\\
     bush.spirit	eye	\textsc{lv}.\textsc{pp}	alright=\textsc{emph}	sago.grub	you	\textsc{fac}-cut-\textsc{fac}	we.\textsc{incl}	break-\textsc{du}.\textsc{a}\\
\glt `The bush spirit saw: ``Alright, you have cut the sago grubs. We'll break them together.'''
\z

\ea\label{ex:nrexbepuki}
Bepu kiniyo kupurapoipno, kuru. Kiniyo banawoko.\\
\gll bepu	kiniyo	k-purapi-i-p-no	kuru	kiniyo	ba-nawe-ko\\
     sago.grub	all	\textsc{sub}-break-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	be.finished	all	\textsc{fac}-use.up-\textsc{fac}\\
\glt `They broke all sago grubs, (then) it was finished. All (the grubs) are used up.'
\z

\ea\label{ex:nrexriwull}
Kopukei ruri bepu parno lolo, rileyo weppuo. Ri wuliyo sakana sowo, dob pop.\\
\gll Kopukei	ruri	bepu	par-no	lolo	rileyo	weppuo	ri	wuli-yo	sakana	sowo	dob	pi-p\\
     Kopukei	child	sago.grub	palm.mat-\textsc{ins}	wrap.\textsc{pp}	above	carry.up.\textsc{pp}	tree	tree.top-\textsc{loc}	secretly	hide.\textsc{pp}	eye	\textsc{lv}-\textsc{pc}\\
\glt `Kopukei's child wrapped the sago grubs in a palm leaf, carried them up, hid himself secretly in the top of the tree, and was watching (what happened on the ground).'
\z

\ea\label{ex:nrexeponoi}
Sukupu Kopukei muelno: ``Nuko bepu epono imali.''\\
\gll sukupu	Kopukei	mueli-no	nuko	bepu	epo-no	i-mali\\
     bush.spirit	Kopukei	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	we.\textsc{incl}	sago.grub	faeces-\textsc{ins}	\textsc{du}.\textsc{a}-fight\\
\glt `The bush spirit said to Kopukei: ``We'll fight with faeces over the sago grubs.'''
\z

\ea\label{ex:nrexbisepn}
Epono kimalko, mi bîsepno imali. Bîsepno kimalko, mi sûno imali. Sûno kimalko, mi rino imali.\\
\gll epo-no	k-i-mali-ko	mi	bîsep-no	i-mali	bîsep-no	k-i-mali-ko	mi	sû-no	i-mali	sû-no	k-i-mali-ko	mi	ri-no	i-mali\\
     faeces-\textsc{ins}	\textsc{sub}-\textsc{du}.\textsc{a}-fight-\textsc{rts}	again	saliva-\textsc{ins}	\textsc{du}.\textsc{a}-fight	saliva-\textsc{ins}	\textsc{sub}-\textsc{du}.\textsc{a}-fight-\textsc{rts}	again	fire-\textsc{ins}	\textsc{du}.\textsc{a}-fight	fire-\textsc{ins}	\textsc{sub}-\textsc{du}.\textsc{a}-fight-\textsc{rts}	again	stick-\textsc{ins}	\textsc{du}.\textsc{a}-fight\\
\glt `Having fought with faeces, they fight with saliva. Having fought with saliva, they fight with fire. Having fought with fire, they (finally) fight with sticks.'
\z

\ea\label{ex:nrexpukeil}
Rino kimalko, sukupu mi Kopukei lelio. Paliya.\\
\gll ri-no	k-i-mali-ko	sukupu	mi	Kopukei	lelio	paliya\\
     stick-\textsc{ins}	\textsc{sub}-\textsc{du}.\textsc{a}-fight-\textsc{rts}	bush.spirit	again	Kopukei	kill.people.\textsc{pp}	be.dead\\
\glt `While fighting with sticks the bush spirit killed Kopukei. He is dead.'
\z

\newpage
\ea\label{ex:nrexuloimo}
Kopukei ako lelio. Rino suloimoina kapiyo. Paliya. Uki akono leliowe.\\
\gll Kopukei	ako	lelio	ri-no	suloimoina	kapiyo	paliya	uki	ako-no	lelio-we\\
     Kopukei	wife	kill.people.\textsc{pp}	stick-\textsc{ins}	extraordinarily	beat.fiercely.\textsc{pp}	be.dead	husband	wife-\textsc{ins}	kill.people.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `(The bush spirit) killed Kopukei's wife. It beat her fiercely and fervently with a stick, then she is dead. It killed both, husband and wife.'\footnote{The wife is introduced only here.}
\z

\ea\label{ex:nrexaikimi}
Ai kimike puenpo, mi epe puenpo.\\
\gll ai	kimike	puenpo	mi	epe	puenpo\\
     father	first	cut.meat.\textsc{pp}	then	mother	cut.meat.\textsc{pp}\\
\glt `First (the spirit) cut the father's flesh. Then it cut the mother's flesh.'
\z

\ea\label{ex:nrexkupuep}
Sukupu epe aino uroyo nisei, ulei, meli yip kepyo.\\
\gll sukupu	epe	ai-no	uro-yo	nisei	ulei	meli	yip	kep-yo\\
     bush.spirit	mother	father-\textsc{ins}	netbag-\textsc{loc}	fill.in	put.inside	carry.\textsc{pl}.\textsc{o}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}\\
\glt `The bush spirit puts (the flesh) of father and mother in a netbag. It puts it inside (and) carries all (the flesh) to its house.'
\z

\ea\label{ex:nrexuyoyek}
Yipyo kemelipno, ruri yeloyo kûno. Boyo ekuyo yek.\\
\gll yip-yo	k-meli-p-no	ruri	yelo-yo	kûno	boyo	ekuyo	yek\\
     house-\textsc{loc}	\textsc{sub}-carry.\textsc{pl}.\textsc{o}-\textsc{pc}-\textsc{co}	child	ground-\textsc{loc}	go.down.\textsc{pp}	later	behind	follow.one's.traces.\textsc{pp}\\
\glt `When (the bush spirit) had carried all the meat home, the boy went down on the ground. Later on he followed the foot steps (of the bush spirit).'
\z

\ea\label{ex:nrexaippew}
Ruri kep aipo pewo. Yilauyo lo. Kiu kep muelien: ``Epe aino sukupu no. Leliowe.''\\
\gll ruri	kep	aipo	pewo	yilau-yo	lo	kiu	kep	mueli-en	epe	ai-no	sukupu	no	lelio-we\\
     child	3\textsc{sg}.\textsc{poss}	taboo.sign	put.upright.\textsc{pp}	village-\textsc{loc}	go.\textsc{pp}	clan	3\textsc{sg}.\textsc{poss}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	mother	father-\textsc{ins}	bush.spirit	eat.\textsc{pp}	kill.people.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `The boy set up a taboo sign, went to the village and said to his clan: ``The bush spirit  ate mother and father. It killed them.'''
\z

\ea\label{ex:nrexlepsap}
Yena ruri Kopukeipi wulino. Monomno wulino. Ruri musiyo: ``De lap, sanap!''\\
\gll yena	ruri	Kopukei-pi	wuli-no	monomno	wuli-no	ruri	musiyo	de	le-p	sa-ne-p\\
     people	child	Kopukei-\textsc{poss}	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}	along.the.path	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}	child	send.\textsc{pp}	you	go-\textsc{imp}	ask-3\textsc{sg}.\textsc{or}-\textsc{imp}\\
\glt `The people followed Kopukei's child. They followed him along the track. They sent  the child (ahead): ``Go (and) ask (the bush spirit)!'''
\z

\ea\label{ex:nrexulakip}
``Sukupu epe aino ulakip! Epe aino kopi, ari?''\\
\gll sukupu	epe	ai-no	u-laki-p	epe	ai-no	ko-pi	ari\\
     bush.spirit	mother	father-\textsc{ins}	\textsc{dfac}-fetch-\textsc{pc}	mother	father-\textsc{ins}	1\textsc{sg}-\textsc{poss}	no\\
\glt `{``}Bush spirit, now release mother and father! My parents, aren't they (here)?'''
\z

\ea\label{ex:nrexohkoar}
``Oh, ko ar reyowe.'' ``Ari. De yipyoro wepulowe. De dupuapi.''\\
\gll oh	ko	ar	reyo-we	ari	de	yip-yo=ro	wepulo-we	de	dupuapi\\
     oh	I	\textsc{neg}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}-\textsc{du}.\textsc{o}	no	you	house-\textsc{loc}=\textsc{emph}	bring.\textsc{pp}-\textsc{du}.\textsc{o}	you	lie\\
\glt `{``}Oh, I didn't see them.'' ``No way, you brought them both into the house. You are lying.'''
\z

\ea\label{ex:nrexsiwepu}
Yipyo lili. Ipiyo siwe. Puaku dupua solo ipiyo lilip. Dû kiniyo banok.\\
\gll yip-yo	lili	ipi-yo	si-we	puaku	dupua	solo	ipi-yo	lili-p	dû	kiniyo	ba-ni-ko\\
     house-\textsc{loc}	be.there	pot-\textsc{loc}	cook-\textsc{du}.\textsc{o}	head	two	only	pot-\textsc{loc}	be.there-\textsc{pc}	meat	all	\textsc{fac}-eat-\textsc{fac}\\
\glt `(Their bodies) are in the house. (The spirit) cooked them in a pot. There were only the two heads  in the pot. All the meat is eaten up.'
\z

\ea\label{ex:nrexneppin}
Puaku dupua epe ainopi sipeiwe. Lûli neppino sipeiwe, lûli no.\\
\gll puaku	dupua	epe	ai-no-pi	sipei-we	lûli	neppi-no	sipei-we	lûli	no\\
     head	two	mother	father-\textsc{ins}-\textsc{poss}	take.out.the.brain-\textsc{du}.\textsc{o}	brain	bush.knife-\textsc{ins}	take.out.the.brain-\textsc{du}.\textsc{o}	brain	eat.\textsc{pp}\\
\glt `(The bush spirit) takes the brains out of the two heads of father and mother. It took the brains out with a knife and ate the brains.'
\z

\ea\label{ex:nrexamneki}
Sukupu buar piyo anno. ``Am ko kûne. De am nekiwep.''\\
\gll sukupu	buar	piyo	an-no	am	ko	kûne	de	am	neki-we-p\\
     bush.spirit	stone.axe	take.\textsc{pp}	hand-\textsc{ins}	first	I	go.down	you	still	stand-\textsc{ter}-\textsc{imp}\\
\glt `The bush spirit took an axe with its hands: ``I come down first. You remain staying there.'''\footnote{The bush spirit thought it had only to fight the boy. The people had tricked the evil spirit in that they had sent the boy ahead.}
\z

\ea\label{ex:nrexwapila}
Yena yip pakiyo nekip, bîskilyo nekip. Enuka nekip, woloyo nekip. Peno wapilayo. Dop kepyo polip, sui. Paliya.\\
\gll yena	yip	pakiyo	neki-p	bîskilyo	neki-p	enuka	neki-p	wolo-yo	neki-p	pe-no	wapilayo	dop	kep-yo	poli-p	sui	paliya\\
     people	house	next.to	stand-\textsc{pc}	underneath	stand-\textsc{pc}	in.the.corner	stand-\textsc{pc}	ladder-\textsc{loc}	stand-\textsc{pc}	arrow-\textsc{ins}	shoot.with.arrows.\textsc{pp}	body	3\textsc{sg}.\textsc{poss}-\textsc{loc}	be.there-\textsc{pc}	die	be.dead\\
\glt `The people were standing around the house. They were standing underneath (it), they were standing at the corners, they were standing on the ladder. They peppered it with arrows (which) stuck in its body. It is dying. (The bush spirit) is dead.'
\z

\ea\label{ex:nrexyeppue}
Anno piyeppuowepu, yipyo panapo.\\
\gll an-no	piyeppue-wepu	yip-yo	panapo\\
     hand-\textsc{ins}	lift.up-\textsc{quant}.\textsc{o}.\textsc{pp}	house-\textsc{loc}	put.thither.\textsc{pp}\\
\glt `They lifted it up with their hands and put it into the house.'
\z

\ea\label{ex:nrexnonori}
Smep bili musi. Sûno noriye. Sû kiniyo yip iseye sukupu roisero.\\
\gll smep	bili	musi	sû-no	noriye	sû	kiniyo	yip	iseye	sukupu	roise=ro\\
     door	opening	lock	fire-\textsc{ins}	fill.in	fire	many	house	char	bush.spirit	together=\textsc{emph}\\
\glt `They lock the opening of the door and start fires. Many fires burn the house to the ground together with the bush spirit.'
\z

\newpage
\ea\label{ex:nrexupuuli}
Sukupu ûli pulwolo.\\
\gll sukupu	ûli	pulwolo\\
     bush.spirit	gall.bladder	burst.\textsc{pp}\\
\glt `The gall bladder of the bush spirit burst.'\footnote{Generally, the burst of the gall bladder is clear evidence that the bush spirit is dead.}
\z

\ea\label{ex:nrexloisim}
Yena yipyo molo. Yilauyo löiso molo.\\
\gll yena	yip-yo	molo	yilau-yo	löi-so	molo\\
     people	house-\textsc{loc}	go.\textsc{pl}.\textsc{pp}	village-\textsc{loc}	löi-\textsc{sim}	go.\textsc{pl}.\textsc{pp}\\
\glt `The people went to (their) houses in the village. They went with shouts of victory.'
\z
