\ea\label{ex:nrexclaudi}
Claudia epul male. Bo sukupupi. Bermepu bo kep so.\\
\gll Claudia	epul	male	bo	sukupu-pi	bermepu	bo	kep	so\\
     Claudia	ear	hear	story	bush.spirit-\textsc{poss}	sago.grub	story	3\textsc{sg}.\textsc{poss}	like\\
\glt `Claudia listen. A story about a bush spirit. The story of the sago grubs (goes) like (this).'
\z

\ea\label{ex:nrexkopuke}
Sukupu dupua nako, Kopukei dupua nako. Sukupu dupua nako, Kopueki dupua nako.\\
\gll sukupu	dupua	nako	Kopukei	dupua	nako	sukupu	dupua	nako	Kopukei	dupua	nako\\
     bush.spirit	two	beget	Kopukei	two	beget	bush.spirit	two	beget	Kopukei	two	beget\\
\glt `The bush spirit placed two (taboo signs). Kopukei placed two (taboo signs). The bush spirit put two and Kopukei put two.'
\z

\ea\label{ex:nrexopukei}
Sukupu kupulapno Kopukei ki popiye, aipo popiye.\\
\gll sukupu	k-pule-p-no	Kopukei	ki	popiye	aipo	popiye\\
     bush.spirit	\textsc{sub}-come-\textsc{pc}-\textsc{co}	Kopukei	\textsc{aph}	take.away	taboo.sign	take.away	\\
\glt `(Later), having come (again), the bush spirit takes away Kopukei's (signs). It takes away the taboo signs.'
\z

\newpage
\ea\label{ex:nrexkpopiy}
Kopopiyapno Kopukei riyepuloro: ``Ah, sukupu asa popiyeko. Aipo ke kopi, bepu ke kopi.''\\
\gll k-popiye-p-no	Kopukei	riye\_pulo=ro	ah	sukupu	asa	popiye-ko	aipo	ke	ko-pi	bepu	ke	ko-pi\\
     \textsc{sub}-take.away-\textsc{pc}-\textsc{co}	Kopukei	see.\textsc{o}.\textsc{inanim}\_come.\textsc{pp}=\textsc{emph}	ah	bush.spirit	how	take.away-\textsc{rts}	taboo.sign	\textsc{aph}	1\textsc{sg}-\textsc{poss}	sago.grubs	\textsc{aph}	1\textsc{sg}-\textsc{poss}\\
\glt `When it had taken away (the taboo signs), Kopukei came and saw (that): ``Ah, why did the bush spirit take them? The taboo signs are mine and the sago grubs are mine.'''
\z

\ea\label{ex:nrexlepnoy}
Klapno yilauyo lo. Ako kep lakiyoro, ruri kep lakiyoro.\\
\gll k-le-p-no	yilau-yo	lo	ako	kep	lakiyo=ro	ruri	kep	lakiyo=ro\\
     \textsc{sub}-go-\textsc{pc}-\textsc{co}	village-\textsc{loc}	go.\textsc{pp}	wife	3\textsc{sg}.\textsc{poss}	fetch.\textsc{pp}=\textsc{emph}	child	3\textsc{sg}.\textsc{poss}	fetch.\textsc{pp}=\textsc{emph}\\
\glt `He left (the bush place). He went to the village and brought on his wife and his son.'
\z

\ea\label{ex:nrexbepumo}
Kupulapno bepu moi. Bepu komoipno yeloyo layo. Yeloyo klayewapno.\\
\gll k-pule-p-no	bepu	mo-i	bepu	k-moi-i-p-no	yelo-yo	layo	yelo-yo	k-laye-we-p-no\\
     \textsc{sub}-come-\textsc{pc}-\textsc{co}	sago.grubs	cut.\textsc{pp}-\textsc{du}.\textsc{a}	sago.grubs	\textsc{sub}-cut-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	ground-\textsc{loc}	lay.\textsc{pp}	ground-\textsc{loc}	\textsc{sub}-lay-\textsc{ter}-\textsc{pc}-\textsc{co}\\
\glt `Having come (there) (father and mother) cut the sago grubs.  Having cut them, they put them on the ground. They had put them there.'
\z

\ea\label{ex:nrexyopuno}
Riyopuno purapi. Kuprapopno, bepu ba kololapno ruri muel: ``De rile ppuewap!''\\
\gll riyopuno	purapi	k-purapi-p-no	bepu	ba	k-lole-p-no	ruri	muel	de	rile	ppue-we-p\\
     then	break	\textsc{sub}-break-\textsc{pc}-\textsc{co}	sago.grubs	other	\textsc{sub}-wrap-\textsc{pc}-\textsc{co}	child	talk.to.\textsc{pp}	you	above	climb-\textsc{ter}-\textsc{imp}\\
\glt `Then they break (the sago grubs). After breaking and wrapping some grubs (the father) said to his son: ``Climb up high (into the tree)!'''
\z

\newpage
\ea\label{ex:nrexrileyo}
Ruri rileyo ppuowe. Bepu par isiye weppuap. Rileyo nakap, dob pop.\\
\gll ruri	rileyo	ppuo-we	bepu	par	isiye	weppue-p	rileyo	nake-p	dob	pi-p\\
     child	above	climb.\textsc{pp}-\textsc{ter}	sago.grubs	pannier	together	carry.up-\textsc{pc}	above	sit-\textsc{pc}	eye	\textsc{lv}-\textsc{pc}\\
\glt `The son climbed up and was carrying up the sago grubs together with the pannier. He was sitting above and looking.'
\z

\ea\label{ex:nrexkupupu}
Sukupu pule. Skupu kupulapno muelne: ``Bepu de bamoiko?''\\
\gll sukupu	pule	sukupu	k-pule-p-no	muel-ne	bepu	de	ba-moi-ko\\
     bush.spirit	come	bush.spirit	\textsc{sub}-come-\textsc{pc}-\textsc{co}	talk.to-3\textsc{sg}.\textsc{or}	sago.grubs	you	\textsc{fac}-cut-\textsc{fac}\\
\glt `The bush spirit comes. Having come it says: ``The sago grubs, you have cut them?'''
\z

\ea\label{ex:nrexupunar}
Upunaro. Duko poi. Kuprapopno muelne: ``Nuko bîsepno imali.''\\
\gll upuna=ro	duko	pi-i	k-purapi-p-no	muel-ne	nuko	bîsep-no	i-mali\\
     alright=\textsc{emph}	we.\textsc{incl}	do-\textsc{du}.\textsc{a}	\textsc{sub}-break-\textsc{pc}-\textsc{co}	talk.to-3\textsc{sg}.\textsc{or}	we.\textsc{incl}	saliva-\textsc{ins}	\textsc{du}.\textsc{s}-fight\\
\glt `All right, we do it together. After breaking the grubs, (the bush spirit) says to him: ``(Now) we are going to fight with saliva.'''
\z

\ea\label{ex:nrexbesepn}
Bîsepno kimalko igo. Bîsep ari. Muelne: ``Nuko rino imali.''\\
\gll bîsep-no	ki	i-mali-ko	i\_go	bîsep	ari	muel-ne	nuko	ri-no	i-mali\\
     saliva-\textsc{ins}	\textsc{aph}	\textsc{du}.\textsc{s}-fight-\textsc{rts}	and.so.on(\textsc{tp})	saliva	no	talk.to-3\textsc{sg}.\textsc{or}	we.\textsc{incl}	stick-\textsc{ins}	\textsc{du}.\textsc{s}-fight\\
\glt `They fought with saliva. They fought and fought, no saliva (left). (The spirit) says: ``(Now) we'll fight with sticks.'''
\z

\ea\label{ex:nrexmaliko}
Rino kimalko. Sukupu pulap, Kopukei pulap. Kimalko, kimalko, kimalko.\\
\gll ri-no	ki	i-mali-ko	sukupu	pulap	Kopukei	pulap	ki	i-mali-ko	ki	i-mali-ko	ki	i-mali-ko\\
     stick-\textsc{ins}	\textsc{aph}	\textsc{du}.\textsc{s}-fight-\textsc{rts}	bush.spirit	abound.in(\textsc{tp})	Kopukei	abound.in(\textsc{tp})	\textsc{aph}	\textsc{du}.\textsc{s}-fight-\textsc{rts}	\textsc{aph}	\textsc{du}.\textsc{s}-fight-\textsc{rts}	\textsc{aph}	\textsc{du}.\textsc{s}-fight-\textsc{rts}\\
\glt `They fought with sticks. The bush spirit beat and Kopukei beat. They  fought and fought and fought.'
\z

\ea\label{ex:nrexwinimk}
Riyopuno sukupu winim, kilim meri.\\
\gll riyopuno	sukupu	winim	kilim	meri\\
     then	bush.spirit	win.\textsc{tr}(\textsc{tp})	kill.\textsc{tr}(\textsc{tp})	woman(\textsc{tp})\\
\glt `Then the bush spirit won and killed the woman.'
\z

\ea\label{ex:nrexmaelip}
Ako kep maelipo. Paliya. Ruri dob pop: epe paliya. Igo. Mi ai kep. Ai kep kimalko rino igo.\\
\gll ako	kep	maeli-po	paliya	ruri	dob\_pi-p	epe	paliya	i\_go	mi	ai	kep	ai	kep	ki	i-mali-ko	ri-no	i\_go\\
     wife	3\textsc{sg}.\textsc{poss}	fell-\textsc{lv}.\textsc{pp}	be.dead	child	eye\_\textsc{lv}-\textsc{pc}	mother	be.dead	and.so.on(\textsc{tp})	then	father	3\textsc{sg}.\textsc{poss}	father	3\textsc{sg}.\textsc{poss}	\textsc{aph}	\textsc{du}.\textsc{s}-fight-\textsc{rts}	stick-\textsc{ins}	and.so.on(\textsc{tp})\\
\glt `It killed (Kopukei's) wife. She is dead. The boy was watching. His mother is dead. Then his father, his father. They fought on with sticks.'
\z

\ea\label{ex:nrexeponom}
Epono, mike epono kimalipno. Ri miriyemon. Mi rino. Ai paliya.\\
\gll epo-no	mike	epo-no	k-i-mali-p-no	ri	mi-riye\_mon	mi	ri-no	ai	paliya\\
     faeces-\textsc{ins}	first	faeces-\textsc{ins}	\textsc{sub}-\textsc{du}.\textsc{s}-fight-\textsc{pc}-\textsc{co}	stick	\textsc{iter}-see.\textsc{o}.\textsc{inanim}\_come.hither.\textsc{pp}	then	stick-\textsc{ins}	father	be.dead\\
\glt `First with faeces. Having fought with faeces, they saw some sticks nearby. So (they fought) with sticks. The father is dead.'
\z

\ea\label{ex:nrexparnol}
Riyopuno parno lolo. Parno kololapno, uki lolapno, yako lolap. Yilau kepyo wel, yip kepyo.\\
\gll riyopuno	par-no	lolo	par-no	k-lole-p-no	uki	lole-p-no	yako	lole-p	yilau	kep-yo	wel	yip	kep-yo\\
     then	pannier-\textsc{ins}	tie.\textsc{pp}	pannier-\textsc{ins}	\textsc{sub}-tie-\textsc{pc}-\textsc{co}	husband	tie-\textsc{pc}-\textsc{co}	woman	tie-\textsc{pc}	village	3\textsc{sg}.\textsc{poss}-\textsc{loc}	carry.\textsc{pp}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}\\
\glt `Then (the bush spirit) tied them on the pannier. After doing that it was lashing up the husband and the woman. It carried them to its place, to its house.'
\z

\ea\label{ex:nrexwepepo}
Wepepo, layo, ako puenpo. Ipino siu. Riyopuno puaku oyo royo, yeniyo. Riyopuno uki puenpo.\\
\gll wepepo	layo	ako	puenpo	ipi-no	siu	riyopuno	puaku	o-yo	royo	yeni-yo	riyopuno	uki	puenpo\\
     carry.on.top.\textsc{pp}	lay.\textsc{pp}	wife	cut.meat.\textsc{pp}	pot-\textsc{ins}	cook.\textsc{pp}	then	head	\textsc{prox}-\textsc{loc}	put.\textsc{pp}	plank-\textsc{loc}	then	husband	cut.meat.\textsc{pp}\\
\glt `It carried (the pannier) up and put it (there). It cut up the wife and cooked (the flesh) in a pot. Then it put the head here, on a plank. Afterwards it cut up the husband.'
\z

\ea\label{ex:nrexruriri}
Ruri riyo kûnoro. Kukûnapno ruri lo. Klapno oki wili.\\
\gll ruri	ri-yo	kûno=ro	k-kûne-p-no	ruri	lo	k-le-p-no	o-ki	wili\\
     child	tree-\textsc{loc}	go.down=\textsc{emph}	\textsc{sub}-go.down-\textsc{pc}-\textsc{co}	child	go.\textsc{pp}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	\textsc{prox}-\textsc{aph}	carry\\
\glt `The boy climbed down from the tree. Having climbed down he left (the place). While walking he carries these (grubs).'
\z

\ea\label{ex:nrexaipope}
Aipo pewono yip kepyo. Aipo pewonapno. Riyopuno dorimaliyo.\\
\gll aipo	pewo-no	yip	kep-yo	aipo	pewo-ne-p-no	riyopuno	dori\_maliyo\\
     taboo.sign	erect-3\textsc{sg}.\textsc{or}.\textsc{pp}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	taboo.sign	erect-3\textsc{sg}.\textsc{or}-\textsc{pc}-\textsc{co}	then	turn.back\_go.home.\textsc{pp}\\
\glt `He erected a taboo sign at his bush hut. After erecting the sign he returned back home.'
\z

\ea\label{ex:nrexloroyi}
Loro yilauyo. Kiu kep muelini: ``Ine pulupepe. Nuko mole. Epe aino sukupu leliowe. Ine awe!''\\
\gll lo=ro	yilau-yo	kiu	kep	mueli-ini	ine	pulupi-p-e	nuko	mole	epe	ai-no	sukupu	lelio-we	ine	awe\\
     go.\textsc{pp}=\textsc{emph}	village-\textsc{loc}	clan	3\textsc{sg}.\textsc{poss}	talk.to-\textsc{nsg}.\textsc{or}	you.\textsc{pl}	come.\textsc{pl}-\textsc{imp}-\textsc{voc}	we.\textsc{incl}	go.\textsc{pl}	mother	father-\textsc{ins}	bush.spirit	kill.\textsc{pp}-\textsc{du}.\textsc{o}	you.\textsc{pl}	come.\textsc{imp}\\
\glt `He went to the village. He tells his kin people: ``You come! Let's go! A bush spirit 	killed my parents. Come!'''
\z

\ea\label{ex:nrexkeppep}
Kiu kep pe pakono kisau romolo. Ruri wulino.\\
\gll kiu	kep	pe	pako-no	kisau	ro=molo	ruri	wuli-no\\
     clan	3\textsc{sg}.\textsc{poss}	arrow	bow-\textsc{ins}	?	\textsc{emph}=go.\textsc{pl}.\textsc{pp}	child	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `His clan took bows and arrows. They went and followed the boy.'
\z

\newpage
\ea\label{ex:nrexriburi}
Ruri buri wonppueno: ``Epe aino wulikûnap!'' Yena yipele baka baka.\\
\gll ruri	buri	woni\_ppue-no	epe	ai-no	wuli\_kûne-p	yena	yip-ele	ba-ka	ba-ka\\
     child	go.ahead	call\_go.up-3\textsc{sg}.\textsc{or}.\textsc{pp}	mother	father-\textsc{ins}	follow\_go.down-\textsc{imp}	people	house-very	other-\textsc{path}	other-\textsc{path}\\
\glt `The boy went ahead and called up to (the bush spirit): ``Come down with my parents 	following!'' The people (surrounded) the house from all sides.'
\z

\ea\label{ex:nrexainoko}
``Epe aino? Ko ke ari. Aska. Ko ar ukepulo.'' Sukupu kaeli, ruri kaeli.\\
\gll epe	ai-no	ko	ke	ari	aska	ko	ar	uke-pulo	sukupu	kaeli	ruri	kaeli\\
     mother	father-\textsc{ins}	I	\textsc{top}	no	none	I	\textsc{neg}	jointly-come.\textsc{pp}	bush.spirit	strong	child	strong\\
\glt `{``}Mother and father? I didn't do anything. No, we didn't come together.''  The bush spirit is unrelenting, and the boy is also unrelenting.'
\z

\ea\label{ex:nrexnapwap}
Kiu nap. Wapilayekûpu, paliya. Yipyo panepo. Sû piyamu. Layepanepo ruri aesino roise isaeaupo.\\
\gll kiu	na-p	wapi\_laye\_kûpu	paliya	yip-yo	pane-po	sû	piyamu	laye\_pane-po	ruri	aesi-no	roise	isaeaupo\\
     clan	go.inside-\textsc{pc}	collect\_lay\_go.down.\textsc{pl}.\textsc{a}.\textsc{pp}	be.dead	house-\textsc{loc}	do.thither-\textsc{lv}.\textsc{pp}	fire	take.hither.\textsc{pp}	lay\_do.thither-\textsc{lv}.\textsc{pp}	child	young-\textsc{ins}	together	let.suckle.\textsc{pp}\\
\glt `The kin people were entering (the house). They peppered (the bush spirit) with arrows. It is dead. They put it into the house. They got fire and set (the house) on fire, together with all the young ones (of the bush spirit), (which) were suckling babies.'
\z

\ea\label{ex:nrexlikepp}
Riyopuno ûli kep pulwolo. ûli kep kupulwolapno yena lo yilauyo. Löi löi roise lo.\\
\gll riyopuno	ûli	kep	pulwolo	ûli	kep	k-pulwole-p-no	yena	lo	yilau-yo	löi\_löi	roise	lo\\
     then	gall.bladder	3\textsc{sg}.\textsc{poss}	burst.\textsc{pp}	gall.bladder	3\textsc{sg}.\textsc{poss}	\textsc{sub}-burst-\textsc{pc}-\textsc{co}	people	go.\textsc{pp}	village-\textsc{loc}	löi.löi	together	go.\textsc{pp}\\
\glt `Then (the bush spirit's) gall bladder burst. When it had burst the people went to the village. They went with shouts of victory.'
\z

\newpage
\ea\label{ex:nrexsepaye}
Yilauyo sepa, yena nakap.\\
\gll yilau-yo	sepa	yena	nake-p\\
     village-\textsc{loc}	arrive	people	live-\textsc{pc}\\
\glt `They arrived at the village and the people lived on.'
\z

\ea\label{ex:nrexbokusepaye}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `End of the story.'
\z
