\ea\label{ex:nrexdiriew}
Diri ewe weliyeulipop. Bi ki kama niulipop; ar namepop. Pili solo nameulipop. Dû ki kama niulipop; ewe ar namepop.\\
\gll diri	ewe	weliye-uli-pi-p	bi	ki	kama	ni-uli-pi-p	ar	name-pi-p	pili	solo	name-uli-pi-p	dû	ki	kama	ni-uli-pi-p	ewe	ar	name-pi-p\\
     younger.brother	older.brother	take.away.from-\textsc{prog}-\textsc{lv}-\textsc{pc}	animal	\textsc{aph}	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	\textsc{neg}	give.3\textsc{sg}.\textsc{or}-\textsc{lv}-\textsc{pc}	skin	only	give.3\textsc{sg}.\textsc{or}-\textsc{prog}-\textsc{lv}-\textsc{pc}	meat	\textsc{aph}	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	older.brother	\textsc{neg}	give.3\textsc{sg}.\textsc{or}-\textsc{lv}-\textsc{pc}\\
\glt `The younger brother always withheld from the older brother. He used to eat the (killed) animals alone; he did not give him anything. He usually gave him only the skin. He ate the meat always alone. He didn't give the older brother any (of it).'
\z

\ea\label{ex:nrexminuim}
Minui, mi riyo le duyo, bi lui, wepule. Mi ki kama ni ruri akono roise, iki kama. Ewe bi pili solo ponamo.\\
\gll mi-nui	mi	ri-yo	le	du-yo	bi	lui	wepule	mi	ki	kama	ni	ruri	ako-no	roise	iki	kama	ewe	bi	pili	solo	ponamo\\
     \textsc{iter}-sleep	again	\textsc{dist}-\textsc{loc}	go	bush-\textsc{loc}	animal	shoot	bring	again	\textsc{aph}	alone	eat	child	wife-\textsc{ins}	with	\textsc{aph}.\textsc{pl}	alone	older.brother	animal	skin	only	give.3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `He sleeps, he goes there to the bush, shoots animals, and brings them (home). He eats them alone together with his children and his wife, they alone. To the older brother he gave only the skin of the animals.'
\z

\ea\label{ex:nrexnekrur}
Ewe umul nek, ``Ruri kopi wo mopi.'' Diri muelino, ``Wor ko uliyewaip, wor dupua ko ukeli duyo.'' ``Wor de ukelip!''\\
\gll ewe	umul\_nek	ruri	ko-pi	wo\_mopi	diri	mueli-no	wor	ko	uliyewai-p	wor	dupua	ko	ukeli	du-yo	wor	de	ukeli-p\\
     older.brother	think.\textsc{pp}	child	1\textsc{sg}-\textsc{poss}	cry	younger.brother	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	dog	I	leave.for-\textsc{imp}	dog	two	I	take.with.oneself	bush-\textsc{loc}	dog	you	take.with.oneself-\textsc{imp}\\
\glt `The older brother thought: ``My children cry.'' He said to the younger brother: ``Leave  me the dogs! I take two dogs with (me) to the bush.''  (The younger brother replied): ``Take the dogs with (you)!'''
\z

\ea\label{ex:nrexeliwor}
Duyo ukeli. Wor bike nopo. Bi lu, bike lu. Puyo layo, aipo po; milo.\\
\gll du-yo	ukeli	wor	bike	nopo	bi	lu	bike	lu	pu-yo	layo	aipo	po	mi-lo\\
     bush-\textsc{loc}	take.with.oneself	dog	cassowary	bark.loudly.\textsc{pp}	pig	shoot.\textsc{pp}	cassowary	shoot.\textsc{pp}	river-\textsc{loc}	put.\textsc{pp}	taboo.sign	make.\textsc{pp}	\textsc{iter}-go.\textsc{pp}\\
\glt `He took (them) to the bush. The dogs barked (and startled) the cassowaries. He shot a pig, and he shot a cassowary. He put them into the river (in order to cool the dead animals) and set a taboo sign. (Then) he went again (to hunt).'
\z

\ea\label{ex:nrexieplub}
Biep lu. Biep kuluipno puyo layo, aipo pewo; milo.\\
\gll biep	lu	biep	k-lui-p-no	pu-yo	layo	aipo	pewo	mi-lo\\
     boar	shoot.\textsc{pp}	boar	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	river-\textsc{loc}	put.\textsc{pp}	taboo.sign	put.upright.\textsc{pp}	\textsc{iter}-go.\textsc{pp}\\
\glt `He shot a boar. When he had shot the boar, he put it into the river and set a taboo sign. (Then) he went again.'
\z

\ea\label{ex:nrexnambat}
Bike lu; wor bike nopo. Kuluipno puyo layo, aipo namba tri pewo. Milo wor ukeli.\\
\gll bike	lu	wor	bike	nopo	k-lui-p-no	pu-yo	layo	aipo	namba\_tri	pewo	mi-lo	wor	ukeli\\
     cassowary	shoot.\textsc{pp}	dog	cassowary	bark.loudly.\textsc{pp}	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	river-\textsc{loc}	put.\textsc{pp}	taboo.sign	third	put.upright.\textsc{pp}	\textsc{iter}-go.\textsc{pp}	dog	take.with.oneself\\
\glt `He shot a cassowary. The dogs (had) barked (and startled) the cassowaries. After shooting one,  he put it into the river and set the third taboo sign. He went off again and took the dogs with (him).'
\z

\ea\label{ex:nrexsemnop}
Bisem nopo. Wor dupua bisem wiyoi, paliya. Puyo layo, aipo pewo; milo.\\
\gll bisem	nopo	wor	dupua	bisem	wiyo-i	paliya	pu-yo	layo	aipo	pewo	mi-lo\\
     groundliving.kangaroo	bark.loudly.\textsc{pp}	dog	two	groundliving.kangaroo	catch.\textsc{pp}-\textsc{du}.\textsc{a}	be.dead	river-\textsc{loc}	put.\textsc{pp}	taboo.sign	put.upright	\textsc{iter}-go.\textsc{pp}\\
\glt `(The dogs) bayed for kangaroos. The two dogs caught a kangaroo; it is dead. He put it into the river and set a taboo sign. (Then) he went on.'
\z

\ea\label{ex:nrexiwiluw}
Biwi lu; wor nopo. Puyo layo, aipo pewo. Miloro.\\
\gll biwi	lu	wor	nopo	pu-yo	layo	aipo	pewo	mi-lo=ro\\
     sow	shoot.\textsc{pp}	dog	bark.loudly.\textsc{pp}	river-\textsc{loc}	put.\textsc{pp}	taboo.sign	put.upright.\textsc{pp}	\textsc{iter}-go.\textsc{pp}=\textsc{emph}\\
\glt `He shot a sow. The dogs (had) bayed (for animals). He put it into the river and set a taboo sign. Off he went again.'
\z

\ea\label{ex:nrexpoehur}
Wor urual nopo. Urual niniyo nuip. Ai epul malo: auna lo, auna loro. Dob po, ``Eh, urual nake!''\\
\gll wor	urual	nopo	urual	nini-yo	nui-p	ai	epul	malo	auna	lo	auna	lo=ro	dob	po	eh	urual	nake\\
     dog	big.lizard	bark.loudly.\textsc{pp}	big.lizard	sun-\textsc{loc}	sleep-\textsc{pc}	father	ear	hear.\textsc{pp}	slowly	go.\textsc{pp}	slowly	go.\textsc{pp}=\textsc{emph}	eye	do.\textsc{pp}	eh	big.lizard	sit\\
\glt `The dogs bayed for big lizards. A big lizard was sleeping in the sun. The father listened (for noises). He went slowly. He approached more slowly and looked: ``Eh, (there) is a big lizard.'''
\z

\ea\label{ex:nrexmaroip}
Ai rimaro ipulka weliana. Pe pulapu; urual dopyo papiko polip. Dop ilewo, pe ba pulapu dopyo.\\
\gll ai	ri\_maro	ipul-ka	weliana	pe	pulapu	urual	dop-yo	papiko	poli-p	dop\_ilewo	pe	ba	pulapu	dop-yo\\
     father	ironwood	buttress-\textsc{path}	hide.oneself.thither.\textsc{pp}	arrow	release.\textsc{pp}	big.lizard	body-\textsc{loc}	hit.\textsc{pp}	be.there-\textsc{pc}	hold.one's.body.\textsc{pp}	arrow	other	release.\textsc{pp}	body-\textsc{loc}\\
\glt `The father hid in the buttress of an ironwood tree. He released an arrow. It stuck in the lizard's body and was stuck. (The lizard) turned (its) body towards (the father).  He released another arrow (aiming) at (the lizard's) body.'
\z

\ea\label{ex:nrexkpiowo}
Urual kipiowopno ono wiyo. Ri roise mo, ono roise suelowe. Dop kep opi kep soso. Lu pulapi ri roise suel. Elyo suelowe, paliya.\\
\gll urual	k-piowo-p-no	ono	wiyo	ri	roise	mo	ono	roise	suelo-we	dop	kep	opi	kep	so-so	lu	pulapi	ri	roise	suel	el-yo	suelo-we	paliya\\
     big.lizard	\textsc{sub}-jump-\textsc{pc}-\textsc{co}	man	hold.\textsc{pp}	tree	together	fell.\textsc{pp}	man	together	cut.\textsc{pp}-\textsc{du}.\textsc{o}	body	3\textsc{sg}.\textsc{poss}	tail	3\textsc{sg}.\textsc{poss}	saw-\textsc{sim}	tooth	strain	tree	together	cut.\textsc{pp}	belly-\textsc{loc}	cut.\textsc{pp}-\textsc{du}.\textsc{o}	be.dead\\
\glt `With a jump the lizard got hold of the man. Together with the tree it threw (the man) over. It cut (the tree) apart together with the man. (The lizard's) body (and) its tail are like a saw. It aimed its teeth. It cut him together with the tree. (The lizard) cut (the man) in two at his belly; he is dead.'
\z

\ea\label{ex:nrexsekeai}
Riyopuno puyo wel. Boyo wor lil puaku kepyo ppaliyo. Wor ba seke suel, seke aipi luyo pana, wel yipyo yilauyo.\\
\gll riyopuno	pu-yo	wel	boyo	wor	lil	puaku	kep-yo	ppaliyo	wor	ba	seke	suel	seke	ai-pi	lu-yo	pana	wel	yip-yo	yilau-yo\\
     then	water-\textsc{loc}	carry.\textsc{pp}	later	dog	blood	head	3\textsc{sg}.\textsc{poss}-\textsc{loc}	rub.\textsc{pp}	dog	other	hair	cut.\textsc{pp}	hair	father-\textsc{poss}	tooth-\textsc{loc}	put.thither.\textsc{pp}	carry.\textsc{pp}	house-\textsc{loc}	village-\textsc{loc}\\
\glt `Then (the bush spirit) carried (the dead man) to the water. Later it rubbed blood on the head of (one of) the dogs.  It cut (off) some hair for the other dog and put the father's hair between its teeth. (The dog) brought it to the house in the village.'
\z

\ea\label{ex:nrexlilppa}
Wor lil ppaliyo kaeau kepyo, wel epeyo. Ba seke wel, ba lil wel.\\
\gll wor	lil	ppaliyo	kaeau	kep-yo	wel	epe-yo	ba	seke	wel	ba	lil	wel\\
     dog	blood	rub.\textsc{pp}	face	3\textsc{sg}.\textsc{poss}-\textsc{loc}	carry.\textsc{pp}	mother-\textsc{loc}	other	hair	carry.\textsc{pp}	other	blood	carry.\textsc{pp}\\
\glt `(The spirit) rubbed blood on the dog, on its face. (The dog) carried it to the mothers. One carried the (father's) hair, and the other one carried the (father's) blood.'
\z

\ea\label{ex:nrexakueye}
Puaku eye roise wel puyo wekûno. Bou dupua roise wekûno, puyo melkûno. Yilau kepyo, yip kepyo wel.\\
\gll puaku	eye	roise	wel	pu-yo	wekûno	bou	dupua	roise	wekûno	pu-yo	melkûno	yilau	kep-yo	yip	kep-yo	wel\\
     head	arm	together	carry.\textsc{pp}	water-\textsc{loc}	carry.down.\textsc{pp}	leg	two	together	carry.down.\textsc{pp}	water-\textsc{loc}	carry.down.\textsc{pl}.\textsc{o}.\textsc{pp}	place	3\textsc{sg}.\textsc{poss}-\textsc{loc}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	carry.\textsc{pp}\\
\glt `(The bush spirit) took head and arms and carried them down to the water. It carried the two legs down to the water. It carried a lot (of human flesh) to its place. It carried it to its house.'
\z

\ea\label{ex:nrexpperpu}
Pper pu epiyo pewo, riyo sepauwo.\\
\gll pper	pu	epiyo	pewo	ri-yo	sepauwo\\
     penis.cover	water	beside	put.upright.\textsc{pp}	\textsc{dist}-\textsc{loc}	fasten.\textsc{pp}\\
\glt `(The bush spirit) put (the man's) penis cover upright beside the water. It fastened it there (on a stick).'
\z

\ea\label{ex:nrexsukeik}
Wor yilauyo iloi. Epe dupua leip. Riyopuno sukei kep burilo yipyo.\\
\gll wor	yilau-yo	i-lo-i	epe	dupua	le-i-p	riyopuno	sukei	kep	burilo	yip-yo\\
     dog	village-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	mother	two	wait-\textsc{du}.\textsc{a}-\textsc{pc}	then	spirit	3\textsc{sg}.\textsc{poss}	go.ahead.\textsc{pp}	house-\textsc{loc}\\
\glt `The dogs went to the village, (where) the two mothers were waiting. Then (the dead man's) spirit went ahead to the house.'
\z


\newpage
\ea\label{ex:nrexalkaun}
Woloyo nekip, ``Ah, al kauna dor kopi, al kauna.'' Yip wolo ppuap, biliyo nekip.\\
\gll wolo-yo	neki-p	ah	al	kauna	dor	ko-pi	al	kauna	yip	wolo	ppue-p	bili-yo	neki-p\\
     ladder-\textsc{loc}	stand-\textsc{pc}	ah	leech	in.large.amounts	foot	1\textsc{sg}-\textsc{poss}	leech	in.large.amounts	house	ladder	go.up-\textsc{pc}	opening-\textsc{loc}	stand-\textsc{pc}\\
\glt `He was standing at the ladder: ``Ah, so many leeches on my feet, countless leeches.''  He climbed the ladder of the house and was standing at the door.'
\z

\ea\label{ex:nrexborook}
Ako dupua smep paliyei: ``Ari, aska. Oke boro? Oke yala sukei kep.''\\
\gll ako	dupua	smep	paliye-i	ari	aska	o-ke	bo=ro	o-ke	yala	sukei	kep\\
     wife	two	door	open-\textsc{du}.\textsc{a}	no	nothing	\textsc{prox}-\textsc{aph}	what=\textsc{emph}	\textsc{prox}-\textsc{aph}	now	spirit	3\textsc{sg}.\textsc{poss}\\
\glt `The two wives open the door: ``No, nothing. (But) what is this? This is (the man's) spirit.'''
\z

\ea\label{ex:nrexulpuso}
Umul puso yopip, umul nekanepoip.\\
\gll umul	pu-so	yopi-p	umul\_nekane-pi-i-p\\
     heart	water-\textsc{sim}	flood-\textsc{pc}	focus.thoughts-\textsc{lv}-\textsc{du}.\textsc{s}-\textsc{pc}\\
\glt `Their hearts were flooding like water. They focused (their) thoughts on (what could have happened).'
\z

\ea\label{ex:nrexsmeppa}
Wor am ar pulupi. Riyopuno wor dupua ipulo yipyo. Epe epul maloip, smep paliyoip.\\
\gll wor	am	ar	pulupi	riyopuno	wor	dupua	i-pulo	yip-yo	epe	epul	male-i-p	smep	paliyo-i\\
     dog	still	\textsc{neg}	come.\textsc{pl}	then	dog	two	\textsc{du}.\textsc{s}-come.\textsc{pp}	house-\textsc{loc}	mother	ear	hear-\textsc{du}.\textsc{a}-\textsc{pc}	door	open.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `The dogs don't come yet. Finally the two dogs came to the house. The mothers were listening and opened the door.'
\z

\ea\label{ex:nrexaelssu}
Ael sûno layei; beri sû beri po. Epe wor besi riyepowepuen: Wor ba lili ppaliyo, wor ba seke wepulo.\\
\gll ael	sû-no	laye-i	beri	sû	beri	po	epe	wor	besi	riye-pi-wepi-en	wor	ba	lil	ppaliyo	wor	ba	seke	wepulo\\
     chip.of.wood	fire-\textsc{ins}	put-\textsc{du}.\textsc{a}	burn	fire	burn	\textsc{lv}.\textsc{pp}	mother	dog	mouth.of.animals	see.\textsc{o}.\textsc{inanim}-\textsc{lv}-\textsc{quant}.\textsc{o}-\textsc{nsg}.\textsc{or}.\textsc{pp}	dog	other	blood	rub.\textsc{pp}	dog	other	hair	bring.\textsc{pp}\\
\glt `They put fire on a chip of wood, it burns, the flame burned. The mothers gazed at the dogs' mouths (with the bad indications). One dog has blood rubbed on it, and the other dog brought hair.'
\z

\ea\label{ex:nrexkaeaun}
Epe dupua riyepoi: Wor besino kaeauno luno riyepoi. Seke luyo polip, lil kaeauyo polip.\\
\gll epe	dupua	riye-po-i	wor	besi-no	kaeau-no	lu-no	riye-po-i	seke	lu-yo	poli-p	lil	kaeau-yo	poli-p\\
     mother	two	see.\textsc{o}.\textsc{inanim}-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}	dog	mouth.of.animals-\textsc{ins}	face-\textsc{ins}	tooth-\textsc{ins}	see.\textsc{o}.\textsc{inanim}-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}	hair	tooth-\textsc{loc}	be.there-\textsc{pc}	blood	face-\textsc{loc}	be.there-\textsc{pc}\\
\glt `The two mothers saw that. They saw the dogs with their mouths and faces and teeth. Hair was between the teeth (of one dog). Blood was on the face (of the other dog).'
\z

\ea\label{ex:nrexkunboo}
Epe wo mopi. Diri yipyo kûno, ``Bo?''\\
\gll epe	wo\_mop-i	diri	yip-yo	kûno	bo\\
     mother	weep.\textsc{pp}-\textsc{du}.\textsc{s}	younger.brother	house-\textsc{loc}	go.down.\textsc{pp}	what\\
\glt `The mothers wept. The younger brother came down to the house: ``What's the matter?'''
\z

\ea\label{ex:nrexlibika}
``De eli bi kama niulipop. De ar namopop bi dû. Uke pili solo ileulipop.''\\
\gll de\_eli	bi	kama	ni-uli-pi-p	de	ar	namo-pi-p	bi	dû	uke	pili	solo	ile-uli-pi-p\\
     you.yourself	meat	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	you	\textsc{neg}	give.3\textsc{sg}.\textsc{or}-\textsc{lv}-\textsc{pc}	animal	meat	we.\textsc{excl}	skin	only	eat.\textsc{pl}.\textsc{a}-\textsc{prog}-\textsc{lv}-\textsc{pc}\\
\glt `You alone used to eat the meat. You didn't give him any meat of the animals. We used to eat only the skin.'
\z

\ea\label{ex:nrexwenomo}
Diri epul kamalapno wo mop, ewe wo mop.\\
\gll diri	epul	k-male-p-no	wo\_mop	ewe	wo\_mop\\
     younger.brother	ear	\textsc{sub}-hear-\textsc{pc}-\textsc{co}	weep.\textsc{pp}	older.brother	weep.\textsc{pp}\\
\glt `When the younger brother had heard (that), he wept. He wept for his older brother.'
\z

\newpage
\ea\label{ex:nrexuleewe}
Riyopuno ya piyo, uroyo ule, ewe lipelip. Lo. Pu riyepulo, pu epi mono lo.\\
\gll riyopuno	ya	piyo	uro-yo	ule	ewe	lipeli-p	lo	pu	riye\_pulo	pu	epi\_mono	lo\\
     then	sago	take.\textsc{pp}	netbag-\textsc{loc}	put.inside.\textsc{pp}	older.brother	seek-\textsc{pc}	go.\textsc{pp}	river	see.\textsc{o}.\textsc{inanim}\_come.\textsc{pp}	river	side.path	go.\textsc{pp}\\
\glt `Then he took sago and put it into a netbag. He was searching for his brother. He left.  Seeing the river he came (near) and walked along the edge of the river.'
\z

\ea\label{ex:nrexulilim}
Dob po: Bike ulili. Milo: Biep ulili. Milo: Bike ba ulili. Milo: Bisem ulili. Milo: Biwi el roise ulili.\\
\gll dob	po	bike	u-lili	mi-lo	biep	u-lili	mi-lo	bike	ba	u-lili	mi-lo	bisem	u-lili	mi-lo	biwi	el	roise	u-lili\\
     eye	\textsc{lv}.\textsc{pp}	cassowary	\textsc{dfac}-be.there	\textsc{iter}-go.\textsc{pp}	boar	\textsc{dfac}-be.there	\textsc{iter}-go.\textsc{pp}	cassowary	other	\textsc{dfac}-be.there	\textsc{iter}-go.\textsc{pp}	kangaroo	\textsc{dfac}-be.there	\textsc{iter}-go.\textsc{pp}	sow	belly	with	\textsc{dfac}-be.there\\
\glt `He looked: Here is a cassowary. He went on: Here is a boar. He went on: Here is another cassowary. He went on: Here is a kangaroo. He went on: Here is a pregnant sow.'
\z

\ea\label{ex:nrexopebap}
Dob powolo, ``Ri ope ba po? Bo suelo?'' Dob sopop, ``Lil kauna walpue. Ewe kopi wiyo, puyo wekûno.''\\
\gll dob	po-wolo	ri	o=pe	ba	po	bo	suelo	dob\_sopi-p	lil	kauna	walpue	ewe	ko-pi	wiyo	pu-yo	wekûno	wo\_mop\\
     eye	\textsc{lv}-go.further.\textsc{pp}	tree	\textsc{prox}=\textsc{q}	what	do.\textsc{pp}	what	cut.\textsc{pp}	realise-\textsc{pc}	blood	in.large.amounts	spread.about	older.brother	1\textsc{sg}-\textsc{poss}	hold.\textsc{pp}	water-\textsc{loc}	carry.down.\textsc{pp}	weep.\textsc{pp}\\
\glt `He looked around: ``This tree here, what happened to it? What cut it through?'' He realised: blood is spread around everywhere. (A bush spirit) took hold of the older brother and carried (the corpse) down to the water. He burst into tears.'
\z

\ea\label{ex:nrexsepauw}
Wo mop. Wo komopopno dob riye: Pper sepauwo, ulili. Yilauyo lo, woise lo.\\
\gll wo\_k-mopi-p-no	dob	riye	pper	sepauwo	u-lili	yilau-yo	lo	woise	lo\\
     tears\_\textsc{sub}-weep-\textsc{pc}-\textsc{co}	eye	see.\textsc{o}.\textsc{nanim}	penis.cover	fasten.\textsc{pp}	\textsc{dfac}-be.there	village-\textsc{loc}	go.\textsc{pp}	with.tears	go.\textsc{pp}\\
\glt `He was weeping. (Then) he sees (something else): (The bush spirit) fastened the penis cover (on a stick). Here it is! He left (the place) for the village; he went crying.'
\z

\ea\label{ex:nrexoisero}
Ako dupua epul maloi: ``Uki koyopi basuiko.'' Diri woise roki pule. Pulo yipyo, paeau, kiyo muelien:\\
\gll ako	dupua	epul	malo-i	uki	koyo-pi	ba-sui-ko	diri	woise	ro-ki	pule	pulo	yip-yo	paeau	kiyo	mueli-en\\
     wife	two	ear	hear.\textsc{pp}-\textsc{du}.\textsc{a}	husband	we.\textsc{du}.\textsc{excl}-\textsc{poss}	\textsc{fac}-die-\textsc{fac}	younger.brother	with.tears	\textsc{prox}.\textsc{emph}-\textsc{aph}	come	come.\textsc{pp}	house-\textsc{loc}	arrive	\textsc{aph}.\textsc{du}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}\\
\glt `The two wives heard (the news): Our husband has died. The younger brother is coming here crying. He came to the house, reached it, and said to them.'
\z

\ea\label{ex:nrexolekiu}
``Urual noro. Sukupu noro, puyo wekûno. Nuko mole.'' Kiu ukeli, moloro liki kepyo, paeau.\\
\gll urual	no=ro	sukupu	no=ro	pu-yo	wekûno	nuko	mole	kiu	ukeli	molo=ro	liki	kep-yo	paeau\\
     big.lizard	eat.\textsc{pp}=\textsc{emph}	bush.spirit	eat=\textsc{emph}	river-\textsc{loc}	carry.down.\textsc{pp}	we.\textsc{incl}	go.\textsc{pl}	clan	take.with.oneself	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	designated.place	3\textsc{sg}.\textsc{poss}-\textsc{loc}	arrive\\
\glt `{``}A big lizard ate (him). A bush spirit ate (him) and carried (him) down to the river.  Let's go.'' He took the clan with him. They went to the (bush spirit's) place and arrived (there).'
\z

\ea\label{ex:nrexpuawom}
Diri ako ewe ako, dupua womopip yena ba roise. Sukupu urual epul malo, ``Yena wo mopi.''\\
\gll diri	ako	ewe	ako	dupua	wo\_mopi-p	yena	ba	roise	sukupu	urual	epul	malo	yena	wo\_mopi\\
     younger.brother	wife	older.brother	wife	two	weep-\textsc{pc}	people	other	together	bush.spirit	big.lizard	ear	hear.\textsc{pp}	people	weep\\
\glt `The younger and older brothers' wives were weeping together with the other people. The bush spirit lizard heard it: ``People are weeping.'''
\z

\newpage
\ea\label{ex:nrexpulebi}
Riyopuno bili paliyo. Yena epul malo, ``Ere pule.'' Bili bapaliyoko. ``Ine pu epiyo poyewap! Yala pinem.''\\
\gll riyopuno	bili	paliyo	yena	epul	malo	ere	pule	bili	ba-paliye-ko	ine	pu	epiyo	poye-we-p	yala	pini-m\\
     then	opening	open.\textsc{pp}	people	ear	hear.\textsc{pp}	here	come	opening	\textsc{fac}-open-\textsc{fac}	you.\textsc{pl}	water	beside	stand.\textsc{pl}-\textsc{ter}-\textsc{imp}	now	come.up.hither-\textsc{pos}\\
\glt `Then it opened the door. The people heard it: ``Here it comes.'' The door is open:  ``You all stand in line beside the water! It will come up now.'''
\z

\ea\label{ex:nrexaliyep}
Smep ba paliyepana; bili ba bapaliyepanoko. ``Ere pule! Ine pe pakono powap!''\\
\gll smep	ba	paliyepana	bili	ba	ba-paliyepane-ko	ere	pule	ine	pe	pako-no	pi-we-p\\
     door	other	open.wide.\textsc{pp}	opening	other	\textsc{fac}-open.wide-\textsc{fac}	here	come	you.\textsc{pl}	arrow	bow-\textsc{ins}	do-\textsc{ter}-\textsc{imp}\\
\glt `(The bush spirit) opened another door. The other door is wide open:  ``Here it comes! You all take bows and arrows.'''
\z

\ea\label{ex:nrexwaplay}
Smep namba tri paliyo, bapaliyoko. Urual olyo lelwolo pu epiyo. Riyopuno yena wapilayekûpu. Urual paliya.\\
\gll smep	namba\_tri	paliyo	ba-paliye-ko	urual	ol-yo	liliwolo	pu	epiyo	riyopuno	yena	
wapilayekûpu	urual	paliya\\
     door	third(\textsc{tp})	open.\textsc{pp}	\textsc{fac}-open-\textsc{fac}	big.lizard	river.bank-\textsc{loc}	be.there.suddenly.\textsc{pp}	water	beside	then	people	shoot.down.hither.with.arrows.\textsc{pl}.\textsc{a}.\textsc{pp}	big.lizard	be.dead\\
\glt `(The spirit) opened the third door. It is open. Suddenly the big lizard was on the river bank. Then the people shot it down with arrows. The big lizard is dead.'
\z

\ea\label{ex:nrexomosop}
Yena puyo kûno, puyo kepue. Yip kepyo paeau. Yena le kep rapiyo, rapiyemayo: Neppi nana dawa ipi, wîl sawono, spet, dipsu walno; oil kerosin petrol. Ipipis leki, yipp lekiro roise. Lam wik blanket siket taul sop omo. Kiniyo.\\
\gll yena	pu-yo	kûno	pu-yo	kepue	yip	kep-yo	paeau	yena	le	kep	rapiyo	rapiye-mayo	neppi	nana	dawa	ipi	wîl	sawo-no	spet	dipsu	wal-no	oil	kerosin	petrol	ipipis	leki	yipp	leki=ro	roise	lam	wik	blanket	siket	taul	sop	omo	kiniyo\\
     people	water-\textsc{loc}	go.down.\textsc{pp}	water-\textsc{loc}	pass.by	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	arrive	people	things	3\textsc{sg}.\textsc{poss}	fetch.\textsc{pp}	fetch.\textsc{pp}-\textsc{mal}.\textsc{pp}	bush.knife	small.knife	axe	pot	dish	scoop-\textsc{ins}	spate(\textsc{tp})	rice	fish-\textsc{ins}	oil	kerosin	petrol	wild.duck	fetch.\textsc{pl}.\textsc{o}	wildfowl	fetch.\textsc{pl}.\textsc{o}=\textsc{emph}	together	lamp	wick	blanket	skirt	towel	soap	detergent	all\\
\glt `The people went down to the water. They are passing by the water and reach its house. They took its belongings, took them away: the bush knife, the small knife, the axe, the pots, the dishes with scoops, the spate, rice and fish, the oil, the petrol; they fetched the wild ducks and the wildfowls; the lamp and the wicks, the blankets, skirts and towels, soap and detergent, everything.'
\z

\ea\label{ex:nrexbakaak}
Yena meli, le meli. Ako dupua le royeen. Baka iki meli, baka ako dupua royeen.\\
\gll yena	meli	le	meli	ako	dupua	le	roye-en	baka	iki	meli	baka	ako	dupua	roye-en\\
     people	carry.\textsc{pl}.\textsc{o}	things	carry.\textsc{pl}.\textsc{o}	wife	two	things	give-\textsc{nsg}.\textsc{or}.\textsc{pp}	half	\textsc{aph}.\textsc{pl}	carry.\textsc{pl}.\textsc{o}	half	wife	two	give-\textsc{nsg}.\textsc{or}.\textsc{pp}\\
\glt `The people carry a lot. They carry (its) belongings. They gave (many of) the belongings to the two wives. They carry one half. The other half they gave to the two wives.'
\z

\ea\label{ex:nrexwepusu}
Yena ri purapuwepu, sû mappo. Sû rino mappo, beri. Urual sûyo royepepo, sûro.\\
\gll yena	ri	purapi-wepu	sû\_mappo	sû	ri-no	mappo	beri	urual	sû-yo	royepepo	sû=ro\\
     people	wood	cut.wood-\textsc{quant}.\textsc{o}.\textsc{pp}	light.a.fire.\textsc{pp}	fire	wood-\textsc{ins}	light.\textsc{pp}	burn	big.lizard	fire-\textsc{loc}	lay.on.top.\textsc{pp}	fire=\textsc{emph}\\
\glt `The people cut a lot of wood. They lit a fire, lit the fire with firewood. It is burning. They put the big lizard in the fire. The fire (blazes).'
\z


\ea\label{ex:nrexlebaul}
Yena yilauyo molo. Le sukupupi meli. Sukupu royekoro, le ba uleko, sukupupi. Sukupu urual le roise molo yilauyo.\\
\gll yena	yilau-yo	molo	le	sukupu-pi	mel	sukupu	roye-ko=ro	le	ba	ule-ko	sukupu-pi	sukupu	urual	le	roise	molo	yilau-yo\\
     people	village-\textsc{loc}	go.\textsc{pl}.\textsc{pp}	things	bush.spirit-\textsc{poss}	carry.\textsc{pl}.\textsc{o}.\textsc{pp}	bush.spirit	lay-\textsc{rts}=\textsc{emph}	things	other	be.there.\textsc{pl}-\textsc{rts}	bush.spirit-\textsc{poss}	bush.spirit	big.lizard	things	together	go.\textsc{pl}.\textsc{pp}	village-\textsc{loc}\\
\glt `The people went to the village. They carried the belongings of the bush spirit. They had put the bush spirit (in the fire), and a few things of the bush spirit were (in the fire, too). They went to the village (loaded) with the belongings of the bush spirit lizard.'
\z

\ea\label{ex:nrexuliboy}
Ûli boyo pulwolko. Yena epul malwolo, yilauyo paeau.\\
\gll ûli	boyo	pulwole-ko	yena	epul	male\_wolo	yilau-yo	paeau\\
     gall.bladder	later	burst-\textsc{rts}	people	ear	hear\_move.further.\textsc{pp}	village-\textsc{loc}	arrive\\
\glt `Later the gall bladder burst. The people heard (the bang) from a distance. They arrived in the village.'
\z

\ea\label{ex:nrexurukob}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `The story is finished.'
\z

\ea\label{ex:nrexsukrur}
Sukupu ba ppili. Sukupu ba akoyo rurino. Sukupu luo bîyo mape, bîyo mapepulupi. Liki sukupupi poli, luo bîyo poli. Sukupu dop sei sowemayo. Ari, leliewepi aska. Yala uke upuna mape. Sukupu basepoloko, basepolowepuko.\\
\gll sukupu	ba	ppili	sukupu	ba	ako-no	ruri-no	sukupu	luo	bî-yo	mape	bî-yo	mape\_pulupi	liki	sukupu-pi	poli	luo	bî-yo	poli	sukupu	dop	sei	sowe-mayo	ari	lelie-wepi	aska	yala	uke	upuna	mape	sukupu	ba-sepole-ko	ba-sepole-wepi-ko\\
     bush.spirit	other	live.single	bush.spirit	other	wife-\textsc{ins}	child-\textsc{ins}	bush.spirit	stone	hole-\textsc{loc}	live.\textsc{pl}	hole-\textsc{loc}	come.to.stay.\textsc{pl}	made.ready.place	bush.spirit-\textsc{poss}	be.there	stone	hole-\textsc{loc}	be.there	bush.spirit	skin	white	hide-\textsc{mal}.\textsc{pp}	no	kill.people-\textsc{quant}.\textsc{o}	none	no	we.\textsc{excl}	alright	live.\textsc{pl}	bush.spirit	\textsc{fac}-disappear-\textsc{fac}	\textsc{fac}-vanish-\textsc{quant}.\textsc{s}-\textsc{fac}\\
\glt `Some bush spirits live single, other bush spirits have wives and children. The bush  spirits live in stone caves. They come to stay in (many) caves. The ready made places  of the bush spirits are there, they are in the stone caves. The bush spirits hid from the whites. No, there is no killing any more. Now we live good. The bush spirits have disappeared, they have totally vanished.'
\z
