\ea\label{ex:18exwalp01}
Ako dari werino uki klokni piyoi. Dari el piamu, weri el aska, weri ppili.\\
\gll ako	dari	weri-no	uki	klokni	piyo-i	dari	el\_piamu	weri	el\_aska	weri	ppili\\
     wife	older.sister	younger.sister-\textsc{ins}	husband	one	take.\textsc{pp}-\textsc{du}.\textsc{a}	older.sister	get.pregnant.\textsc{pp}	younger.sister	not.pregnant	younger.sister	live.single\\
\glt `The women, two sisters, took one husband. The older sister got pregnant. The younger  sister was not pregnant, she lived single.'
\z

\ea\label{ex:18exwalp02}
Uki kiyopi kaikai weliyoulipop. Ki kama noulipop, bi wal ya solo noulipop.\\
\gll uki	kiyo-pi	kaikai	weliye-uli-pi-p	ki	kama	ni-uli-pi-p	bi	wal	ya	solo	ni-uli-pi-p\\
     husband	\textsc{aph}.\textsc{du}-\textsc{poss}	food(\textsc{tp})	take.away.from-\textsc{prog}-\textsc{lv}-\textsc{pc}	\textsc{aph}	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	meat	fish	sago	only	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}\\
\glt `Their husband usually withheld food (from them). He used to eat alone. Only he ate meat, fish, and sago.'
\z

\ea\label{ex:18exwalp03}
Ako dupua dari werino umul nekpamui: ``Nuko ile. Nuko pu riyepoi, pu ki îkori.''\\
\gll ako	dupua	dari	weri-no	umul\_nekpamu-i	nuko	i-le	nuko	pu	riye-pi-i	pu	ki	î-ko=ro\\
     wife	two	older.sister	younger.sister-\textsc{ins}	reflect.\textsc{pp}-\textsc{du}.\textsc{s}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	we.\textsc{incl}	river see.\textsc{o}.\textsc{inanim}-\textsc{lv}-\textsc{du}.\textsc{a}	river	\textsc{aph}	dry.up-\textsc{rts}=\textsc{emph}\\
\glt `The two wives, the sisters, reflected: ``Let's go look at the river. Did the river dry up?'''
\z

\ea\label{ex:18exwalp04}
Ilo. Pu baîko. Luika kililuipno saulno yopi.\\
\gll i-lo	pu	ba-î-ko	luika	k-lilu-i-p-no	saul-no	yopo-i\\
     \textsc{du}.\textsc{s}-go.\textsc{pp}	river	\textsc{fac}-dry.up-\textsc{fac}	downriver	\textsc{sub}-stake.off-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	scooping.container-\textsc{ins}	scoop.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `They went. The river has (almost) dried up. Having staked off (a small basin) downriver, they scooped it (empty) with a \textit{pangal}-shovel.'
\z

\newpage
\ea\label{ex:18exwalp05}
Pu î, yelo solo wal solo mapap, waeus wamo waeripi waesepp weiop. Bi riyei.\\
\gll pu	î	yelo	solo	wal	solo	mape-p	waeus	wamo	waeripi	waesepp	weiop	bi	riye-i\\
     water	dry.up	ground	only	fish	only	stay.\textsc{pl}-\textsc{pc}	shrimps	kind.of.fish	kind.of.fish	kind.of.fish	kind.of.fish	meat	see.\textsc{o}.\textsc{inanim}-\textsc{du}.\textsc{a}\\
\glt `The water dries up. Only the ground (of the river is visible). There were only fish (left), (namely) shrimps, \textit{wamo}-fish, \textit{weripi}-fish, \textit{waesopp}-fish, \textit{weiop}-fish. They see meat.'
\z

\ea\label{ex:18exwalp06}
Wal saul dupua layewoye. Dob bî riyei, walpop dor riyei, dor uppue.\\
\gll wal	saul	dupua	laye-we	dob	bî	riye-i	walpop	dor	riye-i	dor	u-ppue\\
     fish	scooping.container	two	put-\textsc{du}.\textsc{o}	eye	hole	see.\textsc{o}.\textsc{inanim}-\textsc{du}.\textsc{a}	turtle	foot	see.\textsc{o}.\textsc{inanim}-\textsc{du}.\textsc{a}	foot	\textsc{dfac}-go.up\\
\glt `They put two \textit{pangal}-containers (full) of fish (aside). They notice a hole. They see the footprints of a small turtle, right here the traces go straight up.'
\z

\ea\label{ex:18exwalp07}
Riyopuno weri dari muelno: ``De nakowap, de elno. Ko ppue.''  ``Ari, ko ike ppue.''\\
\gll riyopuno	weri	dari	mueli-no	de	nake-we-p	de	el\_no	ko	ppue	ari	ko\_ike	ppue\\
     then	younger.sister	older.sister	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	stay-\textsc{ter}-\textsc{imp}	you	pregnant	I	go.up	no	I.myself	go.up\\
\glt `Then the younger sister said to the older: ``You stay here, you are pregnant. I'll go up.'' (The older sister replied:)  ``No, I go up myself.'''
\z

\ea\label{ex:18exwalp08}
Dari loro sukupu yipyo paeau. Weri kleouro. Weri womop dupuni.\\
\gll dari	lo=ro	sukupu	yip-yo	paeau	weri	ki	le-ou=ro	weri	wo\_mop	dupuni\\
     older.sister	go.\textsc{pp}=\textsc{emph}	bush.spirit	house-\textsc{loc}	arrive	younger.sister	\textsc{aph}	wait-\textsc{frus}=\textsc{emph}	younger.sister	cry.\textsc{pp}	night\\
\glt `So the older sister went to the house of the bush spirit and arrived (there). The younger  sister waited for her in vain, and she cried until evening.'
\z

\ea\label{ex:18exwalp09}
Wal saul dupua puyo unepana, woise yipyo lo.\\
\gll wal	saul	dupua	pu-yo	uneipana-we	woise	yip-yo	lo\\
     fish	scooping.container	two	river-\textsc{loc}	tip.across.\textsc{pp}-\textsc{du}.\textsc{o}	with.tears	house-\textsc{loc}	go.\textsc{pp}\\
\glt `She tipped the two scooping containers across the river and went home crying.'
\z

\ea\label{ex:18exwalp10}
Uki muelno: ``De bi kama noulipop, dari kopi sepolo. Walpop bî luppue, yala sukupu noniri. Ko kleouro. De eli kama noulipop, bi de koyo ar ponienpop.''\\
\gll uki	mueli-no	de	bi	kama	ni-uli-pi-p	dari	ko-pi	sepolo	walpop	bî	luppue	yala	sukupu	no-ni=ro	ko	ke	le-ou=ro	de\_eli	kama	ni-uli-pi-p	bi	de	koyo	ar	ponini-pi-p\\
     husband	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	meat	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	older.sister	1\textsc{sg}-\textsc{poss}	vanish.\textsc{pp}	turtle	hole	go.up	now	bush.spirit	\textsc{aug}-eat=\textsc{emph}	I	\textsc{top}	wait-\textsc{frus}=\textsc{emph}	you.yourself	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	meat	you	we.\textsc{du}.\textsc{excl}	\textsc{neg}	give.\textsc{nsg}.\textsc{or}-\textsc{lv}-\textsc{pc}\\
\glt `She said to her husband: ``You ate the meat always alone. My sister has disappeared. She  went up into the cave of a small turtle. Now a bush spirit will eat her. I waited for her  in vain. You yourself would eat (everything) alone. You didn't give us any meat.'''\footnote{In this context \textit{bi} `meat' refers to protein-rich, valuable food, which the husband did not share with his wives.}
\z

\ea\label{ex:18exwalp11}
Uki womopip. ``De ke wo komopem! De kama noulpop!'' Sukei roki wapono.\\
\gll uki	wo\_mopi-p	de	ke	wo\_k-mopi-m	de	kama	ni-uli-pi-p	sukei	ro-ki	wape-no\\
     husband	cry-\textsc{pc}	you	\textsc{top}	crying\_\textsc{proh}-cry-\textsc{fac}	you	alone	eat-\textsc{prog}-\textsc{lv}-\textsc{pc}	spirit	\textsc{prox}.\textsc{emph}-\textsc{aph}	put.together-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `The husband was crying (and his wife said to him): ``Don't even cry! You alone ate. This bad mindset caused her (trouble).'''
\z

\ea\label{ex:18exwalp12}
Dari loro sukupu yipyo paeau. ``Depe arka pulo?''  ``Ko so ke walpopri!''  ``Walpop ari, kore! Awe, de epi baka nakap!''\\
\gll dari	lo=ro	sukupu	yip-yo	paeau	de=pe	arka	pulo	ko	so	ke	walpop=ro	walpop	ari	ko=ro	awe	de	epi	ba-ka	nake-p\\
     older.sister	go.\textsc{pp}=\textsc{emph}	bush.spirit	house-\textsc{loc}	arrive	you=\textsc{q}	where	come.\textsc{pp}	I	believe	\textsc{aph}	turtle=\textsc{emph}	turtle	no	I=\textsc{emph}	come.\textsc{imp}	you	side	other-\textsc{path}	live-\textsc{imp}\\
\glt `The older sister went to the house of the bush spirit and arrived (there): ``Where did  you come from?'' she was asked.  (The older sister replied:) ``I assumed, it (is) a small turtle.''  ``A small turtle, no, it's me, (the bush spirit). Come, live on the other side (of the house)!'''
\z

\ea\label{ex:18exwalp13}
Ako ruri nako roipi. Sukupu yipyo dob reyo. Wolo dupua poli, wolo ba sukupu kep wolo, ba akoro.\\
\gll ako	ruri	nako	roipi	sukupu	yip-yo	dob	reyo	wolo	dupua	poli	wolo	ba	sukupu	kep	wolo	ba	ako=ro\\
     wife	child	gave.birth	boy	bush.spirit	house-\textsc{loc}	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	ladder	two	be.there	ladder	other	bush.spirit	3\textsc{sg}.\textsc{poss}	ladder	other	wife=\textsc{emph}\\
\glt `The wife gave birth to a child, to a boy, in the bush spirit's house. (The bush spirit) saw her. There are two ladders (there), one (belongs to) the bush spirit('s room), the other one (belongs to) the woman('s room).'
\z

\ea\label{ex:18exwalp14}
Se seku, emi lupi suelo, mi ki no. Mi konopno ekuka seku, mi yeloyo seku.\\
\gll se	seku	emi\_lupi	suelo	mi	ki	no	mi	k-ni-p-no	eku-ka	seku	mi	yelo-yo	seku\\
     placenta	fall.down.\textsc{pp}	umbilical.cord	cut.\textsc{pp}	then	\textsc{aph}	eat.\textsc{pp}	then	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	anus-\textsc{path}	fall.down.\textsc{pp}	then	ground-\textsc{loc}	fall.down.\textsc{pp}\\
\glt `The placenta fell down. (The bush spirit) cut the umbilical cord and ate it readily. Then, when it had eaten it, it fell out of (its) anus and dropped on the ground.'
\z

\ea\label{ex:18exwalp15}
Akoro dob riye: ``Eh, oke sukupu! Ko so onori. Oke sukupu, ko yala koniyoni.''\\
\gll ako=ro	dob	reye	eh	o-ke	sukupu	ko	so	ono=ro	o-ke	sukupu	ko	yala	koniye\_ni\\
     wife=\textsc{emph}	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}	eh	\textsc{prox}-\textsc{aph}	bush.spirit	I	believe	man=\textsc{emph}	\textsc{prox}-\textsc{aph}	bush.spirit	I	soon	devour\\
\glt `The wife realises: ``Oh, this one is a bush spirit! I expected (the man was) human, (but) this  one is a bush spirit. Will it devour me soon?'''\footnote{When food comes out of a person's anus undigested, this is a sign that the person in fact is a bush spirit.}
\z

\ea\label{ex:18exwalp16}
Umul pole roise nakap: ``Yala ko asa pi?''\\
\gll umul\_pole	roise	nake-p	yala	ko	asa	pi\\
     be.sad	with	live-\textsc{pc}	now	I	how	do\\
\glt `She was living with a sad heart: ``What am I going to do now?'''
\z

\ea\label{ex:18exwalp17}
Nakapno duyo lo, ono lure, bi lure, melpulo yipyo.\\
\gll nake-p-no	du-yo	lo	ono	lu=ro	bi	lu=ro	melpulo	yip-yo\\
     live-\textsc{pc}-\textsc{co}	bush-\textsc{loc}	go.\textsc{pp}	person	shoot.\textsc{pp}=\textsc{emph}	animal	shoot.\textsc{pp}=\textsc{emph}	bring.\textsc{pl}.\textsc{o}.\textsc{pp}	house-\textsc{loc}\\
\glt `While she was living (like this), (the bush spirit) went to the bush, shot humans, shot animals, and  brought them to the house.'
\z

\ea\label{ex:18exwalp18}
Bi akoro ponamo. Onope ki no, ono dokomiyo.\\
\gll bi	ako=ro	ponamo	ono=pe	ki	no	ono	d-komiyo\\
     meat	wife=\textsc{emph}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	person=\textsc{q}	\textsc{aph}	eat.\textsc{pp}	person	\textsc{lkh}-hide.\textsc{pp}\\
\glt `It gave the meat (of the animals) to the woman. (But) a human being, did (the spirit) eat it? It surely hid a human being (when eating it).'
\z

\ea\label{ex:18exwalp19}
So poulipop. Ruri ireri, ruri puana, dorno le, ikoi.\\
\gll so	pi-uli-pi-p	ruri	ireri	ruri	puane	dor-no	le	ikoi\\
     as	do-\textsc{prog}-\textsc{lv}-\textsc{pc}	child	crawl	child	stand.up	foot-\textsc{ins}	go	big\\
\glt `(The spirit) was doing like this. (Meanwhile) the child crawls, the child stands up, walks on his feet, is big.'
\z

\ea\label{ex:18exwalp20}
Riyopuno sukupu umul nekpamu: ``Ko yesi wapke.'' Lo seleyo, sele numuelyo polip. Klapno yesi wapo.\\
\gll riyopuno	sukupu	umul\_nekpamu	ko	yesi	wapi-ke	lo	sele-yo	sele	numuelyo	poli-p	k-le-p-no	yesi	wapo\\
     then	bush.spirit	reflect.\textsc{pp}	I	aibika	collect-\textsc{ingr}	go.\textsc{pp}	garden-\textsc{loc}	garden	far.away	be.there-\textsc{pc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	aibika	collect.\textsc{pp}\\
\glt `Then the bush spirit decided: ``I go to collect \textit{aibika}-leaves. It went to  the garden. The garden was far away. Having gone, it collected \textit{aibika}-leaves.'
\z

\ea\label{ex:18exwalp21}
\largerpage[2]
Bipep puloro ako muelo: ``De awe nuko ile! Ereka sukupu yala de noni, de ruri roise. Nuko ile, awe! Ko de ar ni, ko de mekiye.''\\
\gll Bipep	pulo=ro	ako	mueli-no	de	awe	nuko	i-le	ere-ka	sukupu	yala	de	no-ni	de	ruri	roise	nuko	i-le	awe	ko	de	ar	ni	ko	de	mekiye\\
     Bipep	come.\textsc{pp}=\textsc{emph}	wife	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	come.\textsc{imp}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	\textsc{prox}-\textsc{path}	bush.spirit	soon	you	\textsc{aug}-eat	you	child	together	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	come.\textsc{imp}	I	you	\textsc{neg}	eat	I	you	help\\
\glt `(At once) Bipep came and said to the woman: ``Come, we'll leave! The bush spirit will eat you up soon, you together with your child. Let's go, come! I won't eat you up, I will help you.'''\footnote{The man called Bipep is introduced as the positive opponent of the bush spirit without mentioning him before.}
\z

\ea\label{ex:18exwalp22}
Dari ruri roise Bipep wulino. Bipep yipyo inakap: ``Yilau depi roki,'' mosupino.\\
\gll dari	ruri	roise	Bipep	wuli-no	Bipep	yip-yo	i-nake-p	yilau	de-pi	ro-ki	mosupi-no\\
     older.sister	child	together	Bipep	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}	Bipep	house-\textsc{loc}	\textsc{du}.\textsc{s}-live-\textsc{pc}	place	2\textsc{sg}-\textsc{poss}	\textsc{prox}.\textsc{emph}-\textsc{aph}	show-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `The older sister with (her) child followed Bipep, and they were living in Bipep's house.  ``Here is your place'', he showed it to her.'
\z

\ea\label{ex:18exwalp23}
Sukupu pulo, yesi yipyo wepulo. ``De nake?'' ``Ou ko unake,'' Muem muelno.\\
\gll sukupu	pulo	yesi	yip-yo	wepulo	de	nake	ou	ko	u-nake	Muem	mueli-no\\
     bush.spirit	come.\textsc{pp}	aibika	house-\textsc{loc}	bring.\textsc{pp}	you	stay	yes	I	\textsc{dfac}-stay	Muem	talk.to-3\textsc{sg}.\textsc{or}\\
\glt `The bush spirit came and brought the \textit{aibika}-leaves home: ``Are you there?''  ``Yes, I am here,'' said Muem, the White Ant, to it.'\footnote{Muem, the White Ant, behaves as a friendly spirit who helps people against evil bush spirits. Actually, \textit{white ant} is the name for termites.}
\z

\ea\label{ex:18exwalp24}
Bipep Muem muelo: ``De yala wapiyam.'' ``Ou, ko yala wapiyam.'' Muem yip bîyo nakap, Bipep Muem smep musono. ``Koyo ile, de yala nakowonam.'' ``Ou, deyo ilap.''\\
\gll Bipep	Muem	mueli-no	de	yala	wapiye-m	ou	ko	yala	wapiye-m	Muem	yip	bîyo	nake-p	Bipep	Muem	smep	musi-no	koyo	i-le	de	yala	nake-we-ne-m	ou	deyo	i-le-p\\
     Bipep	Muem	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	\textsc{mod}	deceive-\textsc{pos}	yes	I	\textsc{mod}	deceive-\textsc{pos}	Muem	house	inside	stay-\textsc{pc}	Bipep	Muem	door	lock-3\textsc{sg}.\textsc{or}.\textsc{pp}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-go	you	\textsc{mod}	stay-\textsc{ter}-3\textsc{sg}.\textsc{or}-\textsc{pos}	yes	you.\textsc{du}	\textsc{du}.\textsc{s}-go-\textsc{imp}\\
\glt `Bipep said to White Ant: ``You should deceive it.''  ``Yes, I will deceive  (it).''  White Ant was staying in the cave, and Bipep closed the door (behind) Muem:  ``We leave, you will stay here (and wait) for (the bush spirit).''  ``Yes, you two go (now).'''\footnote{According to the proper course of events, Sequence 24 should come before Sequence 23, but the narrator chose this order, which gives a postposed explanation for what is said in the preceding sequence.}
\z

\ea\label{ex:18exwalp25}
Rum kepyo lo, buar piyoko, wepulo: ``Yala Muem puluwei, ko wapiyap. Oh ako ari, balok.''\\
\gll rum	kep-yo	lo	buar	piye-ko	wepulo	yala	Muem	puluwei	ko	wapiye-p	oh	ako	ari	ba-le-ko\\
     room	3\textsc{sg}.\textsc{poss}-\textsc{loc}	go.\textsc{pp}	stone.axe	take-\textsc{rts}	bring.\textsc{pp}	now	white.ant	break.with.force	I	deceive-\textsc{pc}	oh	wife	no	\textsc{fac}-go-\textsc{fac}\\
\glt `(The bush spirit) went into its room, took the stone axe and brought it: ``Now I will break White Ant into pieces, it deceived me. Oh, the woman isn't (here). She has gone.'''
\z

\ea\label{ex:18exwalp26}
Dorka yek: ``Oh, oki woko Bipep yipyo.''\\
\gll dor-ka	yek	oh	o-ki	woko	Bipep	yip-yo\\
     foot-\textsc{path}	follow.one's.traces.\textsc{pp}	oh	\textsc{prox}-\textsc{aph}	go.together.with.\textsc{pp}	Bipep	house-\textsc{loc}\\
\glt `It followed their traces: ``Oh, here he went together (with her), to Bipep's house.'''
\z

\ea\label{ex:18exwalp27}
Sukupu Bipep sano: ``Deyo inake?'' ``Oh koyo pari!'' ``De dupuapi deyoru.'' ``Oh koyo pari!'' ``De ko wapiye, ko bareyeko, ere nake.''\\
\gll sukupu	Bipep	sa-no	deyo	i-nake	oh	koyo	pari	de	dupuapi	deyo=ro	oh	koyo	pari	de	ko	wapiye	ko	ba-reye-ko	ere	nake\\
     bush.spirit	Bipep	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	you.\textsc{du}	\textsc{du}.\textsc{s}-stay	oh	we.\textsc{du}.\textsc{excl}	\textsc{neg}.\textsc{cop}	you	lie	you.\textsc{du}=\textsc{emph}	oh	we.\textsc{du}.\textsc{excl}	\textsc{neg}.\textsc{cop}	you	I	deceive	I	\textsc{fac}-see.\textsc{o}.\textsc{anim}.\textsc{sg}-\textsc{fac}	here	stay\\
\glt `The bush spirit asked Bipep: ``Are you there?''  ``No, we are not.''  ``You are lying, it's you two.'' (Bipep insisted:) ``No, it's not us.'' (The bush spirit counters:) ``You deceive me. I have seen her, she is here.'''
\z

\ea\label{ex:18exwalp28}
Sukupu yipyo ppuo, wor dupua puesu. Yeloyo seku, puana: ko due nup.\\
\gll sukupu	yip-yo	ppuo	wor	dupua	puesu-i	yelo-yo	seku	puana	ko	due\_nui-p\\
     bush.spirit	house-\textsc{loc}	go.up.\textsc{pp}	dog	two	bite.\textsc{pp}-\textsc{du}.\textsc{a}	ground-\textsc{loc}	fall.\textsc{pp}	stand.up.\textsc{pp}	I	sleep-\textsc{pc}\\
\glt `The bush spirit went up to the house. Two dogs bit (it). It fell on the ground, but stood up (again): ``I was sleeping,'' (it thought).'
\z

\newpage
\ea\label{ex:18exwalp29}
Bipep muelno: ``Ko le yipyo, ko yip ikapyo le.'' Dorilo, balok.\\
\gll Bipep	mueli-no	ko	le	yip-yo	ko	yip	ikap-yo	le	dorilo	ba-le-ko\\
     Bipep	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	I	go	house-\textsc{loc}	I	house	1\textsc{sg}.\textsc{poss}.\textsc{emph}-\textsc{loc}	go	go.back.\textsc{pp}	\textsc{fac}-go-\textsc{fac}\\
\glt `(Then the bush spirit) said to Bipep: ``I'll go to the house. I will go (back) to my house.'' It turned back. It has gone.'
\z

\ea\label{ex:18exwalp30}
Bipep akoro muelo: ``Nuko ile yilau depiyo.'' Ilo. Bi ropno bepu ropno, suel parno lolo. Ako wokoro yilau kepyo.\\
\gll Bipep	ako=ro	mueli-no	nuko	i-le	yilau	de-pi-yo	i-lo	bi	rop-no	bepu	rop-no	suel	par-no	lolo	ako	woko=ro	yilau	kep-yo\\
     Bipep	wife=\textsc{emph}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	village	2\textsc{sg}-\textsc{poss}-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	meat	basket-\textsc{ins}	sago.grub	basket-\textsc{ins}	tobacco	bark.mat-\textsc{ins}	wrap.\textsc{pp}	wife	go.together.with.\textsc{pp}=\textsc{emph}	village	3\textsc{sg}.\textsc{poss}-\textsc{loc}\\
\glt `Bipep said to the wife: ``We'll go to your village. They went with meat in the basket  and with sago grubs in the basket. He wrapped tobacco in a bark mat. Together with the woman he went to her village.'
\z

\ea\label{ex:18exwalp31}
Pewo boliyo le royowo, ruri kep roise inekip, dob pop.\\
\gll pewo	boliyo	le	royo-we	ruri	kep	roise	i-neki-p	dob	pi-p\\
     banana	at.the.foot	things	lay.\textsc{pp}-\textsc{ter}	child	3\textsc{sg}.\textsc{poss}	together	\textsc{du}.\textsc{s}-stand-\textsc{pc}	eye	\textsc{lv}-\textsc{pc}\\
\glt `He put (their) belongings at the foot of a banana tree, (where) they were standing together with her child and were looking.'
\z

\ea\label{ex:18exwalp32}
Uki kep sele pop. Ako kep puap pulapana.\\
\gll uki	kep	sele	pi-p	ako	kep	puap	pulapana\\
     husband	3\textsc{sg}.\textsc{poss}	garden	do-\textsc{pc}	wife	3\textsc{sg}.\textsc{poss}	kind.of.fruit release.thither.\textsc{pp}\\
\glt `Her husband was working in the garden. His wife threw \textit{puap}-fruits (at him).'
\z

\newpage
\ea\label{ex:18exwalp33}
Uki dob riye puap: ``Ana pulap?''  Loro pewo boliyo: ``Eh ako kopi,'' anno wiyo ruri kep roise. ``Ko so de kusuikori.''\\
\gll uki	dob	riye	puap	ana	pula-p	lo=ro	pewo	boliyo	ah	ako	ko-pi	an-no	wiyo	ruri	kep	roise	ko	so	de	ke	sui-ko=ro\\
     husband	eye	see.\textsc{o}.\textsc{inanim}	fruit.species	who	release-\textsc{pc}	go.\textsc{pp}=\textsc{emph}	banana	at.the.foot	ah	wife	1\textsc{sg}-\textsc{pos}	arm-\textsc{ins}	hold.\textsc{pp}	child	3\textsc{sg}.\textsc{poss}	together	I	believe	you	\textsc{top}	die-\textsc{rts}=\textsc{emph}\\
\glt `The husband looks around: ``Who was throwing the \textit{puap}-fruits?'' He went to the foot of the banana tree, ``Oh, my wife!'' He put his arms around her and her/his child:  ``I believed you died not long ago.'''
\z

\ea\label{ex:18exwalp34}
``Ko ar sui, ko unake. Ko Bipep mekiyo.''\\
\gll ko	ar	sui	ko	u-nake	ko	Bipep	mekiyo\\
     I	\textsc{neg}	die	I	\textsc{dfac}-live	I Bipep	help.\textsc{pp}\\
\glt `I did not die, I am alive. Bipep helped me.'
\z

\ea\label{ex:18exwalp35}
Le rapiyoi yipyo ilo, le yipyo royo. Weri kep pulo, anno wiyo, mosiyo, womop: ``Ko so de kusuikori!'' ``Ko unake, ko ar sui. Ko Bipep mekiyo.''\\
\gll le	rapiyo-i	yip-yo	i-lo	le	yip-yo	royo	weri	kep	pulo	an-no	wiyo	mosiyo	wo\_mop	ko	so	de	ke	sui-ko=ro	ko	u-nake	ko	ar	sui	ko	Bipep	mekiyo\\
     things	fetch.\textsc{pp}-\textsc{du}.\textsc{a}	house-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	things	house-\textsc{loc}	lay.\textsc{pp}	younger.sister	3\textsc{sg}.\textsc{poss}	come.\textsc{pp}	arm-\textsc{ins}	hold.\textsc{pp}	embrace.\textsc{pp}	weep.\textsc{pp}	I	believe	you	\textsc{top}	die-\textsc{rts}=\textsc{emph}	I	\textsc{dfac}-live	I	\textsc{neg}	die	I	Bipep	help.\textsc{pp}\\
\glt `They fetched the things and went to the house. She put (her) things into the house. Her  younger sister came, embraced her, rejoiced and wept: ``I believed you died.''  ``I am alive, I didn't die. Bipep helped me.'''
\z

\ea\label{ex:18exwalp36}
Kiu muelien: ``Inawe, ine yip kopiyo pulupip! Bo kopi malap! Bipep kikilimpoyam kuwiyoyam, ko mekiyo!''\\
\gll kiu	mueli-en	ine	awe	ine	yip	ko-pi-yo	pulupi-p	bo	ko-pi	male-p	Bipep	k-kilim-pi-ye-m	k-wiye-ye-m	ko	mekiyo\\
     kin.group	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{pl}	come.\textsc{imp}	you.\textsc{pl}	house	1\textsc{sg}-\textsc{poss}-\textsc{loc}	come.\textsc{pl}-\textsc{imp}	word	1\textsc{sg}-\textsc{poss}	hear-\textsc{imp}	Bipep	\textsc{proh}-kill(\textsc{tp})-\textsc{lv}-\textsc{pl}-\textsc{proh}	\textsc{proh}-hold-\textsc{pl}-\textsc{proh}	I	help.\textsc{pp}\\
\glt `She said to the family: ``Come, come to my house! Hear my words! Do not kill Bipep,  do not hold him. He helped me!'''
\z

\ea\label{ex:18exwalp37}
``De uke bamuelienko, epul uke bamalok, boyo uke ar wiye.''\\
\gll de	uke	ba-mueli-ini-ko	epul	uke	ba-male-ko	boyo	uke	ar	wiye\\
     you	we.\textsc{excl}	\textsc{fac}-talk.to-\textsc{nsg}.\textsc{or}-\textsc{fac}	ear	we.\textsc{excl}	\textsc{fac}-hear-\textsc{fac}	later	we.\textsc{excl}	\textsc{neg}	hold\\
\glt `{``}You have spoken to us, and we have understood it. We won't get hold (of him) later (and won't punish him).'''\footnote{Normally a stranger would have been punished or even killed, when he entered a village accompanying a married woman of this village.}
\z

\ea\label{ex:18exwalp38}
Yena monomno molo. Sukupu yipyo moloro, paeau yip sukupupiyo.\\
\gll yena	monomno	molo	sukupu	yip-yo	molo=ro	paeau	yip	sukupu-pi-yo\\
     people	along.the.path	go.\textsc{pl}.\textsc{pp}	bush.spirit	house-\textsc{loc}	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	arrive	house	bush.spirit-\textsc{poss}-\textsc{loc}\\
\glt `The people went along the bush track. They went to the house of the bush spirit. They arrived at the house of the bush spirit.'
\z

\ea\label{ex:18exwalp39}
Wonppueno: ``De nake?'' ``Depe ana?''  Smep paliyo, yeloyo kûno.\\
\gll wonippue-no	de	nake	de=pe	ana	smep	paliyo	yelo-yo	kûno\\
     call.up-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	stay	you=\textsc{q}	who	door	open.\textsc{pp}	ground-\textsc{loc}	go.down.\textsc{pp}\\
\glt `They called up to it: ``Are you there?''  ``Who are you?'' (The spirit) opened the door and  came down to the ground.'
\z

\ea\label{ex:18exwalp40}
Yena epika poyap, peno wapilayoko. Pe kiniyo dop kepyo papulip, paliya.\\
\gll yena	epika	poye-p	pe-no	wapilaye-ko	pe	kiniyo	dop	kep-yo	papuli-p	paliya\\
     people	along	stand.\textsc{pl}-\textsc{pc}	arrow-\textsc{ins}	collect.lay.\textsc{pp}-\textsc{rts}	arrow	all	body	3\textsc{sg}.\textsc{poss}-\textsc{loc}	be.there.\textsc{pl}-\textsc{pc}	be.dead\\
\glt `The people were standing in a line and peppered it with their arrows. All the arrows stuck in its body. It is dead.'
\z

\ea\label{ex:18exwalp41}
Piyapoppuo, yip bîyo panapo. Riyopuno yip sûno noriyo. Yena le roise yilauyo molo.\\
\gll piyape\_ppuo	yip	bîyo	pane\_po	riyopuno	yip	sû-no	noriyo	yena	le	roise	yilau-yo	molo\\
     lift.up\_go.up.\textsc{pp}	house	inside	put.thither.\textsc{pp}	then	house	fire-\textsc{ins}	fill.in.\textsc{pp}	people	things	together	village-\textsc{loc}	go.\textsc{pl}.\textsc{pp}\\
\glt `They lifted (it) up and put it inside the house. Then they set the house on fire. (Afterwards) they went together to the village (loaded) with the belongings (of the bush spirit).'
\z

\ea\label{ex:18exwalp42}
Ûli boyo pulwolo. Yena epul malwolo: Ar nake, ûli kep bapulwoloko.\\
\gll ûli	boyo	pulwolo	yena	epul	male-wolo	ar	nake	ûli	kep	ba-pulwole-ko\\
     gall.bladder	later	burst.\textsc{pp}	people	ear	hear-\textsc{cpl}.\textsc{pp}	\textsc{neg}	live	gall.bladder	3\textsc{sg}.\textsc{poss}	\textsc{fac}-burst-\textsc{fac}\\
\glt' `Later (the spirit's) gall bladder burst. The people heard it loudly. (They knew) it wasn't alive (anymore), its gall bladder has burst.'
\z

\ea\label{ex:18exwalp43}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `The story is finished.'
\z
