\ea\label{ex:nrexhukpii}
Dari werino ilo hukpoipno puyo. Weri huk lilika pop, dari huk luika pop.\\
\gll dari	weri-no	i-lo	huk-pi-i-p-no	pu-yo	weri	huk	lilika	pi-p	dari	huk	luika	pi-p\\
     older.sister	younger.sister-\textsc{ins}	\textsc{du}.\textsc{s}-go.\textsc{pp}	catch.fish(\textsc{tp})-\textsc{lv}-\textsc{du}.\textsc{s}-\textsc{pc}-\textsc{co}	river-\textsc{loc}	younger.sister	catch.fish(\textsc{tp})	upriver	\textsc{lv}-\textsc{pc}	older.sister	catch.fish(\textsc{tp})	downriver	\textsc{lv}-\textsc{pc}\\
\glt `Two sisters went catching fish in the river, the younger one caught fish upriver, the  older one downriver.'
\z

\ea\label{ex:nrexrakida}
Weri ke wal kiniyo raki. Dari kero wal moniseso raki.\\
\gll weri	ke	wal	kiniyo	raki	dari	ke=ro	wal	moniseso	raki\\
     younger.sister	\textsc{aph}	fish	many	catch.fish	older.sister	\textsc{aph}=\textsc{emph}	fish	very.little	catch.fish\\
\glt `The younger sister catches many fish, the older one catches a small amount of fish.'
\z

\ea\label{ex:nrexdoriil}
Riyopuno dari werino doriilo yipyo, wal weloi.\\
\gll riyopuno	dari	weri-no	dori\_i-lo	yip-yo	wal	welo-i\\
     then	older.sister	younger.sister-\textsc{ins}	turn.back\_\textsc{du}.\textsc{s}-go.\textsc{pp}	house-\textsc{loc}	fish	carry.away.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `Then the sisters turned back home and took the fish with them.'
\z

\newpage
\ea\label{ex:nreximpoko}
Dari ki weri muelno: ``De ke wal kiniyo pulimpo. Ko ke wal kiniyo ar pulimpo.''\\
\gll dari	ki	weri	mueli-no	de	ke	wal	kiniyo	pulim-po	ko	ke	wal	kiniyo	ar	pulim-po\\
     older.sister	\textsc{aph}	younger.sister	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	\textsc{aph}	fish	many	pull(\textsc{tp})-\textsc{lv}.\textsc{pp}	I	\textsc{aph}	fish	many	\textsc{neg}	pull(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `The older sister said to the younger: ``You caught many fish, as for me, I didn't catch many.'''
\z

\ea\label{ex:nrexdekepp}
Weri ki muelno: ``De ke ppulaena hukpo.'' ``Ari, ko ke makina hukpop.''\\
\gll weri	ki	mueli-no	de	ke	ppulae-na	huk-po	ari	ko	ke	maki-na	huk-pi-p\\
     younger.sister	\textsc{aph}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	\textsc{top}	bad-\textsc{adv}	catch.fish(\textsc{tp})-\textsc{lv}.\textsc{pp}	no	I	\textsc{top}	good-\textsc{adv}	catch.fish(\textsc{tp})-\textsc{lv}-\textsc{pc}\\
\glt `The younger sister said to her: ``You caught badly.” ``No, I caught well.”'
\z

\ea\label{ex:nrexriiloy}
Riyopuno dari werino midoriilo yipyo.\\
\gll riyopuno	dari	weri-no	mi-dori\_i-lo	yip-yo\\
     then	older.sister	younger.sister-\textsc{ins}	\textsc{iter}-turn.back\_\textsc{du}.\textsc{s}-go.\textsc{pp}	house-\textsc{loc}\\
\glt `Then the two sisters turned back again to go home.'
\z

\ea\label{ex:nrexkerowa}
Dari ki epe bo mosaupono: ``Weri kopi kero wal kiniyo raki. Ko ke kiniyo ar raki.''\\
\gll dari	ki	epe	bo	mosaupo-no	weri	ko-pi	ke=ro	wal	kiniyo	raki	ko	ke	kiniyo	ar	raki\\
     older.sister	\textsc{aph}	mother	word	teach-3\textsc{sg}.\textsc{or}.\textsc{pp}	younger.sister	1\textsc{sg}-\textsc{poss}	\textsc{aph}=\textsc{emph}	fish	many	catch	I	\textsc{top}	many	\textsc{neg}	catch\\
\glt `The older sister informed the mother: ``My younger sister catches many fish. But I don't catch many.'''
\z

\ea\label{ex:nrexonobad}
Riyopuno dari ki weri muelno: ``De kero ono ba de wal namo piye.''\\
\gll riyopuno	dari	ki	weri	mueli-no	de	ke=ro	ono	ba	de	wal	namo	piye\\
     then	older.sister	\textsc{aph}	younger.sister	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	\textsc{aph}=\textsc{emph}	person	some	you	fish	give.2\textsc{sg}.\textsc{or}.\textsc{pp}	take\\
\glt `Then the older sister said to the younger: ``As for you, somebody gave you the fish  to take.”'
\z

\ea\label{ex:nrexikepul}
``Ari. Ko ike pulimpo. Ono ba pari.''\\
\gll ari	ko\_ike	pulim-po	ono	ba	pari\\
     no	I.myself	pull(\textsc{tp})-\textsc{lv}.\textsc{pp}	person	some	\textsc{neg}.\textsc{cop}\\
\glt `{``}No, I myself caught them, there wasn't anybody else.'''
\z

\ea\label{ex:nrexyiprum}
Riyopuno dari werino yip rumyo ina, due inuip.\\
\gll riyopuno	dari	weri-no	yip	rum-yo	i-na	due	i-nui-p\\
     then	older.sister	younger.sister-\textsc{ins}	house	room(\textsc{tp})-\textsc{loc}	\textsc{du}.\textsc{s}-go.inside	sleep	\textsc{du}.\textsc{a}-do.intentionally-\textsc{pc}\\
\glt `Then the two sisters went inside their room in the house and were sleeping.'
\z

\ea\label{ex:nrexialwol}
Punipino pini. Masalai pial bapulok. Woloyo nakap.\\
\gll punipino	pini	masalai	pial	ba-pule-ko	wolo-yo	nake-p\\
     morning	come.up.hither	bush.spirit(\textsc{tp})	snake	\textsc{fac}-come-\textsc{fac}	ladder-\textsc{loc}	sit-\textsc{pc}\\
\glt `The morning comes up. A bush spirit, the snake has come. It was sitting on the ladder.'
\z

\ea\label{ex:nrexmawatk}
Dop kep pialno tanimpo. Woloyo nakap. Nem kep ke Mawatkawi.\\
\gll dop	kep	pial-no	tanim-po	wolo-yo	nake-p	nem	kep	ke	Mawatkawi\\
     skin	3\textsc{sg}.\textsc{poss}	snake-\textsc{ins}	change(\textsc{tp})-\textsc{lv}.\textsc{pp}	ladder-\textsc{loc}	sit-\textsc{pc}	name	3\textsc{sg}.\textsc{poss}	\textsc{aph}	Mawatkawi\\
\glt `It changed its appearance with (that of a) snake, it was sitting (man-like) on the ladder, its name is Mawatkawi.'
\z

\ea\label{ex:nrexdebopu}
Epe aino masalai pial muelnoi: ``De bo pulo?''\\
\gll epe	ai-no	masalai	pial	mueli-no-i	de	bo	pulo\\
     mother	father-\textsc{ins}	bush.spirit(\textsc{tp})	snake	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	you	what	come.\textsc{pp}\\
\glt `Father and mother said to the bush spirit, the snake: ``What did you come for?”'
\z

\newpage
\ea\label{ex:nrexkorurd}
Masalai pial epe aino muelien: ``Ko ke bo ba pule. Ko ruri depi pule.''\\
\gll masalai	pial	epe	ai-no	mueli-en	ko	ke	bo	ba	pule	ko	ruri	de-pi	pule\\
     bush.spirit(\textsc{tp})	snake	mother	father-\textsc{ins}	talk.to-\textsc{n\textsc{sg}}.\textsc{or}.\textsc{pp}	I	\textsc{top}	what	\textsc{neg}.\textsc{emph}	come	I	child	2\textsc{sg}-\textsc{poss}	come\\
\glt `The bush spirit, the snake, said to father and mother: ``I don't come for any goods. I come for your child.'''
\z

\ea\label{ex:nrexnuandu}
Epe aino muelnoi: ``De bo pule? De bo muli? Wor bi yûr wal suo nuan due?'' Ar muli.\\
\gll epe	ai-no	mueli-no-i	de	bo	pule	de	bo	muli	wor	bi	yûr	wal	suo	nuan	due	ar	muli\\
     mother	father-\textsc{ins}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	you	what	come	you	what	want	dog	pig	chicken	fish	coconut	breadfruit	sago	\textsc{neg}	speak\\
\glt `Father and mother said to it: ``What do you come for, what do you want? A dog, a pig,  chicken, fish, coconuts, breadfruit, sago?” (Mawatkawi) doesn't speak.'
\z

\ea\label{ex:nrexorimue}
Epe aino dorimuelnoi: ``De bo muli?''\\
\gll epe	ai-no	dorimueli-no-i	de	bo	muli\\
     mother	father-\textsc{ins}	repeat-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	you	what	want\\
\glt `Father and mother repeated (to Mawatkawi): ``What do you want?”'
\z

\ea\label{ex:nrexakomul}
Masalai pial muelien: ``Ko ke ako muli, ruri depi.''\\
\gll masalai	pial	mueli-en	ko	ke	ako	muli	ruri	de-pi\\
     bush.spirit(\textsc{tp})	snake	talk.to-\textsc{n\textsc{sg}}.\textsc{or}.\textsc{pp}	I	\textsc{top}	wife	want	child	2\textsc{sg}-\textsc{poss}\\
\glt `The bush spirit, the snake, said to them: ``As for me, I want a wife, your daughter.”'
\z

\ea\label{ex:nrexyohukp}
Epe aino dari werino muelien: ``Deyo hukpoipno masalai pial oki pulo.''\\
\gll epe	ai-no	dari	weri-no	mueli-en	deyo	huk-pi-i-p-no	masalai	pial	o-ki	pulo\\
     mother	father-\textsc{ins}	older.sister	younger.sister-\textsc{ins}	talk.to-\textsc{n\textsc{sg}}.\textsc{or}.\textsc{pp}	you.\textsc{du}	catch.fish-\textsc{lv}-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	bush.spirit(\textsc{tp})	snake	\textsc{prox}-\textsc{aph}	come.\textsc{pp}\\
\glt `Father and mother said to the sisters: ``While you were catching fish, a bush spirit, the  snake, came here.”'
\z

\newpage
\ea\label{ex:nrexdeelir}
Dari werino muelno: ``De eliro wal kiniyo rakiulipop. Ko pari. De eli piyap. Ko pari.''\\
\gll dari	weri-no	mueli-no	de\_eli=ro	wal	kiniyo	raki-uli-pi-p	ko	pari	de\_eli	piye-p	ko	pari\\
     older.sister	younger.sister-\textsc{ins}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you.yourself=\textsc{emph}	fish	many	catch-\textsc{prog}-\textsc{lv}-\textsc{pc}	I	\textsc{neg}.\textsc{cop}	you.yourself	take-\textsc{pc}	I	\textsc{neg}.\textsc{cop}\\
\glt `The older sister said to the younger: ``You yourself were catching many fish, not me. You yourself were taking them, not me.”'
\z

\ea\label{ex:nrexpatipi}
Riyopuno epe aino patiponoi; weri ponoip. Na le kiniyo meli. Ri malalpoipno wo mopoi.\\
\gll riyopuno	epe	ai-no	pati-pi-no-i	weri	pi-no-i-p	na	le	kiniyo	meli	ri	malalpi-i-p-no	wo\_mopo-i\\
     then	mother	father-\textsc{ins}	party-\textsc{lv}-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	younger.sister	do-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}-\textsc{pc}	and(\textsc{tp})	things	all	carry.\textsc{pl}.\textsc{o}	wood	cut.firewood-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	cry.\textsc{pp}-\textsc{du}.\textsc{s}\\
\glt `Then father and mother had a party for her. They did it for the younger sister and  carried all the things. While cutting firewood they cried.'\footnote{Now the wedding party takes place at the girl's family's place; then they leave for the river, where the bush spirit lives.}
\z

\ea\label{ex:nrexsuduab}
Pial masalai ki burilo, Kusudua boyo lo. Ekuyo wulino epe aino roise wokoi.\\
\gll pial	masalai	ki	burilo	Kusudua	boyo	lo	ekuyo	wuli-no	epe	ai-no	roise	woko-i\\
     snake	bush.spirit(\textsc{tp})	\textsc{aph}	go.ahead.\textsc{pp}	Kusudua	behind	go.\textsc{pp}	behind	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}	mother	father-\textsc{ins}	with	accompany.\textsc{pp}-\textsc{du}.\textsc{s}\\
\glt `The bush spirit, the snake, went ahead, and Kusudua went behind. She followed it  behind with father and mother, (who) accompanied her.'
\z

\ea\label{ex:nrexduapon}
Pu epiyo nekip. Na pial ki dop kep Kusudua ponamo.\\
\gll pu	epiyo	neki-p	na	pial	ki	dop	kep	Kusudua	ponamo\\
     river	beside	stand-\textsc{pc}	and(\textsc{tp})	snake	\textsc{aph}	skin	3\textsc{sg}.\textsc{poss}	Kusudua	give.3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `They were standing beside the river and the snake gave its skin to Kusudua, (and she changed into a snake).'
\z

\ea\label{ex:nrexbalepi}
Riyopuno pial dop senisimpono na puyo ikûno. Balopisiko.\\
\gll riyopuno	pial	dop	senisim-pi-no	na	pu-yo	i-kûno	ba-le-pisi-ko\\
     then	snake	skin	change(\textsc{tp})-\textsc{lv}-3\textsc{sg}.\textsc{or}.\textsc{pp}	and(\textsc{tp})	river-\textsc{loc}	\textsc{du}.\textsc{s}-go.down.\textsc{pp}	\textsc{fac}-go-\textsc{cpl}-\textsc{fac}\\
\glt `Then the snake changed its appearance, and they went down into the river. They are gone forever.'\footnote{The bush spirit who had come in human appearance changed back into a snake.}
\z

\ea\label{ex:nrexepaiwo}
Epe aino boyo wo mopoi.\\
\gll epe	ai-no	boyo	wo\_mopo-i\\
     mother	father-\textsc{ins}	later	cry.\textsc{pp}-\textsc{du}.\textsc{s}\\
\glt `Later mother and father cried.'
\z

\ea\label{ex:nrexkurbri}
Kuru stori kopi.\\
\gll kuru	stori	ko-pi\\
     be.finished	story	1\textsc{sg}-\textsc{poss}\\
\glt `My story is finished.'
\z
