\ea\label{ex:t2exruriai}
Ruri aino duyo woko. Kiyo kama woko.\\
\gll ruri	ai-no	du-yo	woko	kiyo	kama	woko\\
     child	father-\textsc{ins}	bush-\textsc{loc}	go.together.with.\textsc{pp}	\textsc{aph}.\textsc{du}	alone	go.together.with.\textsc{pp}\\
\glt `Father and son went together to the bush. The two of them went alone.'
\z

\ea\label{ex:t2exkwokep}
Kowokapno duyo, duyo wonu.\\
\gll k-woke-p-no	du-yo	du-yo	wo-nu\\
     \textsc{sub}-go.together.with-\textsc{pc}-\textsc{co}	bush-\textsc{loc}	bush-\textsc{loc}	\textsc{accom}-sleep.\textsc{pp}\\
\glt `Having gone together to the bush, they slept in the bush in (each other's) company.'
\z

\ea\label{ex:t2exriyokn}
Riyo kunuipno ai muele: ``Ko duyo le, ko du pue.''\\
\gll ri-yo	k-nui-p-no	ai	mueli-no	ko	du-yo	le	ko	du	pue\\
     \textsc{dist}-\textsc{loc}	\textsc{sub}-sleep-\textsc{pc}-\textsc{co}	father	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	I	bush-\textsc{loc}	go	I	forest	roam\\
\glt `Having slept there, the father said to (his son): ``I'll go into the bush, I'll roam the forest.'''
\z

\ea\label{ex:t2exdekena}
``De ke nakowo, de riyo ppuowo.''  Riyo ulepoko, ``de dob riyo powe.''\\
\gll de	ke	nake-we	de	ri-yo	ppue-we	ri-yo	ulei-pi-ko	de	dob	ri-yo	pi-we\\
     you	\textsc{top}	stay-\textsc{ter}	you	tree-\textsc{loc}	climb-\textsc{ter}	tree-\textsc{loc}	put.inside-\textsc{lv}-\textsc{rts}	you	eye	tree-\textsc{loc}	\textsc{lv}-\textsc{ter}\\
\glt `{``}(But) you, you stay (here). You climb up a tree.'' The father put him in the (chosen)  tree: ``You watch out from there attentively.'''
\z

\ea\label{ex:t2exkuleip}
Kuleipopno ai duyo lo.\\
\gll k-ulei-pi-p-no	ai	du-yo	lo\\
     \textsc{sub}-put.inside-\textsc{lv}-\textsc{pc}-\textsc{co}	father	bush-\textsc{loc}	go.\textsc{pp}\\
\glt `Having put (him) in (the tree), the father went into the (deep) bush.'
\z

\ea\label{ex:t2exduyovo}
Duyo loro, du puap.\\
\gll du-yo	lo=ro	du	pue-p\\
     bush-\textsc{loc}	go.\textsc{pp}=\textsc{emph}	forest	roam-\textsc{pc}\\
\glt `He went into the bush and was roaming the forest.'
\z

\ea\label{ex:t2exkepmue}
Ruri kep muelkono: ``De dob powe, de rileyo nakeno dob powe.''\\
\gll ruri	kep	mueli-ko-no	de	dob	pi-we	de	rileyo	nake-no	dob	pi-we\\
     child	3\textsc{sg}.\textsc{poss}	talk.to-\textsc{rts}-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	eye	\textsc{lv}-\textsc{ter}	you	above	stay-\textsc{co}	eye	\textsc{lv}-\textsc{ter}\\
\glt `Before he had said to his son: ``You watch attentively, while you are up there. You keep watch attentively!'''
\z

\ea\label{ex:t2exaikefd}
Ai kep duyo lo. Duyo klapno du puap.\\
\gll ai	kep	du-yo	lo	du-yo	k-le-p-no	du	pue-p\\
     father	3\textsc{sg}.\textsc{poss}	bush-\textsc{loc}	go.\textsc{pp}	bush-\textsc{loc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	forest	roam-\textsc{pc}\\
\glt `His father went into the bush. Having left, he was roaming the bush.'
\z

\ea\label{ex:t2exkamari}
Ruri ki kama ri wuliyo nakap.\\
\gll ruri	ki	kama	ri	wuli-yo	nake-p\\
     child	\textsc{aph}	alone	tree	tree.top-\textsc{loc}	sit-\textsc{pc}\\
\glt `The child was sitting alone in the tree top.'
\z

\ea\label{ex:t2exraunwa}
Raunwara yip pakiyo lilip, pu paek.\\
\gll raunwara	yip	pakiyo	lili-p	pu\_paek\\
     pond(\textsc{tp})	house	next.to	be.there-\textsc{pc}	waterhole\\
\glt `There was a pond next to a house, a hole with water.'
\z

\ea\label{ex:t2expupaeq}
Pu paek ikoiele yipyo lilip, raunwara ri epeyo îwoulip.\\
\gll pu\_paek	ikoiele	yip-yo	lili-p	raunwara	ri	epeyo	îwole-p\\
     waterhole	very.big	house-\textsc{loc}	be.there-\textsc{pc}	pond(\textsc{tp})	wood	visible	dry.up-\textsc{pc}\\
\glt `The pond was very big, it was near the (bush)house. The pond was drying up, and (at the bottom) wood was visible.'
\z

\ea\label{ex:t2exmasala}
Masalai riyo nakap, riyo piniro.\\
\gll masalai	ri-yo	nake-p	ri-yo	pini=ro\\
     bush.spirit(\textsc{tp})	\textsc{dist}-\textsc{loc}	live-\textsc{pc}	\textsc{dist}-\textsc{loc}	come.up.hither=\textsc{emph}\\
\glt `A bush spirit was living there (at the bottom of the pond), from there it comes up here.'
\z

\newpage
\ea\label{ex:t2exkepros}
Ruri kepro sano: ``Depe rileyo bo nake, ai de rileyo ulepoko, ai depi arka lo?''\\
\gll ruri	kep=ro	sa-no	de=pe	rileyo	bo	nake	ai	de	rileyo	ulei-pi-ko	ai	de-pi	arka	lo\\
     child	3\textsc{sg}.\textsc{poss}=\textsc{emph}	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	you=Q	above	what	sit	father	you	above	put.inside-\textsc{lv}-\textsc{rts}	father	2\textsc{sg}-\textsc{poss}	where	go.\textsc{pp}\\
\glt `It asked the (father's) son: ``What are you sitting there for, high above, did your father put you up in (the tree)? Where did your father go?'''
\z

\ea\label{ex:t2exaikopf}
``Ai kopi busyo lo. Du pue, yala pulam.''\\
\gll ai	ko-pi	du-yo	lo	du	pue	yala	pule-m\\
     father	1\textsc{sg}-\textsc{poss}	bush-\textsc{loc}	go.\textsc{pp}	forest	roam	soon	come-\textsc{pos}\\
\glt `{``}My father went into the (deep) bush, he roams the forest. He will come (back) soon.'''
\z

\ea\label{ex:t2exdeyoka}
``Deyo kama wopulo?''\\
\gll deyo	kama	wo-pulo\\
     you.\textsc{du}	alone	\textsc{accom}-come.\textsc{pp}\\
\glt `{``}Did the two of you come together alone?'''
\z

\ea\label{ex:t2exoukoyo}
``Ou koyo kama wopulo.''\\
\gll ou	koyo	kama	wo-pulo\\
     yes	we.\textsc{du}.\textsc{excl}	alone	\textsc{accom}-come.\textsc{pp}\\
\glt `{``}Yes, we came alone.'''
\z

\ea\label{ex:t2exyenike}
``Yeni kep ari ope yeni anapi?''\\
\gll yeni	kep	ari	o=pe	yeni	ana-pi\\
     bed	3\textsc{sg}.\textsc{poss}	no	\textsc{prox}=Q	bed	who-\textsc{poss}\\
\glt `{``}His bed, no, whose bed is this?'''
\z

\ea\label{ex:t2exzenoro}
``Yeniro erki kopi, ai kopi yeni kep erki.''\\
\gll yeni=ro	ere-ki	ko-pi	ai	ko-pi	yeni	kep	ere-ki\\
     bed=\textsc{emph}	\textsc{prox}-\textsc{aph}	1\textsc{sg}-\textsc{poss}	father	1\textsc{sg}-\textsc{poss}	bed	3\textsc{sg}.\textsc{poss}	\textsc{prox}-\textsc{aph}\\
\glt `{``}This bed is mine. My father's, his bed is this.'''
\z

\largerpage
\ea\label{ex:t2exmalapi}
``Yeni kep ko malapi. Mak koyopi kopuala dupua. Yala ko dupuni pule, ai so muele.''  ``Ou.''\\
\gll yeni	kep	ko	malapi	mak	koyo-pi	kopuala\_dupua	yala	ko	dupuni	pule	ai	so	mueli-ne	ou\\
     bed	3\textsc{sg}.\textsc{poss}	I	measure	mark	we.\textsc{du}.\textsc{excl}-\textsc{poss}	two.times.the.same	today	I	night	come	father	so	talk.to-3\textsc{sg}.\textsc{or}	yes\\
\glt `{``}I measure his bed. Our measurement is the same, later in the night I'll come. Tell (your) father.''  ``Yes (I will).'''
\z

\ea\label{ex:t2exdukpue}
Ai du kupuapno bike wepulo yipyo.\\
\gll ai	du	k-pue-p-no	bike	wepulo	yip-yo\\
     father	forest	\textsc{sub}-roam-\textsc{pc}-\textsc{co}	cassowary	bring.\textsc{pp}	house-\textsc{loc}\\
\glt `Having roamed the forest, the father brought a cassowary to the (bush)house.'
\z

\ea\label{ex:t2exbikekl}
Bike kuluipno, kowepulapno yipyo, kuwopuenpopno, aplayo woroyepepo, sû mappoppue.\\
\gll bike	k-lui-p-no	k-wepule-p-no	yip-yo	k-wo-puenpi-p-no	apla-yo	wo-royepepo	sû	mappeppue\\
     cassowary	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	\textsc{sub}-bring-\textsc{pc}-\textsc{co}	house-\textsc{loc}	\textsc{sub}-\textsc{accom}-cut.meat-\textsc{pc}-\textsc{co}	plank-\textsc{loc}	\textsc{accom}-lay.on.top.\textsc{pp}	fire	light.go.up\\
\glt `He shot the cassowary and brought it to the (bush)house. Then, having cut the meat together, they laid it on top of planks, and the burning fire is going up.'
\z

\ea\label{ex:t2exsukwom}
Sû kowomappoppuepno, punino sû womappoppuo.\\
\gll sû	k-wo-mappeppue-p-no	puni-no	sû	wo-mappeppuo\\
     fire	\textsc{sub}-\textsc{accom}-light.go.up-\textsc{pc}-\textsc{co}	evening-\textsc{ins}	fire	\textsc{accom}-light.go.up.\textsc{pp}\\
\glt `Having lit the fire, they kept the fire lit during the evening.'
\z

\ea\label{ex:t2exqaikai}
Kaikai bilil kwesipno kwonopno kiyo kama dupua inakap.\\
\gll kaikai	bilil	k-wo-si-p-no	k-wo-ni-p-no	kiyo	kama	dupua	i-nake-p\\
     food(\textsc{tp})	pieces.of.meat	\textsc{sub}-\textsc{accom}-cook-\textsc{pc}-\textsc{co}	\textsc{sub}-\textsc{accom}-eat-\textsc{pc}-\textsc{co}	\textsc{aph}.\textsc{du}	alone	two	\textsc{du}.\textsc{s}-sit-\textsc{pc}\\
\glt `Having cooked and eaten food and pieces of meat, the two of them were sitting alone.'
\z

\ea\label{ex:t2exkvonip}
Kwonopno due wonui ai muele:\\
\gll k-wo-ni-p-no	due	wo-nui	ai	mueli-ne\\
     \textsc{sub}-\textsc{accom}-eat-\textsc{pc}-\textsc{co}	sleep	\textsc{accom}-do.intentionally	father	talk.to-3\textsc{sg}.\textsc{or}\\
\glt `After the meal they go to bed together, (and the son) says to (his) father:'
\z

\newpage
\ea\label{ex:t2exonobaf}
``Ono ba pu ipiyo pin. Ana ari. Puyo pin, ana ari. Yeni depi malapop.''\\
\gll ono	ba	pu	ipi-yo	pin	ana	ari	pu-yo	pin	ana	ari	yeni	de-pi	malapi-p\\
     man	other	water	clay-\textsc{loc}	come.up.hither.\textsc{pp}	who	no	water-\textsc{loc}	come.up.hither.\textsc{pp}	who	no	bed	2\textsc{sg}-\textsc{poss}	measure-\textsc{pc}\\
\glt `{``}Somebody came up here from the water bottom. (I do) not (know) who he is. He came up here from the water bottom. (I do) not (know) who he is. He was measuring your bed.'''
\z

\ea\label{ex:t2exkonize}
``Yala ko puni pini, yala ko puni koniye,'' so muel.\\
\gll yala	ko	puni	pini	yala	ko	puni	koniye	so	muel\\
     today	I	night	come.up.hither	today	I	night	swallow	so	say.\textsc{pp}\\
\glt `{``}Today at night I'll come up. Today at night I will devour (your father),'' so he said.'
\z

\ea\label{ex:t2exdeqedu}
``De ke dupuapi , mopsi bo mui! Bike ulili.'' ``Yala ko asa ni.'' ``Bike de oso ni, dob de riyoworo.''\\
\gll de	ke	dupuapi	mopsi	bo	mui	bike	u-lili	yala	ko	asa	ni	bike	de	oso	ni	dob	de	riye-we=ro\\
     you	\textsc{top}	lie	nonsense	word	talk	cassowary	\textsc{dfac}-be.there	now	I	how	eat	cassowary	you	more	eat	eye	you	see.\textsc{o}.\textsc{inanim}-\textsc{ter}=\textsc{emph}\\
\glt `{``}You are lying, you are talking nonsense. Here is cassowary (meat).'' ``How do I eat now?'' ``You eat more of the cassowary, (tomorrow) you watch out attentively.'''
\z

\ea\label{ex:t2exarifuk}
``Ari, dukiru so muel.''\\
\gll ari	duki=ro	so	muel\\
     no	true=\textsc{emph}	so	say.\textsc{pp}\\
\glt `{``}No, it is true, so he said.'''
\z

\ea\label{ex:t2exarifek}
``Ari, de ke giamanpi, de ke dupuapi!''\\
\gll ari	de	ke	giaman-pi	de	ke	dupuapi\\
     no	you	\textsc{top}	lie(\textsc{tp})-\textsc{lv}	you	\textsc{top}	lie\\
\glt `{``}No, you are lying, you are lying.'''
\z


\ea\label{ex:t2exdupuaf}
``Dupuapi ari, bo dukiru. Dob ko reyo, ono bekulu pu ipiyo pin. Dob ko reyo.''\\
\gll dupuapi	ari	bo	duki=ro	dob	ko	reyo	ono	bekulu	pu	ipi-yo	pin	dob	ko	reyo\\
     lie	no	word	true=\textsc{emph}	eye	I	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	man	huge	water	clay-\textsc{loc}	come.up.hither.\textsc{pp}	eye	I	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}\\
\glt `{``}I am not lying, it is true. I saw him. A huge man came up here from the bottom  of the water, I saw him.'''
\z

\ea\label{ex:t2exbikeli}
``Bike lili!''  ``Yala ko asa ni?''  ``Bike de apaina ni!''\\
\gll bike	lili	yala	ko	asa	ni	bike	de	apai-na	ni\\
     cassowary	be.there	now	I	how	eat	cassowary	you	greedy-\textsc{adv}	eat\\
\glt `{``}There is the cassowary (meat).'' ``How can I eat now?'' ``You eat the cassowary with a good appetite.'''
\z

\ea\label{ex:t2exuqunar}
``Upru!''\\
\gll upuna=ro\\
     alright=\textsc{emph}\\
\glt `{``}Alright, alright.'''
\z

\ea\label{ex:t2exvonuro}
Wonuero puni pin. Puni kaepul an sayo. Kaepul kasayapno pu paek ke mi dopyo lili.\\
\gll wo-nu=ro	puni	pin	puni	kaepul	an	saye	kaepul	k-saye-p-no	pu\_paek	ke	mi	dopyo	lili\\
     \textsc{accom}-sleep.\textsc{pp}=\textsc{emph}	night	come.up.hither.\textsc{pp}	night	knee	hand	touch	knee	\textsc{sub}-touch-\textsc{pc}-\textsc{co}	waterhole	\textsc{aph}	again	next.to	be.there\\
\glt `They slept together. At night (the bush spirit) came up. At night the (father's) hand touches his knee. While he touches his knee, the waterhole is nearby.'\footnote{This gesture indicates that the father is sleeping soundly.}
\z

\ea\label{ex:t2expunikp}
Puni kipinipno ai kep koniyo.\\
\gll puni	k-pini-p-no	ai	kep	koniyo\\
     night	\textsc{sub}-come.up.hither-\textsc{pc}-\textsc{co}	father	3\textsc{sg}.\textsc{poss}	swallow.\textsc{pp}\\
\glt `When (the bush spirit) came up (from the bottom of the pond) at night, it swallowed up (the child's) father.'
\z

\ea\label{ex:t2exzoloko}
Ai solo koniyoro, yeni rupuolo.\\
\gll ai	solo	koniyo=ro	yeni	rupue-wolo\\
     father	ony	swallow.\textsc{pp}=\textsc{emph}	bed	break-\textsc{cpl}.\textsc{pp}\\
\glt `Only the father it swallowed up. (His) bed (the bush spirit) broke completely.'
\z

\ea\label{ex:t2exroiser}
Yeni roise rupuap. Ruri ba koniyo, ruri nakap.\\
\gll yeni	roise	rupue-p	ruri	ba	koniyo	ruri	nake-p\\
     bed	together	break-\textsc{pc}	child	\textsc{neg}.\textsc{emph}	swallow.\textsc{pp}	child	stay-\textsc{pc}\\
\glt `It was breaking father and bed, (but) the child it didn't swallow. The child was staying (alive).'
\z

\ea\label{ex:t2exrippul}
Ruri ppulaena puana: Masalai ai kep koniyo.\\
\gll ruri	ppulae-na	puana	masalai	ai	kep	koniyo\\
     child	bad-\textsc{adv}	rise.\textsc{pp}	bush.spirit(\textsc{tp})	father	3\textsc{sg}.\textsc{poss}	swallow.\textsc{pp}\\
\glt `The boy rose (in a) bad (state of mind): the bush spirit devoured his father.'
\z

\ea\label{ex:t2exkkoniz}
Kokoniyapno puyo kûno. Ruri puni klapno ri ipulyo nakap duruwa.\\
\gll k-koniye-p-no	pu-yo	kûno	ruri	puni	k-le-p-no	ri\_ipul-yo	nake-p	duruwa\\
     \textsc{sub}-swallow-\textsc{pc}-\textsc{co}	water-\textsc{loc}	go.down.\textsc{pp}	child	night	\textsc{sub}-go-\textsc{pc}-\textsc{co}	tree.buttress-\textsc{loc}	sit-\textsc{pc}	early.morning\\
\glt `Having swallowed him up, (the bush spirit) went down into the water. After running away at night the child was sitting between the buttresses of a tree until early morning.'
\z

\ea\label{ex:t2exsekimp}
Ai sekimpo: Piaune ari.\\
\gll ai	sekim-po	piaune	ari\\
     father	seek(\textsc{tp})-\textsc{lv}.\textsc{pp}	python	no\\
\glt `He searched for the father. A python? No.'
\z

\ea\label{ex:t2exaimaza}
Ai masalai koniyo, piaune ari. Oke ono sukupu bakoniyok.\\
\gll ai	masalai	koniyo	piaune	ari	o-ke	ono	sukupu	ba-koniye-ko\\
     father	bush.spirit(\textsc{tp})	swallow.\textsc{pp}	python	no	\textsc{prox}-\textsc{aph}	man	bush.spirit	\textsc{fac}-swallow-\textsc{fac}\\
\glt `(The child thinks:) ``A bush spirit swallowed up (my) father, not a python. This was a man. A bush spirit has swallowed him up.'''
\z

\ea\label{ex:t2exrizopu}
Riyopuno ruri woise lo yilauyo.\\
\gll riyopuno	ruri	woise	lo	yilau-yo\\
     then	child	with.tears	go.\textsc{pp}	village-\textsc{loc}\\
\glt `Then the child went tearfully to the village.'
\z

\ea\label{ex:t2exmuelie}
Yena kiniyo ki muelen epe kep roise:\\
\gll yena	kiniyo	ki	mueli-en	epe	kep	roise\\
     people	all	\textsc{aph}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	mother	3\textsc{sg}.\textsc{poss}	together\\
\glt `He said to all the people including his mother:'
\z

\ea\label{ex:t2exnukomo}
``Ai kopi masalai bakoniyok. Ine awe, nuko mole!\\
\gll ai	ko-pi	masalai	ba-koniye-ko	ine	awe	nuko	mole\\
     father	1\textsc{sg}-\textsc{poss}	bush.spirit(\textsc{tp})	\textsc{fac}-swallow-\textsc{fac}	you.\textsc{pl}	come.\textsc{imp}	we.\textsc{incl}	go.\textsc{pl}\\
\glt `{``}A bush spirit has swallowed up my father. You all come, we'll all go!'''
\z

\ea\label{ex:t2exdepifa}
Masalai ono yeni depi bamalapoko.\\
\gll masalai	ono	yeni	de-pi	ba-malapi-ko\\
     bush.spirit(\textsc{tp})	man	bed	2\textsc{sg}-\textsc{poss}	\textsc{fac}-measure-\textsc{fac}\\
\glt `{``}The bush spirit, the man has measured your bed.'''
\z

\ea\label{ex:t2exmopsip}
De ke mopsi bo mui. Bike lili, de noro!\\
\gll de	ke	mopsi	bo	mui	bike	lili	de	ni=ro\\
     you	\textsc{top}	nonsense	word	talk	cassowary	be.there	you	eat=\textsc{emph}\\
\glt `{``}You are talking nonsense. There is some cassowary, you eat.'''
\z

\ea\label{ex:t2exfikelu}
Ai kopi bike luro, koyo wesip. Mi ko so muelpup:\\
\gll ai	ko-pi	bike	lu=ro	koyo	wo-si-p	mi	ko	so	mueli-ipi-p\\
     father	1\textsc{sg}-\textsc{poss}	cassowary	shoot.\textsc{pp}=\textsc{emph}	we.\textsc{du}.\textsc{excl}	\textsc{accom}-cook-\textsc{pc}	then	I	so	talk.to-1\textsc{sg}.\textsc{or}-\textsc{pc}\\
\glt `{``}My father shot a cassowary, (and) we were cooking it together. Then he was talking  to me like this.'''
\z

\ea\label{ex:t2exlilife}
Bike lili, de noro, de ke dupuapi!\\
\gll bike	lili	de	ni=ro	de	ke	dupuapi\\
     cassowary	be.there	you	eat=\textsc{emph}	you	\textsc{top}	lie\\
\glt `{``}There is cassowary meat, you just eat. You are lying.'''
\z

\ea\label{ex:t2exdeboqo}
De bo kopi epul ar male, masalai de puni koniye.''\\
\gll de	bo	ko-pi	epul	ar	male	masalai	de	puni	koniye\\
     you	word	1\textsc{sg}-\textsc{poss}	ear	\textsc{neg}	hear	bush.spirit(\textsc{tp})	you	night	swallow\\
\glt `(I replied to him:) ``You don't listen to my words, the bush spirit will swallow you up at night.'''
\z

\ea\label{ex:t2exikiqum}
Riyopuno iki kumune molo. Kukepulapno pu riye.\\
\gll riyopuno	iki	kumune	molo	k-uke-pule-p-no	pu	riye\\
     then	\textsc{aph}.\textsc{pl}	all.\textsc{coll}	go.\textsc{pl}.\textsc{pp}	\textsc{sub}-jointly-come-\textsc{pc}-\textsc{co}	water	see.\textsc{o}.\textsc{inanim}\\
\glt `Then they all left. Having come (there) together they see the water(hole).'
\z

\ea\label{ex:t2exkzopip}
Kiyopipno, riyopuno pu îwolo yelo solo.\\
\gll k-yopi-p-no	riyopuno	pu	îwolo	yelo	solo\\
     \textsc{sub}-scoop-\textsc{pc}-\textsc{co}	then	water	dry.up.\textsc{pp}	ground	only\\
\glt `They scooped (the pond) ... Then the water dried up, there was only the ground.'
\z

\ea\label{ex:t2exepeyoi}
Ri epeyo îwoulip riyo nakap. Yena kumune peno lu.\\
\gll ri	epeyo	îwole-p	ri-yo	nake-p	yena	kumune	pe-no	lu\\
     wood	visible	dry.up-\textsc{pc}	\textsc{dist}-\textsc{loc}	live-\textsc{pc}	people	all.\textsc{coll}	arrow-\textsc{ins}	shoot.\textsc{pp}\\
\glt `Wood is visible (because the water) was drying up. There it was living. All the people jointly shot (it) with arrows.'
\z

\ea\label{ex:t2exqluipn}
Kuluipno paliya. Masalai roke baluko.\\
\gll k-lui-p-no	paliya	masalai	ro-ke	ba-lui-ko\\
     \textsc{sub}-shoot-\textsc{pc}-\textsc{co}	be.dead	bush.spirit(\textsc{tp})	\textsc{prox}.\textsc{emph}-\textsc{aph}	\textsc{fac}-shoot-\textsc{fac}\\
\glt `Having shot at (the bush spirit) it was dead. They have killed this bush spirit.'
\z

\ea\label{ex:t2exklayep}
Kelayapno neppno suko, iki yilauyo molo.\\
\gll k-laye-p-no	neppi-no	suko	iki	yilau-yo	molo\\
     \textsc{sub}-lay-\textsc{pc}-\textsc{co}	bush.knife-\textsc{ins}	cut.\textsc{pp}	\textsc{aph}.\textsc{pl}	village-\textsc{loc}	go.\textsc{pl}.\textsc{pp}\\
\glt `Having laid it (down), they cut (it to pieces) with (their) bush knives. (Then) they went to the village.'
\z

\ea\label{ex:t2exemtaso}
Em tasol liklik stori.\\
\gll em	tasol	liklik	stori\\
     \textsc{aph}(\textsc{tp})	only(\textsc{tp})	small(\textsc{tp})	story(\textsc{tp})\\
\glt `That's it, the little story.'
\z
