\ea\label{ex:nrexseleyi}
Uki akono sele yipyo ilo.\\
\gll uki	ako-no	sele	yip-yo	i-lo\\
     husband	wife-\textsc{ins}	garden	house-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}\\
\glt `A husband and a wife went to (their) garden house.'
\z

\ea\label{ex:nrexdeoyon}
Uki ako muelno: ``De oyo nakewe, sele yipyo. Ko duyo le.''\\
\gll uki	ako	mueli-no	de	o-yo	nake-we	sele	yip-yo	ko	du-yo	le\\
     husband	wife	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	\textsc{prox}-\textsc{loc}	stay-\textsc{ter}	garden	house-\textsc{loc}	I	bush-\textsc{loc}	go\\
\glt `The husband said to his wife: ``You stay here in the garden house, I'll go into the  bush.'''
\z

\ea\label{ex:nrexduyokl}
Ke duyo klapno, bi kuluipno, wepulo yipyo.\\
\gll ke	du-yo	k-le-p-no	bi	k-lui-p-no	wepulo	yip-yo\\
     \textsc{aph}	bush-\textsc{loc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	animal	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	bring.\textsc{pp}	house-\textsc{loc}\\
\glt `He went to the bush. Having shot animals, he brought them to the house.'
\z

\newpage
\ea\label{ex:nrexplayor}
Kupuenpoipno aplayo royepepoi. Sû mappeppuo aplayo. Bi smokimpoi.\\
\gll k-puenpi-i-p-no	apla-yo	royepepo-i	sû\_mappeppuo	apla-yo	bi	smokim-po-i\\
     \textsc{sub}-cut.meat-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	plank-\textsc{loc}	lay.on.top.\textsc{pp}-\textsc{du}.\textsc{a}	light.a.fire.up.\textsc{pp}	plank-\textsc{loc}	meat	smoke-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `Having cut the meat, they laid it on top of planks. They lit a fire (going) upward and, on the planks, they smoked the meat.'
\z

\ea\label{ex:nrexksmoki}
Kosmokimpoipno ako due nuip. Pial ploulipopno, ar saupo. Due nuipop.\\
\gll k-smokim-pi-i-p-no	ako	due\_nui-p	pial	pule-uli-pi-p-no	ar	saupo	due\_nui-pi-p\\
     \textsc{sub}-smoke-\textsc{lv}-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	wife	sleep-\textsc{pc}	snake	come-\textsc{prog}-\textsc{lv}-\textsc{pc}-\textsc{co}	\textsc{neg}	know	sleep-\textsc{lv}-\textsc{pc}\\
\glt `They were smoking (the meat). The wife was sleeping. (Then), while the snake was coming, she did not sense it. She was sleeping.'
\z

\ea\label{ex:nrexnuipop}
Ako due nuipop, ar saupo.\\
\gll ako	due\_nui-pi-p	ar	saupo\\
     wife	sleep-\textsc{lv}-\textsc{pc}	\textsc{neg}	know\\
\glt `The wife was sleeping, she did not know it.'
\z

\ea\label{ex:nrexialbar}
Uki kep ar reyepo. Uki ke duyo ploulipop. Pial ba reyo.\\
\gll uki	kep	ar	reye-po	uki	ke	du-yo	pue-uli-pi-p	pial	ba	reyo\\
     husband	3\textsc{sg}.\textsc{poss}	\textsc{neg}	see.\textsc{o}.\textsc{anim}.\textsc{sg}-\textsc{lv}.\textsc{pp}	husband	\textsc{aph}	forest-\textsc{loc}	roam-\textsc{prog}-\textsc{lv}-\textsc{pc}	snake	\textsc{neg}.\textsc{emph}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}\\
\glt `Her husband didn't see it. The husband was roaming the forest. He did not see the snake.'
\z

\ea\label{ex:nrexwalwul}
Kuso luwalwulpop.\\
\gll kuso	luwalwuli-pi-p\\
     always	have.sexual.intercourse-\textsc{lv}-\textsc{pc}\\
\glt `(The bush spirit-snake and the wife) had sex the whole time.'
\z

\ea\label{ex:nrexalwuli}
Luwalwulpopno uki kep duyo ploulipop. Sakana plouliponap.\\
\gll luwalwuli-pi-p-no	uki	kep	du-yo	pue-uli-pi-p	sakana	pule-uli-pi-ne-p\\
     have.sexual.intercourse-\textsc{lv}-\textsc{pc}-\textsc{co}	husband	3\textsc{sg}.\textsc{poss}	forest-\textsc{loc}	roam-\textsc{prog}-\textsc{lv}-\textsc{pc}	secretly	come-\textsc{prog}-\textsc{lv}-3\textsc{sg}.\textsc{or}-\textsc{pc}\\
\glt `While they were having sex, her husband was roaming the bush. (The bush spirit-snake) was coming secretly for her.'
\z

\ea\label{ex:nrexloples}
Ako el kipiyamipno. Bi kosmokimpoipno tupela i laik i go long ples.\\
\gll ako	el	k-piyami-p-no	bi	k-smokim-pi-i-p-no	tupela	i	laik	i	go	long	ples\\
     wife	belly	\textsc{sub}-become.pregnant-\textsc{pc}-\textsc{co}	meat	\textsc{sub}-smoke-\textsc{lv}-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	\textsc{aph}.\textsc{du}(\textsc{tp})	\textsc{pred}(\textsc{tp})	\textsc{nfut}(\textsc{tp})	\textsc{pred}(\textsc{tp})	go(\textsc{tp})	\textsc{prep}(\textsc{tp})	village(\textsc{tp})\\
\glt `When the wife had become pregnant ... Having smoked the meat, the two  of them wanted to go to the village.'
\z

\ea\label{ex:nrexopnana}
Iki solo yilauyo ilo. Ako el bekulu; ruri dupua onona pialna.\\
\gll iki	solo	yilau-yo	i-lo	ako	el	bekulu	ruri	dupua	ono-na	pial-na\\
     \textsc{aph}.\textsc{pl}	only	village-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	wife	belly	huge	child	two	man-\textsc{aff}	snake-\textsc{aff}\\
\glt `They just went to the village. The wife had a big belly, there were two children, a man-like one and a snake-like one.'
\z

\ea\label{ex:nrexilaiki}
Klostu i laik i karim tupela i go. Bi sali ropyo niskûno.\\
\gll klostu	i	laik	i	karim	tupela	i	go	bi	sali	rop-yo	niskûno\\
     near(\textsc{tp})	\textsc{pred}(\textsc{tp})	\textsc{nfut}(\textsc{tp})	\textsc{pred}(\textsc{tp})	give.birth(\textsc{tp})	\textsc{aph}.\textsc{du}(\textsc{tp})	\textsc{pred}(\textsc{tp})	go(\textsc{tp})	meat	dry	basket-\textsc{loc}	fill.in.\textsc{pp}\\
\glt `Not long before giving birth the two of them went and filled dried meat into the basket.'
\z

\ea\label{ex:nrexlaikig}
Tupela i laik i go long ples. Draipela ren i kam.\\
\gll tupela	i	laik	i	go	long	ples	draipela	ren	i	come\\
     \textsc{aph}.\textsc{du}(\textsc{tp})	\textsc{pred}(\textsc{tp})	\textsc{nfut}(\textsc{tp})	\textsc{pred}(\textsc{tp})	go(\textsc{tp})	\textsc{prep}(\textsc{tp})	village(\textsc{tp})	heavy(\textsc{tp})	rain(\textsc{tp})	\textsc{pred}(\textsc{tp})	come(\textsc{tp})\\
\glt `They wanted to go to the village, (but) a heavy rain came.'
\z

\ea\label{ex:nrexpuripa}
Pu bekulu pulo. Pu ripapno ruwepuloro.\\
\gll pu	bekulu	pulo	pu	ripap-no	ruwepulo=ro\\
     rain	huge	come.\textsc{pp}	rain	storm-\textsc{ins}	come.to.break.in.\textsc{pp}=\textsc{emph}\\
\glt `A huge rain came, rain and storm broke loose.'\footnote{This sequence repeats the second part of Sequence \ref{ex:nrexlaikig} in Kilmeri.}
\z


\ea\label{ex:nrexpasimt}
Pasim tupela i stap. Ba ilo yilauyo. Kiyo inakeko yipyo. So ilap. Ari. Pu ke kinikuro.\\
\gll pasim	tupela	i	stap	ba	i-lo	yilau-yo	kiyo	i-nake-ko	yip-yo\\
     block(\textsc{tp})	\textsc{aph}.\textsc{du}(\textsc{tp})	\textsc{pred}(\textsc{tp})	\textsc{dur}(\textsc{tp})	\textsc{neg}.\textsc{emph}	\textsc{du}.\textsc{s}-go.\textsc{pp}	village-\textsc{loc}	\textsc{aph}.\textsc{du}	\textsc{du}.\textsc{s}-stay-\textsc{rts}  small	house-\textsc{loc}\\
\glt `(The rain) prevented them from going. They did not go to the village. They remained in the (garden)house.'
\z

\ea\label{ex:nrexilapar}
So ilap. Ari. Pu ke kinikuro. So ilap. Ari. Pu ke kinikuro.\\
\gll so	i-le-p	ari	pu	ke	kinikuro	so	i-le-p	ari	pu	ke	kinikuro\\
     so	\textsc{du}.\textsc{s}-go-\textsc{pc}	no	rain	\textsc{aph}	cover.everything.\textsc{pp}	so	\textsc{du}.\textsc{s}-go-\textsc{pc}	no	rain	\textsc{aph}	cover.everything.\textsc{pp}\\
\glt `So they would go. No. The rain covered everything. So they would go. No. The rain covered everything.'
\z

\ea\label{ex:nrexkeprom}
Uki ako kepro muelno: ``De nakap. De ke pulenakap. De oyo nui. Ko solo le. Ko kama yilauyo nui.''\\
\gll uki	ako	kep=ro	mueli-no	de	nake-p	de	ke	pule	nake-p	de	o-yo	nui	ko	solo	le	ko	kama	yilau-yo	nui\\
     husband	wife	3\textsc{sg}.\textsc{poss}=\textsc{emph}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	stay-\textsc{imp}	you	\textsc{top}	come	stay-\textsc{imp}	you	\textsc{prox}-\textsc{loc}	sleep	I	only	go	I	alone	village-\textsc{loc}	sleep\\
\glt `The husband said to his wife: ``You stay, you really stay, you sleep here. Only I will go. I alone will sleep in the village.'''
\z

\ea\label{ex:nrexnaonon}
Ako ke ruri boyo nako, pialna onona.\\
\gll ako	ke	ruri	boyo	nako	pial-na	ono-na\\
     wife	\textsc{aph}	child	later	gave.birth	snake-\textsc{aff}	man-\textsc{aff}\\
\glt `The wife later gave birth to the children, a snake-like (child) and a man-like (child).'
\z

\ea\label{ex:nrexyonukn}
Uki yilauyo nuknoko. Punipino riyo puloro. Ako ruri banakeko, ba isaeaupop.\\
\gll uki	yilau-yo	nuknoko	punipino	ri-yo	pulo=ro	ako	ruri	ba-nake-ko	ba\_isaeaupi-p\\
     husband	village-\textsc{loc}	sleep.for.some.nights.\textsc{pp}	morning	\textsc{dist}-\textsc{loc}	come.\textsc{pp}=\textsc{emph}	wife	child	\textsc{fac}-give.birth-\textsc{fac}	suckle-\textsc{pc}\\
\glt `The husband slept in the village for some nights. One morning he came (back) there. The wife has given birth to the children, and she let them suckle.'
\z

\ea\label{ex:nrexbaisae}
Ba isaeaupopno ako kep muelno: ``Nuko yilauyo ile.''\\
\gll ba\_isaeaupi-p-no	ako	kep	mueli-no	nuko	yilau-yo	i-le\\
     suckle-\textsc{pc}-\textsc{co}	wife	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	we.\textsc{incl}	village-\textsc{loc}	\textsc{du}.\textsc{s}-go\\
\glt `After the breastfeeding he said to his wife: ``We'll go to the village.'''
\z

\ea\label{ex:nrexlnarop}
Ako ruri dupua woko. Pialna ropyo ule, onona wepulyo ule.\\
\gll ako	ruri	dupua	woko	pial-na	rop-yo	ule	ono-na	wepul-yo	ule\\
     wife	child	two	go.together.with.\textsc{pp}	snake-\textsc{aff}	basket-\textsc{loc}	put.into.\textsc{pp}	man-\textsc{aff}	baby.sling-\textsc{loc}	put.into.\textsc{pp}\\
\glt `The wife went together with the two children. The snake-like one she put into a basket, and the man-like one she put into the baby sling.'
\z

\ea\label{ex:nreximpowe}
Ilo yilauyo. Yena muelien: ``Ako ruri dupua karimpowe, onona pialna.''\\
\gll i-lo	yilau-yo	yena	mueli-en	ako	ruri	dupua	karim-po-we	ono-na	pial-na\\
     \textsc{du}.\textsc{s}-go.\textsc{pp}	village-\textsc{loc}	people	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	wife	child	two	give.birth(\textsc{tp})-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{o}	man-\textsc{aff}	snake-\textsc{aff}	\\
\glt `They went to the village. (There) he said to the people: ``My wife gave birth to two  children, a man-like one and a snake-like one.'''
\z

\ea\label{ex:nrexlukaut}
Ruri ba pialna kamappo, ba onona kamappo. Tupela em lukautim, ki iseaupoko.\\
\gll ruri	ba	pial-na	kamap-po	ba	ono-na	kamap-po	tupela	em	lukautim	ki	isaeaupi-ko\\
     child	other	snake-\textsc{aff}	become(\textsc{tp})-\textsc{lv}.\textsc{pp}	other	man-\textsc{aff}	become(\textsc{tp})-\textsc{lv}.\textsc{pp}	\textsc{aph}.\textsc{du}(\textsc{tp})	\textsc{aph}(\textsc{tp})	take.care(\textsc{tp})	\textsc{aph}	let.suckle-\textsc{rts}\\
\glt `One child became a snake-like (creature), the other one became a man-like (creature).  She took care of the two (children). She let them suckle.'
\z

\ea\label{ex:nrexepikoi}
Pialna nowap, onona nowap. Ikoi.\\
\gll pial-na	nowe-p	ono-na	nowe-p	ikoi\\
     snake-\textsc{aff}	grow-\textsc{pc}	man-\textsc{aff}	grow-\textsc{pc}	big\\
\glt `The snake-like one was growing, the man-like one was growing. (They are) big.'
\z

\ea\label{ex:nrexlowedu}
Kikoipoipno. Riyopuno bopap sepalowe, dupua.\\
\gll k-ikoi-po-i-p-no	riyopuno	bopap	sepalo-we	dupua\\
     \textsc{sub}-big-\textsc{lv}-\textsc{du}.\textsc{s}-\textsc{pc}-\textsc{co}	then	pig.trap	fence.\textsc{pp}-\textsc{du}.\textsc{o}	two\\
\glt `They have become big. Then they fenced pig traps (for hunting pigs), two (traps).'
\z

\ea\label{ex:nrexbopaps}
Ononaro bopap sepalowe, dupua. Pialnaro sepalonoro, ki sepaloro.\\
\gll ono-na=ro	bopap	sepalo-we	dupua	pial-na=ro	sepale-no=ro	ki	sepalo=ro\\
     man-\textsc{aff}=\textsc{emph}	pig.trap	fence.\textsc{pp}-\textsc{du}.\textsc{o}	two	snake-\textsc{aff}=\textsc{emph}	fence-3\textsc{sg}.\textsc{or}.\textsc{pp}=\textsc{emph}	\textsc{aph}	fence.\textsc{pp}=\textsc{emph}\\
\glt `The man-like one fenced two traps. He (also) fenced one for the snake-like (brother). He made the fences.'
\z

\ea\label{ex:nrexpulwis}
Wîs mono pule. Wîs ke apuli. Wîs mono pulilayeya.\\
\gll wîs	mono	pule	wîs	ke	a-puli	wîs	mono	pulilaye=ya\\
     moon	path	come	moon	\textsc{aph}	\textsc{imp}3-shine	moon	path	shine.brightly=\textsc{emph}\\
\glt `The path of the moonlight is coming. The moon has to shine. (Now) the moon shines brightly on the path.'
\z

\ea\label{ex:nrexiloewe}
Riyopuno kumune ilo. Ewe kep bopap nisep. Diri kep bopap nisep.\\
\gll riyopuno	kumune	i-lo	ewe	kep	bopap	nise-p	diri	kep	bopap	nise-p\\
     then	all.\textsc{coll}	\textsc{du}.\textsc{s}-go.\textsc{pp}	older.brother	3\textsc{sg}.\textsc{poss}	pig.trap	lie.in.wait-\textsc{pc}	younger.brother	3\textsc{sg}.\textsc{poss}	pig.trap	lie.in.wait-\textsc{pc}\\
\glt `Then they went together. The older brother was lying in wait at the pig trap, and his younger brother was lying in wait at the pig trap.'
\z

\ea\label{ex:nrexmikebu}
Pial mike klapno. Bi lu. Bi kuluipno kimike buri.\\
\gll pial	mike	k-le-p-no	bi	lu	bi	k-lui-p-no	kimike	buri\\
     snake	first	\textsc{sub}-go-\textsc{pc}-\textsc{co}	pig	shoot.\textsc{pp}	pig	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	first	go.ahead\\
\glt `The snake, having gone first, shot a pig. After having shot the pig it went  ahead first (and left the place of the trap).'
\z

\newpage
\ea\label{ex:nrexwilkal}
Kupulapno ri wili kaliyeulip.\\
\gll k-pule-p-no	ri\_wili	kaliye-uli-p\\
     \textsc{sub}-come-\textsc{pc}-\textsc{co}	log	lay.horizontally-\textsc{prog}-\textsc{pc}\\
\glt `Having come, (the snake) was laying a log horizontally.'\footnote{The snake-like brother must have reached a place where he could hide and later attack his man-like brother at the bush-camp.}
\z

\ea\label{ex:nrexklilul}
Kililulipno sikilyo, puaku kepro imiyo pepo. Onoro lewonap, ewe kepro.\\
\gll k-liluli-p-no	sikilyo	puaku	kep=ro	imiyo	pepo	ono=ro	lewo-ne-p	ewe	kep=ro\\
     \textsc{sub}-curl.up-\textsc{pc}-\textsc{co}	under	head	3\textsc{sg}.\textsc{poss}=\textsc{emph}	outside	put.on.top.\textsc{pp}	man=\textsc{emph}	wait.for-3\textsc{sg}.\textsc{or}-\textsc{pc}	older.brother	3\textsc{sg}.\textsc{poss}=\textsc{emph}\\
\glt `After curling up beneath it put its head outside on top. (So) it was waiting for the man, its very own brother.'
\z

\ea\label{ex:nrexbiklui}
Bi kuluipno, onoro kupulapno liki kiyopi nakap. Wopiyoro.\\
\gll bi	k-lui-p-no	ono=ro	k-pule-p-no	liki	kiyo-pi-yo	nake-p	wopiyo=ro\\
     pig	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	man=\textsc{emph}	\textsc{sub}-come-\textsc{pc}-\textsc{co}	made.ready.place	\textsc{aph}.\textsc{du}-\textsc{poss}-\textsc{loc}	stay-\textsc{pc}	straighten.\textsc{pp}=\textsc{emph}\\
\glt `Having shot a pig and come, the man was staying at their ready made place and straightened it.'
\z

\ea\label{ex:nrexkiekun}
Ki eku nakeno. El kep rika wiana pialro.\\
\gll ki	eku	nake-no	el	kep	ri-ka	wiana	pial=ro\\
     \textsc{aph}	behind	sit-\textsc{co}	belly	3\textsc{sg}.\textsc{poss}	\textsc{dist}.\textsc{emph}-\textsc{path}	turn.thither.\textsc{pp}	snake=\textsc{emph}\\
\glt `While sitting he turned his belly there, (towards) the snake.'
\z

\ea\label{ex:nrexpepeul}
Pial puaku ri wili pepeulip. Niseulip. Pauwepiyo sikapyo pisiyo. Onoro paliya.\\
\gll pial	puaku	ri\_wili	pepe-uli-p	nise-uli-p	pauwe-pi-yo	sikapyo	pisiyo	ono=ro	paliya\\
     snake	head	log	put.on.top-\textsc{prog}-\textsc{pc}	watch-\textsc{prog}-\textsc{pc}	breast-\textsc{poss}-\textsc{loc}	shoot.arrows.\textsc{pp}	finish.\textsc{pp}	man=\textsc{emph}	be.dead\\
\glt `The snake was putting its head on top of the log and was lurking. It shot arrows  at his breast, it ended it. The man is dead.'
\z

\ea\label{ex:nrexlewepy}
Ki onoro kisikapyapno, ki yilauyo kelewep, yipyo.\\
\gll ki	ono=ro	k-sikapye-p-no	ki	yilau-yo	kelewep	yip-yo\\
     \textsc{aph}	man=\textsc{emph}	\textsc{sub}-shoot.arrows-\textsc{pc}-\textsc{co}	\textsc{aph}	village-\textsc{loc}	go.home.\textsc{pp}	house-\textsc{loc}\\
\glt `After having shot the man, (the snake) (went) home to the village, to (their) house.'
\z

\ea\label{ex:nrexroewek}
Onoro ewe kep kikilimpopno, ki yilauyo yipyo lo.\\
\gll ono=ro	ewe	kep	k-kilim-pi-p-no	ki	yilau-yo	yip-yo	lo\\
     man=\textsc{emph}	older.brother	3\textsc{sg}.\textsc{poss}	\textsc{sub}-kill\textsc{(tp)}-\textsc{lv}-\textsc{pc}-\textsc{co}	\textsc{aph}	village-\textsc{loc}	house-\textsc{loc}	go.\textsc{pp}\\
\glt `When the older brother had killed (the younger brother), he went to the village, to (his) house.'
\z

\ea\label{ex:nrexrumkaa}
Rum kepyo lupuana. Kililulipno rum kepyo sipulyo.\\
\gll rum	kep-yo	lupuana	k-liluli-p-no	rum	kep-yo	sipul-yo\\
     room(\textsc{tp})	3\textsc{sg}.\textsc{poss}-\textsc{loc}	enter.thither.\textsc{pp}	\textsc{sub}-curl.up-\textsc{pc}-\textsc{co}	room(\textsc{tp})	3\textsc{sg}.\textsc{poss}-\textsc{loc}	floor-\textsc{loc}\\
\glt `He entered his room. When (the snake) had curled up in its room, on the floor …'
\z

\ea\label{ex:nreximipep}
Kililulipno liki kepyo puaku imiyo pepo.\\
\gll k-liluli-p-no	liki	kep-yo	puaku	imiyo	pepo\\
     \textsc{sub}-curl.up-\textsc{pc}-\textsc{co}	made.ready.place	3\textsc{sg}.\textsc{poss}-\textsc{loc}	head	outside	put.on.top.\textsc{pp}\\
\glt `After curling up at its place, it put its head on top (of its body).'
\z

\largerpage[2]
\ea\label{ex:nrexanupua}
Du baruweko. Ai epe ikepro kana riyeiya. Nini bapulilayeko. Anupua ba ipuane, diri eweno.\\
\gll du	ba-ruwe-ko	ai	epe	ikep=ro	kana	riye-i=ya	nini	ba-pulilaye-ko	anupua	ba	i-puane	diri	
ewe-no\\
     darkness	\textsc{fac}-break-\textsc{fac}	father	mother	3\textsc{pl}.\textsc{poss}=\textsc{emph}	quickly	see.\textsc{o}.\textsc{inanim}-\textsc{du}.\textsc{a}=\textsc{emph}	sun	\textsc{fac}-shine.brightly-\textsc{fac}	two	\textsc{neg}.\textsc{emph}	\textsc{du}.\textsc{s}-rise	younger.brother	older.brother-\textsc{ins}\\
\glt `The darkness is over. Their father and mother quickly check (the situation). The sun shines brightly, (but) the two don't rise, the younger and the older brother.'
\z

\ea\label{ex:nrexpapriy}
Onona pari. Epe ai muelkono: ``Bopap riyekap!''\\
\gll ono-na	pari	epe	ai	mueli-ko-no	bopap	riye-ke-p\\
     man-\textsc{aff}	\textsc{neg}.\textsc{cop}	mother	father	talk.to-\textsc{rts}-3\textsc{sg}.\textsc{or}.\textsc{pp}	pig.trap	see.\textsc{o}.\textsc{inanim}-\textsc{ingr}-\textsc{imp}\\
\glt `The man-like brother is not there. (Therefore) the mother had said to the father: ``Go to check the pig trap.'''
\z

\ea\label{ex:nrexaiklen}
Ai kana lenou. Pialna oke mi ewe kep kilimpo.\\
\gll ai	kana	le-ne-ou	pial-na	o-ke	mi	ewe	kep	kilim-po\\
     father	quickly	wait.for-3\textsc{sg}.\textsc{or}-\textsc{frus}	snake-\textsc{aff}	\textsc{prox}-\textsc{aph}	again	older.brother	3\textsc{sg}.\textsc{poss}	kill(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `The father is quickly frustrated in waiting for (the man-like brother). The snake-like one, this one (must have) killed its older brother.'
\z

\ea\label{ex:nrexbapapi}
Ewe kep barireko. Paliya. Eye bouno barireko. Oh, bapapoko. Ples bapapok.\\
\gll ewe	kep	ba-rire-ko	paliya	eye\_bou-no	ba-rire-ko	oh	ba-papi-ko	ples	ba-papi-ko\\
     older.brother	3\textsc{sg}.\textsc{poss}	\textsc{fac}-be.stiff-\textsc{fac}	be.dead	limbs-\textsc{ins}	\textsc{fac}-be.stiff-\textsc{fac}	oh	\textsc{fac}-make.\textsc{pl}.\textsc{o}-\textsc{fac}	maggots	\textsc{fac}-make.\textsc{pl}.\textsc{o}-\textsc{fac}\\
\glt `Its older brother is already stiff. He is dead, the limbs are stiff. Oh, (they) have already infested (the body), the maggots have infested (the body).'
\z

\ea\label{ex:nrexonawep}
Riyopuno wepulo. Ai kep onona wepulo.\\
\gll riyopuno	wepulo	ai	kep	ono-na	wepulo\\
     then	bring.\textsc{pp}	father	3\textsc{sg}.\textsc{poss}	man-\textsc{aff}	bring.\textsc{pp}\\
\glt `Then he brought him. His father brought the man-like one (to the parent's house).'
\z

\ea\label{ex:nrexnaroew}
Pialnaro ewe kep kilimpo. Ewe kep pialnaro kilimpo.\\
\gll pial-na=ro	ewe	kep	kilim-po	ewe	kep	pial-na=ro	kilim-po\\
     snake-\textsc{aff}=\textsc{emph}	older.brother	3\textsc{sg}.\textsc{poss}	kill\textsc{(tp)}-\textsc{lv}.\textsc{pp}	older.brother	3\textsc{sg}.\textsc{poss}	snake-\textsc{aff}=\textsc{emph}	kill\textsc{(tp)}-\textsc{lv}.\textsc{pp}\\
\glt `The snake-like one killed its older brother, the older brother was killed by the snake-like brother.'\footnote{Grammatical comment: The English passive is chosen as translation of the subject focus in the second clause; the first clause shows object focus.}
\z

\ea\label{ex:nrexakikoi}
Epe aino ri krapiyeipno kilikoi pialro. Rino pakikoi, pakikoi. Yipyo pakikoi.\\
\gll epe	ai-no	ri	k-rapiye-i-p-no	kili-ko-i	pial=ro	ri-no	paki-ko-i	paki-ko-i	yip-yo	paki-ko-i\\
     mother	father-\textsc{ins}	stick	\textsc{sub}-fetch-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	beat.fiercely-\textsc{rts}-\textsc{du}.\textsc{a}	snake=\textsc{emph}	stick-\textsc{ins}	beat-\textsc{rts}-\textsc{du}.\textsc{a}	beat-\textsc{rts}-\textsc{du}.\textsc{a}	house-\textsc{loc}	beat-\textsc{rts}-\textsc{du}.\textsc{a}\\
\glt `When mother and father had fetched sticks, they beat the snake fiercely with the sticks. They beat and beat, they had beat it out in the house.'
\z

\ea\label{ex:nrexowoyel}
Yeloyo piowo. Yeloka kikilipno, pialnaro paliya.\\
\gll yelo-yo	piowo	yeloka	k-kili-p-no	pial-na=ro	paliya\\
     ground-\textsc{loc}	jump.\textsc{pp}	outside	\textsc{sub}-beat.fiercely-\textsc{pc}-\textsc{co}	snake-\textsc{aff}=\textsc{emph}	be.dead\\
\glt `(The snake-like brother) jumped on the ground, and after they had fiercely beaten it outside, the snake is dead.'
\z

\ea\label{ex:nrexsusulo}
Ri kupurapoipno sû suloimoina. Sû suloimoina mappoi.\\
\gll ri	k-purapi-i-p-no	sû	suloimoina	sû	suloimoina	mappo-i\\
     wood	\textsc{sub}-cut.wood-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	fire	extraordinary	fire	extraordinary	light.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `Having cut wood, they lit a colossal fire, a huge fire.'
\z

\ea\label{ex:nrexpoises}
Pial sûyo pepoi. Kili solo po, ise solo.\\
\gll pial	sû-yo	pepo-i	kili	solo	po	ise	solo\\
     snake	fire-\textsc{loc}	lay.on.top.\textsc{pp}-\textsc{du}.\textsc{a}	bone	only	\textsc{lv}.\textsc{pp}	char	only\\
\glt `They put the snake on top of the fire. (Soon) it was bones only, charred only.'
\z

\ea\label{ex:nrexupubis}
Ononaro bî kraripno matmatyo pakûpu. Bî sowelayo.\\
\gll ono-na=ro	bî	k-rari-p-no	matmat-yo	pakûpu	bî	sowelayo\\
     man-\textsc{aff}=\textsc{emph}	hole	\textsc{sub}-dig-\textsc{pc}-\textsc{co}	burial.place(\textsc{tp})-\textsc{loc}	throw.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	hole	cover.thoroughly.\textsc{pp}\\
\glt `Having dug a hole for the man-like brother at the burial place, they threw him down and covered the hole thoroughly.'
\z

\ea\label{ex:nrexsolten}
Em tasol. Tenkyu.\\
\gll em	tasol	tenk\_yu\\
     \textsc{aph}(\textsc{tp})	only(\textsc{tp})	thank.you(\textsc{tp})\\
\glt `That's it, thank you.'
\z
