\ea\label{ex:nrexnembil}
Nem bilong me Andrew Wapi. Mi laik stori long tumbuna stori.\\
\gll nem	bilong	mi	Andrew	Wapi	mi	laik	stori	long	tumbuna	stori\\
     name	\textsc{poss}(\textsc{tp})	I(\textsc{tp})	Andrew	Wapi	I(\textsc{tp})	like(\textsc{tp})	tell(\textsc{tp})	\textsc{rel}(\textsc{tp})	ancestor(\textsc{tp})	story(\textsc{tp})\\
\glt `My name is Andrew Wapi. I like to tell an ancestral story.'
\z

\ea\label{ex:nrexkonodu}
Uki akono duyo ilo. Duyo ipuap.\\
\gll uki	ako-no	du-yo	i-lo	du-yo	i-pue-p\\
     husband	wife-\textsc{ins}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-roam-\textsc{pc}\\
\glt `A husband and a wife went to the bush. They were roaming the bush.'
\z

\ea\label{ex:nrexsupoun}
Duyo ipuapno bi luwapoi. Ako rop supouno po. Uki pe parno lolo.\\
\gll du-yo	i-pue-p-no	bi	luwapo-i	ako	rop	supou-no	po	uki	pe	par-no	lolo\\
     bush-\textsc{loc}	\textsc{du}.\textsc{s}-roam-\textsc{pc}-\textsc{co}	animal	catch.\textsc{pp}-\textsc{du}.\textsc{a}	wife	basket	support-\textsc{ins}	make.\textsc{pp}	husband	arrow	pannier-\textsc{ins}	tie.\textsc{pp}\\
\glt `While roaming the bush they caught animals. The wife attached a supporting part to the basket. The husband tied the arrows he had not used on the pannier (to take them home).'
\z

\ea\label{ex:nrexparnkk}
Parno klolapno, sû kamappoipno, ya keniminiipno, elro ki melipuloiwap ropno.\\
\gll par-no	k-lole-p-no	sû\_k-mappe-i-p-no	ya	k-ni\_mini-i-p-no	el=ro	ki	melipulo-i-we-p	rop-no\\
     pannier-\textsc{ins}	\textsc{sub}-tie-\textsc{pc}-\textsc{co}	fire\_\textsc{sub}-light.a.fire-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	sago	\textsc{sub}-eat\_come.hither-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	game=\textsc{emph}	\textsc{aph}	bring.\textsc{pl}.\textsc{o}.\textsc{pp}-\textsc{du}.\textsc{a}-\textsc{ter}-\textsc{pc}	basket-\textsc{ins}\\
\glt `He tied (the killed animals) on the pannier. Having lit a fire, eaten some sago, they transported the game in baskets.'
\z

\ea\label{ex:nrexolouli}
Ki melipuloiro, ippuoro ol ouli.\\
\gll ki	melipulo-i=ro	i-ppuo=ro	ol	ouli\\
     \textsc{aph}	bring.\textsc{pl}.\textsc{o}.\textsc{pp}-\textsc{du}.\textsc{a}=\textsc{emph}	\textsc{du}.\textsc{s}-go.up.\textsc{pp}=\textsc{emph}	hill	ridge\\
\glt `They brought plenty (of meat) and climbed the riverbank.'
\z

\ea\label{ex:nrexukibur}
Kukûnoipno puyo. Puyo uki buri.\\
\gll k-kûne-i-p-no	pu-yo	pu-yo	uki	buri\\
     \textsc{sub}-go.down-\textsc{du}.\textsc{s}-\textsc{pc}-\textsc{co}	river-\textsc{loc}	river-\textsc{loc}	husband	go.ahead\\
\glt `Having gone down to the river, the husband went ahead into the water.'
\z

\ea\label{ex:nrexpuppoy}
Uki kimike pu siamu. Amaka puppoyo nakap.\\
\gll uki	kimike	pu	siamu	amaka	puppo-yo	nake-p\\
     husband	first	river	cross.hither.\textsc{pp}	over.there	sand-\textsc{loc}	stay-\textsc{pc}\\
\glt `The husband crossed the river first and was staying over there on the sand.'
\z

\ea\label{ex:nrexyosian}
Ako boyo siana rop roise.\\
\gll ako	boyo	siana	rop	roise\\
     wife	behind	cross.thither.\textsc{pp}	basket	together\\
\glt `The wife crossed (the river) behind him, together with the basket.'
\z

\ea\label{ex:nrexkppuee}
Kuppuapno pu pokoyo urai riyo wiyo pu pokoyo.\\
\gll k-ppue-p-no	pu	pokoyo	urai	ri-yo	wiyo	pu	pokoyo\\
     \textsc{sub}-go.up-\textsc{pc}-\textsc{co}	river	in.the.middle	crocodile	\textsc{dist}-\textsc{loc}	catch.\textsc{pp}	river	in.the.middle\\
\glt `When the wife reached the middle of the river, a crocodile caught her there, (right) in the middle of the river.'
\z

\ea\label{ex:nrexrairoo}
Urairo oke onoso. Urai ono sukei.\\
\gll urai=ro	o-ke	ono-so	urai	ono	sukei\\
     crocodile=\textsc{emph}	\textsc{prox}-\textsc{aph}	man-\textsc{sim}	crocodile	man	spirit\\
\glt `But the crocodile was man-like. The crocodile had the spirit of a human being.'\footnote{Here it is explicitly said that bush spirits appearing as dangerous animals have a human mind.}
\z

\ea\label{ex:nrexkwiyen}
Kwiyapno yip kepyo wena bi rop roisero. Roise wel, roise wiyo, roise wena.\\
\gll k-wiye-p-no	yip	kep-yo	wena	bi	rop	roise=ro	roise	wel	roise	wiyo	roise	wena\\
     \textsc{sub}-catch-\textsc{pc}-\textsc{co}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	carry.inside	meat	basket	together=\textsc{emph}	together	carry.\textsc{pp}	together	catch.\textsc{pp}	together	carry.inside\\
\glt `Having caught (the wife) it carried her inside its house together with the basket  of meat. It carried (all) together, held everything and put it inside.'
\z

\ea\label{ex:nrexyorosu}
Urai yip kepyoro sukupu. Kimike dop uraiso po.\\
\gll urai	yip	kep-yo=ro	sukupu	kimike	dop	urai-so	po\\
     crocodile	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}=\textsc{emph}	bush.spirit	before	skin	crocodile-\textsc{sim}	\textsc{lv}.\textsc{pp}\\
\glt `In its house the crocodile is a bush spirit. Before it had a crocodile's skin.'
\z

\ea\label{ex:nrexukiso}
Uki solo. Ki kama lo yilauyo. Uki wo roise lo, loro yilauyo.\\
\gll uki	solo	ki	kama	lo	yilau-yo	uki	wo\_roise	lo	lo=ro	yilau-yo\\
     husband	only	\textsc{aph}	alone	go.\textsc{pp}	village-\textsc{loc}	husband	crying	go.\textsc{pp}	go.\textsc{pp}=\textsc{emph}	village-\textsc{loc}\\
\glt `Only the husband. He went alone to the village. He went crying. He went to the village.'
\z

\ea\label{ex:nrexmuelak}
Yena kumuelinopno: ``Ako kopi kero urai wiyo elro roise. Wel, roise wiyo.\\
\gll yena	k-muel-ini-p-no	ako	ko-pi	ke=ro	urai	wiyo	el=ro	roise	wel	roise	wiyo\\
     people	\textsc{sub}-talk.to-\textsc{nsg}.\textsc{or}-\textsc{pc}-\textsc{co}	wife	1\textsc{sg}-\textsc{poss}	\textsc{aph}=\textsc{emph}	crocodile	catch.\textsc{pp}	game=\textsc{emph}	together	carry.\textsc{pp}	together	catch.\textsc{pp}\\
\glt `He told the people: ``My wife, a crocodile caught her together with the  game. It took (both). It caught her together (with the game).'''
\z

\ea\label{ex:nrexemssuk}
Yala so kunuipno em. So ukepule puyo, puyo kepue.''\\
\gll yala	so	k-nui-p-no	em	so	uke-pule	pu-yo	pu-yo	kepue\\
     now	so	\textsc{sub}-sleep-\textsc{pc}-\textsc{co}	tomorrow	so	jointly-come	river-\textsc{loc}	river-\textsc{loc}	dive\\
\glt `(The husband said:) ``Well, after having slept, tomorrow (morning) we'll come to the river together. We'll dive into the river.'''
\z

\ea\label{ex:nrexnoklkk}
Ukelipelup yakume urai wiyoro. Ono klokni mike kûno.\\
\gll uke-lipeli-p	yakume	urai	wiyo=ro	ono	klokni	mike	kûno\\
     jointly-seek-\textsc{pc}	woman.\textsc{sg}	crocodile	catch.\textsc{pp}=\textsc{emph}	man	one	first	go.down.\textsc{pp}\\
\glt `They were searching together for the woman (whom) the crocodile caught. One man went down first.'
\z

\ea\label{ex:nrexkowwwe}
Kukûnapno masalai yip riyo bili koweweno, masalai sano:\\
\gll k-kûne-p-no	masalai	yip	ri-yo	bili	kowe-we-no	masalai	sa-no\\
     \textsc{sub}-go.down-\textsc{pc}-\textsc{co}	bush.spirit(\textsc{tp})	house	\textsc{dist}-\textsc{loc}	opening	knock-\textsc{ter}-3\textsc{sg}.\textsc{or}.\textsc{pp}	bush.spirit(\textsc{tp})	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Having gone down to the bush spirit's house there, he knocked on its door and asked the bush spirit.'
\z

\ea\label{ex:nrexouunnn}
``Ako onopiro nake?''  ``Ou unake ou.''\\
\gll ako	ono-pi=ro	nake	ou	u-nake	ou\\
     wife	man-\textsc{poss}=\textsc{emph}	sit	yes	\textsc{dfac}-sit	yes\\
\glt `{``}Is the man's wife is staying there?'' ``Yes, she is here, yes.'''
\z

\ea\label{ex:nrexyalaar}
``De ko yala ar ûliyo name?'' ``De lap!''\\
\gll de	ko	yala	ar	ûliyo	name	de	le-p\\
     you	I	now	\textsc{neg}	inside	give.2\textsc{sg}.\textsc{or}	you	go-\textsc{imp}\\
\glt `{``}Won't you give her back to me from your house?''  ``You better go!'''
\z

\ea\label{ex:nrexnakuss}
Yena ki so solo onoro kûnapro. Sosolo sosolo sosoloro.\\
\gll yena	ki	so	solo	ono=ro	kûne-p=ro	sosolo	sosolo	sosolo=ro\\
     people	\textsc{aph}	like.this	only	man=\textsc{emph}	go.down-\textsc{pc}=\textsc{emph}	like.this	like.this	like.this=\textsc{emph}\\
\glt `The people (did nothing) but this. A man went down, like this, like this, like this.'
\z

\ea\label{ex:nrexlassma}
Las man boyoro kukûnapno riyopuno muelno:\\
\gll las	man	boyo=ro	k-kûne-p-no	riyopuno	mueli-no\\
     last(\textsc{tp})	man(\textsc{tp})	later=\textsc{emph}	\textsc{sub}-go.down-\textsc{pc}-\textsc{co}	then	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `When later the last man had gone down, he finally said to (the bush spirit):'
\z

\ea\label{ex:nrexkoakok}
``Uki kepro romusiyepo. Ko ako kepro lakiye.''\\
\gll uki	kep=ro	ro=musiye-po	ko	ako	kep=ro	lakiye\\
     husband	3\textsc{sg}.\textsc{poss}=\textsc{emph}	\textsc{emph}=send-\textsc{lv}.\textsc{pp}	I	wife	3\textsc{sg}.\textsc{poss}=\textsc{emph}	fetch\\
\glt `{``}Her husband sent me. I will fetch his wife.'''
\z

\ea\label{ex:nrexkialak}
Ou ako kep kero. De ki muelnap: ``Ki akûne. Ki alakiye, ki alakiyeke.''\\
\gll ou	ako	kep	ke=ro	de	ki	mueli-ne-p	ki	a-kûne	ki	a-lakiye	ki	a-lakiye-ke\\
     yes	wife	3\textsc{sg}.\textsc{poss}	\textsc{aph}=\textsc{emph}	you	\textsc{aph}	talk.to-3\textsc{sg}.\textsc{or}-\textsc{imp}	\textsc{aph}	\textsc{imp}3-go.down	\textsc{aph}	\textsc{imp}3-fetch	\textsc{aph}	\textsc{imp}3-fetch-\textsc{ingr}\\
\glt `{``}Yes, his wife, she ... Tell him: He (himself) should go down, he should fetch her! He  should go to fetch her.'''
\z

\ea\label{ex:nrexkkunoo}
Uki kepro ki muelno. Uki kepro kukûnapno muelne:\\
\gll uki	kep=ro	ki	mueli-no	uki	kep=ro	k-kûne-p-no	mueli-ne\\
     husband	3\textsc{sg}.\textsc{poss}=\textsc{emph}	\textsc{aph}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	husband	3\textsc{sg}.\textsc{poss}=\textsc{emph}	\textsc{sub}-go.down-\textsc{pc}-\textsc{co}	talk.to-3\textsc{sg}.\textsc{or}\\
\glt `(The last man) said (that) to her husband. Having gone down her husband says to (the bush spirit):'
\z

\ea\label{ex:nrexkaprka}
``Ko ako ikapro. De wepulo, de wepuloro yipyo. Nake?''\\
\gll ko	ako	ikap=ro	de	wepulo	de	wepulo=ro	yip-yo	nake\\
     I	wife	1\textsc{sg}.\textsc{poss}.\textsc{emph}=\textsc{emph}	you	bring.\textsc{pp}	you	bring.\textsc{pp}=\textsc{emph}	house-\textsc{loc}	sit\\
\glt `{``}I ... my wife ... You brought her. You brought her to your house. Is she there?'''
\z

\ea\label{ex:nrexhapsol}
``Ari. Ko banok. Hap solo lili. Hap solo ko parno lolo.\\
\gll ari	ko	ba-ni-ko	hap	solo	lili	hap	solo	ko	par-no	lolo\\
     no	I	\textsc{fac}-eat-\textsc{fac}	half(\textsc{tp})	only	be.there	half(\textsc{tp})	only	I	pannier-\textsc{ins}	tie.\textsc{pp}\\
\glt `{``}No, I have eaten her. There is only half (left), half only. I tied (her remains) on the pannier.'
\z

\ea\label{ex:nrexbakabk}
Yala ko de ponameipe. De yala wili baka, baka ko banok.''\\
\gll yala	ko	de	poname-ipe	de	yala	wili	baka	baka	ko	ba-ni-ko\\
     now	I	you	give.2\textsc{sg}.\textsc{or}-\textsc{ant}	you	now	carry	half	half	I	\textsc{fac}-eat-\textsc{fac}\\
\glt `Now I give her to you before (you go up again). You carry one half, the other half I have eaten up.'''
\z

\newpage
\ea\label{ex:nrexwiliwi}
Uki kep kukûnapno muelne: ``Ko wili.''\\
\gll uki	kep	k-kûne-p-no	mueli-ne	ko	wili\\
     husband	3\textsc{sg}.\textsc{poss}	\textsc{sub}-go.down-\textsc{pc}-\textsc{co}	talk.to-3\textsc{sg}.\textsc{or}	I	carry\\
\glt `Having come down her husband replies: ``I will carry (her up).'''
\z

\ea\label{ex:nrexparroi}
Par roise ponamo. Baka apulyo kisepamu. Baka solo kisepamu. Baka ki banok. Weppuo olyo.\\
\gll par	roise	ponamo	baka	apulyo	kisepamu	baka	solo	kisepamu	baka	ki	ba-ni-ko	weppuo	ol-yo\\
     pannier	together	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	half	in.the.middle	cut.off.hither.\textsc{pp}	half	only	cut.off.hither.\textsc{pp}	half	\textsc{aph}	\textsc{fac}-eat-\textsc{fac}	carry.up.\textsc{pp}	hill-\textsc{loc}\\
\glt `(The bush spirit) gave (her remains) to him together with the pannier. Half it (had) cut off in the middle. Only half it (had) cut off and eaten up. (The husband) carried (her remains) up to the river bank.'
\z

\ea\label{ex:nrexolywor}
Yena uki kepro wo roise kuppuapno olyo. Yena wo roise wel.\\
\gll yena	uki	kep=ro	wo\_roise	k-ppue-p-no	ol-yo	yena	wo\_roise	wel\\
     people	husband	3\textsc{sg}.\textsc{poss}=\textsc{emph}	crying	\textsc{sub}-go.up-\textsc{pc}-\textsc{co}	hill-\textsc{loc}	people	crying	carry.\textsc{pp}\\
\glt `The people and the husband climbed up to the river bank in tears,  some people carried (her remains) crying.'\footnote{Apparently several men are still down at the bush spirit's house in the river. Now they climb up to the river bank.}
\z

\ea\label{ex:nrexnoroha}
Masalai noro. Hap solo baka banok. Hap solo welro yilauyo.\\
\gll masalai	no=ro	hap	solo	baka	ba-ni-ko	hap	solo	wel=ro	yilau-yo\\
     bush.spirit(\textsc{tp})	eat=\textsc{emph}	half(\textsc{tp})	only	half	\textsc{fac}-eat-\textsc{fac}	half(\textsc{tp})	only	carry.\textsc{pp}=\textsc{emph}	village-\textsc{loc}\\
\glt `The bush spirit ate (half). Only half (they carried). One half it has eaten up,  only the (other) half they carried to the village.'
\z

\ea\label{ex:nrexplanim}
Mi yilauyo planimpo. Lupi solo kwilipno, mi yala rar.\\
\gll mi	yilau-yo	planim-po	lupi	solo	k-wili-p-no	mi	yala	rar\\
     then	village-\textsc{loc}	bury-\textsc{lv}.\textsc{pp}	piece	only	\textsc{sub}-carry-\textsc{pc}-\textsc{co}	then	now	dig.\textsc{pp}\\
\glt `Then they buried her in the village. Having carried only a part of her they dug now (a hole for burying).'
\z

\ea\label{ex:nrexemtatu}
Em tasol tumbuna stori.\\
\gll em	tasol	tumbuna	stori\\
     \textsc{aph}(\textsc{tp})	only(\textsc{tp})	ancestor(\textsc{tp})	story(\textsc{tp})\\
\glt `That's it, the ancestor's story.'
\z 
