\ea\label{ex:nrexmilaik}
Mi laik storim stori wantaim Jeffrey na David.\\
\gll mi	laik	storim	stori	wantaim	Jeffrey	na	David\\
     I(\textsc{tp})	want(\textsc{tp})	tell(\textsc{tp})	story(\textsc{tp})	with(\textsc{tp})	Jeffrey	and(\textsc{tp})	David\\
\glt `I want to tell a story (connected) with Jeffrey and David.'\footnote{Jeffrey is the second son of Lis Osi's first wife Helen; his older brother Paul went to Port Moresby and settled there. So Jeffrey is Lis Osi's successor in the village. David is the oldest son of Lis Osi's second wife Margaret. When mentioning both sons, Jeffrey and David, Margaret wants to reassure their equal rank in the family. Actually, David was promoting his public influence in the village by leading debates and monitoring quarrels. How exactly the story is connected with Jeffrey and David in practice remained obscure.}
\z

\ea\label{ex:nrexnanpak}
Wau nana pako puyo. Nana puyo seku.\\
\gll Wau	nana	pako	pu-yo	nana	pu-yo	seku\\
     Wau	small.knife	throw.\textsc{pp}	river-\textsc{loc}	small.knife	river-\textsc{loc}	fall.\textsc{pp}\\
\glt `Wau threw a small knife into the river. The knife fell into the river.'
\z

\ea\label{ex:nrexnanrok}
Nanaro kûnoro, ono pauweyo lelko, ono bepi.\\
\gll nana=ro	kûno=ro	ono	pauwe-yo	lili-ko	ono	bepi\\
     small.knife=\textsc{emph}	go.down.\textsc{pp}=\textsc{emph}	man	breast-\textsc{loc}	be.there-\textsc{rts}	man	old\\
\glt `The small knife sank down. (Now) it was on a man's breast, an old man's (breast).'
\z

\ea\label{ex:nrexwaubep}
Ono bepiro nana anno piyowe, nana ono bapi.\\
\gll ono	bepi=ro	nana	an-no	piyo-we	nana	ono	ba-pi\\
     man	old=\textsc{emph}	small.knife	hand-\textsc{ins}	take.\textsc{pp}-\textsc{ter}	small.knife	man	other-\textsc{poss}\\
\glt `The old man took the knife in his hand, the knife of another man.'\footnote{Meanwhile Wau went back to his place and village where he is asked about the knife.}
\z

\ea\label{ex:nrexauenan}
``Waue, nana ari?'' ``Nana puyo seku.'' ``De minalap, piyekap!''\\
\gll Wau-e	nana	ari	nana	pu-yo	seku	de	mi-inale-p	piye-ke-p\\
     Wau-\textsc{voc}	small.knife	no	small.knife	river-\textsc{loc}	fall.\textsc{pp}	you	\textsc{iter}-hurry-\textsc{imp}	take-\textsc{ingr}-\textsc{imp}\\
\glt `{``}Wau, the small knife, (is it there or) not?'' -- ``The small knife fell into the river.'' -- ``You hurry up again and go to get it!'''
\z

\ea\label{ex:nrexyallor}
``Yala ko asa le rope?'' Pu ikoipo, umul nekpamu.\\
\gll yala	ko	asa	le	ro-pe	pu	ikoi-po	umul\_nekpamu\\
     now	I	how	go	\textsc{prox}.\textsc{emph}-\textsc{q}	river	big-\textsc{lv}.\textsc{pp}	ponder.\textsc{pp}\\
\glt `{``}What am I going to do now because of this?'' (thinks Wau) The river was deep and he thought back and forth.'
\z

\ea\label{ex:nrexileiil}
Klapno puap moiko, puap ilei ileimainu.\\
\gll k-le-p-no	puap	moi-ko	puap	ilei	ileimainu\\
     \textsc{sub}-go-\textsc{pc}-\textsc{co}	wild.limbum	cut-\textsc{rts}	wild.limbum	long	very.long\\
\glt `He went away. He cut (a liana) of wild \textit{limbum}. A long liana, a very long one.'
\z

\ea\label{ex:nrexwepwep}
Komoipno wepulo, kwepulapno pako puyo.\\
\gll k-moi-p-no	wepulo	k-wepule-p-no	pako	pu-yo\\
     \textsc{sub}-cut-\textsc{pc}-\textsc{co}	bring.\textsc{pp}	\textsc{sub}-bring-\textsc{pc}-\textsc{co}	throw.\textsc{pp}	river-\textsc{loc}\\
\glt `Having cut it, he took it (to the river). Having brought it, he threw it into the river.'
\z

\ea\label{ex:nrexupuppu}
``O, upuna, ko yala upuna kûne.''\\
\gll o	upuna	ko	yala	upuna	kûne\\
     oh	well	I	now	well	go.down\\
\glt `{``}Oh, well, now I (can) go down properly.'''
\z

\ea\label{ex:nrexkewipp}
Lupi baka kewipp buripolip.\\
\gll lupi	baka	kewipp	buri\_poli-p\\
     end	other	be.vertically.tightened.above.the.water	go.ahead\_be.there-\textsc{pc}\\
\glt `The other end (of the rope) was tightened vertically above the water. There it starts.'\footnote{The liana rope that Wau fastened to a tree reminds one of ``magic ropes'' which spirits use. These ropes are tied up to a tree and secure the spirits’ healthy return to their place. (cf. \cite[44; 57]{Kiki:1969tx})}
\z

\ea\label{ex:nrexunorok}
Riyopuno Wau kûno, kûnoro kûnoro kûnoro.\\
\gll riyopuno	Wau	kûno	kûno=ro	kûno=ro	kûno=ro\\
     then	Wau	go.down.\textsc{pp}	go.down.\textsc{pp}=\textsc{emph}	go.down.\textsc{pp}=\textsc{emph}	go.down.\textsc{pp}=\textsc{emph}\\
\glt `Then Wau went down, further down, further down, further down.'
\z

\ea\label{ex:nrexayobia}
Epul kep an sayo, bialil ulili.\\
\gll epul	kep	an	sayo	bia\_lil	u-lili\\
     ear	3\textsc{sg}.\textsc{poss}	hand	touch.\textsc{pp}	corpse.blood	\textsc{dfac}-be.there\\
\glt `(His) hand touched his ear: The (magic) blood of a 
corpse is there.'
\z

\ea\label{ex:nrexigodau}
Em i go daun, kupuapno pulupako.\\
\gll em	i	go	daun	k-pue-p-no	pulupako\\
     \textsc{aph}(\textsc{tp})	\textsc{pred}(\textsc{tp})	go(\textsc{tp})	down(\textsc{tp})	\textsc{sub}-roam-\textsc{pc}-\textsc{co}	spit.out.\textsc{pp}\\
\glt `He went down. While he was gliding (through the water), he spit (the magic blood) out.'\footnote{He did that in order to make the water opaque that he cannot be seen.}
\z

\ea\label{ex:nrexbepire}
Ono bepi reye paeau.\\
\gll ono	bepi	reye	paeau\\
     man	old	see.\textsc{o}.\textsc{anim}.\textsc{sg}	arrive\\
\glt `He becomes aware of the old man and arrives (at his place).'
\z

\ea\label{ex:nrexnanlip}
``A de asa pulo?'' ``A ko nana lipeli.''\\
\gll a	de	asa	pulo	a	ko	nana	lipeli\\
     ah	you	how	come.\textsc{pp}	ah	I	small.knife	seek\\
\glt `{``}Ah, why did you come?'' -- ``Ah, I am looking for the small knife.'''
\z

\ea\label{ex:nrexnanaul}
``Nana ko piyowe, ou. Nana ulili, we. De minalap! Yena wokyo molo, yena mi yala dupulupi. De inalap!\\
\gll nana	ko	piyo-we	ou	nana	u-lili	we	de	mi-inale-p	yena	wok-yo	molo	yena	mi	yala	d-pulupi	de	inale-p\\
     small.knife	I	take.\textsc{pp}-\textsc{ter}	yes	small.knife	\textsc{dfac}-be.there	look	you	\textsc{iter}-hurry-\textsc{imp}	people	work-\textsc{loc}	go.\textsc{pl}.\textsc{pp}	people	again	\textsc{mod}	\textsc{lkh}-come.\textsc{pl}	you	hurry-\textsc{imp}\\
\glt `{``}The small knife, I took it. Yes, here it is, the small knife, look! Hurry up again! The people went to work, but they will soon come again. Hurry up!'''\footnote{There seems to be a sort of underwater settlement in the river, where some people live and move back and forth for work. The nature of these people is not known except they are a threat for the diving man Wau. The `under-water-world' wasn't described any further by the narrator.}
\z

\ea\label{ex:nrexriapya}
Yala de kilimpi, de minalap, hariapya! De inappuap, yala de suellui.''\\
\gll yala	de	kilim-pi	de	mi-inale-p	hariap=ya	de	inappue-p	yala	de	sueli\_lui\\
     now	you	kill(\textsc{tp})-\textsc{lv}	you	\textsc{iter}-quickly.go-\textsc{imp}	hurry.up(\textsc{tp})=\textsc{emph}	you	climb.up.quickly-\textsc{imp}	now	you	cut\_kill\\
\glt `{``}They will kill you. Hurry up, hurry up! Go up quickly! They'll kill you by cutting (the rope).'''
\z

\ea\label{ex:nrexwaupon}
Ono bepi sû Wau ponamo.\\
\gll ono	bepi	sû	Wau	ponamo\\
     man	old	fire	Wau	give.3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `The old man gave Wau a firestick.'
\z

\ea\label{ex:nrexwausur}
Wau ppuero sû roise. Sû bepue solo, sû bamuriko. Mi kuppuero, kuppuero, kuppuero. Kamap olyo, paeau, eppinoyap.\\
\gll Wau	ppue-ro	sû	roise	sû	bepue	solo	sû	ba-muri-ko	mi	k-ppue=ro	k-ppue=ro	k-ppue=ro	kamap	ol-yo	paeau	eppi\_noye-p\\
     Wau	go.up-\textsc{emph}	fire	together	fire	smouldering.piece.of.firewood	only	fire	\textsc{fac}-extinguish-\textsc{fac}	again	\textsc{aph}-go.up=\textsc{emph}	\textsc{aph}-go.up=\textsc{emph}	\textsc{aph}-go.up=\textsc{emph}	reach(\textsc{tp})	hill-\textsc{loc}	arrive	rest-\textsc{pc}\\
\glt `Wau is going up with the fire, with a glowing piece of firewood only. The flame has  extinguished. He is going up again, going up, going up. He reaches the river bank,  arrives (there). He was resting.'
\z

\ea\label{ex:nrexwauupi}
Ri lupi sukerpop: ``O Wau upini.''\\
\gll ri	lupi	suker\_pi-p	o	Wau	u-pini\\
     tree	end	creak-\textsc{pc}	oh	Wau	\textsc{dfac}-come.up.hither\\
\glt `The end of the tree was creaking: ``Oh, Wau is right coming up here.'''\footnote{This is the inference of his not yet introduced companion who watched him from the riverbank.}
\z

\newpage
\ea\label{ex:nrexwausan}
Kenekipno yipyo lo, sano:\\
\gll k-neki-p-no	yip-yo	lo	sa-no\\
     \textsc{sub}-stand-\textsc{pc}-\textsc{co}	house-\textsc{loc}	go.\textsc{pp}	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Having stood there (for a while), (Wau) went (back) to the house. (There) they asked him.'
\z

\ea\label{ex:nrexwauede}
``Waue, de nana bapiyeko?'' ``Ko bapiyeko, we, nana mi oki.''\\
\gll Wau-e	de	nana	ba-piye-ko	ko	ba-piye-ko	we	nana	mi	o-ki\\
     Wau-\textsc{voc}	you	small.knife	\textsc{fac}-take-\textsc{fac}	I	\textsc{fac}-take-\textsc{fac}	look	small.knife	again	\textsc{prox}-\textsc{aph}\\
\glt `{``}Wau, you have got the small knife?'' -- ``I have got it, look, the small knife, here it is again.'''
\z

\ea\label{ex:nrexkurwau}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `End of the story.'
\z
