\ea\label{ex:nrexyukume}
Yukume sele ikoi po, sele bekulu yorer.\\
\gll yukume	sele	ikoi	po	sele	bekulu	yorer\\
     man.\textsc{sg}	garden	big	do.\textsc{pp}	garden	huge	very.far.away\\
\glt `A man laid out a big garden. A large garden, stretching far.'
\z

\ea\label{ex:nrexpaisol}
Pewo layewepu pasok pailili paisol panpal paipines klapua pamapup paipopo. Suel dui yasiyo, ap yasiyo, war yasiyo, opse pues elo puel kupp kaukau.\\
\gll pewo	laye-wepu	pasok	pailili	paisol	panpal	paipines	klapua	pamapup	paipopo	suel	dui	yasiyo	ap	yasiyo	war	yasiyo	opse	pues	elo	puel	kupp	kaukau\\
     banana	lay-\textsc{quant}.\textsc{o}.\textsc{pp}	kind.of.banana	kind.of.banana	kind.of.banana	kind.of.banana	kind.of.banana	kind.of.banana	kind.of.banana	kind.of.banana	tobacco	shoot	plant.\textsc{pp}	spinach	plant.\textsc{pp}	wild.sugar.cane	plant.\textsc{pp}	stick.taro	kind.of.taro	sugar.cane	betel.nut	yam	sweet.potato\\
\glt `He laid banana (plantations) out, \textit{pasok}, \textit{pailili}, \textit{paisol}, \textit{panpal}, \textit{paipines}, \textit{klapua}, \textit{pamapup}, \textit{paipopo}. He planted tobacco shoots, and he planted spinach. He planted wild sugar cane, stick taro, sugar cane, betel nuts, yams, and sweet potatos.'
\z

\ea\label{ex:nrexrapuek}
Rapue kiniyo maki powepu.\\
\gll rapue	kiniyo	maki	pi-wepu\\
     vegetables	all	good	\textsc{lv}-\textsc{quant}.\textsc{s}.\textsc{pp}\\
\glt `All vegetables were good (for harvesting).'
\z

\ea\label{ex:nrexwisyak}
Wîs yako pulo. Sele riyoulipop, ``Eh pewo bamakiwepuko.  Bamakipoko.''\\
\gll wîs	yako	pulo	sele	riye-uli-pi-p	eh	pewo	ba-maki-wepi-ko	ba-maki-pi-ko\\
     moon	woman	come.\textsc{pp}	garden	see.\textsc{o}.\textsc{inanim}-\textsc{prog}-\textsc{lv}-\textsc{pc}	eh	banana	\textsc{fac}-good-\textsc{quant}.\textsc{s}-\textsc{fac}	\textsc{fac}-good-\textsc{lv}-\textsc{fac}\\
\glt `The moon woman came and was looking at the garden: ``Eh, the bananas are all good.  They have become ripe.'''
\z

\ea\label{ex:nrexropkun}
Yipyo lo, neppi wepulo. Pewo sukewepu, yipyo mel. Midorilo, ba rapiyoko. Suelno apno yesi roisero, puel war ropyo niskûno, yipyo mel.\\
\gll yip-yo	lo	neppi	wepulo	pewo	suke-wepu	yip-yo	mel	mi-dorilo	ba	rapiye-ko	suel-no	ap-no	yesi	roise=ro	puel	war	rop-yo	niskûno	yip-yo	mel\\
     house-\textsc{loc}	go.\textsc{pp}	bush.knife	bring.\textsc{pp}	banana	cut-\textsc{quant}.\textsc{o}.\textsc{pp}	house-\textsc{loc}	carry.\textsc{pl}.\textsc{o}.\textsc{pp}	\textsc{iter}-go.back	other	fetch.\textsc{pp}-\textsc{rts}	tobacco-\textsc{ins}	spinach-\textsc{ins}	aibika	together=\textsc{emph}	betel.nut	wild.sugar.cane	basket-\textsc{loc}	fill.in.\textsc{pp}	house-\textsc{loc}	carry.\textsc{pl}.\textsc{o}.\textsc{pp}\\
\glt `She went to (her) house, took a bush knife, cut plenty of bananas, and carried them home. She went back (to the garden) and fetched other (food plants). With tobacco, spinach, and \textit{aibika} (she left). She filled betel nuts and wild sugar cane in baskets and carried everything home.'
\z

\ea\label{ex:nrexyipbis}
Rapue bakesiyowolewepuko. Yip bî solo polip.\\
\gll rapue	ba-kesiye-wole-wepi-ko	yip	bî	solo	poli-p\\
     food	\textsc{fac}-use.up-\textsc{cpl}-\textsc{quant}.\textsc{s}-\textsc{fac}	house	hole	only	be.there-\textsc{pc}\\
\glt `The food was completely consumed. The house was empty.'
\z

\ea\label{ex:nrexsakkan}
Yukume seleyo lo sele kepyo. ``Eh pewo kopi ana suko? Yala ko asa pi? Am ko nake. Ko dob powe, ko sowewe. Ono yala ana puleke? Saka ana piulipi?''\\
\gll yukume	sele-yo	lo	sele	kep-yo	eh	pewo	ko-pi	ana	suko	yala	ko	asa	pi	am	ko	nake	ko	dob	pi-we	ko	sowe-we	ono	yala	ana	pule-ke	saka	ana	pi-uli-pi\\
     man.\textsc{sg}	garden-\textsc{loc}	go.\textsc{pp}	garden	3\textsc{sg}.\textsc{poss}-\textsc{loc}	eh	banana	1\textsc{sg}-\textsc{poss}	who	cut.\textsc{pp}	now	I	how	do	still	I	stay	I	eye	\textsc{lv}-\textsc{ter}	I	hide-\textsc{ter}	person	now	who	come-\textsc{ingr}	secretly	who	do-\textsc{prog}-\textsc{lv}\\
\glt `The man went to the garden, to his garden: ``Eh, who cut my bananas? What am I going to do now? I'll stay on. I will watch closely and hide well. What man is going to come now? Who does (this) secretly?'''
\z

\ea\label{ex:nrexpasokp}
Yakoro puni pulo seleyo. Yukumero pasok boliyo nekip. ``Eh yakumero lero pewo riyepi.'' Pewo rodupua rodupua suko, pasok klokni, an kiniyo royowe yeloyo.\\
\gll yako=ro	puni	pulo	sele-yo	yukume=ro	pasok	boliyo	neki-p	eh	yakume=ro	le=ro	pewo	riye-pi	pewo	rodupua\_rodupua	suko	pasok	klokni	an	kiniyo	royo-we	yelo-yo\\
     woman=\textsc{emph}	night	come.\textsc{pp}	garden-\textsc{loc}	man.\textsc{sg}=\textsc{emph}	kind.of.banana	at.the.foot	stand-\textsc{pc}	eh	woman.\textsc{sg}=\textsc{emph}	go=\textsc{emph}	banana	see.\textsc{o}.\textsc{inanim}-\textsc{lv}	banana	four	cut.\textsc{pp}	kind.of.banana	one	hand	all	lay.\textsc{pp}-\textsc{ter}	ground-\textsc{loc}\\
\glt `At night a woman came in the garden. The man was standing at the foot of a \textit{pasok}-banana: ``Eh, a woman is going (around) looking for the bananas.'' She cut four bananas and a \textit{pasok}-banana, five (bunches of bananas). She put them on the ground.'
\z

\ea\label{ex:nrexkikipi}
Royowe yeloyo. Yukume nekipno an pulapu, anno wiyo, kikipiyo.\\
\gll royo-we	yelo-yo	yukume	neki-p-no	an	pulapu	an-no	wiyo	kikipiyo\\
     lay.\textsc{pp}-\textsc{ter}	ground-\textsc{loc}	man.\textsc{sg}	stand-\textsc{pc}-\textsc{co}	hand	strain.\textsc{pp}	hand-\textsc{ins}	hold.\textsc{pp}	hold.strong.\textsc{pp}\\
\glt `She put them on the ground, while the man was standing (there). He reached out (his) arms and grabbed her with his hands. He held her tightly.'
\z

\ea\label{ex:nrexwiswiy}
Wîs yako wiyo anno, ``Ere ko de bareyeko! Saka de piulipop. Yalaka dedukoyo ile yipyo.''\\
\gll wîs	yako	wiyo	an-no	ere	ko	de	ba-reye-ko	saka	de	pi-uli-pi-p	yalaka	dedukoyo	i-le	yip-yo\\
     moon	woman	hold.\textsc{pp}	hand-\textsc{ins}	now	I	you	\textsc{fac}-see.\textsc{o}.\textsc{anim}.\textsc{sg}-\textsc{fac}	secretly	you	\textsc{lv}-\textsc{prog}-\textsc{lv}-\textsc{pc}	now	we.\textsc{du}.\textsc{incl}	\textsc{du}.\textsc{s}-go	house-\textsc{loc}\\
\glt `He got hold of the moon woman with (both) hands: ``Now I have seen you. You did (that) always secretly. Now we two go together to the house.'''
\z

\ea\label{ex:nrexwisyip}
Wîs yako muelno: ``Ari, yalaka yip ikapyo ilei.'' Yip kepyo iloi. Yip kepyo paeau, inakap.\\
\gll wîs	yako	mueli-no	ari	yalaka	yip	ikap-yo	i-le-i	yip	kep-yo	i-lo-i	yip	kep-yo	paeau	i-nake-p\\
     moon	woman	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	no	now	house	1\textsc{sg}.\textsc{poss}.\textsc{emph}-\textsc{loc}	\textsc{du}.\textsc{s}-go-\textsc{du}.\textsc{s}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	arrive	\textsc{du}.\textsc{s}-live-\textsc{pc}\\
\glt `The moon woman said to him: ``No, we'll go to my house.'' They went to her house. They arrived at her house and were living there.'
\z

\ea\label{ex:nrexwiseln}
Inakero. Wîs yako elno. Mi yilau ukipiyo iloi. Euryo uliyoko, ako ipei kep muelnap:\\
\gll i-nake=ro	wîs	yako	el\_no	mi	yilau	uki-pi-yo	i-lo-i	eur-yo	uliyo-ko	ako	ipei	kep	mueli-ne-p\\
     \textsc{du}.\textsc{s}-live=\textsc{emph}	moon	woman	pregnant	then	village	husband-\textsc{poss}-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	toilet-\textsc{loc}	leave.behind.\textsc{pp}-\textsc{rts}	wife	first.ranking	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}-\textsc{pc}\\
\glt `They live there, and the moon woman gets pregnant. Then they went to the man's village. He had left her at the toilet-house and (then) was talking to his first wife.'
\z

\ea\label{ex:nrexeuryoy}
``De muli? Ko ako ba wepulo.'' ``De aryo uliyoko?'' ``Ko uliyoko. Ou, ko de samap.'' ``Ko muli ou. Lakiyekap! Koyo kumune inake.''\\
\gll de	muli	ko	ako	ba	wepulo	de	aryo	uliye-ko	ko	uliye-ko	ou	ko	de	sa-me-p	ko	muli	ou	lakiye-ke-p	koyo	kumune	i-nake\\
     you	want	I	wife	other	bring.\textsc{pp}	you	where	leave.behind-\textsc{rts}	I	leave.behind-\textsc{rts}	yes	I	you	ask-2\textsc{sg}.\textsc{or}-\textsc{pc}	I	want	yes	fetch.people-\textsc{ingr}-\textsc{imp}	we.\textsc{du}.\textsc{excl}	all.\textsc{coll}	\textsc{du}.\textsc{s}-live\\
\glt `{``}Do you agree (that) I brought another wife?'' -- ``Where did you leave her?'' -- ``I left her (over there). Yes, I wanted to ask you.'' -- ``I agree, yes. Go get her! We'll live together.'''
\z

\ea\label{ex:nrexipeiaw}
Lakiyoko, yipyo woko. Ako ipeiro muelno: ``Awe, de yo nakap! Ko ika nake, de rika nake. Inakap.''\\
\gll lakiye-ko	yip-yo	woko	ako	ipei=ro	mueli-no	awe	de	yo	nake-p	ko	i-ka	nake	de	ri-ka	nake	i-nake-p\\
     fetch.people-\textsc{rts}	house-\textsc{loc}	accompany.\textsc{pp}	wife	first.ranking=\textsc{emph}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	come.\textsc{imp}	you	location	sit-\textsc{imp}	I	\textsc{dist}-\textsc{path}	sit	you	\textsc{dist}.\textsc{emph}-\textsc{path}	sit	\textsc{du}.\textsc{s}-live-\textsc{pc}\\
\glt `He took her to the house, accompanied her. The first wife said to her: ``Come! You live in this place. I live here, you live there.'' (So) they lived on.'
\z

\ea\label{ex:nrexrurduk}
Wîs yako ruri nako, roipi nako. Nakoro. Duku po, ekuno wolip, ireri.\\
\gll wîs	yako	ruri	nako	roipi	nako	nako=ro	duku\_po	eku-no	woli-p	ireri\\
     moon	woman	child	gave.birth	boy	gave.birth	gave.birth=\textsc{emph}	laugh\_\textsc{lv}.\textsc{pp}	behind-\textsc{ins}	sit-\textsc{pc}	crawl\\
\glt `The moon woman gave birth to a child, to a boy. She gave birth to a boy. He laughed, he was able to sit, he crawled.'
\z

\ea\label{ex:nrexwisyae}
Uki ako ipei roise yaeauyo iloi. ``Nuko ako ppusi muelnei.'' ``Ari yipyo kranake. Nuko kama mole yaeauyo.''\\
\gll uki	ako	ipei	roise	yaeau-yo	i-lo-i	nuko	ako	ppusi	mueli-ne-i	ari	yip-yo	kra-nake	nuko	kama	mole	yaeau-yo\\
     husband	wife	first.ranking	with	traditional.feast-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	we.\textsc{incl}	wife	second.ranking	talk.to-3\textsc{sg}.\textsc{or}-\textsc{du}.\textsc{a}	no	house-\textsc{loc}	\textsc{niv}-stay	we.\textsc{incl}	alone	go.\textsc{pl}	traditional.feast-\textsc{loc}\\
\glt `The man went to a feast together with his first wife. (The first wife said:) ``We'll tell the second wife.'' (The husband replied:) ``No, let her stay at home. We'll go alone to the feast.'''
\z

\ea\label{ex:nrexipeimu}
Wîs yako ako ipei muelno: ``Ruri ko uliyewayap, ruri depi! Ruri kopiro de wokap!''\\
\gll wîs	yako	ako	ipei	mueli-no	ruri	ko	uliyewaye-p	ruri	de-pi	ruri	ko-pi=ro	de	woke-p\\
     moon	woman	wife	first.ranking	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	child	I	leave.for-\textsc{imp}	child	2\textsc{sg}-\textsc{poss}	child	1\textsc{sg}-\textsc{poss}=\textsc{emph}	you	go.together.with-\textsc{imp}\\
\glt `The moon woman told the first wife: ``Leave me the child, your child! Go together with my child!'''
\z

\ea\label{ex:nrexiwawel}
Smep musikono. Riyopuno iwa puyo wel, pu ipiyo yipyo wel. Yip sipul pusapopo yenino.\\
\gll smep	musi-ko-ne	riyopuno	iwa	pu-yo	wel	pu	ipiyo	yip-yo	wel	yip	sipul	pusapopo	yeni-no\\
     door	lock-\textsc{rts}-3\textsc{sg}.\textsc{or}	then	bucket	river-\textsc{loc}	carry.\textsc{pp}	water	bucket.\textsc{pp}	house-\textsc{loc}	carry.\textsc{pp}	house	floor	wash.\textsc{pl}.\textsc{o}.\textsc{pp}	board-\textsc{ins}\\
\glt `She had locked the door behind her. Then she took buckets to the river and carried the buckets of water to the house. She cleaned the floor and the boards in the house.'
\z

\ea\label{ex:nrexepkewi}
Wolo luliwepu, wolo luliwepu appyo. App moniseso poli. Wepul pino: wepul ba ekuyo, wepul ba elyo, wepul ba waliyo. Wîs yako ruri epkewi papeno.\\
\gll wolo	luli-wepu	wolo	luli-wepu	app-yo	app	moniseso	poli	wepul	pi-no	wepul	ba	eku-yo	wepul	ba	el-yo	wepul	ba	wali-yo	wîs	yako	ruri	epkewi\_pape-no\\
     ladder	fence-\textsc{quant}.\textsc{o}.\textsc{pp}	ladder	fence-\textsc{quant}.\textsc{o}.\textsc{pp}	sky-\textsc{loc}	sky	very.little	be.there	baby.sling	make-3\textsc{sg}.\textsc{or}.\textsc{pp}	baby.sling	other	behind-\textsc{loc}	baby.sling	other	belly-\textsc{loc}	baby.sling	other	neck-\textsc{loc}	moon	woman	child	carry.piggyback-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `She made a long ladder. She made a ladder up to the sky. The sky is very close (now). She made a baby sling for the child: a sling in the back, a sling in the front, and a sling around the neck. The moon woman carried the child piggyback.'
\z

\ea\label{ex:nrexaisuso}
Weppuo wolo mono. Dob seku, uki wonkûneno: ``Ai ai ai, ko ke ule.'' Sûso popana.\\
\gll weppuo	wolo	mono	dob	seku	uki	wonikûne-no	ai	ai	ai	ko	ke	u-le	sû-so	popana\\
     carry.up.\textsc{pp}	ladder	path	eye	fall.\textsc{pp}	husband	call.down-3\textsc{sg}.\textsc{or}.\textsc{pp}	ai	ai	ai	I	\textsc{top}	\textsc{dfac}-go	fire-\textsc{sim}	take.away.thither.\textsc{pp}\\
\glt `She carried him up along the ladder, looked down, and called down to her husband: ``Ai, ai, ai, here I go.'' She vanished like fire.'
\z

\ea\label{ex:nrexwisrle}
``Ako kopi rleya. Umul ko asa poro? Ko umul pole.''\\
\gll ako	ko-pi	r-le=ya	umul	ko	asa	pi=ro	ko	umul\_pole\\
     wife	1\textsc{sg}-\textsc{poss}	\textsc{dist}.\textsc{emph}-go=\textsc{emph}	heart	I	how	do=\textsc{emph}	I	be.sad\\
\glt `(The husband lamented:) ``My wife goes away over there. How does my heart feel? I am sad.'''
\z

\ea\label{ex:nrexmukari}
``Ko muelmouro nuko kumune mole yaeauyo. De mi ba muli. De umul ppulae po. De pusampi. Yako maki, yako mukari. Nuko yako pari, yako yilau bayanapi. Rope de asa pono? Balok.''\\
\gll ko	mueli-me-ou=ro	nuko	kumune	mole	yaeau-yo	de	mi	ba	muli	de	umul	ppulae\_po	de	pusampi	yako	maki	yako	mukari	nuko	yako	pari	yako	yilau	bayana-pi	rope	de	asa	po-no	ba-le-ko\\
     I	talk.to-2\textsc{sg}.\textsc{or}-\textsc{frus}=\textsc{emph}	we.\textsc{incl}	all.\textsc{coll}	go.\textsc{pl}	traditional.feast-\textsc{loc}	you	again	\textsc{neg}.\textsc{emph}	want	you	heart	bad\_\textsc{lv}.\textsc{pp}	you	keep.a.woman.away.from.people	woman	good	woman	young.and.pretty	we.\textsc{incl}	woman	\textsc{neg}.\textsc{cop}	woman	place	different-\textsc{poss}	why	you	how	do.\textsc{pp}-3\textsc{sg}.\textsc{or}.\textsc{pp}	\textsc{fac}-go-\textsc{fac}\\
\glt `(The first wife replied): ``I said in vain to you (that) we go to the feast together. You did not want (it). You're coldhearted. You kept her away from other people. The woman is kind(hearted), and she is young and pretty. She is not woman of us, she is the woman of some different place. Why did you do this to her? She has gone (for good).'''
\z

\ea\label{ex:nrexawauwi}
Wîs appyo pin. Appyo banakepisiko ruri kep roise, dari ruri Awau.\\
\gll wîs	app-yo	pin	app-yo	ba-nake-pisi-ko	ruri	kep	roise	dari	ruri	Awau\\
     moon	sky-\textsc{loc}	come.up.hither.\textsc{pp}	sky-\textsc{loc}	\textsc{fac}-stay-\textsc{cpl}-\textsc{fac}	child	3\textsc{sg}.\textsc{poss}	with	older.sister	child	Awau\\
\glt `The moon came up hither to the sky. (The moon woman) stays forever in the sky together with her child. (It's) the child of the older sister, (its name is) Awau.'\footnote{The first wife is regarded as the older sister of the moon woman. The child Awau might refer to the evening star. Occasionally, moon and Venus appear close together in the night sky. Yet this hypothesis could not be verified.}
\z

\ea\label{ex:nrexwismon}
Ani wîs ar puli. Dupuni wîs puli. Wîs maki, ono mono poname. Wîs duwaniso makina puli. Mono de makina riye, de makina le.\\
\gll ani	wîs	ar	puli	dupuni	wîs	puli	wîs	maki	ono	mono	poname	wîs	duwani-so	maki-na	puli	mono	de	maki-na	riye	de	maki-na	le\\
     day	moon	\textsc{neg}	shine	night	moon	shine	moon	good	man	path	give.3\textsc{sg}.\textsc{or}	moon	daylight-\textsc{sim}	good-\textsc{adv}	shine	path	you	good-\textsc{adv}	see.\textsc{o}.\textsc{inanim}	you	good-\textsc{adv}	go\\
\glt `During the day the moon doesn't shine. It shines in the night. The moon is (something) good. It gives people the path (to walk). The moon is like daylight, it shines brightly. You see (your) path well and walk safely.'
\z

\ea\label{ex:nrexisboku}
Wîs bo kuru.\\
\gll wîs	bo	kuru\\
     moon	story	be.finished\\
\glt `The moon story is finished.'
\z