\ea\label{ex:nrexwisduk}
Kimike dukiro wîs appyo ar nakoulipop. Yena nini solo riyepop.\\
\gll kimike	duki=ro	wîs	app-yo	ar	nake-uli-pi-p	yena	nini	solo	riye-pi-p\\
     before	true=\textsc{emph}	moon	sky-\textsc{loc}	\textsc{neg}	sit-\textsc{prog}-\textsc{lv}-\textsc{pc}	people	sun	only	see.\textsc{o}.\textsc{inanim}-\textsc{lv}-\textsc{pc}\\
\glt `A long time ago the moon was not in the sky. The people saw only the sun.'
\z

\ea\label{ex:nrexwoknin}
Yena wok ninino solo poulipop. Yena wal hukno ninino solo poulipop. Yena due ninino solo poulipop.\\
\gll yena	wok	nini-no	solo	pi-uli-pi-p	yena	wal	huk-no	nini-no	solo	pi-uli-pi-p	yena	due	nini-no	solo	pi-uli-pi-p\\
     people	work(\textsc{tp})	sun-\textsc{ins}	only	do-\textsc{prog}-\textsc{lv}-\textsc{pc}	people	fish	hook.fish(\textsc{tp})-\textsc{ins}	sun-\textsc{ins}	only	do-\textsc{prog}-\textsc{lv}-\textsc{pc}	people	sago	sun-\textsc{ins}	only	do-\textsc{prog}-\textsc{lv}-\textsc{pc}\\
\glt `People did (their) work only under the sun. People hooked fish only during daytime, and they processed sago only during daytime.'
\z

\ea\label{ex:nrexyakpun}
Riyopuno yakume wok puni poulipop. Huk puni poulipop.\\
\gll riyopuno	yakume	wok	puni	pi-uli-pi-p	huk	puni	pi-uli-pi-p\\
     then	woman.\textsc{sg}	work	night	do-\textsc{prog}-\textsc{lv}-\textsc{pc}	hook.fish(\textsc{tp})	night	do-\textsc{prog}-\textsc{lv}-\textsc{pc}\\
\glt `Then a certain woman was working at night. She was hooking fish at night.'
\z

\ea\label{ex:nrexikiwak}
Yena iki wako muellayepo: ``Yakume oke wok puni poulipi.''\\
\gll yena	iki	wako	muel-laye-po	yakume	o-ke	wok	puni	pi-uli-pi\\
     people	\textsc{aph}.\textsc{pl}	amongst	talk.to-\textsc{recp}-\textsc{lv}.\textsc{pp}	woman.\textsc{sg}	\textsc{prox}-\textsc{aph}	work(\textsc{tp})	night	do-\textsc{prog}-\textsc{lv}\\
\glt `The people talked among each other: ``This woman always works at night.'''
\z

\ea\label{ex:nrexwisany}
Yena yakume wulino. Yakume ki wîs anyo laye. Du kla, dupuni piye.\\
\gll yena	yakume	wuli-no	yakume	ki	wîs	an-yo	laye	du	kla	dupuni	piye\\
     people	woman.\textsc{sg}	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}	woman.\textsc{sg}	\textsc{aph}	moon	hand-\textsc{loc}	lay	darkness	transparent	night	take\\
\glt `The people followed the woman. (This) woman puts the moon into her hand. The darkness (becomes) light. She takes the night (away).'
\z

\ea\label{ex:nrexwisriy}
Yakume ki mi wîs riyo layowe. Ki wok sele pop.\\
\gll yakume	ki	mi	wîs	ri-yo	layo-we	ki	wok	sele	pi-p\\
     woman.\textsc{sg}	\textsc{aph}	then	moon	tree-\textsc{loc}	lay.\textsc{pp}-\textsc{ter}	\textsc{aph}	work(\textsc{tp})	garden	do-\textsc{pc}\\
\glt `Then the woman put the moon up in a tree and was doing (her) garden work.'
\z

\ea\label{ex:nrexwismay}
Yena kikeno muelnoro: ``Wîs balso, piyemaye.''\\
\gll yena	kike-no	mueli-no=ro	wîs	bal-so	piye-maye\\
     people	run-\textsc{co}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}=\textsc{emph}	moon	ball-\textsc{sim}	take-\textsc{mal}\\
\glt `While the people were running after her they said: ``The moon is like a ball, we'll steal it.'''
\z

\ea\label{ex:nrexwispiy}
Yakume ki piyowe.\\
\gll yakume	ki	piyo-we\\
     woman.\textsc{sg}	\textsc{aph}	take.\textsc{pp}-\textsc{ter}\\
\glt `The woman held it firmly.'
\z

\ea\label{ex:nrexwisdus}
Riyopuno yena wîs piyemayo dusoipo. Appyo banakepisiko.\\
\gll riyopuno	yena	wîs	piye-mayo	dusoi-po	app-yo	ba-nake-pisi-ko\\
     then	people	moon	take-\textsc{mal}.\textsc{pp}	let.go.off-\textsc{lv}.\textsc{pp}	sky-\textsc{loc}	\textsc{fac}-sit-\textsc{cpl}-\textsc{fac}\\
\glt `Then the people stole the moon. (But) they let it go away. It is in the sky forever.'
\z

\ea\label{ex:nrexispuli}
Yala nuko muli: ``Wîs puli.''\\
\gll yala	nuko	muli	wîs	puli\\
     now	we.\textsc{incl}	say	moon	shine\\
\glt `Now we say: ``The moon shines.'''
\z

\ea\label{ex:nrexkubbri}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `End of the story.'
\z