\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Kimike dukiro wîs appyo ar nakoulipop. Yena nini solo riyepop. Yena wok ninino solo poulipop. Yena wal hukno ninino solo poulipop. Yena due ninino solo poulipop. Riyopuno yakume wok puni poulipop. Huk puni poulipop. Yena iki wako muellayepo: ``Yakume oke wok puni poulipi.'' Yena yakume wulino. Yakume ki wîs anyo laye. Du kla, dupuni piye. Yakume ki mi wîs riyo layo\-we. Ki wok sele pop. Yena kikeno muelnoro: ``Wîs balso, piyemaye.'' Yaku\-me ki piyowe. Riyopuno yena wîs piyemayo, dusoipo. Appyo banakepisiko. Yala nuko muli: ``Wîs puli.''}}
    \ParallelRText{A long time ago the moon was not in the sky. The people saw only the sun. People did their work only under the sun. People hooked fish only during day time, and people made sago only during daytime. Then a certain woman was working at night. She was hooking fish at night. The people talked among each other: ``This woman works always at night.'' The people followed her. The woman grasps the moon, and the darkness becomes light. She takes the night away. Then the woman puts the moon up in a tree and was doing her garden work. While the people were running after her they said: ``The moon is like a ball, we'll steal it.'' The woman held it fast. But finally the people stole the moon. But they let it go away, and so it is in the sky forever. Now we say: ``The moon shines.''}
\end{Parallel}
    \medskip
\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Bo kuru.}}
    \ParallelRText{End of the story.}
\end{Parallel}
    \medskip

