\ea\label{ex:nrexiwanpe}
Iwan ke bike muelno: ``Depe asa nake rileyo riyo? De yala ar nake rileyo. De yala yeloyo k没ni.''\\
\gll iwan	ke	bike	muel-no	de-pe	asa	nake	rileyo	ri-yo	de	yala	ar	nake	rileyo	de	yala	yelo-yo	k没ni\\
     hornbill	\textsc{aph}	cassowary	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	2\textsc{sg}-\textsc{q}	how	sit	above	tree-\textsc{loc}	you	now	\textsc{neg}	sit	above	you	now	ground-\textsc{loc}	come.down.hither\\
\glt `The hornbill said to the cassowary: ``How do you live above in the trees? Now you won't live above (anymore). Now you come down here to the ground.'''
\z

\ea\label{ex:nrexiwanpa}
Iwan paki bikepi weno. Bike yeloyo k没no.\\
\gll iwan	paki	bikepi	we-no	bike	yelo-yo	k没no\\
     hornbill	wing	cassowary-\textsc{poss}	break-3\textsc{sg}.\textsc{or}.\textsc{pp}	cassowary	ground-\textsc{loc}	come.down.hither.\textsc{pp}\\
\glt `The hornbill broke the cassowary's wings. The cassowary came down here to the ground.'
\z

\ea\label{ex:nrexiwanri}
Ri ini rupuo. Bike yeloyo ri roise seku. Iwan ri rupueno.\\
\gll ri	ini	rupuo	bike	yelo-yo	ri	roise	seku	iwan	ri	rupue-no\\
     tree	branch	break.\textsc{pp}	cassowary	ground-\textsc{loc}	wood	together	fall.\textsc{pp}	hornbill	tree	break-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `(The hornbill) broke the branch of the tree. The cassowary fell to the ground together with the branch. The hornbill broke the tree branch (where it was sitting).'
\z

\ea\label{ex:nrexiwanki}
Bike paki kana wayou. Iwan paki bikepi baweneko, asa lam.\\
\gll bike	paki	kana	waye-ou	iwan	paki	bike-pi	ba-we-ne-ko	asa	le-m\\
     cassowary	wing	quickly	try-\textsc{frus}	hornbill	wing	cassowary-\textsc{poss}	\textsc{fac}-break-3\textsc{sg}.\textsc{or}-\textsc{fac}	how	go-\textsc{pos}\\
\glt `Quickly the cassowary tried (to move) his wings, but in vain. (The hornbill) has broken the cassowary's wings. It cannot fly (anymore).'
\z

\ea\label{ex:nrexiwanwo}
Bike yeloyo nakeko. Bike paki wewolono.\\
\gll bike	yelo-yo	nake-ko	bike	paki	we-wole-no\\
     cassowary	ground-\textsc{loc}	sit-\textsc{rts}	cassowary	wing	break-\textsc{cpl}-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `The cassowary sat on the ground since then. (The hornbill) broke its wings completely.'
\z

\ea\label{ex:nrexiwanna}
Iwan rileyo lo. Riyo nakeko.\\
\gll iwan	rileyo	lo	ri-yo	nake-ko\\
     hornbill	above	go.\textsc{pp}	tree-\textsc{loc}	sit-\textsc{rts}\\
\glt `The hornbill went above and sat in the tree since then.'
\z

\ea\label{ex:nrexiwanba}
Riyopuno banakepisiko. Bike liki iwanpi piyo, iwan liki bikepi piyo.\\
\gll riyopuno	ba-nake-pisi-ko	bike	liki	iwan-pi	piyo	iwan	liki	bike-pi	piyo\\
     then	\textsc{fac}-live-\textsc{cpl}-\textsc{fac}	cassowary	designated.place	hornbill-\textsc{poss}	take.\textsc{pp}	hornbill	designated.place	cassowary-\textsc{poss}	take.\textsc{pp}\\
\glt `Then they finally live for ever (in the following way): The cassowary took over the hornbill's place, and the hornbill took over the cassowary's place.'
\z