\ea\label{ex:t7exbiepdu}
Epe aino kopi duyo ilo. Duyo inakap, bi lipeliwepup. Fopela lu, bike dupua biep dupua, bisem rodupua rokini, bipuel an kinika. Kuru, ba ar lu.\\
\gll epe	ai-no	ko-pi	du-yo	i-lo	du-yo	i-nake-p	bi	lipeli-wepi-p	fopela	lu	bike	dupua	biep	dupua	bisem	rodupua\_rokini	bipuel	an\_kinika	kuru	ba	ar	lu\\
     mother	father-\textsc{ins}	1\textsc{sg}-\textsc{poss}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-stay-\textsc{pc}	animal	seek-\textsc{quant}.\textsc{o}-\textsc{pc}	four(\textsc{tp})	shoot.\textsc{pp}	cassowary	two	boar	two	kangaroo	three	tree.kangaroo	five	be.finished	other	\textsc{neg}	shoot.\textsc{pp}\\
\glt `My parents went to the bush. They were staying in the bush and were looking for  	game. (My father) shot four (big animals), two cassowaries, and two boars. (Then he shot) three kangaroos and five tree kangaroos. (That's) enough, other (animals) he didn't shoot.'
\z

\ea\label{ex:t7exaplmer}
Aplayo royewepu wapo. Sû mappo bi slau po. Ba mererpi ba mererile.\\
\gll apla-yo	roye-wepu	wapo	s{û}\_mappo	bi	slau	po	ba	mererpi	ba	mereri\_le\\
     plank-\textsc{loc}	put-\textsc{quant}.\textsc{o}.\textsc{pp}	put.side.by.side.\textsc{pp}	light.fire.\textsc{pp}	meat	dry	\textsc{lv}.\textsc{pp}	some	cut.with.sharp.bone.\textsc{pl}.\textsc{o}	some	cut.with.sharp.bone.\textsc{pl}.\textsc{o}\_go\\
\glt `(They) put (the animals) on planks. They put (the planks) side by side and lit fires (underneath the planks). The meat became dry. Some they cut up with a sharp bone (to eat in the bush). Some they cut up with a sharp bone to go (and bring home).'
\z

\ea\label{ex:t7eximiyue}
Imiyu epul malo: ``Duyo inake. Apai duyo nake.'' Imiyu molo, dob reyekûpo. Dorimolo.\\
\gll imiyu	epul	malo	du-yo	i-nake	Apai	du-yo	nake	imiyu	molo	dob	reye\_kûpo	dorimolo\\
     sorcerer	ear	hear.\textsc{pp}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-stay	Apai	bush-\textsc{loc}	stay	sorcerer	go.\textsc{pl}.\textsc{pp}	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\_go.down.\textsc{pl}.\textsc{a}.\textsc{pp}	go.back.\textsc{pl}.\textsc{pp}	\\
\glt `The sorcerers heard (something): ``The two stay in the bush. Apai is staying in the bush.'' The sorcerers went (there) and looked down on him. (Then) they went back.'
\z

\ea\label{ex:t7eximiyud}
Riyopuno imiyu doripulup. Puyo lelio pul mopip. Epe kopi dob ar reyo, sakana lelio.\\
\gll riyopuno	imiyu	doripulup	pu-yo	lelio	pul\_mopi-p	epe	ko-pi	dob	ar	reyo	sakana	lelio\\
     then	sorcerer	come.back.\textsc{pl}.\textsc{pp}	river-\textsc{loc}	kill.\textsc{pp}	bathe-\textsc{pc}	mother	1\textsc{sg}-\textsc{poss}	eye	\textsc{neg}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	secretly	kill.\textsc{pp}\\
\glt `Then the sorcerers came back. They killed (my father) at the river. He was bathing. My mother did not see him (ensorcelled). They killed him secretly.'
\z

\ea\label{ex:t7eximiyur}
Imiyu rauyo bia lilno. Puana. Ai kopi puana, yipyo lo.\\
\gll imiyu	rauyo	bia\_lil-no	puana	ai	ko-pi	puana	yip-yo	lo\\
     sorcerer	shake.awake.\textsc{pp}	magic.blood-\textsc{ins}	stand.up.\textsc{pp}	father	1\textsc{sg}-\textsc{poss}	stand.up.\textsc{pp}	house-\textsc{loc}	go.\textsc{pp}\\
\glt `The sorcerers rouse him by means of the blood of a dead person. He stood up. 	My father got up and went to (his) (bush)house.'\footnote{A dead person's blood has various powers. Cf. \textref{sc0308}, Sequence \ref{ex:nrexayobia}.}
\z

\ea\label{ex:t7exnakepw}
Nakap. Wik klokni nakap. Riyopuno mar, ikoina mar.\\
\gll nake-p	wik	klokni	nake-p	riyopuno	mar	ikoi-na	mar\\
     stay-\textsc{pc}	week	one	stay-\textsc{pc}	then	be.sick.\textsc{pp}	big-\textsc{adv}	be.sick.\textsc{pp}\\
\glt `He was staying on. For one week he was staying on. Then he became sick. He was very sick.'
\z

\ea\label{ex:t7exsipyar}
Sîp Yar eme umul nek: ``Yala dusui. Ko woke kanuno Yar yilauyo.''\\
\gll Sîp	Yar	eme	umul\_nek	yala	d-sui	ko	woke	kanu-no	Yar	yilau-yo\\
     Sîp	Yar	place.of.origin	think.\textsc{pp}	soon	\textsc{lkh}-die	I	go.together.with	canoe(\textsc{tp})-\textsc{ins}	Yar	village-\textsc{loc}\\
\glt `Sîp, a man from Ninggera, thought: ``He will die soon. I'll go together with him   	to Ninggera by canoe.'''\footnote{The village of Yar is now commonly called Ninggera.}
\z

\largerpage
\ea\label{ex:t7exdueklo}
Due klokni Yar yilauyo nuro duruwa. Punipino woko kanuno, puwayo. Paeau.\\
\gll due	klokni	Yar	yilau-yo	nu-ro	duruwa	punipino	woko	kanu-no	puwa-yo	paeau\\
     night	one	Yar	village-\textsc{loc}	sleep.\textsc{pp}-\textsc{emph}	daybreak	morning	go.together.with.\textsc{pp}	canoe(\textsc{tp})-\textsc{ins}	mangrove.swamp-\textsc{loc}	arrive\\
\glt `One night he slept in Ninggera until daybreak. In the morning he paddled together with  	him through the mangrove swamp. They arrived (in Ninggera).'
\z

\ea\label{ex:t7exkanula}
Kanu layeko, wakayo puo, yipyo woko, yipyo uliyowe.\\
\gll kanu	laye-ko	waka-yo	puo	yip-yo	woko	yip-yo	uliyo-we\\
     canoe(\textsc{tp})	put-\textsc{rts}	shoulder-\textsc{loc}	roam	house-\textsc{loc}	go.together.with.\textsc{pp}	house-\textsc{loc}	leave.behind.\textsc{pp}-\textsc{ter}\\
\glt `He left the canoe and walked (carrying) him on the shoulders. (Sîp) went with him to  	(his) house and left him in the house.'
\z

\ea\label{ex:t7exsiploo}
Sîp lo, opo lipelip. Nuni kopi reyana, muelno: ``Ai Margaretpi mari. Mari ikoina pi. Yala dusuiro.''\\
\gll Sîp	lo	opo	lipeli-p	nuni	ko-pi	reyana	mueli-no	ai	Margaret-pi	mari	mari	ikoi-na	pi	yala	d-sui=ro\\
     Sîp	go.\textsc{pp}	car	seek-\textsc{pc}	maternal.uncle	1\textsc{sg}-\textsc{poss}	meet.\textsc{pp}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	father	Margaret-\textsc{poss}	be.sick	be.sick	big-\textsc{adv}	\textsc{lv}	soon	\textsc{lkh}-die=\textsc{emph}\\
\glt `Sîp left and was looking for a car. He met my uncle and said to him: ``Margaret's father is sick. He is very sick and may just die.'''
\z

\ea\label{ex:t7exopovan}
Opo Vanimoyo lo. Opo lo puwayo. Ai kopi epe kopi disei kopi lekiko, ukepulo Pasiyo.\\
\gll opo	Vanimo-yo	lo	opo	lo	puwa-yo	ai	ko-pi	epe	ko-pi	disei	ko-pi	leki-ko	uke-pulo	Pasi-yo\\
     car	Vanimo-\textsc{loc}	go.\textsc{pp}	car	go.\textsc{pp}	mangrove.swamp-\textsc{loc}	father	1\textsc{sg}-\textsc{poss}	mother	1\textsc{sg}-\textsc{poss}	brother	1\textsc{sg}-\textsc{poss}	fetch.\textsc{pl}.\textsc{o}-\textsc{rts}	jointly-come.\textsc{pp}	Pasi-\textsc{loc}\\
\glt `The car came from Vanimo. The car went along the mangrove swamp. They picked up my father, my mother, and my brother and came together to Pasi.'
\z

\ea\label{ex:t7exyoeppi}
Kiniyo eppi noyo. Op mek kep nikiyemayo, muelien: ``Imiyu ba leliekûpo. Yala suem. Yilauyo wokeyep.''\\
\gll kiniyo	eppi\_noyo	Op	mek	kep	nikiye-mayo	mueli-en	imiyu	ba	lelie\_kûpo	yala	sui-m	yilau-yo	woke-yep\\
     all	rest.\textsc{pp}	Op	mouth	3\textsc{sg}.\textsc{poss}	smell-\textsc{mal}.\textsc{pp}	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	sorcerer	other	kill\_go.down.\textsc{pl}.\textsc{a}.\textsc{pp}	soon	die-\textsc{pos}	village-\textsc{loc}	go.together.with-\textsc{imp}.\textsc{pl}\\
\glt `All rested. Op, (Margaret's uncle), smelled his mouth and said to them: ``Some sorcerers went down killing him. He may die. All of you go together to the village.'''
\z

\ea\label{ex:t7exoponow}
Opono wepulup. Monoyo suiouro. Mipuana, mi pu namo.\\
\gll opo-no	wepulup	mono-yo	sui-ou=ro	mi-puana	mi	pu	namo\\
     car-\textsc{ins}	bring.\textsc{pl}.\textsc{a}.\textsc{pp}	road-\textsc{loc}	die-\textsc{frus}=\textsc{emph}	\textsc{iter}-rise.\textsc{pp}	again	water	give.3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `They brought him by car. He must not die on the road. He rose again (a bit), 	and they gave him some water again.'
\z

\ea\label{ex:t7exyipkep}
Yip kepyo wepulup. Opoyo lakiyepakûpo yipyo woko. Yeni kepyo nuiweyo.\\
\gll yip	kep-yo	wepulup	opo-yo	lakiye\_pakûpo	yip-yo	woko	yeni	kep-yo	nuiweyo\\
     house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	bring.\textsc{pl}.\textsc{pp}	car-\textsc{loc}	fetch\_put.down.\textsc{pl}.\textsc{a}.\textsc{pp}	house-\textsc{loc}	go.together.with.\textsc{pp}	bed	3\textsc{sg}.\textsc{poss}-\textsc{loc}	bed.\textsc{pp}\\
\glt `They drove him to his house. (There) they put him down and went together to the house. 	They bedded him on his bed.'
\z

\ea\label{ex:t7expunisu}
Puni sui. Bo ruri kep muelneyep: ``Ai kep basuiko.''\\
\gll puni	sui	bo	ruri	kep	mueli-ne-yep	ai	kep	ba-sui-ko\\
     night	die	word	child	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}-\textsc{imp}.\textsc{pl}	father	3\textsc{sg}.\textsc{poss}	\textsc{fac}-die-\textsc{fac}\\
\glt `During the night he dies. (People said:) ``Tell his daughter that her father has died.'''
\z

\ea\label{ex:t7exkomuel}
Ko muelien: ``Pu ikoina yopo. Ko ruri moniseso. Ko punipino pule kanuno. Ko kanuno siami.''\\
\gll ko	mueli-en	pu	ikoi-na	yopo	ko	ruri	moniseso	ko	punipino	pule	kanu-no	ko	kanu-no	siami\\
     I	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	river	big-\textsc{adv}	rise.\textsc{pp}	I	child	very.small	I	morning	come	canoe(\textsc{tp})-\textsc{ins}	I	canoe(\textsc{tp})-\textsc{ins}	cross.hither\\
\glt `I told them: ``The river rose high. I have a baby. I'll come in the morning by canoe.  I will cross (the river) by canoe.'''
\z

\ea\label{ex:t7exwoikoi}
Ko wo mop, wo ikoina mop. Yena bî rar. Yeni wopiyo, yeni pop. Matres roise piyeppuo, yeniyo uleilayekûpu.\\
\gll ko	wo\_mop	wo\_ikoina\_mop	yena	bî	rar	yeni	wopiyo	yeni	pi-p	matres	roise	piyeppuo	yeni-yo	uleilayekûpu\\
     I	cry.\textsc{pp}	cry.much.\textsc{pp}	people	hole	dig.\textsc{pp}	coffin	get.ready.\textsc{pp}	coffin	make-\textsc{pc}	mattress(\textsc{tp})	together	lift.up.\textsc{pp}	coffin-\textsc{loc}	put.inside.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}\\
\glt `I cried. I cried desperately. The people dug a hole and got a coffin ready. They were making it. (Then) they lifted him together with the mattress and put him down into the coffin.'
\z

\ea\label{ex:t7exmatmat}
Matmatyo wel bili kumau. Bî pakûpo, sowo. Ri wapo, ri royo, sowo. Basoweko. Yipyo molowe.\\
\gll matmat-yo	wel	bili	kumau	bî	pakûpo	sowo	ri	wapo	ri	royo	sowo	ba-sowe-ko	yip-yo	molo-we\\
     place.of.burial(\textsc{tp})-\textsc{loc}	carry.\textsc{pp}	opening	shut	hole	throw.down.\textsc{pl}.\textsc{a}.\textsc{pp}	cover.\textsc{pp}	stick	put.together.\textsc{pp}	stick	lay.\textsc{pp}	cover.\textsc{pp}	\textsc{fac}-cover-\textsc{fac}	house-\textsc{loc}	go.\textsc{pl}.\textsc{pp}-\textsc{ter}\\
\glt `They carried him to the burial place and shut the opening (of the coffin). They let it down into the hole, covered it. (Then) they put sticks together (as a small fence). They covered (the grave). It is covered, and they went to (their) houses.'
\z

\ea\label{ex:t7exaikuru}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `The story is finished.'
\z
