\ea\label{ex:t7exdueyol}
Epe dueyo lo ruri moni roise. Wepulyo ule, dueyo woko. Due wosinip. Ruri nuweyewe yeniyo, epe due sinip.\\
\gll epe	due-yo	lo	ruri	moni	roise	wepul-yo	ule	due-yo	woko	due	wo-sini-p	ruri	nuweye-we	yeni-yo	epe	due	sini-p\\
     mother	sago-\textsc{loc}	go.\textsc{pp}	child	little	together	cloth-\textsc{loc}	put.into.\textsc{pp}	sago-\textsc{loc}	accompany.\textsc{pp}	sago	\textsc{accom}-pound.sago.pith-\textsc{pc}	child	lay.down.for.resting-\textsc{ter}	platform-\textsc{loc}	mother	sago	pound.sago.pith-\textsc{pc}\\
\glt `The mother went to the sago swamp together with the little child. She put it in the  baby sling. They went together to the sago swamp. She was pounding sago in company  (of the child). (Then) she laid the child down on a platform to sleep. The mother continued pounding sago.'
\z

\ea\label{ex:t7ex}
Yei wopiyo. Due nek wena, yeiyo lul. Ruri due nuip, nek lulip. Nek aska, milupiko. Miwemon, lul, due lul. Nek mi ke kuru. Milo, milupiko, miwepulo. Sosoli nakap. Riyopuno nek kesiyo. Soruo kopue roise pusiyo.\\
\gll yei	wopiyo	due	nek	wena	yei-yo	lul	ruri	due\_nui-p	nek	luli-p	nek	aska	mi-lupi-ko	mi-wemon	lul	due	lul	nek	mi	ke	kuru	mi-lo	mi-lupi-ko	mi-wepulo	sosoli	nake-p	riyopuno	nek	kesiyo	soruo	kopue	roise	pusiyo\\
     kneading.trough.for.sago.washing	get.ready.\textsc{pp}	sago	sago.pith	carry.inside.\textsc{pp}	kneading.trough.for.sago.washing-\textsc{loc}	mix.with.water.\textsc{pp}	child	sleep-\textsc{pc}	sago.pith	mix.with.water-\textsc{pc}	sago.pith	none	\textsc{iter}-shovel.by.hand-\textsc{rts}	\textsc{iter}-bring.hither.\textsc{pp}	mix.with.water.\textsc{pp}	sago	mix.with.water.\textsc{pp}	sago.pith	again	\textsc{aph}	be.finished	\textsc{iter}-go.\textsc{pp}	\textsc{iter}-shovel.by.hand-\textsc{rts}	\textsc{iter}-bring.\textsc{pp}	like.this	stay-\textsc{pc}	then	sago.pith	use.up.\textsc{pp}	filter.net	trough	together	wash.\textsc{pp}\\
\glt `She got the trough for kneading and washing the sago ready. She carried sago pith to 	(the washing place) and washed it in the kneading trough. The child was sleeping, and she was washing the sago pith. (There is) no pith left (anymore). Again she reshovelled some, brought it and washed it. She washed the sago pith. The pith is finished again. She went again, reshovelled and brought more. (So) she was busy like this. Then she (had) used up the sago pith and cleaned the filtering net and the trough.'
\z

\ea\label{ex:t7exekunak}
Eku nakap. Due ep amalûpi. Due ep balûpiko. Ep sipamu, dû piyamu. Dû sepue roise piyamu rop soruo roise. Dû ropyo pokûno.\\
\gll eku	nake-p	due	ep	am-a-lûpi	due	ep	ba-lûpi-ko	ep	sipamu	dû	piyamu	dû	sepue	roise	piyamu	rop	soruo	roise	dû	rop-yo	pokûno\\
     behind	sit-\textsc{pc}	sago	rinse.water	\textsc{grad}-\textsc{imp}3-clean	sago	rinse.water	\textsc{fac}-clean-\textsc{fac}	rinse.water	tip.away.hither.\textsc{pp}	flour	take.hither.\textsc{pp}	flour	trough	together	take.hither.\textsc{pp}	basket	filter.net	together	flour	basket-\textsc{loc}	fill.in.\textsc{pp}\\
\glt `She sat down (for a short while). First the rinse water (of the sago pith) has to be clean. The rinse water has become clean. She tipped out the dirty rinse water and took the (sago) flour. She took the flour and the trough together, and the basket with the filtering net. (Finally) she poured the flour into the basket.'
\z

\ea\label{ex:t7expuanaw}
Ruri puana, wo mop, ba wo mop. Epe ba namo, ruri ba nep. Wepul moniyo laek, ruri wepulyo ule. Due dû piapo rop roise. Rop wakayo puakuyo puo. Yipyo woko.\\
\gll ruri	puana	wo\_mop	ba	wo\_mop	epe	ba	namo	ruri	ba	ni-p	wepul	moni-yo	laek	ruri	wepul-yo	ule	due\_dû	piapo	rop	roise	rop	waka-yo	puaku-yo	puo	yip-yo	woko\\
     child	wake.up.\textsc{pp}	cry.\textsc{pp}	breast	cry.\textsc{pp}	mother	breast	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	child	breast	eat-\textsc{pc}	cloth	neck-\textsc{loc}	put.around.\textsc{pp}	child	cloth-\textsc{loc}	put.into.\textsc{pp}	sago.flour	lift.up.\textsc{pp}	basket	together	basket	shoulder-\textsc{loc}	head-\textsc{loc}	walk.around.\textsc{pp}	house-\textsc{loc}	accompany.\textsc{pp}\\
\glt `The child woke up and cried. She cried for her breast. The mother gave her her breast, and the child was drinking. (The mother) put the cloth around her neck and the child into the baby sling. She lifted up the sago flour together with the baskets. She carried the baskets on shoulders and head. (Then) they went in (each other's) company (back) to the house.'
\z

\ea\label{ex:t7exepekek}
Buai, Epek. Epek dob reyepue ruri moniro. Yipyo wokoro. Dû rop isiye wolako, due rop.\\
\gll Buai	Epek	Epek	dob	reyepue	ruri	moni=ro	yip-yo	woko=ro	dû	rop	isiye	wolako	due	rop\\
     Buai	Epek	Epek	eye	watch.\textsc{o}.\textsc{anim}.\textsc{sg}	child	little=\textsc{emph}	house-\textsc{loc}	accompany.\textsc{pp}=\textsc{emph}	flour	basket	together	set.down.\textsc{pp}	sago	basket\\
\glt `Buai, Epek (the ancestral spirits). Epek watches her. He accompanies the little child to the house. (The mother) set down the flour with the baskets, the sago baskets.'
\z

\ea\label{ex:t7exmamoun}
Riyopuno yaup yowo sûyo. Yaup mol. Due piyo wîlyo, ya mappo. Ya sui. Ya lulpamu mamouno. Supopo rupueyo. Rapue si. Rapue bareko. Ipi isiye puliyo, wîlyo rupopo.\\
\gll riyopuno	yaup	yowo	sû-yo	yaup	mol	due	piyo	wîl-yo	ya\_mappo	ya\_sui	ya	lulpamu	mamou-no	supopo	rupue-yo	rapue	si	rapue	ba-re-ko	ipi	isiye	puliyo	wîl-yo	rupopo\\
     then	water.for.cooking	heat.\textsc{pp}	fire-\textsc{loc}	water.for.cooking	boil.\textsc{pp}	sago	take.\textsc{pp}	dish-\textsc{loc}	stir.sago.\textsc{pp}	sago.be.ready	sago	portion.sago.\textsc{pp}	wooden.sticks-\textsc{ins}	cut.into.pieces.\textsc{pp}	leaf-\textsc{loc}	vegetables	cook	vegetables	\textsc{fac}-be.done-\textsc{fac}	pot	together	take.off.\textsc{pp}	plate-\textsc{loc}	distribute.\textsc{pl}.\textsc{o}.\textsc{pp}\\
\glt `Then she heated water on the fire. The water boiled. She took sago in a dish and stirred it in. The sago is ready. She portioned the sago with wooden sticks and cut it into pieces. (She put the portions) on leaves. She cooked vegetables. The vegetables are done. She took them together with the pot (from the fire) and distributed them on the plates.'
\z

\ea\label{ex:t7exroyeen}
Ruri kep royeen ai kep roise, kimiso. Iki kiniyo ilo ya roise. Baileko.\\
\gll ruri	kep	roye-en	ai	kep	roise	ki-miso	iki	kiniyo	ilo	ya	roise	ba-ile-ko\\
     child	3\textsc{sg}.\textsc{poss}	give-\textsc{nsg}.\textsc{or}.\textsc{pp}	father	3\textsc{sg}.\textsc{poss}	together	\textsc{aph}-also	\textsc{aph}.\textsc{pl}	all	eat.\textsc{pl}.\textsc{a}.\textsc{pp}	sago	together	\textsc{fac}-eat.\textsc{pl}.\textsc{a}-\textsc{fac}\\
\glt `She gave (food) to her children and their father. And also to herself. They all ate (the vegetables) with sago. They have eaten up (all the food).'
\z

\ea\label{ex:t7exbanamo}
Ruri moni wo mop. Epe ba namo. Ba ar no, pusayepop. Ba pusayap.\\
\gll ruri	moni	wo\_mop	epe	ba	namo	ba	ar	no	pusaye-pi-p	ba	pusaye-p\\
     child	little	cry.\textsc{pp}	mother	breast	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	breast	\textsc{neg}	eat.\textsc{pp}	reject-\textsc{lv}-\textsc{pc}	breast	reject-\textsc{pc}\\
\glt `The little child cried. The mother gave her her breast, but (the baby) didn't take her breast. She was rejecting it. She was rejecting the breast.'
\z

\ea\label{ex:t7exepekdo}
Epek dob reyepuo. Ep yeniyo royewolap, yeni epiyo wapewolap. Pael pulupup.\\
\gll Epek	dob	reyepuo	ep	yeni-yo	roye\_wole-p	yeni	epiyo	wape\_wole-p	pael	pulu-pi-p\\
     Epek	eye	watch.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	kind.of.leaf	bed-\textsc{loc}	lay\_move.further-\textsc{pc}	bed	at.the.edge	put.together\_move.further-\textsc{pc}	chewed.bark.of.tree	spit-\textsc{lv}-\textsc{pc}\\
\glt `(The ancestral spirit) Epek watched (the child). (The parents) put \textit{ep}-leaves around the bed. They were putting them at the edges of the bed. (Then) they were spitting chewed bark on the child.'
\z

\ea\label{ex:t7exsineik}
Ruri wo basneiko. Epek dob reyepuo. Bamolko. Wo sina, monemno nake. Ba ere ni, upuna nake.\\
\gll ruri	wo	ba-sinei-ko	Epek	dob	reyepuo	ba-mole-ko	wo	sina	monemno	nake	ba	ere	ni	upuna	nake\\
     child	crying	\textsc{fac}-become.silent-\textsc{fac}	Epek	eye	watch.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	\textsc{fac}-go.\textsc{pl}-\textsc{fac}	crying	become.silent.\textsc{pp}	quiet	sit-\textsc{pc}	breast	now	eat	alright	sit\\
\glt `The child has become silent. The ancestral spirits (had) watched it. (Now) they have gone. The crying stopped. (The baby) stays quiet. Now she takes the breast and is fine.'\footnote{It is not clear whether Epek is a single spirit or a group of disturbing ancestral spirits. In Sequence (\ref{ex:t7exsineik}) we have the plural form \textit{bamolko} whose subject is Epek.}
\z