\ea\label{ex:t5exkoyodu}
Koyo dueyo iloi, epe ko ikap roise. Koyo due wosonip. Epe kopi sonip, ko eku nakap, dob ponap.\\
\gll koyo	due-yo	i-lo	epe	ko\_ikap	roise	koyo	due	wo-soni-p	epe	ko-pi	soni-p	ko	eku	nake-p	dob	pi-ne-p\\
     we.\textsc{du}.\textsc{excl}	sago-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	mother	I.myself	with	we.\textsc{du}.\textsc{excl}	sago	\textsc{accom}-pound.sago.pith-\textsc{pc}	mother	1\textsc{sg}-\textsc{poss}	pound.sago.pith-\textsc{pc}	I	behind	sit-\textsc{pc}	eye	\textsc{lv}-3\textsc{sg}.\textsc{or}-\textsc{pc}\\
\glt `We went to the sago swamp, I together with my mother. We pounded sago pith together. My mother was pounding. I was sitting down (on a palm rib) and watching her.'
\z

\ea\label{ex:t3exnekkau}
Nek kau ropyo lupi. Ropyo lupuwapo, yeiyo meli. Sepue royepana pekolyo. Sepue apulyo lelweip.\\
\gll nek	kau	rop-yo	lupi	rop-yo	lupiwapo	yei-yo	meli	sepue	royepana	pekol-yo	sepue	apulyo	lili-we-p\\
     sago.pith	full	basket-\textsc{loc}	shovel.by.hand	basket-\textsc{loc}	shovel.by.hand.a.lot.\textsc{pp}	kneading.trough.for.sago.washing-\textsc{loc}	carry.\textsc{pl}.\textsc{o}	trough	put.thither.\textsc{pp}	circle.of.wooden.sticks-\textsc{loc}	trough	in.the.middle	be.there-\textsc{ter}-\textsc{pc}\\
\glt `(There is) plenty of sago pith which she shovels into a basket. She shovelled (pith) into several baskets and carried it to the kneading trough for sago washing. She put the trough in a circle of wooden sticks. The trough was stably placed in the middle of it.'
\z

\ea\label{ex:t3exsaryow}
Riyopuno saryo wapo. Due pul sepueyo kûno. Pu ipiyo nekyo sipepo. Sipako, due lulip. Kululilpno nek kesiyowolo. Nek lupuapoko, melipulo, yeiyo lulip.\\
\gll riyopuno	sar-yo	wapo	due	pul	sepue-yo	kûno	pu	ipiyo	nek-yo	sipepo	sipako	due	luli-p	k-luli-p-no	nek	kesiye-wolo	nek	lupiwapo-ko	melipulo	yei-yo	luli-p\\
     then	palm.rib.container-\textsc{loc}	put.together.\textsc{pp}	sago	fruit	trough-\textsc{loc}	go.down.\textsc{pp}	water	bucket.\textsc{pp}	sago.pith-\textsc{loc}	pour.on.top.\textsc{pp}	pour.down.into.\textsc{pp}	sago	mix.with.water-\textsc{pc}	\textsc{sub}-mix.with.water-\textsc{pc}-\textsc{co}	sago.pith	use.up-\textsc{cpl}.\textsc{pp}	sago.pith	shovel.by.hand.a.lot-\textsc{rts}	bring.\textsc{pl}.\textsc{o}.\textsc{pp}	kneading.trough.for.sago.washing-\textsc{loc}	mix.with.water-\textsc{pc}\\
\glt `Then she put (sago pith)  into a palm rib container. The pith went down into the  	trough. She bucketed water and poured it on the sago pith. She poured it down and was washing the sago. Having washed (for some time), the sago pith was used up. (Then) she shovelled more sago pith, brought it, and was washing it in the trough.'
\z

\ea\label{ex:t3exelsuik}
Due lulip. Due lulip. Ko nakap. Ko unakap. Ko nakoro, epe kopiro ko muelno: ``Ko el sui. Ko lelo piu lipeli, riyepi.''\\
\gll due	luli-p	due	luli-p	ko	nake-p	ko	u-nake-p	ko	nako=ro	epe	ko-pi=ro	ko	muel-no
ko	el\_sui	ko	lelo	piu	lipeli	riye-pi\\
     sago	mix.with.water-\textsc{pc}	sago	mix.with.water-\textsc{pc}	I	sit-\textsc{pc}	I	\textsc{dfac}-sit-\textsc{pc}	I	sit.\textsc{pp}=\textsc{emph}	mother	1\textsc{sg}-\textsc{poss}=\textsc{emph}	I	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp} I	be.hungry	I	gecko	frog	seek	see.\textsc{o}.\textsc{inanim}-\textsc{lv}\\
\glt `She was washing sago ... She was washing sago. I was sitting. I was just sitting. I sat (there) and said to my mother: ``I feel hungry. I will seek geckos and frogs. I will look for them.'''
\z

\ea\label{ex:t3exkonumo}
Ko lo, ko numomo ppuo. Ko piu riyepo lelo roise. Ko ppue dob seku: piu unake. Ko wiyo, ko pako yeloyo.\\
\gll ko	lo	ko	numomo	ppuo	ko	piu	riye-po	lelo	roise	ko	ppuo	dob	seku	piu	u-nake	ko	wiyo	ko	pako	yelo-yo\\
     I	go.\textsc{pp}	I	sago.palm.species	climb.\textsc{pp}	I	frog	see.\textsc{o}.\textsc{inanim}-\textsc{lv}.\textsc{pp}	gecko	together	I	climb.\textsc{pp}	eye	fall.\textsc{pp}	frog	\textsc{dfac}-sit	I	hold.\textsc{pp}	I	throw.\textsc{pp}	ground-\textsc{loc}\\
\glt `I went. I climbed a \textit{numomo}-sago palm and looked for frogs and geckos. I climbed up and looked down (into a palm rib): Here is a frog. I caught it and threw it on the ground.'\footnote{Grammatical comment: In standard linguistic semantics, constructions with \textit{seek} are considered ambiguous due to a scope distinction: in the specific reading of \textit{x seeks y}, $y$ has wide scope over \textit{seek}, meaning that there is some (specific) $y_i$ such that $x$ seeks $it_i$. The non-specific reading has this order reversed, in the sense of $x$ seeking some or other $y$ with a certain property. In Kilmeri, by contrast, there is no scope ambiguity involved at all. Instead, in the non-specific reading of some or other frog being sought the feature [-\textsc{anim}] is assigned via the verb form \textit{riye} in the second clause, and once a specific frog is talked about, we find the feature [+\textsc{anim}] via the choice of the existential-postural verb \textit{nake}, in the fourth clause of Sequence (\ref{ex:t3exkonumo}).}
\z

\ea\label{ex:t3exkweepi}
Numomo kwe epi baka ko dob seku: lelo unake. Ko wiyo, ko pako yeloyo.\\
\gll numomo	kwe	epi	baka	ko	dob	seku	lelo	u-nake	ko	wiyo	ko	pako	yelo-yo\\
     sago.palm.species	palm.rib	side	other	I	eye	fall.\textsc{pp}	gecko	\textsc{dfac}-sit	I	hold.\textsc{pp}	I	throw.\textsc{pp}	ground-\textsc{loc}\\
\glt `I looked down into the \textit{numomo}-palm rib on the other side: Here is a gecko. I caught it and threw it on the ground.'
\z

\ea\label{ex:t3exkokuno}
Ko kûno yeloyo. Ko piu lelo piyowe, rupueno lolowe.\\
\gll ko	kûno	yelo-yo	ko	piu	lelo	piyo-we	rupue-no	lolo-we\\
     I	go.down.\textsc{pp}	ground-\textsc{loc}	I	frog	gecko	take.\textsc{pp}-\textsc{du}.\textsc{o}	leaf-\textsc{ins}	wrap.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `I went down to the ground. I took the frog and the gecko and wrapped them in a leaf.'
\z

\ea\label{ex:t3exmomoba}
Ko numomo ba ppuo. Numomo kwe dob seku: piu unake. Ko wiyo, yeloyo pako.\\
\gll ko	numomo	ba	ppuo	numomo	kwe	dob	seku	piu	u-nake	ko	wiyo	yelo-yo	pako\\
     I	sago.palm.species	other	climb.\textsc{pp}	sago.palm.species	palm.rib	eye	fall.\textsc{pp}	frog	\textsc{dfac}-sit	I	hold.\textsc{pp}	ground-\textsc{loc}	throw.\textsc{pp}\\
\glt `I climbed another sago palm and looked down into the \textit{numomo}-palm rib: Here is a frog. I caught it and threw it on the ground.'
\z

\ea\label{ex:t3exkwebak}
Kwe epi baka dob seku: piu dupua inakap. Ko wiyowe, ko yeloyo pakowe.\\
\gll kwe	epi	baka	dob	seku	piu	dupua	i-nake-p	ko	wiyo-we	ko	yelo-yo	pako-we\\
     palm.rib	side	other	eye	fall.\textsc{pp}	frog	two	\textsc{du}.\textsc{s}-sit-\textsc{pc}	I	hold.\textsc{pp}-\textsc{du}.\textsc{o}	I	ground-\textsc{loc}	throw.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `I looked down into the palm rib on the other side: Two frogs were sitting (there). I caught them and threw them on the ground.'
\z

\ea\label{ex:t3exyeloyo}
Ko kûno yeloyo. Piu rondupua rokini rupueno lolo, meli yeiyo.\\
\gll ko	kûno	yelo-yo	piu	rondupua\_rokini	rupue-no	lolo	meli	yei-yo\\
     I	go.down.\textsc{pp}	ground-\textsc{loc}	frog	three	leaf-\textsc{ins}	wrap.\textsc{pp}	carry.\textsc{pl}.\textsc{o}	kneading.trough.for.sago.washing-\textsc{loc}\\
\glt `I went down to the ground. I wrapped the three frogs in a leaf to carry them to the trough for sago washing.'
\z

\ea\label{ex:t3expiulel}
``Piu lelo de luwapo?'' ``Ko baluwapoko.''  Ko sû mappoipe. Ko rupueno re, piyo sûyo. Sûyo rap.\\
\gll piu	lelo	de	luwapo	ko	ba-luwapi-ko	ko	sû\_mappo-ipe	ko	rupue-no	re	piyo	sû-yo	sû-yo	re-p\\
     frog	gecko	you	catch.\textsc{pp}	I	\textsc{fac}-catch-\textsc{fac}	I	light.a.fire.\textsc{pp}-\textsc{ant}	I	leaf-\textsc{ins}	get.done	take.\textsc{pp}	fire-\textsc{loc}	fire-\textsc{loc}	get.done-\textsc{pc}\\
\glt `{``}Did you catch frogs and geckos?''  ``I have caught some.'' I lit a fire first. (Then) I took them in the leaves to get done in the fire. They were roasting in the fire.'
\z

\ea\label{ex:t3expuliyo}
Ko puliyo sûyo. Ko piu rupue nepe. Epe kopi piu ko namo, dupua. Ko ike piu dupua leloyo. Koyo inoi.\\
\gll ko	puliyo	sû-yo	ko	piu	rupue	nepe	epe	ko-pi	piu	ko	namo	dupua	ko\_ike	piu	dupua	lelo-yo	koyo	i-no\\
     I	take.off.\textsc{pp}	fire-\textsc{loc}	I	frog	leaf	take.off.\textsc{pp}	mother	1\textsc{sg}-\textsc{poss}	frog	I	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	two	I.myself	frog	two	gecko-\textsc{loc}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{a}-eat.\textsc{pp}\\
\glt `I took them off the fire. I took away the leaves (from) the frogs.  I gave frogs to my mother, two (of them). I myself (had) two frogs and the gecko. We ate.'
\z

\ea\label{ex:t3exduebap}
Due bapusiyoko. Due dû ep sipamu. Sepue piyamu due dû roise. Ropyo pokûno. Due dû rop roise piapo, puakuyo wakayo puo.\\
\gll due	ba-pusiye-ko	due\_dû	ep	sipamu	sepue	piyamu	due\_dû	roise	rop-yo	pokûno	due\_dû	rop	roise	piapo	puaku-yo	waka-yo	puo\\
     sago	\textsc{fac}-wash-\textsc{fac}	sago.flour	rinse.water	tip.away.hither.\textsc{pp}	trough	take.hither.\textsc{pp}	sago.flour	with	basket-\textsc{loc}	fill.in.\textsc{pp}	sago.flour	basket	together	lift.up.\textsc{pp}	head-\textsc{loc}	shoulder-\textsc{loc}	walk.\textsc{pp}\\
\glt `The sago is washed. (My mother) tipped out the rinse water of the sago flour. She took the trough together with the sago flour. (Then) she filled (the flour) in  baskets and lifted the baskets along with the sago flour up on her head and shoulders. We walked (back).'
\z

\ea\label{ex:t3exanikul}
Ani kululipno yipyo iloi. Yilauyo paeau. Yipyo wolomno ppuo, due dû wapoyo wolo.\\
\gll ani	k-luli-p-no	yip-yo	i-lo	yilau-yo	paeau	yip-yo	wolom-no	ppuo	due\_d{û}	wapo-yo	wolo\\
     day	\textsc{sub}-mix.with.water-\textsc{pc}-\textsc{co}	house-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	village-\textsc{loc}	reach	house-\textsc{loc}	ladder-\textsc{ins}	go.up.\textsc{pp}	sago.flour	porch-\textsc{loc}	set.\textsc{pp}\\
\glt `Having washed sago (all) day we went to (our) house. We reached the village. (My mother) climbed the ladder and set the sago flour on the porch.'
\z

\ea\label{ex:t3exepesum}
Epe sû mappo, yaup yowo sûyo. Yaup mol. Epe kopi due piyo wîlyo, ya mappo. Ya kamappapno ya sui. Ya kusuipno, rupueyo supopo.\\
\gll epe	sû\_mappo	yaup	sû-yo	yowo	yaup	mol	epe	ko-pi	due	piyo	wîl-yo	ya\_mappo	
ya	k-mappe-p-no	ya\_sui	ya	k-sui-p-no	rupue-yo	supopo\\
     mother	light.a.fire.\textsc{pp}	water.for.cooking	fire-\textsc{loc}	boil.\textsc{pp}	water.for.cooking	boil.\textsc{pp}	mother	1\textsc{sg}-\textsc{poss}	sago	take.\textsc{pp}	dish-\textsc{loc}	stir.sago.\textsc{pp}	sago	\textsc{sub}-stir.sago-\textsc{pc}-\textsc{co}	sago.become.ready	sago	\textsc{sub}-sago.become.ready-\textsc{pc}-\textsc{co}	leaf-\textsc{loc}	cut.in.pieces.\textsc{pl}.\textsc{o}.\textsc{pp}\\
\glt `(My) mother lit a fire and boiled water on the fire. The water boiled. My mother put the sago in a dish and stirred it. After stirring in (the water) the sago is ready. When it was ready, she cut it in pieces on leaves.'
\z

\ea\label{ex:t3expaepsi}
Lelpanapno bese ipino si. Bese kisipno krapno, puliyo. Paepu si. Ipiyo sikûno, sûyo yowo. Paepu bese roise baroko.\\
\gll lilipane-p-no	bese	ipi-no	si	bese	k-si-p-no	k-re-p-no	puliyo	paepu	si	ipi-yo	sikûno	sû-yo	yowo	paepu	bese	roise	ba-re-ko\\
     be.there.do.thither-\textsc{pc}-\textsc{co}	tulip	pot-with	cook.\textsc{pp}	tulip	\textsc{sub}-cook-\textsc{pc}-\textsc{co}	\textsc{sub}-get.done-\textsc{pc}-\textsc{co}	take.out.\textsc{pp}	mushroom	cook.\textsc{pp}	pot-\textsc{loc}	fill.in.\textsc{pp}	fire-\textsc{loc}	boil.\textsc{pp}	mushroom	tulip	with	\textsc{fac}-get.done-\textsc{fac}\\
\glt `While the sago (portions) are (put) aside, she cooked \textit{tulip} in a pot. When the \textit{tulip} was cooked and done, she took it out (of the pot). (Then) she cooked mushrooms. She placed them in a pot and cooked them over the fire. (Soon) the mushrooms and the \textit{tulip}-vegetable are done.'
\z

\ea\label{ex:t3exwilroy}
Puliyo, wîl royo, rupopo. Uke roinen. Uke ilo. Uke bailoko rapue kesiyo. Mapap.\\
\gll puliyo	wîl	royo	rupopo	uke	roye-en	uke	ilo	uke	ba-ile-ko	rapue	kesiyo	mape-p\\
     take.out.\textsc{pp}	plate	put.\textsc{pp}	distribute.\textsc{pl}.\textsc{o}.\textsc{pp}	we.\textsc{excl}	give-\textsc{nsg}.\textsc{or}.\textsc{pp}	we.\textsc{excl}	eat.\textsc{pl}.\textsc{a}.\textsc{pp}	we.\textsc{excl}	\textsc{fac}-eat.\textsc{pl}.\textsc{a}-\textsc{fac}	food	use.up.\textsc{pp}	sit.\textsc{pl}-\textsc{pc}\\
\glt `She took them out (of the pot), put the plates (on the floor), distributed (the vegetables) and gave them to us. We ate. We have eaten. The food was consumed, and we were sitting (and resting).'
\z

\ea\label{ex:t3exfinish}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `The story is finished.'
\z