\ea\label{ex:nrexaikopi}
Ai kopi lopos mael, suko. Ba mo, suel boliyo wuliyo. Ba mo boliyo wuliyo. Ba mo boliyo wuliyo. An baka klokni lopos wapoko, royoko. Yipyo lo. Due klokni eppi noyo.\\
\gll ai	ko-pi	lopos	mael	suko	ba	mo	suel	boli-yo	wuli-yo	ba	mo	boli-yo	wuli-yo	ba	mo	boli-yo	wuli-yo	an\_baka	klokni	lopos	wapo-ko	royo-ko	yip-yo	lo	due	klokni	eppi\_noyo\\
     father	1\textsc{sg}-\textsc{poss}	post	cut.\textsc{pp}	cut.\textsc{pp}	other	fell.\textsc{pp}	cut.\textsc{pp}	origin-\textsc{loc}	tree.top-\textsc{loc}	other	fell.\textsc{pp}	origin-\textsc{loc}	tree.top-\textsc{loc}	other	fell.\textsc{pp}	origin-\textsc{loc}	tree.top-\textsc{loc}	five	one	post	put.together.\textsc{pp}-\textsc{rts}	lay.\textsc{pp}-\textsc{rts}	house-\textsc{loc}	go.\textsc{pp}	night	one	rest.\textsc{pp}\\
\glt `My father cut posts. He cut them (to size). He felled other (trees) and cut them (in shape between) trunk and top, felled more (and cut them in shape between) trunk and top, felled (still more) (and cut them in shape between) trunk and top. He put together six posts. He laid them down (on the ground). (Then) he went to the house and rested for one day.'
\z

\newpage
\ea\label{ex:nrexumulne}
Umul nek, nekpamu, ``Nuko mole lopos memiye.'' Uke molo ko epe aino ikap roise. Uke paeau loposyo ulapno. Uke riyo paeau. Uke mapap sû mappo.\\
\gll umul\_nek	nekpamu	nuko	mole	lopos	memiye	uke	molo	ko	epe	ai-no	ikap	roise	uke	paeau	lopos-yo	ule-p-no	uke	ri-yo	paeau	uke	mape-p	sû\_mappo\\
     think.\textsc{pp}	ponder.\textsc{pp}	we.\textsc{incl}	go.\textsc{pl}.\textsc{pp}	post	carry	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}	I	mother	father-\textsc{ins}	1\textsc{sg}.\textsc{poss}.\textsc{emph}	with	we.\textsc{excl}	arrive	post-\textsc{loc}	be.there.\textsc{pl}-\textsc{pc}-\textsc{co}	we.\textsc{excl}	\textsc{dist}-\textsc{loc}	arrive	we.\textsc{excl}	sit.\textsc{pl}-\textsc{pc}	light.fire.\textsc{pp}\\
\glt `He thought (about what to do). He thought thoroughly: ``We'll go and carry the posts.'' We went off, me together with my parents. We arrived where the posts were. We arrived there, were staying and lit a fire.'
\z

\ea\label{ex:nrexkopilo}
Ai kopi lopos ba wel, yipyo layeko. Mipulo dori, mi ba laliko, wel yilau yipyo. Midoripulo, ba piyo, wel, yipyo layeko. Kuso sosoli nakap. Kuso ponake.\\
\gll ai	ko-pi	lopos	ba	wel	yip-yo	laye-ko	mi-pulo	dori	mi	ba	lali-ko	wel	yilau	yip-yo	mi-dori\_pulo	ba	piyo	wel	yip-yo	laye-ko	kuso	sosoli	nake-p	kuso	po-nake\\
     father	1\textsc{sg}-\textsc{poss}	post	some	carry.\textsc{pp}	house-\textsc{loc}	lay-\textsc{rts}	\textsc{iter}-come.\textsc{pp}	return	again	other	carry.on.shoulder.\textsc{pp}-\textsc{rts}	carry.\textsc{pp}	place	house-\textsc{loc}	\textsc{iter}-come.back.\textsc{pp}	other	take.\textsc{pp}	carry.\textsc{pp}	house-\textsc{loc}	lay-\textsc{rts}	always	like.this	stay-\textsc{pc}	always	do.\textsc{pp}-\textsc{dur}\\
\glt `My father carried posts, he laid them down at (the place of) the (new) house. He came again,  took others, carried them to the place of the house, and returned. He took another one, carried it to the (place of the) house, and laid it (down) there. He stayed on like this. He did that once and again.'\footnote{Besides the six stilts that carry the floor one meter above the ground, many more lighter posts are in need.}
\z

\ea\label{ex:nrexpospia}
Lopos piapo, laliko, yipyo layeko. Midoripulo, mi ba piapo. Mi ba lal, yipyo layeko. Midoripulo.\\
\gll lopos	piapo	lali-ko	yip-yo	laye-ko	mi-dori\_pulo	mi	ba	piapo	mi	ba	lal	yip-yo	laye-ko	mi-dori\_pulo\\
     post	lift.up.\textsc{pp}	carry.on.shoulder.\textsc{pp}-\textsc{rts}	house-\textsc{loc}	lay.\textsc{pp}-\textsc{rts}	\textsc{iter}-come.back.\textsc{pp}	again	other	lift.up.\textsc{pp}	again	other	carry.on.shoulder.\textsc{pp}	house-\textsc{loc}	lay-\textsc{rts}	\textsc{iter}-come.back.\textsc{pp}\\
\glt `He lifted a post, carried it, and laid it down at (the future place of) the house. (Then) he came back, lifted another one, carried it, laid it down, and came back.'
\z

\ea\label{ex:nrexknisol}
Klokni solo lilip, ba aska. Piapo, lal, burilo. Uke boyo wulino. ``Ine boyo pulupip! Ko lopos burilaliwe.''\\
\gll klokni	solo	lili-p	ba	aska	piapo	lal	buri\_lo	uke	boyo	wuli-no	ine	boyo	pulupi-p	ko	lopos	buri\_lali-we\\
     one	only	be.there-\textsc{pc}	other	none	lift.up.\textsc{pp}	carry.on.shoulder.\textsc{pp}	go.ahead.\textsc{pp}	we.\textsc{excl}	behind	follow-3\textsc{sg}.\textsc{or}.\textsc{pp}	you.\textsc{pl}	behind	come.\textsc{pl}-\textsc{imp}	I	post	go.ahead\_carry.on.shoulder-\textsc{ter}\\
\glt `Only one was left. No other one. He lifted it, carried it, and went ahead. We went behind him: ``You come behind! I'll carry the post ahead.'''
\z

\ea\label{ex:nrexkimike}
Kimike yipyo nakap. Uke boyo molo, paeau. Uke eppi noyo, mapap. Kwerno epe kopi yaup yowo. Yaup mol.\\
\gll kimike	yip-yo	nake-p	uke	boyo	molo	paeau	uke	eppi\_noyo	mape-p	kwerno	epe	ko-pi	yaup	yowo	yaup	mol\\
     before	house-\textsc{loc}	sit-\textsc{pc}	we.\textsc{excl}	behind	go.\textsc{pl}.\textsc{pp}	arrive	we.\textsc{excl}	rest.\textsc{pp}	sit.\textsc{pl}-\textsc{pc}	afternoon	mother	1\textsc{sg}-\textsc{poss}	water.for.cooking	boil.\textsc{pp}	water.for.cooking	boil.\textsc{pp}\\
\glt `He was sitting in the (old) house first. We went behind and arrived there. We rested and were sitting. In the afternoon my mother boiled water. The water boiled.'
\z

\ea\label{ex:nrexmappoy}
Ya mappo. Ya sui. Lulpamu, rupueyo supopo. Bese si paepu roise. Bese paepu roise wi. Rap. Bareko. Puliyo, sipulyo wolowe.\\
\gll ya\_mappo	ya\_sui	lulpamu	rupue-yo	supopo	bese	si	paepu	roise	bese	paepu	roise	wi	re-p	ba-re-ko	puliyo	sipul-yo	wolo-we\\
     stir.sago.\textsc{pp}	sago.be.ready	portion.sago.\textsc{pp}	leaf-\textsc{loc}	cut.into.pieces.\textsc{pp}	tulip	cook.\textsc{pp}	mushroom	together	tulip	mushroom	together	turn	get.done-\textsc{pc}	\textsc{fac}-get.done-\textsc{fac}	take.off.\textsc{pp}	floor-\textsc{loc}	set.\textsc{pp}-\textsc{ter}\\
\glt `She stirred sago in. The sago is ready. She portioned it and cut it into pieces on leaves. She cooked \textit{tulip}-vegetables with mushrooms. She turned vegetables and mushrooms over. They are getting done, they are done. She took (the pot) off (the fire) and set it on the floor.'
\z

\ea\label{ex:nrexamanis}
Amanisiipe. Pupuol pi, banisiko. Wîlyo rupopo. Ai kopi ponamo, ko powa, disei kopi ponamo. Ki misero no. Uke ilap yano. Baileko. Uke bakesiyewepuko.\\
\gll am-a-nisi-ipe	pupuol	pi	ba-nisi-ko	wîl-yo	rupopo	ai	ko-pi	ponamo	ko	powa	disei	ko-pi	ponamo	ki	misero	no	uke	ile-p	ya-no	ba-ile-ko	uke	ba-kesiye-wepi-ko\\
     \textsc{grad}-\textsc{imp}3-become.cool-\textsc{ant}	heat	\textsc{lv}	\textsc{fac}-become.cool-\textsc{fac}	plate-\textsc{loc}	distribute.\textsc{o}.\textsc{pl}.\textsc{pp}	father	1\textsc{sg}-\textsc{poss}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	I	give.1\textsc{sg}.\textsc{or}.\textsc{pp}	brother	1\textsc{sg}-\textsc{poss}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	\textsc{aph}	also	eat.\textsc{pp}	we.\textsc{excl}	eat.\textsc{pl}.\textsc{a}-\textsc{pc}	sago-\textsc{ins}	\textsc{fac}-eat.\textsc{pl}.\textsc{a}-\textsc{fac}	we.\textsc{excl}	\textsc{fac}-use.up-\textsc{quant}.\textsc{o}-\textsc{fac}\\
\glt `It has to cool first. It is hot. (Now) it has cooled down. She distributed it on plates. She gave (one plate) to my father, gave (one) to me, and gave (one) to my brother. She also ate. We were eating it with sago. We have eaten up (the food). We have finished all of it.'
\z

\ea\label{ex:nrexdupuni}
Uke mapap. Du kwer mui. Dupuni bapuniko. Ai kopi lo. Iplîpyo lo, riyo nu. Ko ike epe roise disei roise due nu. Due nuro. Uke nuro. Yûr bo mu bo moliyo. Yûr kiniyo. Duruwe ani du baruweko punipino.\\
\gll uke	mape-p	du\_kwer	mui	dupuni	ba-puni-ko	ai	ko-pi	lo	iplîp-yo	lo	ri-yo	nu	ko\_ike	epe	roise	disei	roise	due\_nu	due\_nu=ro	uke	nu=ro	yûr	bo	mu	bo	moliyo	yûr	kiniyo	duruwe	ani	du	ba-ruwe-ko	punipino\\
     we.\textsc{excl}	stay.\textsc{pl}-\textsc{pc}	dusk	talk	night	\textsc{fac}-be.dark-\textsc{fac}	father	1\textsc{sg}-\textsc{poss}	go.\textsc{pp}	men's.house-\textsc{loc}	go.\textsc{pp}	\textsc{dist}-\textsc{loc}	sleep.\textsc{pp}	I.myself	mother	with	brother	with	sleep.\textsc{pp}	sleep.\textsc{pp}=\textsc{emph}	we.\textsc{excl}	sleep.\textsc{pp}=\textsc{emph}	bird	sound	talk.\textsc{pp}	sound	talk.\textsc{pl}.\textsc{pp}	bird	all	dawn	daylight	darkness	\textsc{fac}-break-\textsc{fac}	morning\\
\glt `We were sitting (and resting). The dusk makes its noises. Night. It is completely dark. My father left. He went to the men's house and slept there. I myself slept with mother and brother. We were fast asleep, fast asleep. (Then) the birds sang, all of them sang. Dawn, daylight, the darkness is over, morning.'
\z

\ea\label{ex:nrexrararp}
Ai kopi pulo. Bî raripo. Lopos bî rararpo. Lopos pewaupo. Lopos dupua apulyo pewowe, ba dupua kipika pewowe, ba dupua elka pewowe. Bapewaupoko.\\
\gll ai	ko-pi	pulo	bî	rari-po	lopos	bî	rararpo	lopos	pewaupo	lopos	dupua	apulyo	pewo-we	ba	dupua	kipika	pewo-we	ba	dupua	elka	pewo-we	ba-pewaupe-ko\\
     father	1\textsc{sg}-\textsc{poss}	come.\textsc{pp}	hole	dig-\textsc{lv}.\textsc{pp}	post	hole	dig.\textsc{pl}.\textsc{o}.\textsc{pp}	post	erect.\textsc{pl}.\textsc{o}.\textsc{pp}	post	two	in.the.middle	erect.\textsc{pp}-\textsc{du}.\textsc{o}	other	two	at.the.back	erect.\textsc{pp}-\textsc{du}.\textsc{o}	other	two	at.the.front	erect.\textsc{pp}-\textsc{du}.\textsc{o}	\textsc{fac}-erect.\textsc{pl}.\textsc{o}-\textsc{fac}\\
\glt `My father came. He dug a hole. He dug holes for the posts. He erected the posts. Two he erected in the middle, two at the back, two at the front. (Now) they are erected.'
\z

\ea\label{ex:nrexyaliro}
Yali royepane. Baroyoko. Wono nopiwepu. Wono banopiwepiwepuko. Yip yali wono nopiwepiwepu.\\
\gll yali	roye\_pane	ba-roye-ko	wo-no	nopiwepu	wo-no	ba-nopiwepi-wepi-ko	yip	yali	wo-no	nopiwepi-wepu\\
     supporting.beam	lay.thither	\textsc{fac}-lay-\textsc{fac}	rope-\textsc{ins}	tie.together.\textsc{pp}	rope-\textsc{ins}	\textsc{fac}-tie.together-\textsc{quant}.\textsc{o}-\textsc{fac}	house	supporting.beam	rope-\textsc{ins}	tie.together-\textsc{quant}.\textsc{o}.\textsc{pp}\\
\glt `(On the posts) he lays along the supporting beams. They are laid. He tied them together with rope. Many of them are tied together.  He tied together all the supporting beams of the house with rope.'
\z

\ea\label{ex:nrexyippik}
Yip piki royepana wono nopiwepu. Banopiwepiwepuko. Akar wusi wono nopiwepu. Banopiwepiko.\\
\gll yip	piki	roye\_pana	wo-no	nopiwepu	ba-nopiwepi-wepi-ko	akar	wusi	wo-no	nopiwepu	ba-nopiwepi-ko	sipul	roye\_pana	yali-yo	yip	sipul	wo-no	nopiwepu	ba-nopiwepi-ko\\
     house	crossbeam	lay.thither.\textsc{pp}	rope-\textsc{ins}	tie.together.\textsc{pp}	\textsc{fac}-tie.together-\textsc{quant}.\textsc{o}-\textsc{fac}	roof.joist	construct	rope-\textsc{ins}	tie.together.\textsc{pp}	\textsc{fac}-tie.together-\textsc{fac}	floor	lay.thither.\textsc{pp}	supporting.beam-\textsc{loc}	house	floor	rope-\textsc{ins}	tie.together.\textsc{pp}	\textsc{fac}-tie.together-\textsc{fac}\\
\glt `He placed the crossbeams and tied them together with rope. They are all tied together. 	He tied the roof joints together with rope. They are tied together.  (Then) he placed the floor (planks) on the supporting beams. He tied together the floor (planks) of the house with rope. They are tied together.'
\z

\ea\label{ex:nrexpalopo}
Palo po. Basepiko. Palo royepepo wono nopiwepippuo. Baka bapok.\\
\gll palo	po	ba-sepi-ko	palo	royepepo	wo-no	nopiwepi\_ppuo	baka	ba-pi-ko\\
     sago.thatches	make.\textsc{pp}	\textsc{fac}-sew-\textsc{fac}	sago.thatches	lay.on.top.\textsc{pp}	rope-\textsc{ins}	tie.together\_go.up.\textsc{pp}	half	\textsc{fac}-do-\textsc{fac}\\
\glt `He made the sago thatches. They are sown. He placed the sago thatches on top (of the roof beams) and tied everything together going up. (Then) half (of the roof) is made.'
\z

\ea\label{ex:nrexkwerwu}
Kwer wui dupuni duruwa. Yip epi baka po. Epi baka palo kiniyo po. Kiniyo bapok. Puele lul. Puele ukelul yena ba roise. Kiniyo ukelul, yena mekiyo. Balulko.\\
\gll kwer	wui	dupuni	duruwa	yip	epi	baka	po	epi	baka	palo	kiniyo	po	kiniyo	ba-pi-ko	puele	lul	puele	uke-lul	yena	ba	roise	kiniyo	uke-lul	yena	mekiyo	ba-luli-ko\\
     dusk	answer	night	dawn	house	side	half	make.\textsc{pp}	side	half	sago.thatches	all	make.\textsc{pp}	all	\textsc{fac}-make-\textsc{fac}	wall	fence.\textsc{pp}	wall	jointly-fence.\textsc{pp}	people	other	together	all	jointly-fence.\textsc{pp}	people	help.\textsc{pp}	\textsc{fac}-fence-\textsc{fac}\\
\glt `The dusk sounds. The night (comes). The dawn (comes). He made the other half (of the roof). He made all the sago thatches of the other side. They are made. (Then) he fenced the walls. He fenced (the walls) together with many other people. People helped (him), and (the walls) are fenced.'
\z

\ea\label{ex:nrexbaesli}
Baes lili. Par sipulyo royo. Yelo rapiyo, rapiyewapo. Yelo dorno piyelayewepu. Pu moni sipakewepu yelo baesyo. Rino seseliwepu. Baes yala sike pi. Baes balelko.\\
\gll baes	lili	par	sipul-yo	royo	yelo	rapiyo	rapiye\_wapo	yelo	dor-no	piyelaye-wepu	pu	moni	sipake-wepu	yelo	baes-yo	ri-no	seseli-wepu	baes	yala	sike	pi	baes	ba-lili-ko\\
     fireplace	be.there	palm.mat	floor-\textsc{loc}	lay.\textsc{pp}	soil	fetch.\textsc{pp}	fetch\_put.together.\textsc{pp}	soil	foot-\textsc{ins}	trample-\textsc{quant}.\textsc{o}.\textsc{pp}	water	little	pour.down.into-\textsc{quant}.\textsc{o}.\textsc{pp}	soil	fireplace-\textsc{loc}	stick-\textsc{ins}	smooth.out-\textsc{quant}.\textsc{o}.\textsc{pp}	fireplace	now	hard	\textsc{lv}	fireplace	\textsc{fac}-be.there-\textsc{fac}\\
\glt `There is the fireplace (left to be done). He put a palm mat on the floor and fetched soil. He fetched it, put it (on the palm mat), and trampled the soil with (his) feet. He poured down a little water and smoothed out the soil on the fireplace with a stick. Now the fireplace becomes hard. The fireplace is there.'
\z

\ea\label{ex:nrexsmepbi}
Smep bili po, bili klokni. Bî sepeiwepu. Bili ba ar sepeiwepu. Kiniyo lul. Bî yip pakika solo polip, sûli lena. Bîom yip sûli rka le, bîom yipka.\\
\gll smep	bili	po	bili	klokni	bî	sepei-wepu	bili	ba	ar	sepei-wepu	kiniyo	lul	bî	yip	paki-ka	solo	poli-p	sûli	le-na	bîom	yip	sûli	r(i)-ka	le	bîom	yip-ka\\
     door	opening	make.\textsc{pp}	opening	one	hole	cut.out-\textsc{quant}.\textsc{o}.\textsc{pp}	opening	other	\textsc{neg}	cut.out-\textsc{quant}.\textsc{o}.\textsc{pp}	all	fence.\textsc{pp}	hole	house	side-\textsc{path}	only	be.there-\textsc{pc}	smoke	go-\textsc{purp}	smoke.outlet	house	smoke	\textsc{dist}-\textsc{path}	go	smoke.outlet	house-\textsc{path}\\
\glt `He made the door, one opening (only). He cut out a hole. Other openings he didn't cut out. He (had) fenced all (the sides of the house). There was only a hole towards (one) side of the house for the smoke to go, the smoke outlet of the house. The smoke goes out that way, (namely) through the smoke outlet of the house.'\footnote{Traditional houses had only one small door for entering the house in a crouched position. They didn't have windows. So it was quite dark inside the house.}
\z

\ea\label{ex:nrexsoswop}
Sos wopiyanewepu. Dupuni ono asa paliyam? Imiyu asa paliyam. Wolo po wono nopiwepu. Banopiwepuko.\\
\gll sos	wopiyane-wepu	dupuni	ono	asa	paliye-m	imiyu	asa	paliye-m	wolo	po	wo-no	nopiwepu	ba-nopiwepi-ko\\
     bolt	fasten.thither-\textsc{quant}.\textsc{o}.\textsc{pp}	night	person	how	open-\textsc{pos}	sorcerer	how	open-\textsc{pos}	ladder	make.\textsc{pp}	rope-\textsc{ins}	tie.together.\textsc{pp}	\textsc{fac}-tie.together-\textsc{fac}\\
\glt `He fastened bolts (on the door). Nobody can open (the door) at night. A sorcerer cannot open it. He made the ladder. He tied it together with rope. (The ladder) is tied together.'
\z

\ea\label{ex:nrexemsouk}
Em so: Uke yeni yasupiwepu. Yeni uke bawopiyewepuko. Uke le rapiyekûpu yip sukuyo. Uke yip pueneyo melip. Le kiniyo: neppi dawa ipi rop due dû, kiniyo.\\
\gll em	so	uke	yeni	yasupi-wepu	yeni	uke	ba-wopiye-wepi-ko	uke	le	rapiye\_kûpu	yip	suku-yo	uke	yip	puene-yo	meli-p	le	kiniyo	neppi	dawa	ipi	rop	due\_dû	kiniyo\\
     tomorrow	like	we.\textsc{excl}	board	place.\textsc{pl}.\textsc{o}-\textsc{quant}.\textsc{o}.\textsc{pp}	board	we.\textsc{excl}	\textsc{fac}-get.ready-\textsc{quant}.\textsc{o}-\textsc{fac}	we.\textsc{excl}	belongings	fetch.down.hither.\textsc{pl}.\textsc{pp}	house	old-\textsc{loc}	we.\textsc{excl}	house	new-\textsc{loc}	carry.\textsc{pl}.\textsc{o}-\textsc{pc}	belongings	all	bush.knife	axe	pot	basket	sago.flour	all\\
\glt `The next day (goes) like this: We placed several boards. We have gotten the boards ready.  	(Then) we took down (all our) belongings from the old house and carried them to the  	new house. All the belongings: the bush knifes, the axe, the pots, the baskets, the sago flour, everything.'
\z

\ea\label{ex:nrexyippue}
Uke mapap yip pueneyo. Ai kopi eppi makina noyo, numuelna. Eppi noyenakap wîs klokni.\\
\gll uke	mape-p	yip	puene-yo	ai	ko-pi	eppi	maki-na	noyo	numuelna	eppi\_noye-nake-p	wîs	klokni\\
     we.\textsc{excl}	stay.\textsc{pl}-\textsc{pc}	house	new-\textsc{loc}	father	1\textsc{sg}-\textsc{poss}	rest	good-\textsc{adv}	rest.\textsc{pp}	for.a.long.time	rest-\textsc{dur}-\textsc{pc}	moon	one\\
\glt `(From now on) we were staying in the new house. My father took a good rest. He rested for a long time, (for) one month.'
\z
