\ea\label{ex:t6exyenadu}
Yena duyo molo. Diri eweno ilo ako ikep roise. Duyo mapap bi lipel. Lipeliwepu, puenpowepu. Diri baka royenamo, ewe ki baka. Boyopuno diri miso biep lu. Inakap.\\
\gll yena	du-yo	molo	diri	ewe-no	i-lo	ako	ikep	roise	du-yo	mape-p	bi	lipel	lipeli-wepu	puenpi-wepu	diri	baka	roye\_namo	ewe	ki	baka	boyopuno	diri	miso	biep	lu	i-nake-p\\
     people	bush-\textsc{loc}	go.\textsc{pl}.\textsc{pp}	younger.brother	older.brother-\textsc{ins}	\textsc{du}.\textsc{s}-go.\textsc{pp}	wife	3\textsc{pl}.\textsc{poss}	with	bush-\textsc{loc}	mape-p	animal	seek.\textsc{pp}	find-\textsc{quant}.\textsc{o}.\textsc{pp}	cut.meat-\textsc{quant}.\textsc{o}.\textsc{pp}	younger.brother	half	lay\_give.3\textsc{sg}.\textsc{or}.\textsc{pp}	older.brother	\textsc{aph}	half	then	younger.brother	also	boar	shoot.\textsc{pp}	\textsc{du}.\textsc{s}-stay-\textsc{pc}\\
\glt `Some people went to the bush. Two brothers went together with their wives. They were staying in the bush and looked for animals. They found many and cut up the meat (of the shot animals). (The older brother) gave half (of the meat) to his younger brother, and he (took the other) half. Then the younger brother also shot a boar. They remained (in the bush).'
\z

\ea\label{ex:t6exdirmar}
Diri mar diri nomar. Due an kinika so mar. Ewe umul nek, ``Diri kopi suloimoina mari. Yala ko asa pi? Yala ko ba pi?'' Dob reyo, ``Oke yala suem.'' Due an bakapuno diri kep sui. Basuiko. Paliya.\\
\gll diri	mar	diri	nomar	due	an\_kinika	so	mar	ewe	umul\_nek	diri	ko-pi	suloimoina	mari	yala	ko	asa	pi	yala	ko	ba	pi	dob	reyo	o-ke	yala	sui-m	due	an\_bakapuno	diri	kep	sui	ba-sui-ko	paliya\\
     younger.brother	be.sick.\textsc{pp}	younger.brother	be.very.sick.\textsc{pp}	night	five	five	be.sick.\textsc{pp}	older.brother	think.\textsc{pp}	younger.brother	1\textsc{sg}-\textsc{poss}	extraordinarily	be.sick	now	I	how	do	now	I	what	do	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	\textsc{prox}-\textsc{aph}	\textsc{mod}	die-\textsc{pos}	night	after.five.more	younger.brother	3\textsc{sg}.\textsc{poss}	die	\textsc{fac}-die-\textsc{fac}	be.dead\\
\glt `The younger brother was sick. He was very sick for five days. Five  	(days) sick (already). The older brother thought: ``My brother is seriously ill. What am I going to do now? What should I do?'' He looked at him (and thought): ``This one will possibly die.'' After five more nights the younger brother dies. He has died. He is dead.'
\z

\ea\label{ex:t6exewewoe}
Ewe wo. Ewe dob pul seku wo mop. Yala ko asa pi? Yeni rileyo po. Yol apla rileyo po. Ako kep muelnoro: ``Uki depi yala nuko parno lole, parno panepue.'' Wono lolo, yolyo layo. Sû mappo.\\
\gll ewe	wo	ewe	dob	pul	seku	wo\_mop	yala	ko	asa	pi	yeni	rileyo	po	yol	apla	rileyo	po	ako	kep	mueli-no=ro	uki	de-pi	yala	nuko	par-no	lole	par-no	pane\_pue	wo-no	lolo	yol-yo	layo	sû\_mappo\\
     older.brother	crying	older.brother	eye	liquid	fall.\textsc{pp}	cry.\textsc{pp}	now	I	how	do	platform	above	do.\textsc{pp}	grid	plank	above	do.\textsc{pp}	wife	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}=\textsc{emph}	husband	2\textsc{sg}-\textsc{poss}	now	we.\textsc{incl}	bark.mat-\textsc{ins}	wrap	bark.mat-\textsc{ins}	lay.around	rope-\textsc{ins}	tie.\textsc{pp}	grid-\textsc{loc}	lay.\textsc{pp}	light.fire.\textsc{pp}\\
\glt `The older brother cried. His tears fell down, he cried: ``What am I going to do now?'' He put a platform up (in a tree). He put a grid of planks up in the tree. (Then) he said to his brother's wife: ``Now we wrap your husband in a bark mat and put him there with a bark mat around.'' They tied (the mat) with a rope and put (the body) on the grid. (Then) they lit a fire.'
\z

\ea\label{ex:t6exsukmap}
Sû kamappapno sû beri rap. Ako kep muelnoro: ``Nuko ukeli yilauyo.'' Diri sukei wonino: ``Ko ruri depi ako depi ko yilauyo ukeli. De boyo wulimonpep!''\\
\gll sû	k-mappe-p-no	sû	beri	re-p	ako	kep	mueli-no=ro	nuko	ukeli	yilau-yo	diri	sukei	woni-no	ko	ruri	de-pi	ako	de-pi	ko	yilau-yo	ukeli	de	boyo	wulimini-ipi-p\\
     fire	\textsc{sub}-light-\textsc{pc}-\textsc{co}	fire	blaze	get.done-\textsc{pc}	wife	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}=\textsc{emph}	we.\textsc{incl}	take.with.oneself	village-\textsc{loc}	younger.brother	spirit	call-3\textsc{sg}.\textsc{or}.\textsc{pp}	I	child	2\textsc{sg}-\textsc{poss}	wife	2\textsc{sg}-\textsc{poss}	I	village-\textsc{loc}	take.with.oneself	you	later	follow-1\textsc{sg}.\textsc{or}-\textsc{imp}\\
\glt `Having lit the fire, the blazing fire was burning (the body). He said to (the younger brother's) wife: ``We take ourselves to the village.'' To his brother's spirit he called out: ``I take your children and your wife to the village. You follow me later!'''
\z

\ea\label{ex:t6expaeauw}
Yilauyo paeau wo roise. ``De ba po? De aso?'' ``Diri kopi basuik. Ko ke layepaneko duyo. Du yipyo lili. Parno layo yolyo.''\\
\gll yilau-yo	paeau	wo	roise	de	ba	po	de	aso	diri	ko-pi	ba-sui-ko	ko	ke	laye\_pane-ko	du-yo	du	yip-yo	lili	par-no	layo	yol-yo\\
     village-\textsc{loc}	reach	crying	with	you	what	do.\textsc{pp}	you	how	younger.brother	1\textsc{sg}-\textsc{poss}	\textsc{fac}-die-\textsc{fac}	I	\textsc{top}	lay.aside-\textsc{rts}	bush-\textsc{loc}	bush	house-\textsc{loc}	be.there	bark.mat-\textsc{ins}	lay.\textsc{pp}	grid-\textsc{loc}\\
\glt `They reached the village crying (and were asked): ``What did you do? How are you?''  ``My younger brother has died and I left (his body) in the bush, at the bush house. I put it with a mat (wrapped around it) on a grid.'''
\z

\ea\label{ex:t6exkepsum}
Ewe kep sû mappap. Sû solo mappeulipop. Ewe sû mimappeke. Sosoli nakap.\\
\gll ewe	kep	sû\_mappe-p	sû	solo	mappe-uli-pi-p	ewe	sû	mi-mappe-ke	sosoli	nake-p\\
     older.brother	3\textsc{sg}.\textsc{poss}	light.fire-\textsc{pc}	fire	only	light-\textsc{prog}-\textsc{lv}-\textsc{pc}	older.brother	fire	\textsc{iter}-light-\textsc{ingr}	this.way	stay-\textsc{pc}\\
\glt `His brother tended the fire. He was tending the fire. The older brother went to light the fire anew (when it had extinguished). Tending the fire kept him busy.'
\z

\ea\label{ex:t6exkepbis}
Diri kep biso baslaupoko. Klapno par nepeipana. Am ari, nikip. Midorilolo, wono layeko, midorilayeko.\\
\gll diri	kep	bi-so	ba-slau-pi-ko	k-le-p-no	par	nepei\_pana	am	ari	niki-p	mi-dori\_lolo	wo-no	laye-ko	mi-dori\_laye-ko\\
     younger.brother	3\textsc{sg}.\textsc{poss}	meat-\textsc{sim}	\textsc{fac}-dry-\textsc{lv}-\textsc{fac}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	bark.mat	take.off.\textsc{pp}	yet	no	stink-\textsc{pc}	\textsc{iter}-wrap.again.\textsc{pp}	rope-\textsc{ins}	lay-\textsc{rts}	\textsc{iter}-lay.back-\textsc{rts}\\
\glt `The younger brother's (corpse) has dried like meat. When (the older brother) had gone (there), he took off the bark mat (and looked): not yet.  (The corpse) was (still) smelling. He wrapped him again with the rope and put him (there). He put him back again (on the lattice).'
\z

\ea\label{ex:t6exmiyila}
Mi yilauyo pulo. Nakero, due an dupua dor dupua. Makina amanikiipe. Mi boyo lo. Am due lakoipe.\\
\gll mi	yilau-yo	pulo	nake=ro	due	an\_dupua	dor\_dupua	maki-na	am-a-niki-ipe	mi	boyo	lo	am	due	lako-ipe\\
     again	village-\textsc{loc}	come.\textsc{pp}	sit=\textsc{emph}	night	ten	ten	good-\textsc{adv}	\textsc{grad}-\textsc{imp}3-smell-\textsc{ant}	again	later	go.\textsc{pp}	yet	night	count.\textsc{pp}-\textsc{ant}\\
\glt `He came back to the village. He stays (there) for twenty days. Well, before that (the corpse) still would have smelled. Later he went again, yet he counted the days first.'
\z

\ea\label{ex:t6exmiriye}
Klapno miriyeko. Par minepeipana. Diri kep banikiko. Kili solo ulap, puaku kili roise sappi kili roise.\\
\gll k-le-p-no	mi-riye-ko	par	mi-nepei\_pana	diri	kep	ba-niki-ko	kili	solo	ule-p	puaku	kili	roise	sappi	kili	roise\\
     \textsc{sub}-go-\textsc{pc}-\textsc{co}	\textsc{iter}-see.\textsc{o.inanim}-\textsc{rts}	bark.mat	\textsc{iter}-take.off.\textsc{pp}	younger.brother	3\textsc{sg}.\textsc{poss}	\textsc{fac}-smell-\textsc{fac}	bone	only	be.there.\textsc{pl}-\textsc{pc}	head	bone	together	jaw	bone	together\\
\glt `When he had come, he glanced (at the corpse) again. He took off the bark mat again. His brother had smelled. (But now there isn't any smell). There were only the bones, together with the skull and with the jaw bone.'
\z

\ea\label{ex:t6exkrapiy}
Krapiyapno ropyo niskûno, mel yilauyo. Yilauyo paeau. Yip bîyo wena. Laliyowe ki yo nuipno.\\
\gll k-rapiye-p-no	rop-yo	niskûno	mel	yilau-yo	yilau-yo	paeau	yip	bîyo	wena	laliyo-we	ki	yo	nui-p-no\\
     \textsc{sub}-fetch-\textsc{pc}-\textsc{co}	basket-\textsc{loc}	fill.in.\textsc{pp}	carry.\textsc{pl}.\textsc{o}.\textsc{pp}	village-\textsc{loc}	village-\textsc{loc}	arrive	house	inside	carry.inside.\textsc{pp}	hang.\textsc{pp}-\textsc{ter}	\textsc{aph}	location	sleep-\textsc{pc}-\textsc{co}\\
\glt `Having gathered (the bones), he put them in a basket and carried them to the village. He arrived in the village. He carried them into the house. He hung them up, where he slept.'
\z

\ea\label{ex:t6eximiyup}
Imiyu pulupi rilina. Sukei kep yala ewe kep muelne: ``De due kunuem! Imiyu ere pulupi. De puanap! De nakap dob pep!''\\
\gll imiyu	pulupi	rili-na	sukei	kep	yala	ewe	kep	mueli-ne	de	due	k-nui-m	imiyu	ere	pulupi	de	puane-p	de	nake-p	dob	pi-p\\
     sorcerer	come.\textsc{pl}	see.\textsc{o}.\textsc{anim}.\textsc{pl}-\textsc{purp}	spirit	3\textsc{sg}.\textsc{poss}	now	older.brother	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}	you	night	\textsc{proh}-sleep-\textsc{proh}	sorcerer	here	come.\textsc{pl}	you	wake.up-\textsc{imp}	you	sit-\textsc{imp}	eye	do-\textsc{imp}\\
\glt `Sorcerers might come to look for (the older brother's family). (Then the younger brother's) spirit says to his older brother: ``You must not sleep! The sorcerers come here. Wake up, stay (awake) and watch out!'''
\z

\ea\label{ex:t6exkaepul}
Riyopuno kaepul kaepi uroyo ule, klokni piyo. Ani yala rino pue. Monomno le. Du mono leipe. Epueyo puine, epuemno le.\\
\gll riyopuno	kaepul	kaepi	uro-yo	ule	klokni	piyo	ani	yala	ri-no	pue	monomno	le	du	mono	le-ipe	epue-yo	puine	epuemno	le\\
     then	knee	kneecap	netbag-\textsc{loc}	put.inside.\textsc{pp}	one	take.\textsc{pp}	day	now	\textsc{dist}-\textsc{ins}	walk.around	along.the.path	go	bush	path	go-\textsc{ant}	undergrowth-\textsc{loc}	branch.off.thither	through.undergrowth	go\\
\glt `Then he put the kneecap into (his) netbag. He took one (bone). During the day he walks around with it. He walks along a path. First he walks along the bush track, (then) he branches off in the undergrowth and walks through the undergrowth.'
\z

\ea\label{ex:t6exsepiye}
Boyopuno uro kep dob riye: ``Uro kopi sepiye. Imiyu pulupi. Ko pauiyo laliye.'' Mono bayana lo. Mi epuemno lo yilauyo, paeau. Ako kep muelno: ``Uro kopi sepiyo. Ko ro riyo. Mi ko ro pulo. Ko epemna pulo. Bi ko lipelou, mi ko ar lu.''\\
\gll boyopuno	uro	kep	dob	riye	uro	ko-pi	sepiye	imiyu	pulupi	ko	paui-yo	laliye	mono	bayana	lo	mi	epuemno	lo	yilau-yo	paeau	ako	kep	mueli-no	uro	ko-pi	sepiyo	ko	ro	riyo	mi	ko	ro	pulo	ko	epemna	pulo	bi	ko	lipeli-ou	mi	ko	ar	lu\\
     later	netbag	3\textsc{sg}.\textsc{poss}	eye	see.\textsc{o}.\textsc{inanim}	netbag	1\textsc{sg}-\textsc{poss}	shake	sorcerer	come.\textsc{pl}	I	shoulder-\textsc{loc}	hang	path	different	go.\textsc{pp}	again	through.undergrowth	go.\textsc{pp}	village-\textsc{loc}	arrive	wife	3\textsc{sg}.\textsc{poss}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	netbag	1\textsc{sg}-\textsc{poss}	shake.\textsc{pp}	I	\textsc{prox}.\textsc{emph}	see.\textsc{o}.\textsc{inanim}.\textsc{pp}	again	I	\textsc{prox}.\textsc{emph}	come.\textsc{pp}	I	quickly	come.\textsc{pp}	pig	I	seek-\textsc{frus}	again	I	\textsc{neg}	shoot.\textsc{pp}\\
\glt `Later he looks at his netbag: ``My netbag is shaking. Sorcerers are coming. I'll hang it over my shoulder.'' He went along another path. He went again through the undergrowth to the village, arrived (there) and said to his wife: ``My netbag shook, I saw this. I came (back) here. I walked fast. A pig I sought in vain, again I did not shoot (anything).'''
\z