\ea\label{ex:nredirrrix}
Diri eweno duyo ilo. Duyo ipuap.\\
\gll diri	ewe-no	du-yo	i-lo	du-yo	i-pue-p\\
     younger.brother	older.brother-\textsc{ins}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-roam-\textsc{pc}	\\
\glt `Two brothers went to the bush. They were roaming the bush.'
\z

\ea\label{ex:nrexurualr}
Urual reyepuloi. Urual rileyo ppuo, riyo nakap.\\
\gll urual	reyepulo-i	urual	rileyo	ppuo	ri-yo	nake-p\\
     big.lizard	come.to.see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}-\textsc{du}.\textsc{a}	big.lizard	above	go.up.\textsc{pp}	tree-\textsc{loc}	sit-\textsc{pc}\\
\glt `They came to see a big lizard. The big lizard climbed up high and was sitting in a tree.'
\z

\ea\label{ex:nrexewwedi}
Ewe diri muelno: ``De yeloyo nakap! Ko ppuene.''\\
\gll ewe	diri	mueli-no	de	yelo-yo	nake-p	ko	ppue-ne\\
     older.brother	younger.brother	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	ground-\textsc{loc}	stay-\textsc{imp}	I	go.up-3\textsc{sg}.\textsc{or}\\
\glt `The older brother said to the younger brother: ``You stay on the ground!   I'll climb up after (the lizard).'''
\z

\ea\label{ex:nrexmelipp}
Pe pakono melippueno. Eku makina kanakapno urual lu.\\
\gll pe	pako-no	melippue-no	eku	maki-na	k-nake-p-no	urual	lu\\
     arrow	bow-\textsc{ins}	carry.up.\textsc{pl}.\textsc{o}-3\textsc{sg}.\textsc{or}.\textsc{pp}	behind	good-\textsc{adv}	\textsc{sub}-sit-\textsc{pc}-\textsc{co}	big.lizard	shoot.\textsc{pp}\\
\glt `He took bow and arrows (climbing) up after it. Having gotten a good position, he shot at the big lizard.'
\z

\ea\label{ex:nrexpeklok}
Pe klokni ponamo lu. Urual dopyo polip.\\
\gll pe	klokni	ponamo	lu	urual	dop-yo	poli-p\\
     arrow	one	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	shoot.\textsc{pp}	big.lizard	body-\textsc{loc}	be.there-\textsc{pc}\\
\glt `He gave (the lizard) one arrow. He shot. (The arrow) stuck in the body of the lizard.'
\z

\ea\label{ex:nrexpebaka}
Mi pe ba pulapu. Pe ba pulapu, pe ba pulapu. An baka kini. Pe kuru.\\
\gll mi	pe	ba	pulapu	pe	ba	pulapu	pe	ba	pulapu	an\_baka	kini	pe	kuru\\
     again	arrow	other	release.\textsc{pp}	arrow	other	release.\textsc{pp}	arrow	other	release.\textsc{pp}	five	one.\textsc{part}	arrow	be.finished\\
\glt `He released another arrow. He released the third arrow. He released the fourth arrow. Six arrows (alltogether). (Now) the arrows are finished.'
\z

\ea\label{ex:nrexanpues}
Urual anyo puesu. Boyo waliyo puesu. Wali sueli. Lu kep ileiele, saripso. Rileyo sui, paliya.\\
\gll urual	an-yo	puesu	boyo	wali-yo	puesu	wali	sueli	lu	kep	ileiele	sarip-so	rileyo	sui	paliya\\
     big.lizard	hand-\textsc{loc}	bite.\textsc{pp}	later	neck-\textsc{loc}	bite.\textsc{pp}	neck	cut	tooth	3\textsc{sg}.\textsc{poss}	very.long	blade(\textsc{tp})-\textsc{sim}	above	die	be.dead\\
\glt `The big lizard bit (his) hand. Then it bit (his) neck. It cut the man's neck. Its teeth are very long, like a blade. The man dies above (in the tree). He is dead.'
\z

\ea\label{ex:nrexuruall}
Urual ri ini bayana piyana. Nakap.\\
\gll urual	ri	ini	bayana	piyana	nake-p\\
     big.lizard	tree	branch	different	jump.\textsc{pp}	sit-\textsc{pc}	\\
\glt `The lizard jumped onto another branch of the tree and was sitting (there).'
\z

\ea\label{ex:nrexwolosw}
Diri wo los walirue, weppuo rileyo. Ewe wepulno yeloyo wekûno.\\
\gll diri	wo\_los	walirue	weppuo	rileyo	ewe	wepul-no	yelo-yo	wekûno\\
     younger.brother	kind.of.creeper	twine	carry.up.\textsc{pp}	above	older.brother	sling-\textsc{ins}	ground-\textsc{loc}	carry.down.\textsc{pp}\\
\glt `The brother twined up a liana and took it up. (Then) he carried the older brother down to the ground by means of a (liana) sling.'
\z

\ea\label{ex:nrexloswal}
Yilauyo wel. Yena sano: ``Ewe ba po?''\\
\gll yilau-yo	wel	yena	sa-no	ewe	ba	po\\
     village-\textsc{loc}	carry.\textsc{pp}	people	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	older.brother	what	do.\textsc{pp}\\
\glt `He brought him to the village. The people asked him: ``What happened to the older brother?'''
\z

\newpage
\ea\label{ex:nrexurpues}
``Urual puesu. Paliya. Basuiko.''\\
\gll urual	puesu	paliya	ba-sui-ko\\
     big.lizard	bite.\textsc{pp}	be.dead	\textsc{fac}-die-\textsc{fac}\\
\glt `{``}A huge lizard bit him. He is dead. He has died.'''
\z

\ea\label{ex:nrexuknoko}
Yipyo wel. Yip bîyo nuknoko. Yena dupuni klokni nuknoko duruwa.\\
\gll yip-yo	wel	yip	bîyo	nuknoko	yena	dupuni	klokni	nuknoko	duruwa\\
     house-\textsc{loc}	carry.\textsc{pp}	house	inside	having.spent.some.nights	people	night	one	having.spent.some.nights	dawn\\
\glt `He carried him to the house. Inside the house they held a wake (for the dead). One night the people held the wake, until dawn.'
\z

\ea\label{ex:nrexbiabiy}
Bî kraripno bia bîyo wel, pakûpu. Bî sowo.\\
\gll bî	k-rari-p-no bia	bî-yo	wel	pakûpu	bî	sowo\\
     hole	\textsc{sub}-dig-\textsc{pc}-\textsc{co}	corpse	hole-\textsc{loc}	carry.\textsc{pp}	throw.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	hole	cover.\textsc{pp}\\
\glt `Then they had dug a hole, carried the corpse to the hole, and threw it inside. (Then) they covered the hole.'
\z

\ea\label{ex:nrexmoirik}
Bî kosowapno yena molo ri moi. Ri komoipno yeloyo ye.\\
\gll bî	k-sowe-p-no	yena	molo	ri	moi	ri	k-moi-p-no	yelo-yo	ye\\
     hole	\textsc{sub}-cover-\textsc{pc}-\textsc{co}	people	go.\textsc{pl}.\textsc{pp}	tree	cut	tree	\textsc{sub}-cut-\textsc{pc}-\textsc{co}	ground-\textsc{loc}	fall.over\\
\glt `Having covered the hole the people went to cut the tree (where the goanna had been). Having cut that tree, it fell over to the ground (along with the goanna).'
\z

\ea\label{ex:nrexurrual}
Urual peno kuluipno rino waliyo paku. Paliya.\\
\gll urual	pe-no	k-lui-p-no	ri-no	wali-yo	paku	paliya\\
     big.lizard	arrow-\textsc{ins}	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	stick-\textsc{ins}	neck-\textsc{loc}	beat.\textsc{pp}	be.dead\\
\glt `They shot the big lizard with arrows, they beat (its) neck with sticks. It is dead.'
\z

\ea\label{ex:nrexrualpu}
Ri komoipno sû mappo, urual puenpop.\\
\gll ri	k-moi-p-no	sû\_mappo	urual	puenpi-p\\
     wood	\textsc{sub}-cut-\textsc{pc}-\textsc{co}	light.a.fire.\textsc{pp}	big.lizard	cut.meat-\textsc{pc}\\
\glt `Having cut wood, they lit a fire. (Then) they were cutting the flesh of the big lizard.'
\z

\ea\label{ex:nrexnapnop}
Kusukapno sûyo royepepo. Krapno mi sûyo napno puliyoro.\\
\gll k-suke-p-no	sû-yo	royepepo	k-re-p-no	mi	sû-yo	nap-no	puliyo=ro\\
     \textsc{sub}-cut-\textsc{pc}-\textsc{co}	fire-\textsc{loc}	lay.on.top.\textsc{pp}	\textsc{sub}-be.done-\textsc{pc}-\textsc{co}	then	fire-\textsc{loc}	bamboo.tongs-\textsc{ins}	take.out.\textsc{pp}=\textsc{emph}\\
\glt `After cutting it in pieces, they put it on the fire. It was done, and they took the pieces (of meat) off the fire with bamboo tongs.'
\z

\ea\label{ex:nrexduurua}
Dû urualpi yena ilo, dipiyepayo. Urual dû yano ilo.\\
\gll dû	urual-pi	yena	ilo	d-piyepayo	urual	dû	ya-no	ilo\\
     meat	big.lizard-\textsc{poss}	people	eat.\textsc{pl}.\textsc{a}.\textsc{pp}	\textsc{lkh}-share.\textsc{pp}	big.lizard	meat	sago-\textsc{ins}	eat.\textsc{pl}.\textsc{a}.\textsc{pp}\\
\glt `The people ate the meat of the big lizard. They were expected to share it among themselves. They ate the lizard's meat with sago.'
\z

\ea\label{ex:nrexrukruk}
Bo kuru.\\
\gll bo	kuru\\
     story	be.finished\\
\glt `End of the story'
\z

\ea\label{ex:nrexineryo}
Urual kauna walpue. Urual puane, dob kosoupi: ``Eh ono woryo.'' De ineryo luip! Pewai poname urual sui.\\
\gll urual	kauna	walpue	urual	puane	dob\_kosoupi	eh	ono	wor-yo	de	iner-yo	lui-p	pewai	poname	urual	sui\\
     big.lizard	numerous	move.about	big.lizard	stand.up	look.around.alertly	eh	man	dog-\textsc{loc}	you	armpit-\textsc{loc}	shoot-\textsc{imp}	deadly.spot	give.2\textsc{sg}.\textsc{or}	big.lizard	die\\
\glt `Numerous big lizards move about. The lizard stands up and looks around alertly: ``Eh, a man and a dog.'' Shoot it in the pit of its forelimbs! You hit it at the deadly spot, and the big lizard will die.'
\z
