\ea\label{ex:nrexdiriiw}
Diri eweno ki kama inakap yilau kepyo.\\
\gll diri	ewe-no	ki	kama	i-nake-p	yilau	kep-yo\\
     younger.brother	older.brother-\textsc{ins}	\textsc{aph}	alone	\textsc{du}.\textsc{s}-stay-\textsc{pc}	place	3\textsc{sg}.\textsc{poss}-\textsc{loc}\\
\glt `The older and the younger brother were staying alone at their place.'
\z

\ea\label{ex:nrexyilayi}
Yilau kepyo yana inakap.\\
\gll yilau	kep-yo	yana	i-nake-p\\
     place	3\textsc{sg}.\textsc{poss}-\textsc{loc}	different	\textsc{du}.\textsc{s}-stay-\textsc{pc}\\
\glt `They were staying at another place, (namely the place, where they had built their bush camp.)'
\z

\ea\label{ex:nrexdiewak}
Diri ewe ako wonakap yilauyo. Ewe kepro busyo lo.\\
\gll diri	ewe	ako	wo-nake-p	yilau-yo	ewe	kep=ro	bus-yo	lo\\
     younger.brother	older.brother	wife	\textsc{accom}-stay-\textsc{pc}	place-\textsc{loc}	older.brother	3\textsc{sg}.\textsc{poss}=\textsc{emph}	bush(\textsc{tp})-\textsc{loc}	go.\textsc{pp}\\
\glt `The younger brother was staying with the older brother's wife at the bush camp. His older brother went off into the bush.'
\z

\ea\label{ex:nrexewword}
Duyo klapno ewe wor duyo ukeli.\\
\gll du-yo	k-le-p-no	ewe	wor	du-yo	ukeli\\
     bush-\textsc{loc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	older.brother	dog	bush-\textsc{loc}	take.with.oneself\\
\glt `When leaving for the (deep) bush the older brother took dogs (with him) into the bush.'
\z

\ea\label{ex:nrexkaworb}
Wor aska. Wor ba ukeli. Ki kama lo. Wor aska.\\
\gll wor	aska	wor	ba	ukeli	ki	kama	lo	wor	aska\\
     dog	none	dog	\textsc{neg}.\textsc{emph}	take.with.oneself	\textsc{aph}	alone	go.\textsc{pp}	dog	none\\
\glt `(No, there were) no dogs. He did not take any dogs. He went alone. (There were) no dogs.'
\z

\ea\label{ex:nrexlippue}
Bipuel rilippueko dupua. Mi yilauyo klapno diri muelno:\\
\gll bipuel	rilippue-ko	dupua	mi	yilau-yo	k-le-p-no	diri	mueli-no\\
     tree.kangaroo	look.upward.\textsc{o}.\textsc{anim}.\textsc{pl}-\textsc{rts}	two	again	place-\textsc{loc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	younger.brother	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Looking up he had seen tree kangaroos, two (of them). Then, having come (back) to the camp, he said to his younger brother:'
\z

\ea\label{ex:nrexrilipp}
``Bipuel ko barilippueko. De awe! Nuko ile. Nuko riyo bipuel ppuenei, rile inake.''\\
\gll bipuel	ko	ba-rilippue-ko	de	awe	nuko	i-le	nuko	ri-yo	bipuel	ppue-ne-i	rile	i-nake\\
     tree.kangaroo	I	\textsc{fac}-look.upward.\textsc{o}.\textsc{anim}.\textsc{pl}-\textsc{fac}	you	come.\textsc{imp}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	we.\textsc{incl}	tree-\textsc{loc}	tree.kangaroo	go.up-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	above	\textsc{du}.\textsc{s}-sit\\
\glt `{``}I have seen tree kangaroos. Come, let us go! We'll climb up the tree for the kangaroos, they are sitting above.'''
\z

\ea\label{ex:nrexsskaka}
Riyo iloro. Wor kiyopi aska. Ki solo ki kama inake.\\
\gll ri-yo	i-lo=ro	wor	kiyo-pi	aska	ki	solo	ki	kama	i-nake\\
     tree-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}=\textsc{emph}	dog	\textsc{aph}.\textsc{du}-\textsc{poss}	none	\textsc{aph}	only	\textsc{aph}	alone	\textsc{du}.\textsc{s}-stay\\
\glt `They went to the tree. There were no dogs with them. Only they. They are alone.'
\z

\ea\label{ex:nrexbipuel}
Klapno ewe ki muelne diri: ``De nakap! Ko bipuel ppuene. De nakap yeloyo.''\\
\gll k-le-p-no	ewe	ki	mueli-ne	diri	de	nake-p	ko	bipuel	ppue-ne	de	nake-p	yelo-yo\\
     \textsc{sub}-go-\textsc{pc}-\textsc{co}	older.brother	\textsc{aph}	talk.to-3\textsc{sg}.\textsc{or}	younger.brother	you	stay-\textsc{imp}	I	tree.kangaroo	go.up-3\textsc{sg}.\textsc{or}	you	stay-\textsc{imp}	ground-\textsc{loc}\\
\glt `Having come the older brother said to (his) younger brother: ``You stay (down here). I'll climb for the tree kangaroos. You stay on the ground!'''
\z

\ea\label{ex:nrexworrdo}
Diri mi wor dop lo.\\
\gll diri	mi	wor	dop	lo\\
     younger.brother	then	dog	skin	go.\textsc{pp}\\
\glt `Then the younger brother went dog-skinned.'\footnote{From now on the younger brother has the appearance of a dog and behaves like a dog. So the hunt can proceed successfully.}
\z

\ea\label{ex:nrexlsenek}
Ewe umul senek, ``Diri ke onona nakero. Dop kep worso po. Worna kamappo.''\\
\gll ewe	umul\_senek	diri	ke	ono-na	nake=ro	dop	kep	wor-so	po	wor-na	kamap-po\\
     older.brother	think.\textsc{pp}	younger.brother	\textsc{aph}	man-\textsc{aff}	stay=\textsc{emph}	skin	3\textsc{sg}.\textsc{poss}	dog-\textsc{sim}	\textsc{lv}.\textsc{pp}	dog-\textsc{aff}	become(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `The older brother thought: ``My younger brother is a human being. (But) his body (looks) like that of a dog. He became dog-like.'''
\z

\ea\label{ex:nrexkosiye}
Ewe ppuenoro rileyo. Bipuel riyo kosiyepakowe.\\
\gll ewe	ppue-no=ro	rileyo	bipuel	ri-yo	kosiyepako-we\\
     older.brother	go.up-3\textsc{sg}.\textsc{or}.\textsc{pp}=\textsc{emph}	above	tree.kangaroo	tree-\textsc{loc}	push.down.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `The older brother climbed high up after (the kangaroos). There he pushed them down from the tree (by shooting arrows at them).'
\z

\ea\label{ex:nrexworkam}
Diri ke mi loro. Dop kep tanimpo. Mi wor kamappo.\\
\gll diri	ke	mi	lo=ro	dop	kep	tanim-po	mi	wor	kamap-po\\
     younger.brother	\textsc{aph}	again	go.\textsc{pp}=\textsc{emph}	skin	3\textsc{sg}.\textsc{poss}	change(\textsc{tp})-\textsc{lv}.\textsc{pp}	then	dog	become(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `The younger brother came. His skin (had) changed. He became a dog.'
\z

\ea\label{ex:nrexbipued}
Riyopuno bipuel ppueno rileyo. Bipuel dupua dipiowo.\\
\gll riyopuno	bipuel	ppue-no	rileyo	bipuel	dupua	d-piowo\\
     then	tree.kangaroo	go.up-3\textsc{sg}.\textsc{or}.\textsc{pp}	above	tree.kangaroo	two	\textsc{lkh}-jump.\textsc{pp}\\
\glt `Then he climbed high up after the tree kangaroos. The two tree kangaroos must have  jumped (on another branch).'\footnote{Sequences (\ref{ex:nrexworkam}) and (\ref{ex:nrexbipued}) repeat and paraphrase what has already been said in (\ref{ex:nrexworrdo}) and (\ref{ex:nrexkosiye}).}
\z

\ea\label{ex:nrexlimpow}
Diri mi worna kamappo. Riyopuno bipuel dupua piyowe. Dupua kilimpowe.\\
\gll diri	mi	wor-na	kamap-po	riyopuno	bipuel	dupua	piyo-we	dupua	kilim-po-we\\
     younger.brother	again	dog-\textsc{aff}	become(\textsc{tp})-\textsc{lv}.\textsc{pp}	then	tree.kangaroo	two	take.\textsc{pp}-\textsc{du}.\textsc{o}	two	kill(\textsc{tp})-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `The younger brother became dog-like. Then he grabbed the two tree kangaroos and killed them both.'
\z

\ea\label{ex:nrexbipueb}
Bipuel dupua piowo, yeloyo seku. Dupua kilimpowe, uki akono.\\
\gll bipuel	dupua	piowo	yelo-yo	seku	dupua	kilim-po-we	uki	ako-no\\
     tree.kangaroo	two	jump.\textsc{pp}	ground-\textsc{loc}	fall.\textsc{pp}	two	kill-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{o}	husband	wife-\textsc{ins}\\
\glt `The two tree kangaroos jumped and fell on the ground. He killed the two, a male one und a female one.'\footnote{In sequences (\ref{ex:nrexbipued}) and (\ref{ex:nrexbipueb}) there is no subject dual for the explicitly dual subject \textit{bipuel dupua}. The narrator Andrew Wapi only uses the object dual.}
\z

\ea\label{ex:nrexdopnal}
Bipuel dupua wiyowe. Diri mi wor dopna lo.\\
\gll bipuel	dupua	wiyo-we	diri	mi	wor	dop-na	lo\\
     tree.kangaroo	two	hold.\textsc{pp}-\textsc{du}.\textsc{o}	younger.brother	again	dog	skin-\textsc{aff}	go.\textsc{pp}\\
\glt `He killed the two tree kangaroos. The younger brother was still a dog.'
\z

\ea\label{ex:nrexehbipu}
Ewe riyo kûnoro, muelno: ``Eh bipuel?''\\
\gll ewe	ri-yo	kûno=ro	mueli-no	eh	bipuel\\
     older.brother	tree-\textsc{loc}	go.down.\textsc{pp}=\textsc{emph}	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	eh	tree.kangaroo\\
\glt `The older brother climbed down from the tree and said to him: ``Eh, (what about) the kangaroos?'''
\z

\ea\label{ex:nrexsenisi}
``Ko bawiyewoko.'' ``De ke dop elep asa senisimpo?''\\
\gll ko	ba-wiye-we-ko	de	ke	dop	elep	asa	senisim-po\\
     I	\textsc{fac}-kill-\textsc{du}.\textsc{o}-\textsc{fac}	you	\textsc{top}	skin	2\textsc{sg}.\textsc{poss}.\textsc{emph}	how	change(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `{``}I have killed them both.'' ``(But) you, how did you change your own skin?'''
\z

\ea\label{ex:nrexdledle}
``Ko mi worso po. Ko so mi bipuel yala dele. Mi dupua ko roki wiyowe.''\\
\gll ko	mi	wor-so	po	ko	so	mi	bipuel	yala	d-le	mi	dupua	ko	ro-ki	wiyo-we\\
     I	again	dog-\textsc{sim}	\textsc{lv}.\textsc{pp}	I	believe	again	tree.kangaroo	\textsc{mod}	\textsc{lkh}-go	again	two	I	\textsc{prox}.\textsc{emph}-\textsc{aph}	kill.\textsc{pp}-\textsc{du}.\textsc{o}\\
\glt `{``}I became a dog. I thought, (otherwise) the kangaroos will certainly escape. Indeed, the two (ones), I killed them here.'''
\z

\ea\label{ex:nrexelimik}
``Eli mi ko makina ya. Nuko ile yipyo.''\\
\gll eli	mi	ko	maki-na	ya	nuko	i-le	yip-yo\\
     you.\textsc{emph}	again	I	good-\textsc{adv}	take.care.of.\textsc{pp}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	house-\textsc{loc}\\
\glt `{``}You really took great care of me. Let's go (back) to the house.'''
\z

\ea\label{ex:nrexwoworr}
Yipyo welppuo. Kupuenpoipno ako muelno: ``Diri kopi ke wor dop lo.''\\
\gll yip-yo	welppuo	k-puenpi-i-p-no	ako	mueli-no	diri	ko-pi	ke	wor	dop	lo\\
     house-\textsc{loc}	carry.up.\textsc{pp}	\textsc{sub}-cut.meat-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	wife	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	younger.brother	1\textsc{sg}-\textsc{poss}	\textsc{aph}	dog	skin	go.\textsc{pp}\\
\glt `He carried (the dead tree kangaroos) up to the house. While cutting the meat together, he said to his wife: ``My younger brother went as a dog.'''
\z

\ea\label{ex:nrexworlor}
Wor dop ki loro. Bipuel dupua oki wiyowe. Yala nuko asa poi?\\
\gll wor	dop	ki	lo=ro	bipuel	dupua	o-ki	wiyo-we	yala	nuko	asa	pi-i\\
     dog	skin	\textsc{aph}	go=\textsc{emph}	tree.kangaroo	two	\textsc{prox}-\textsc{aph}	kill.\textsc{pp}-\textsc{du}.\textsc{o}	now	we.\textsc{incl}	how	do-\textsc{du}.\textsc{a}\\
\glt `{``}He went as a dog. He killed these two tree kangaroos. (But) what are we going to do now?'''
\z

\ea\label{ex:nrexdopplo}
Diri kopi wor dop lo. Diri kopi klokniro.\\
\gll diri	ko-pi	wor	dop	lo	diri	ko-pi	klokni=ro\\
     younger.brother	1\textsc{sg}-\textsc{poss}	dog	skin	go.\textsc{pp}	younger.brother	1\textsc{sg}-\textsc{poss}	one=\textsc{emph}\\
\glt `{``}My younger brother changed into a dog, my only younger brother.'''
\z

\newpage
\ea\label{ex:nrexilnosu}
Ya yala nuko wîlno suelnei. Kimike nuko rupueno suelwilneip.\\
\gll ya	yala	nuko	wîl-no	sueli-ne-i	kimike	nuko	rupue-no	suelwili-ne-i-p\\
     sago	now	we.\textsc{incl}	plate-\textsc{ins}	cut-3\textsc{sg}.\textsc{or}-\textsc{du}.\textsc{a}	before	we.\textsc{incl}	leaf-\textsc{ins}	cut.and.carry-3\textsc{sg}.\textsc{or}-\textsc{du}.\textsc{a}-\textsc{pc}\\
\glt `{``}Now we cut him the sago pudding on a plate. Before we brought it to him cut into a leaf.'''\footnote{People used to eat sago pudding served on big leaves, or wrapped into leaves as provisions.}
\z

\ea\label{ex:nrexastaim}
Pastaim rupueno suelwiliponoip. Yala ke mi wîlno suelnei.\\
\gll pastaim	rupue-no	suelwili-pi-ne-i-p	yala	ke	mi	wîl-no	sueli-ne-i\\
     before(\textsc{tp})	leaf-\textsc{ins}	cut.and.carry-\textsc{lv}-3\textsc{sg}.\textsc{or}-\textsc{du}.\textsc{a}-\textsc{pc}	now	\textsc{aph}	again	plate-\textsc{ins}	cut-3\textsc{sg}.\textsc{or}-\textsc{du}.\textsc{a}\\
\glt `{``}Before we brought it to him in a leaf. Now we cut it into a dish for him.'''
\z

\ea\label{ex:nrexnukyai}
Nuko yaii. Nuko makina yaii. Nuko makina lukautimpoi.\\
\gll nuko	yai-i	nuko	maki-na	yai-i	nuko	maki-na	lukautim-po-i\\
     we.\textsc{incl}	take.care.of-\textsc{du}.\textsc{a}	we.\textsc{incl}	good-\textsc{adv}	take.care.of-\textsc{du}.\textsc{a}	we.\textsc{incl}	good-\textsc{adv}	take.care(\textsc{tp})-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `{``}We take care (of him). We take great care. We took great care.'''
\z

\ea\label{ex:nrexoplolo}
Diri kopi ke wor dop lo. Nuko makina lukautimpoi.''\\
\gll diri	ko-pi	ke	wor	dop	lo	nuko	maki-na	lukautim-po-i\\
     younger.brother	1\textsc{sg}-\textsc{poss}	\textsc{aph}	dog	skin	go.\textsc{pp}	we.\textsc{incl}	good-\textsc{adv}	take.care(\textsc{tp})-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `{``}My younger brother changed into a dog. We took great care (of him).'''
\z

\ea\label{ex:nrexdiriim}
Boyopuno riyo nakero. Diri miro nakero.\\
\gll boyopuno	ri-yo	nake=ro	diri	mi=ro	nake=ro\\
     later	\textsc{dist}-\textsc{loc}	stay=\textsc{emph}	younger.brother	again=\textsc{emph}	stay=\textsc{emph}\\
\glt `Later he stays there. The younger brother stays (with us).'
\z

\ea\label{ex:nrexmikimi}
Mi kimike ki so nakapno. Mi ono dop kamappo.\\
\gll mi	kimike	ki	so	nake-p-no	mi	ono	dop	kamap-po\\
     again	before	\textsc{aph}	like	stay-\textsc{pc}-\textsc{co}	then	man	skin	become(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `While first staying like this, he then became human again.'
\z

\ea\label{ex:nrexkowors}
Yalaka ko mi bakamappoko onona. Pastaim ko worso kamappo.\\
\gll yalaka	ko	mi	ba-kamap-pi-ko	ono-na	pastaim	ko	wor-so	kamap-po\\
     now	I	again	\textsc{fac}-become(\textsc{tp})-\textsc{lv}-\textsc{fac}	man-\textsc{aff}	before(\textsc{tp})	I	dog-\textsc{sim}	become(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `(Finally the younger brother said to the others:) ``From now on I have become man-like again. Before I had changed into a dog.'''
\z

\ea\label{ex:nrextasass}
Em tasol tenk yu.\\
\gll em	tasol	tenk	yu\\
     \textsc{aph}(\textsc{tp})	only(\textsc{tp})	thank(\textsc{tp})	you(\textsc{tp})\\
\glt `The end, thank you.'
\z
