\ea\label{ex:nrexberpap}
Pper pina sele kepine pper pul seliye. Pul pisisi, pei kiniyo poli. Ili kep wali kep le. Pele pi, pper pele ppipe papi. Pul mini. Pul kiminipno ikoi po, papulip.\\
\gll pper	pi-na	sele	kepine	pper	pul	seliye	pul	pisisi	pei	kiniyo	poli	ili	kep	wali	kep	le	pele	pi	pper	pele	ppipe	papi	pul	mini	pul	k-mini-p-no	ikoi	po	papuli-p\\
     penis.gourd	do-\textsc{purp}	garden	prepare	gourd	seed	sow	seed	sprout	sprout	many	be.there	bunch	3\textsc{sg}.\textsc{poss}	stem	3\textsc{sg}.\textsc{poss}	go	leaf	do	gourd	leaf	flower	get.ready.\textsc{pl}.\textsc{o}	fruit	come.hither	fruit	\textsc{sub}-come.hither-\textsc{pc}-\textsc{co}	big	do.\textsc{pp}	be.there.\textsc{pl}-\textsc{pc}\\
\glt `For making penis gourds one prepares a garden and sows gourd seeds. The seeds  build sprouts, there are many (of them). Their bunches and stems spread. They make leaves, the gourds get ready leaves and flowers. Fruits come. When the fruits have come, they became big. There they were.'
\z

\ea\label{ex:nrexsikedo}
Yukume klapno seleyo pper kep riyeko. Bapok sike. Dob riyepo, pper kiniyo bamonwepko. Rondupua rokini pper lapo, yipyo meli. Royowe.\\
\gll yukume	k-le-p-no	sele-yo	pper	kep	riye-ko	ba-pi-ko	sike	dob	riye-po	pper	kiniyo	ba-mini-wepi-ko	rondupua\_rokini	pper	lapo	yip-yo	meli	royo-we\\
     man.\textsc{sg}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	garden-\textsc{loc}	gourd	3\textsc{sg}.\textsc{poss}	see.\textsc{o}.\textsc{inanim}-\textsc{rts}	\textsc{fac}-do-\textsc{fac}	strong	eye	see.\textsc{o}.\textsc{inanim}-\textsc{lv}.\textsc{pp}	gourd	many	\textsc{fac}-come.hither-\textsc{quant}.\textsc{s}-\textsc{fac}	three	gourd	harvest.\textsc{pp}	house-\textsc{loc}	carry.\textsc{pl}.\textsc{o}	lay.\textsc{pp}-\textsc{ter}\\
\glt `Having gone to the garden, a man looked at his gourds. They are strong indeed. He looked, many have come up. He would harvest three gourds and carry them to the house. He kept them ready (for making penis gourds).'
\z

\ea\label{ex:nrexlepapi}
Bî sepei, pper pul popiwepu nanano. Sû bopoyo wapo. Klokni piyo, nanano lop, lepapo, lepapuwepu. Rondupua rokini wapowe, uroyo niswe. Laliyo yip bîyo. Laliyowe uro roise.\\
\gll bî	sepei	pper	pul	popi-wepu	nana-no	sû	bopo-yo	wapo	klokni	piyo	nana-no	lop	lepapo	lepapi-wepu	rondupua\_rokini	wapo-we	uro-yo	nis-we	laliyo	yip	bîyo	laliyo-we	uro	roise\\
     hole	drill	gourd	seed	take.out-\textsc{quant}.\textsc{o}.\textsc{pp}	small.knife-\textsc{ins}	fire	ashes-\textsc{loc}	put.together.\textsc{pp}	one	take.\textsc{pp}	small.knife-\textsc{ins}	scratch.\textsc{pp}	make.a.pattern.\textsc{pl}.\textsc{o}.\textsc{pp}	make.a.pattern-\textsc{quant}.\textsc{o}.\textsc{pp}	three	put.together.\textsc{pp}-\textsc{ter}	netbag-\textsc{loc}	put.in.\textsc{pp}-\textsc{ter}	hang.\textsc{pp}	house	inside	hang.\textsc{pp}-\textsc{ter}	netbag	together\\
\glt `He drilled a hole, took out the gourd seeds with a knife, and put them on the ashes  of the fire. He took one, scratched it with the knife, made a pattern, designed it with many scratches.  Three (gourds) altogether he treated (like this) and put them into a netbag. He hung  (the netbag) inside the house. He hung up (the penis gourds) with the netbag.'
\z

\ea\label{ex:nrexdiewok}
Pi kepro ikoi ari. Moniseso upuna papi. Die wok kep ikoiele.\\
\gll pi	kep=ro	ikoi	ari	moniseso	upuna	papi	die	wok	kep	ikoiele\\
     do	3\textsc{sg}.\textsc{poss}=\textsc{emph}	big	no	very.small	alright	produce	grass.skirt	work	3\textsc{sg}.\textsc{poss}	very.big\\
\glt `Their making is not big (work). It is very small (work). It is alright to produce them. Grass skirts, (by contrast), their production is labourious.'
\z