\ea\label{ex:nrexpuersu}
Ko puer s没ni bese puer. Ko die s没ni. Ko die puer nopi.\\
\gll ko	puer	s没ni	bese	puer	ko	die	s没ni	ko	die	puer	nopi\\
     I	thread	roll	tulip	thread	I	grass.skirt	roll	I	grass.skirt	thread	produce\\
\glt `I roll a thread, a \textit{tulip}-thread. I roll the grass skirt. I produce a thread for the grass skirt.'
\z

\ea\label{ex:nrexdiewiy}
Ko die wiyi. Ko die wiyi pueryo powole.\\
\gll ko	die	wiyi	ko	die	wiyi	puer-yo	powole\\
     I	grass.skirt	string	I	grass.skirt	string	thread-\textsc{loc}	keep.doing\\
\glt `I string the grass skirt. I string (strips for) the grass skirt on the thread, going on and on.'
\z

\ea\label{ex:nrexwayeup}
Elka pono kipika po. Ko bapok. Ko malapi. Ko waye, upuna. Ko upuna po.\\
\gll el-ka	pi-no	kipi-ka	po	ko	ba-pi-ko	ko	malapi	ko	waye	upuna	ko	upuna	po\\
     belly-\textsc{path}	do-\textsc{co}	back-\textsc{path}	do.\textsc{pp}	I	\textsc{fac}-do-\textsc{fac}	I	measure	I	try	alright	I	well	do.\textsc{pp}\\
\glt `After doing it along the belly, I did it along the back. I have done it. I measure. I try, alright, I did it well.'
\z

\newpage
\ea\label{ex:nrexyulpon}
Ko poniye die elka poli yul kipika pi. Kipika ko poro. Ko lumi elka komiye, yul woleno.\\
\gll ko	poniye	die	el-ka	poli	yul	kipi-ka	pi	kipi-ka	ko	pi=ro	ko	lumi	el-ka	komiye	yul	wole-no\\
     I	wrap	grass.skirt	belly-\textsc{path}	be.there	joint	back-\textsc{path}	do	back-\textsc{path}	I	do=\textsc{emph}	I	knot	belly-\textsc{path}	hide	joint	move.further-\textsc{co}\\
\glt `I wrap it. The grass skirt goes along the belly, and I join it along the back. I do it along the back. I hide the knot at the front, going further as joint.'
\z

\ea\label{ex:nrexponpue}
Ko baponiyeko. Dop kopiyo poli. Ko poniyepue.\\
\gll ko	ba-poniye-ko	dop	ko-pi-yo	poli	ko	poniye\_pue\\
     I	\textsc{fac}-wrap-\textsc{fac}	body	1\textsc{sg}-\textsc{poss}-\textsc{loc}	be.there	I	wrap\_walk.around\\
\glt `I have wrapped it, and it stays on my body. I walk around wrapped up (in the grass skirt).'
\z

\ea\label{ex:nrexdiekob}
Mi ko ba pi. Die ko ba powe. Ba mi koro poniyena. Ko layowe.\\
\gll mi	ko	ba	pi	die	ko	ba	pi-we	ba	mi	ko=ro	poniye-na	ko	laye-we\\
     again	I	other	do	grass.skirt	I	other	do-\textsc{ter}	other	again	I=\textsc{emph}	wrap-\textsc{purp}	I	lay-\textsc{ter}\\
\glt `I make again one more. I make another grass skirt. Again one more, for wrapping me. I keep it ready.'
\z

\ea\label{ex:nrexdiesuk}
Die suku pono ppulae. Ko nepei, die suku. Ko die ba poniye, puene. Ko baponiyeko.\\
\gll die	suku	pi-no	ppulae	ko	nepei	die	suku	ko	die	ba	poniye	puene	ko	ba-poniye-ko\\
     grass.skirt	old	\textsc{lv}-\textsc{co}	bad	I	take.off	grass.skirt	old	I	grass.skirt	other	wrap	new	I	\textsc{fac}-wrap-\textsc{fac}\\
\glt `When the grass skirt gets old, it is bad. I take it off, the old grass skirt. I wrap another grass skirt (around me), a new (one). I have wrapped it.'
\z

\ea\label{ex:nrexdieroy}
Mi ko yala ba pi. Ko bapapko. Kiniyo ko royowe.\\
\gll mi	ko	yala	ba	pi	ko	ba-papi-ko	kiniyo	ko	roye-we\\
     again	I	now	other	do	I	\textsc{fac}-make.\textsc{pl}.\textsc{o}-\textsc{fac}	many	I	lay-\textsc{ter}\\
\glt `Now I make again one more. I have got it ready. I keep ready many.'
\z

\newpage
\ea\label{ex:nrexdiewap}
De kuso ponake. Die papinake. Kiniyo wapowe.\\
\gll de	kuso	pi-nake	die	papi-nake	kiniyo	wape-we\\
     you	always	make-\textsc{dur}	grass.skirt	produce.\textsc{pl}.\textsc{o}-\textsc{dur}	many	put.for.one's.advantage-\textsc{ter}\\
\glt `You always make them. You keep getting ready grass skirts. Many you put in stock.'
\z

\ea\label{ex:nrexdieepu}
Ko puyo le. Die puene ko anno wili. Ko die suku popi. Ppulae po. Ko epueyo piyi.\\
\gll ko	pu-yo	le	die	puene	ko	an-no	wili	ko	die	suku	popi	ppulae	pi	ko	epue-yo	piyi\\
     I	river-\textsc{loc}	go	grass.skirt	new	I	hand-\textsc{ins}	carry	I	grass.skirt	old	take.away	bad	\textsc{lv}	I	weed-\textsc{loc}	throw\\
\glt `I go to the river. The new grass skirt I carry in my hands. The old grass skirt I take  away. It is bad. I throw it into the weeds.'
\z

\ea\label{ex:nrexpuldie}
Ko pulmopi. Pul komopopno ko olyo pini. Ko die puene poniye. Yipyo le, nake.\\
\gll ko	pulmopi	pul\_ki-mopi-p-no	ko	ol-yo	pini	ko	die	puene	poniye	yip-yo	le	nake\\
     I	bathe	bathe\_\textsc{sub}-bathe-\textsc{pc}-\textsc{co}	I	hill-\textsc{loc}	come.up.hither	I	grass.skirt	new	wrap	house-\textsc{loc}	go	sit\\
\glt `I bathe (in the river). Having bathed I climb the river bank. I wrap the new grass skirt, go to the house and stay (there).'
\z

\ea\label{ex:nrexdukdie}
Duku die duam die maki. Die ar sukewole. Upuna poli.\\
\gll duku	die	duam	die	maki	die	ar	suke-wole	upuna	poli\\
     kind.of.sago.palm	grass.skirt	kind.of.sago.palm	grass.skirt	good	grass.skirt	\textsc{neg}	tear-\textsc{cpl}	alright	be.there\\
\glt `\textit{Duku}-grass skirts and \textit{duam}-grass skirts are good. They don't tear. They are fine.'
\z

\largerpage
\ea\label{ex:nrexdiekop}
Die ko papino. Pi kepro ikoiele. Mi dop kemiye.\\
\gll die	ko	papi-no	pi	kep=ro	ikoiele	mi-dop\_kemiye\\
     grass.skirt	I	produce.\textsc{pl}.\textsc{o}-\textsc{co}	do	3\textsc{sg}.\textsc{poss}=\textsc{emph}	very.big	\textsc{iter}-feel.tired\\
\glt `I produce many grass skirts. Their making is a lot of work. I feel very tired (of doing it) over and again.'
\z
