\ea\label{ex:nrexriwies}
Ko ri wies lipeli numuelyo.\\
\gll ko	ri\_wies	lipeli	numuelyo\\
     I	kind.of.tree	seek	far.away\\
\glt `I seek a \textit{wies}-tree, deep in the bush.'
\z

\ea\label{ex:nrexnokomi}
Ri wies nokomiyouli. Ko dob powole. Yala aryo neki?\\
\gll ri\_wies	nokomiye-uli	ko	dob	pi\_wole	yala	aryo	neki\\
     kind.of.tree	hide.well-\textsc{prog}	I	eye	\textsc{lv}\_move.further	now	where	stand\\
\glt `The \textit{wies}-trees are quite hidden. I look around: Where does (one) stand now?'
\z

\ea\label{ex:nrexwiesun}
Ri wies uneki! Ko pele wapi.\\
\gll ri\_wies	u-neki	ko	pele	wapi\\
     kind.of.tree	\textsc{dfac}-stand	I	leaf	collect\\
\glt `Here stands a \textit{wies}-tree. I collect leaves (of the tree).'
\z

\ea\label{ex:nrexwieswa}
Bawapko. Ko yipyo wili.\\
\gll ba-wapi-ko	ko	yip-yo	wili\\
     \textsc{fac}-collect-\textsc{fac}	I	house-\textsc{loc}	carry\\
\glt `They are collected. I bring them to the house.'
\z

\ea\label{ex:nrexwiessi}
Ko wies pele ipi kloknino si, die ino pina.\\
\gll ko	wies	pele	ipi	klokni-no	si	die	i-no	pi-na\\
     I	kind.of.tree	leaf	pot	one-\textsc{ins}	cook	grass.skirt	\textsc{dist}-\textsc{ins}	make-\textsc{purp}\\
\glt `I cook the \textit{wies}-leaves in one (big) pot for dyeing the grass skirts in the liquid.'
\z

\ea\label{ex:nrexdieaep}
Die royepake. Die mi yala aeppu pi.\\
\gll die	royepake	die	mi	yala	aeppu	pi\\
     grass.skirt	lay.down	grass.skirt	then	now	red	\textsc{lv}\\
\glt `I put the grass skirts into (the  pot), then the grass skirts will become red now.'
\z

\ea\label{ex:nrexwiesyo}
Yena yako die melipulupi, wiesyo royepake.\\
\gll yena	yako	die	meli\_pulupi	wies-yo	royepake\\
     people	women	grass.skirt	carry.\textsc{pl}.\textsc{o}\_come.\textsc{pl}.\textsc{s}	kind.of.tree-\textsc{loc}	lay.down\\
\glt `The people, the women, bring (their) grass skirts and put them into the \textit{wies}(-leach).'
\z

\ea\label{ex:nrexkwerra}
Kwerno rapiyekûpe, yipyo meli.\\
\gll kwerno	rapiye\_kûpe	yip-yo	meli\\
     afternoon	fetch\_go.down.\textsc{pl}.\textsc{a}	house-\textsc{loc}	carry.\textsc{pl}.\textsc{o}\\
\glt `In the afternoon they take (the grass skirts) out and carry them to (their) houses.'
\z

\ea\label{ex:nrexdienin}
Die niniyo roye.\\
\gll die	nini-yo	roye\\
     grass.skirt	sun-\textsc{loc}	lay\\
\glt `They put the grass skirts in the sun.'
\z

\ea\label{ex:nrexdiedie}
Die kopi aeppu bapok.\\
\gll die	ko-pi	aeppu	ba-po-ko\\
     grass.skirt	1\textsc{sg}-\textsc{poss}	red	\textsc{fac}-\textsc{lv}-\textsc{fac}\\
\glt `My grass skirts have become red.'
\z