\ea\label{ex:nrexkooilp}
Ko oil pi sekena aeuna. Aeu ko ino pi boyo. Ba sekena ba aeuna.\\
\gll ko	oil	pi	seke-na	aeu-na	aeu	ko	i-no	pi	boyo	ba	seke-na	ba	aeu-na\\
     I	oil	make	hair-\textsc{aff}	sago.pancake-\textsc{aff}	sago.pancake	I	\textsc{dist}-\textsc{ins}	make	later	other	hair-\textsc{aff}	other	sago.pancake-\textsc{aff}\\
\glt `I make oil for the hair and for sago pancakes. Sago pancakes I make with it later. Some for the hair, some for sago pancakes.'
\z

\ea\label{ex:nrexsuoyuk}
Ko suo rapiye, yuki. Ko oil pi.\\
\gll ko	suo	rapiye	yuki	ko	oil	pi\\
     I	coconut	fetch	shred	I	oil	make\\
\glt `I fetch coconuts and shred them. I make oil.'
\z

\ea\label{ex:nrexsuosek}
Ko suo sekapi, suo yani ipiyo. Ko disyo pokûne emna.\\
\gll ko	suo	sekapi	suo	yani	ipi-yo	ko	dis-yo	pokûne	em-na\\
     I	coconut	press	coconut	knead.in.water	pot-\textsc{loc}	I	dish-\textsc{loc}	pour	tomorrow-\textsc{aff}\\
\glt `I press the copra and knead the copra in a pot (with some water). (Then) I pour (the liquid) in a dish for (further use) tomorrow.'
\z

\ea\label{ex:nrexoilmap}
Mi yala ko apoyo pi. Ko sûyo yowe, sû mappane.\\
\gll mi	yala	ko	apo-yo	pi	ko	sû-yo	yowe	sû\_mappane\\
     then	now	I	pan-\textsc{loc}	make	I	fire-\textsc{loc}	heat	light.a.fire.thither\\
\glt `Then I process (the copra liquid) in a pan. I heat it on the fire. I light a fire.'
\z

\ea\label{ex:nrexsuoyan}
Sû beri pianpane. Ko suo yani moli.\\
\gll sû	beri	piane\_pane	ko	suo\_yani	moli\\
     fire	flame	blaze.up	I	copra.milk	make.boil\\
\glt `The flame of the fire blazes up. I let the copra milk boil.'
\z

\ea\label{ex:nrexsuosre}
Suo yani sre. Oil bapok.\\
\gll suo\_yani	sre	oil	ba-pi-ko\\
     copra.milk	boil.away	oil	\textsc{fac}-make-\textsc{fac}\\
\glt `The copra milk boils away. The oil is made.'
\z

\ea\label{ex:nrexsuonis}
Ko puliyepami, layowe. Amanisipe. Banisko.\\
\gll ko	puliyepami	laye-we	am-a-nisi-ipe	ba-nisi-ko\\
     I	take.off.hither	put-\textsc{ter}	\textsc{grad}-\textsc{imp}3-become.cool-\textsc{ant}	\textsc{fac}-become.cool-\textsc{fac}\\
\glt `I take it off (the fire) and put it aside. (The oil) has to cool down first. It has cooled.'
\z

\ea\label{ex:nrexkospun}
Ko spunno noriyepake. Ko botolyo sipake. Oil maki.\\
\gll ko	spun-no	noriyepake	ko	botol-yo	sipake	oil	maki\\
     I	spoon\textsc{(tp)}-\textsc{ins}	fill.in.down	I	bottle-\textsc{loc}	pour.down.into	oil	good\\
\glt `I bottle it with a spoon. I pour it into a bottle. It's good oil.'
\z

\ea\label{ex:nrexoilmak}
Oil ko ikoina po, ko maketyo laye.\\
\gll oil	ko	ikoina	po	ko	maket-yo	laye\\
     oil	I	much	make.\textsc{pp}	I	market-\textsc{loc}	lay\\
\glt `(When) I made a lot of oil, I offer it on the market (for selling).'
\z

\ea\label{ex:nrexsuoaeu}
Ko ikoina ar po, monina po, ko aeu ino pi. Ko ar laye maketyo moniseso.\\
\gll ko	ikoina	ar	po	moni-na	po	ko	aeu	i-no	pi	ko	ar	laye	maket-yo	moniseso\\
     I	much	\textsc{neg}	make.\textsc{pp}	little-\textsc{adv}	make.\textsc{pp}	I	sago.pancake	\textsc{dist}-\textsc{ins}	make	I	\textsc{neg}	lay	market-\textsc{loc}	very.little\\
\glt `(In case) I didn't make a lot, but only a little, I bake sago pancakes with it. I don't offer it on the market, it is too little.'
\z

\ea\label{ex:nrexaeuino}
Klokni solo, ko ar laye. Aeu ino pi.\\
\gll klokni	solo	ko	ar	laye	aeu	i-no	pi\\
     one	only	I	\textsc{neg}	lay	sago.pancake	\textsc{dist}-\textsc{ins}	make\\
\glt `(In case) it's only one (bottle), I don't offer it (for selling). I make sago pancakes with it.'
\z