\ea\label{ex:t7exulmoro}
Ko ul more. Ri malalpore, ul roise wepulo yipyo, wapo.\\
\gll ko	ul	mo=ro	ri	malalpo=ro	ul	roise	wepulo	yip-yo	wapo\\
     I	bamboo	cut.\textsc{pp}=\textsc{emph}	wood	cut.fire.wood.\textsc{pp}=\textsc{emph}	bamboo	together	bring.\textsc{pp}	house-\textsc{loc}	put.together.\textsc{pp}\\
\glt `I cut bamboo. I cut firewood and brought it to the house together with the bamboo. I put (everything) together.'
\z

\ea\label{ex:t7exkoripe}
Ko ri pesoupo, ri ba kaliyo. Ko pu ulyo nororpo.\\
\gll ko	ri	pesoupo	ri	ba	kaliyo	ko	pu	ul-yo	nororpo\\
     I	wood	put.upright.\textsc{pp}	wood	other	lay.horizontally.\textsc{pp}	I	water	bamboo-\textsc{loc}	fill.up.\textsc{pl}.\textsc{o}.\textsc{pp}\\
\glt `I put some pieces of wood vertically, some I laid horizontally. I filled up the water into the bamboo tubes.'
\z

\ea\label{ex:t7exkosuyo}
Ko sûyo wauna, sû mappo.\\
\gll ko	sû-yo	wauna	sû\_mappo\\
     I	fire-\textsc{loc}	put.side.by.side.\textsc{pp}	light.a.fire.\textsc{pp}\\
\glt `I put them (the bamboo tubes) side by side on the fire. I lit the fire.'
\z

\ea\label{ex:t7exkoduew}
Ko due wîlyo piyo. Yaup mol, yaup bamolwepuko.\\
\gll ko	due	wîl-yo	piyo	yaup	mol	yaup	ba-moli-wepi-ko\\
     I	sago	dish-\textsc{loc}	take.\textsc{pp}	water.for.cooking	boil.\textsc{pp}	water.for.cooking	\textsc{fac}-boil-\textsc{quant}.\textsc{s}-\textsc{fac}	\\
\glt `I took sago in a dish. The water boiled. The water (in the bamboo tubes) boiled well.'
\z

\ea\label{ex:t7exyaupul}
Yaup ul isiye ko sipako. Ko due wîl roise uleipiyo. Ko due nonopiyei, due dipini. Due dû puso po.\\
\gll yaup	ul	isiye	ko	sipako	ko	due	wîl	roise	uleipiyo	ko	due	nonopiye	due	d-pini	due	dû	pu-so	po\\
     water.for.cooking	bamboo	together	I	pour.down.into.\textsc{pp}	I	sago	dish	with	toss.gently.\textsc{pp}	I	sago	toss.heavily	sago	\textsc{lkh}-go.up.hither	sago	flour	water-\textsc{sim}	\textsc{lv}\\
\glt `I poured the water down (from) the bamboo (into the dish). I gently tossed the sago in  the dish. (If) I toss the sago heavily, the sago is likely to spill over. (Then) the sago is like water.'
\z

\ea\label{ex:t7expisulw}
Ko due sul popi, sul wisi polino dueyo.\\
\gll ko	due	sul	popi	sul	wisi	poli-no	due-yo\\
     I	sago	crumb	take.away.\textsc{pl}.\textsc{o}	crumb	black	be.there-\textsc{co}	sago-\textsc{loc}\\
\glt `I take away the sago crumbs, the black crumbs that are in the sago.'
\z

\ea\label{ex:t7exyaupro}
Ul yaup roise ko piyo, sipako dueyo. Ko lakwole rondupua rondupua. Ko sipakowepu dueyo.\\
\gll ul	yaup	roise	ko	piyo	sipako	due-yo	ko	lakiwole	rondupua\_rondupua	ko	sipake-wepu	due-yo\\
     bamboo	water.for.cooking	together	I	take.\textsc{pp}	pour.down.into.\textsc{pp}	sago-\textsc{loc}	I	count.go.further	four	I	pour.down.into-\textsc{quant}.\textsc{o}.\textsc{pp}	sago-\textsc{loc}\\
\glt `I took the bamboo with the water and poured it down on the sago. I count four (tubes). I pour them down on the sago.'
\z

\ea\label{ex:t7exsuibas}
Ya sui basuiko. Ko rupueyo supopo.\\
\gll ya\_sui	ba-sui-ko	ko	rupue-yo	supopo\\
     sago.become.ready	\textsc{fac}-die-\textsc{fac}	I	leaf-\textsc{loc}	cut.into.pieces.\textsc{pl}.\textsc{o}.\textsc{pp}\\
\glt `The sago jelly gets ready. It is ready. I cut it into pieces on leaves.'
\z

\ea\label{ex:t7exulyono}
Pu ulyo noriyo. Mi ko bese si. Bese kisipno mi ko sûyo wolana ul isiye.\\
\gll pu	ul-yo	noriyo	mi	ko	bese	si	bese	k-si-p-no	mi	ko	sû-yo	wolana	ul	isiye\\
     water	bamboo-\textsc{loc}	fill.in.\textsc{pp}	again	I	tulip	cook	tulip	\textsc{sub}-cook-\textsc{pc}-\textsc{co}	again	I	fire-\textsc{loc}	set.thither.\textsc{pp}	bamboo	together\\
\glt `I filled water into the bamboo tubes. Now I cook the \textit{tulip}-vegetables. Having cooked the \textit{tulip}-vegetables, I set (water) on the fire again (filled in) bamboo tubes.'
\z

\ea\label{ex:t7exsuyora}
Sûyo rap komolipno bese ro. Baroko. Ko puliyo ul roise.\\
\gll sû-yo	rap	k-moli-p-no	bese	ro	ba-re-ko	ko	puliyo	ul	roise\\
     fire-\textsc{loc}	leafy.vegetable	\textsc{sub}-boil-\textsc{pc}-\textsc{co}	tulip	get.done.\textsc{pp}	\textsc{fac}-get.done-\textsc{fac}	I	take.off.\textsc{pp}	bamboo	together\\
\glt `The \textit{rap}-vegetables on the fire boiled, the \textit{tulip}-vegetables (also) got done. The vegetables are done. I took them off the fire together with the bamboo.'
\z

\ea\label{ex:t7exkonepp}
Ul ko neppino rino kiso. Ko kiseipno wîlyo unepako.\\
\gll ul	ko	neppi-no	ri-no	kiso	ko	kisei-p-no	wîl-yo	uneipako\\
     bamboo	I	bush.knife-\textsc{ins}	stick-\textsc{ins}	split.lengthwise.\textsc{pp}	I	split.lengthwise-\textsc{pc}-\textsc{co}	dish-\textsc{loc}	pour.down.\textsc{pp}\\
\glt `I split the bamboo tubes lengthwise with the bush knife or with a stick. After splitting it I poured the vegetables down into the dish.'
\z

\ea\label{ex:t7exkoyano}
Ko yano ni. Ko bakesiyoko, kuru.\\
\gll ko	ya-no	ni	ko	ba-kesiye-ko	kuru\\
     I	sago-\textsc{ins}	eat	I	\textsc{fac}-use.up-\textsc{fac}	be.finished\\
\glt `I eat it with sago. I have eaten it up, and (the meal is) finished.'
\z