\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Ko ul more. Ri malalpore, ul roise wepulo yipyo, wapo. Ko ri pesoupo, ri ba kaliyo. Ko pu ulyo nororpo. Ko sûyo wauna, sû mappo. Ko due wîlyo piyo. Yaup mol, yaup bamolwepuko. Yaup ul isiye ko sipako. Ko due wîl roise uleipiyo. Ko due nonopiyei, due dipini. Due dû puso po. Ko due sul popi, sul wisi polino dueyo. Ul yaup roise ko piyo, sipako dueyo. Ko lakwole rondupua rondupua. Ko sipakowepu dueyo. Ya sui basuiko. Ko rupueyo supopo.}}
    \ParallelRText{I cut bamboo. I cut firewood and brought it to the house together with the bamboo. I put everything together. I put some pieces of wood vertically, some I laid horizontally. I filled up the water into the bamboo tubes. I put the bamboo tubes side by side in the fire. I lit the fire. I took sago in a dish. The water boiled. The water in the bamboo tubes boiled well. Now I poured the water from the bamboo into the dish. I gently tossed the sago in the dish. If I toss the sago heavily, the sago is likely to spill over. Then the sago is like water. I take away the sago crumbs, the black crumbs that are in the sago. I took the bamboo with the water and poured it down on the sago. I count four tubes. I pour them down on the sago. The sago jelly gets ready. It is ready. I cut it into pieces on leaves.}
\end{Parallel}
    \medskip
\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Pu ulyo noriyo. Mi ko bese si. Bese kisipno mi ko sûyo wolana ul isiye. Sûyo rap komolipno bese ro. Baroko. Ko puliyo ul roise. Ul ko neppino rino kiso. Ko kiseipno wîlyo unepako. Ko yano ni. Ko bakesiyoko, kuru.}}
    \ParallelRText{I filled again water into the bamboo tubes. Now I cook the \textit{tulip}-vegetables. Having cooked the \textit{tulip}-vege\-tables, I boil more water. Again I fill bamboo tubes and put them on the fire. The \textit{rap}-vegetable on the fire boiled, the \textit{tulip}-vegetable also got done. The vegetables are done. I took them off the fire together with the bamboo. I split the bamboo tubes lengthwise with the bush knife or with a stick. After splitting it I poured the vegetables into the dish. I eat them with sago. I have eaten up everything, and (the meal is) finished.}
\end{Parallel}
    \medskip