\ea\label{ex:nrexyembua}
Uki kopi yem buai pakono lu iwan misoru. Buai pakono luiwepu.\\
\gll uki	ko-pi	yem	buai	pako-no	lu	iwan	misoru	buai	pako-no	lui-wepu\\
     husband	1\textsc{sg}-\textsc{poss}	crowned.pigeon	white.people	bow-\textsc{ins}	shoot.\textsc{pp}	hornbill	also	white.people	bow-\textsc{ins}	shoot-\textsc{quant}.\textsc{o}.\textsc{pp}\\
\glt `My husband	shot crowned pigeons with a gun, hornbills also. With the gun he  	shot them efficiently.'
\z

\ea\label{ex:nrexinidop}
Ri ini dopyo nake, yûr riyo nake, ko pe pako weri. Ko pulapi, ko lui.\\
\gll ri	ini	dopyo	nake	yûr	ri-yo	nake	ko	pe	pako	weri	ko	pulapi	ko	lui\\
     tree	branch	close	sit	bird	\textsc{dist}-\textsc{loc}	sit	I	arrow	bow	nock.an.arrow	I	draw.a.bow	I	shoot\\
\glt `If (the bird) sits on a branch close by, if the bird sits there, I nock an arrow. (Then) I draw the bow and shoot.'
\z

\ea\label{ex:nrexyursek}
Yûr seki, yûr basekiko. Yeloyo lili. Ko yûr piyeppue.\\
\gll yûr	seki	yûr	ba-seki-ko	yelo-yo	lili	ko	yûr	piye\_ppue\\
     bird	fall	bird	\textsc{fac}-fall-\textsc{fac}	ground-\textsc{loc}	be.there	I	bird	take\_go.up\\
\glt `The bird falls, the bird has fallen (down). It lies on the ground. I pick the bird up.'
\z

\ea\label{ex:nrexyuamai}
Yûr ri ini numuelyo nake, amainuyo nake, ko buai pakono pulapi. Ko pe pako bîyo pane. Ko pulapi.\\
\gll yûr	ri	ini	numuelyo	nake	amainuyo	nake	ko	buai	pako-no	pulapi	ko	pe	pako	bîyo	pane	ko	pulapi\\
     bird	tree	branch	far.away	sit	aloft	sit	I	white.people	bow-\textsc{ins}	train.a.gun	I	cartridge	gun	inside	put.thither	I	train.a.gun\\
\glt `If the bird sits on a tree branch far away, if it sits aloft, I aim at it with a gun. I put the cartridge inside the gun. I aim the gun (at it).'
\z