\ea\label{ex:nrexsekbop}
Seke suo roise yasiye. Seke bopyo pane, luo ron roise lole.\\
\gll seke	suo	roise	yasiye	seke	bop-yo	pane	luo	ron	roise	lole\\
     hair	coconut	together	plant	hair	kind.of.mussel-\textsc{loc}	do.thither	breadfruit	sap	together	tie\\
\glt `(I) insert hair and coconut fibre together. (I) put the hairy ingredients into a shell and glue it together with breadfruit sap.'
\z

\ea\label{ex:nrexkomarr}
Ko komiyewep. Uki kep yala ar riyei.\\
\gll ko	komiye-we-p	uki	kep	yala	ar	riyei\\
     I	hide-\textsc{ter}-\textsc{pc}	husband	3\textsc{sg}.\textsc{poss}	\textsc{mod}	\textsc{neg}	see.\textsc{o}.\textsc{inanim}\\
\glt `I would hide it well. Her husband will not see it.'
\z

\ea\label{ex:nrexpuluel}
Ko ri pul piyeke, wepule. Ko ako suel ino pine.\\
\gll ko	ri\_pul	piye-ke	wepule	ko	ako	suel	i-no	pi-ne\\
     I	liquid.of.tree	take-\textsc{ingr}	bring	I	wife	tobacco	\textsc{dist}-\textsc{ins}	do-3\textsc{sg}.\textsc{or}\\
\glt `I go take the liquid of a certain tree and bring it (home). I mix the wife's tobacco with it.'
\z

\ea\label{ex:nrexaeunem}
Aeune mi yala ar pi. Lil mupiye. Yala ruri pi.\\
\gll aeune	mi	yala	ar	pi	lil	mupiye	yala	ruri	pi\\
     menstruation	then	\textsc{mod}	\textsc{neg}	do	blood	reduce	now	child	do\\
\glt `Then the menstruation doesn't come. The blood is blocked. Now there will be a child.'
\z

\ea\label{ex:nrexsuippu}
Ruri elyo sui ppul seki. Ruri ppul seki. Maki ari.\\
\gll ruri	el-yo	sui	ppul	seki	ruri	ppul	seki	maki	ari\\
     child	belly-\textsc{loc}	die	fetus	fall	child	fetus	fall	good	no\\
\glt `(If) the child dies in the belly, the fetus falls. The fetal child falls (out). (This) is not good.'
\z

\ea\label{ex:nrexraripw}
Bî rarip, wilikûpi bîyo. Bî sowe.\\
\gll bî	rari-p	wilikûpi	bî-yo	bî	sowe\\
     hole	dig-\textsc{pc}	carry.down.hither.\textsc{pl}.\textsc{a}	hole-\textsc{loc}	hole	cover\\
\glt `(The women) would dig a hole and bring the fetus into the hole. (Then) they cover the hole.'
\z