\ea\label{ex:nrexsusuki}
Ko uki Sepik uki piyo.\\
\gll ko	uki	Sepik	uki	piyo\\
    I	husband	Sepik	husband	take.\textsc{pp}\\
\glt `I took a husband from the Sepik.'
\z

\ea\label{ex:nrexyilsep}
Ko yilau kepyo lo.\\
\gll ko	yilau	kep-yo	lo\\
    I	village	3\textsc{sg}.\textsc{poss}-\textsc{loc}	go.\textsc{pp}\\
\glt `I went to his village.'
\z

\ea\label{ex:nrexpuepiy}
Ko yip ikoiyo nake pu epiyo.\\
\gll ko	yip	ikoi-yo	nake	pu	epiyo\\
    I	house	big-\textsc{loc}	live	river	next.to\\
\glt `I live in a big house on the banks of the river.'
\z

\ea\label{ex:nrexyipppo}
Uki kopi yip po.\\
\gll uki	ko-pi	yip	po\\
    Husband	1\textsc{sg}-\textsc{poss}	house	make.\textsc{pp}\\
\glt `My husband made the house.'
\z

\ea\label{ex:nrexkanuka}
Uki kopi kanu po, kanu kiniyo po.\\
\gll uki	kopi	kanu	po	kanu	kiniyo	po\\
    husband	1\textsc{sg}-\textsc{poss}	canoe	make.\textsc{pp}	canoe	many	make.\textsc{pp}\\
\glt `My husband made a canoe, he made many canoes.'
\z

\ea\label{ex:nrexkanpul}
Uki kopi pul po, kanu pul po.\\
\gll uki	kopi	pul	po	kanu	pul	po\\
    husband	1\textsc{sg}-\textsc{poss}	paddle(\textsc{tp})	make.\textsc{pp}	canoe	paddle(\textsc{tp})	make.\textsc{pp}\\
\glt `My husband made paddles, he made canoe paddles.'
\z

\ea\label{ex:nrexuraiga}
Uki kopi garamut po. Urai po garamutyo.\\
\gll uki	ko-pi	garamut	po	urai	po	garamut-yo\\
    husband	1\textsc{sg}-\textsc{poss}	slit.drum(\textsc{tp})	make.\textsc{pp}	crocodile	make.\textsc{pp}	slit.drum(\textsc{tp})-\textsc{loc}\\
\glt `My husband made a slit drum. He carved crocodiles in the slit drum.'
\z

\ea\label{ex:nrexgaramu}
Ko garamut kowe, yena kiniyo pulupi.\\
\gll ko	garamut	kowe	yena	kiniyo	pulupi\\
    I	slit.drum(\textsc{tp})	beat	people	all	come.\textsc{pl}\\
\glt `When I beat the slid drum, all the people come.'
\z