\ea\label{ex:nrexwaplai}
    Yala ko laip stori ikap mosaupi. Storipi moniya ko nakapno.\\
    \gll yala	ko	laip	stori	ikap	mosaupi	storim-pi	moni=ya	ko	nake-p-no\\
    now	I	life(\textsc{tp})	story(\textsc{tp})	1\textsc{sg}.\textsc{poss}.\textsc{emph}	tell	tell(\textsc{tp})-\textsc{LV}	small=\textsc{emph}(\textsc{tp})	I	sit-\textsc{pc}-\textsc{co}\\
    \glt `Now I will tell my life story. Telling (about the time) when I was little.'
\z

\newpage
\ea\label{ex:nrexnemnem}
    Nem bilong mi, nem kopi Andrew Wapi. Yala ko laip ikap storipi.\\
    \gll nem	bilong	mi	nem	ko-pi	Andrew	Wapi	yala	ko	laip	ikap	storim-pi\\
    name	\textsc{poss}(\textsc{tp})	I(\textsc{tp})	name	1\textsc{sg}-\textsc{poss}	Andrew	Wapi	now	I	life(\textsc{tp})	1\textsc{sg}.\textsc{poss}.\textsc{emph}	tell(\textsc{tp})-\textsc{lv}\\
    \glt `My name, my name is Andrew Wapi. Now I'll tell my life story.'
\z

\ea\label{ex:nrextaimko}
    Taim ko moniya nakapno epe ko karimporo. Ko moniseso.\\
    \gll taim	ko	moni=ya	nake-p-no	epe	ko	karim-po=ro	ko	moniseso\\
    when(\textsc{tp})	I	small=\textsc{emph}(\textsc{tp})	sit-\textsc{pc}-\textsc{co}	mother	I	carry(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	I	very.small\\
    \glt `When I was small, (my) mother carried me. I was very small.'
\z

\ea\label{ex:nrexsuiwer}
    Epe aino kopiro suiwero. Ko moniseso uliyepaneko. Ko boyo noweko.\\
    \gll epe	ai-no	ko-pi=ro	sui-we=ro	ko	moniseso	uliyepane-ko	ko	boyo	nowe-ko\\
    mother	father-\textsc{ins}	1\textsc{sg}-\textsc{poss}=\textsc{emph}	die-\textsc{du}.\textsc{s}=\textsc{emph}	I	very.small	leave.behind-\textsc{rts}	I	later	grow-\textsc{rts}\\
    \glt `My parents died. I was very small, and they had left me behind. Afterwards I grew up.'
\z

\ea\label{ex:nrexbilauk}
    Ko boyo nowo. Boyo ko noworo. Bilau ko lukautimpo.\\
    \gll ko	boyo	nowo	boyo	ko	nowo=ro	Bilau	ko	lukautim-po\\
    I	later	grow.\textsc{pp}	later	I	grow.\textsc{pp}=\textsc{emph}	Bilau	I	take.care.of(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
    \glt `Then I grew up. Later I grew up, and (my brother) Bilau took care of me.'
\z

\ea\label{ex:nrexkomoni}
    Ko moniya. Ko lukautimporo ko ikoi. Ko moniya ko nowapno ikoi.\\
    \gll ko	moni=ya	ko	lukautim-po=ro	ko	ikoi	ko	moni=ya	ko	nowe-p-no	ikoi\\
    I	small=\textsc{emph}(\textsc{tp})	I	take.care.of(\textsc{tp})-\textsc{lv}=\textsc{emph}	I	big	I	small=\textsc{emph}(\textsc{tp})	I	grow-\textsc{pc}-\textsc{co}	big	\\
    \glt `I was small, and he took care of me. I am big. I was small, and I was growing big.'
\z

\ea\label{ex:nrexmibuai}
    Ko mi buai ilewiyo lo, ko boyo noweko.\\
    \gll ko	mi	buai	ilewi-yo	lo	ko	boyo	nowe-ko\\
    I	then	white.people	place-\textsc{loc}	go.\textsc{pp}	I	later	grow-\textsc{rts}\\
    \glt `Then I went to the white people's place. I had grown up.'
\z

\ea\label{ex:nrexkominu}
    Ko mi nuni ko ikap lukautim. Boyo ya. Ko boyo ya.\\
    \gll ko	mi	nuni	ko\_ikap	lukautim	boyo	ya	ko	boyo	ya\\
    I	then	maternal.uncle	1\textsc{sg}.\textsc{poss}.\textsc{emph}	take.care.of(\textsc{tp})	later	take.care.of.\textsc{pp}	I	later	take.care.of.\textsc{pp}\\
    \glt `Then my uncle, my mother's brother, looks after me. Later he took care. He took care of me.'\footnote{Presumably, Andrew's time with his uncle preceded his contract work and stay at the plantation.}
\z

\ea\label{ex:nrexilewiy}
    Ki mi buai ilewiyo lo. Uke boyo molo.\\
    \gll ki	mi	buai	ilewi-yo	lo	uke	boyo	molo\\
    \textsc{aph}	then	white.people	place-\textsc{loc}	go.\textsc{pp}	we.\textsc{excl}	later	go.\textsc{pl}.\textsc{pp}\\
    \glt `Then (my brother) went to the white people's place. Later we went (together).'
\z

\ea\label{ex:nrexikimip}
    Iki mi pinisporo uke yar dupua. Uke pulupiro yar dupua. Boyo pinispo.\\
    \gll iki	mi	pinis-po=ro	uke	yar	dupua	uke	pulup=ro	yar	dupua	boyo	pinis-po\\
    \textsc{aph}.\textsc{pl}	then	finish(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	we.\textsc{excl}	we.\textsc{excl}	two	we.\textsc{excl}	come.\textsc{pl}.\textsc{pp}=\textsc{emph}	year	two	later	finish(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
    \glt `Then they finished (our contract). We (spent there) two years. We came for two years, then it was finished.'
\z

\ea\label{ex:nrexkrisma}
    Krismas kupuapno, uke boyo kupulupiwap yilauyo.\\
    \gll krismas	k-pue-p-no	uke	boyo	k-pulupi-we-p	yilau-yo\\
    krismas	\textsc{sub}-walk-\textsc{pc}-\textsc{co}	we.\textsc{excl}	later	\textsc{sub}-come.\textsc{pl}-\textsc{ter}-\textsc{pc}	village-\textsc{loc}\\
    \glt `After Christmas, then we were finally coming back to the village.'
\z

\ea\label{ex:nrexyoukek}
    Yilauyo uke kupulapno, ko mi ako piyepulo, ako ipei.\\
    \gll yilau-yo	uke	k-pule-p-no	ko	mi	ako	piye\_pulo	ako	ipei\\
    village-\textsc{loc}	we.\textsc{excl}	\textsc{sub}-come-\textsc{pc}-\textsc{co}	I	then	wife	take\_come.\textsc{pp}	wife	first\\
    \glt `Having come (back) to the village I took a wife, my first wife.'
\z

\ea\label{ex:nrexyakomi}
    Ako mi ko kama ari. Mi uke kumune.\\
    \gll ako	mi	ko	kama	ari	mi	uke	kumune\\
    wife	then	I	alone	no	then	we.\textsc{excl}	all.\textsc{coll}\\
    \glt `(With) a wife, then I wasn't alone. We were together.'
\z

\ea\label{ex:nrexyakoro}
    Akoro mi basuik. Ko mi kama nakap. Kama ko nakoro.\\
    \gll ako=ro	mi	ba-sui-ko	ko	mi	kama	nake-p	kama	ko	nako=ro\\
    wife=\textsc{emph}	then	\textsc{fac}-die-\textsc{fac}	I	then	alone	stay-\textsc{pc}	alone	I	stay.\textsc{pp}=\textsc{emph}\\
    \glt `My wife has died. Then I was living alone. I was alone.'\footnote{Possibly, Andrew's first wife died in childbirth together with the child. Only the second wife is said to have born children.}
\z

\ea\label{ex:nrexsispel}
    Ko mi ako ba boyo piyo Anna. Boyo ko kipiyapno ruwaesi koyo karimpoip sispela.\\
    \gll ko	mi	ako	ba	boyo	piyo	Anna	boyo	ko	k-piye-p-no	ruwaesi	koyo	karim-po-i-p	sispela\\
    I	then	wife	other	later	take.\textsc{pp}	Anna	later	I	\textsc{sub}-take-\textsc{pc}-\textsc{co}	children	we.\textsc{du}.\textsc{excl}	carry(\textsc{tp})-\textsc{lv}.\textsc{pp}-\textsc{du}.\textsc{a}-\textsc{pc}	six(\textsc{tp})\\
    \glt `Then I took another wife, Anna. Having married her, we had children, six.'
\z

\ea\label{ex:nrexbuassi}
    Buas Simon, Sipop Vincent, Paio Anton, Ise Barbara, Yamap Samuel, Iman Rodes. Ruwaesi kopiro kuru. Sispela solo kuru.\\
    \gll Buas	Simon	Sipop	Vincent	Paio	Anton	Ise	Barbara	Yamap	Samuel	Iman	Rodes	ruwaesi	ko-pi=ro	kuru	sispela	solo	kuru\\
    Buas	Simon	Sipop	Vincent	Paio	Anton	Ise	Barbara	Yamap	Samuel	Iman	Rodes	children	1\textsc{sg}-\textsc{poss}=\textsc{emph}	be.finished	six	only	be.finished\\
    \glt `Buas Simon, Sipop Vincent, Paio Anton, Ise Barbara, Yamap Samuel, Iman Rodes. (These are) my children. Six only, no more.'
\z

\ea\label{ex:nrexsipipv}
    Em tasol tenkyu.\\
    \gll em	tasol	tenkyu\\
    \textsc{aph}(\textsc{tp})	only(\textsc{tp})	thank.you(\textsc{tp})\\
    \glt `That's it, thank you.'
\z
