\ea\label{ex:nrexepnako}
    Epe kopi nako 1942. Ko moniseso, wepulyo nakap. Ko ireri, nakero, puana, ko dorno lo.\\
    \gll epe	ko-pi	nako	1942	ko	moniseso	wepul-yo	nake-p	ko	ireri	nake=ro	puana	ko	dor-no	lo\\
    mother	1\textsc{sg}-\textsc{poss}	gave.birth	1942	I	very.small	baby.sling-\textsc{loc}	sit-\textsc{pc}	I	crawl	sit=\textsc{emph}	stand.up.\textsc{pp}	I	foot-\textsc{ins}	go.\textsc{pp}\\
    \glt `My mother gave birth (to me) in 1942. I was very small and staying in the baby sling. I crawled for a while, (then) I stood up, and walked on my feet.'
\z

\newpage
\ea\label{ex:nrexwoapol}
    Woa polip. Epe kopi muelpop monem. Riyopuno woa kuru.\\
    \gll woa	poli-p	epe	ko-pi	mueli-ipi-p	monem	riyopuno	woa	kuru\\
    war	be.there-\textsc{pc}	mother	1\textsc{sg}-\textsc{poss}	talk.to-1\textsc{sg}.\textsc{or}-\textsc{pc}	silent	then	war	be.finished\\
    \glt `There was war. My mother was saying to me: ``Be quiet”. Finally the war was over.'\footnote{The second world war is a vivid memory of Margaret, especially the Japanese soldiers who used to roam the forest.}
\z

\ea\label{ex:nrexkonowo}
    Ko nowo ikoiele. Ko Silaso. Ko Silaso kupuapno mi ko Nancyso. Ko Nancyso nakap boyo ko Deiso.\\
    \gll ko	nowo	ikoiele	ko	Sila-so	ko	Sila-so	k-pue-p-no	mi	ko	Nancy-so	ko	Nancy-so	nake-p	boyo	ko	Dei-so\\
    I	grow.\textsc{pp}	very.big	I	Sila-\textsc{sim}	I	Sila-\textsc{sim}	\textsc{sub}-roam-\textsc{pc}-\textsc{co}	then	I	Nancy-\textsc{sim}	I	Nancy-\textsc{sim}	sit-\textsc{pc}	later	I	Dei-\textsc{sim}\\
    \glt `I grew up. I was big like Sila. After I walked around like Sila, I became like Nancy. I was staying like Nancy. Later I was like Dei.'\footnote{Margaret compares her growing body size with children and young girls of the neighbourhood in the year 2004.}
\z

\ea\label{ex:nrexdeisok}
    Deiso ko ikoiele. Ko yili melipop.\\
    \gll Dei-so	ko	ikoiele	ko	yili	meli-pi-p\\
    Dei-\textsc{sim}	I	very.big	I	weight	carry.\textsc{pl}.\textsc{o}-\textsc{lv}-\textsc{pc}\\
    \glt `(When) I was as big as Dei, I carried heavy (burdens).'
\z

\ea\label{ex:nrexppilin}
    Ko ppili nakap epe aino ikap roise Omoiyo.\\
    \gll ko	ppili	nake-p	epe	ai-no	ikap	roise	Omoi-yo\\
    I	be.single	sit-\textsc{pc}	mother	father-\textsc{ins}	1\textsc{sg}.\textsc{poss}.\textsc{emph}	with	Omoi-\textsc{loc}\\
    \glt `I was living unmarried with my parents in Omoi.'
\z

\ea\label{ex:nrexmekiye}
    Ko due mekiyap. Diri kopi weri kopi ko epe ikap mekiyap, ko ai ikap mekiyap. Ai ikap bi luipop. Bi sali ropno wepulap yilauyo epe aino ikap roise.\\
    \gll ko	due	mekiye-p	diri	ko-pi	weri	ko-pi	ko	epe	ikap	mekiye-p	ko	ai	ikap	mekiye-p	ai	ikap	bi	lui-pi-p	bi	sali	rop-no	wepule-p	yilau-yo	epe	ai-no	ikap	roise\\
    I	sago	help-\textsc{pc}	younger.brother	1\textsc{sg}-\textsc{poss}	younger.sister	1\textsc{sg}-\textsc{poss}	I	mother	1\textsc{sg}.\textsc{poss}.\textsc{emph}	help-\textsc{pc}	I	father	1\textsc{sg}.\textsc{poss}.\textsc{emph}	help-\textsc{pc}	father	1\textsc{sg}.\textsc{poss}.\textsc{emph}	animal	shoot-\textsc{lv}-\textsc{pc}	meat	dried	basket-\textsc{ins}	bring-\textsc{pc}	village-\textsc{loc}	mother	father-\textsc{ins}	1\textsc{sg}.\textsc{poss}.\textsc{emph}	with\\
    \glt `I would help with (processing) sago. My younger brothers, my younger sisters, and I helped my mother and my father. My father (often) shot animals. (Then) I carried the dried meat in baskets to the village, together with my parents.'
\z

\ea\label{ex:nrexboyonu}
    Boyo nuni ko ikap roise inakap, nuni Esau. Ko uki am ar piyap, ppili nakap.\\
    \gll boyo	nuni	ko\_ikap	roise	i-nake-p	nuni	Esau	ko	uki	am	ar	piye-p	ppili	nake-p\\
    later	maternal.uncle	1\textsc{sg}.\textsc{poss}.\textsc{emph}	with	\textsc{du}.\textsc{s}-live-\textsc{pc}	maternal.uncle	Esau	I	husband	yet	\textsc{neg}	take-\textsc{pc}	be.single	live-\textsc{pc}\\
    \glt `Later I was living at my uncle's (place), at Uncle Esau's. I did not have a husband yet. I was unmarried.'
\z

\ea\label{ex:nrexmemiko}
    Ppili nakap memi ko ikap roise. Mi ko memiyo lo, memi kopi Lipi. Ko riyo nakap.\\
    \gll ppili	nake-p	memi	ko\_ikap	roise	mi	ko	memi-yo	lo	memi	ko-pi	Lipi	ko	ri-yo	nake-p\\
    be.single	live-\textsc{pc}	grandparent	1\textsc{sg}.\textsc{poss}.\textsc{emph}	with	so	I	grandparent-\textsc{loc}	go.\textsc{pp}	grandparent	1\textsc{sg}-\textsc{poss}	Lipi	I	\textsc{dist}-\textsc{loc}	live-\textsc{pc}\\
    \glt `I was living single together with my grandmother. I went to my grandmother, to my grandmother Lipi. I was living there.'
\z

\ea\label{ex:nrexsauliw}
    Saul Iwan Bopule polip. Bo klokni solo. Mi ari kuru. Yena layepana.\\
    \gll Saul	Iwan	Bopule	poli-p	bo	klokni	solo	mi	ari	kuru	yena	layepana\\
    Saul	Iwan	Bopule	be.there-\textsc{pc}	word	one	only	again	no	be.finished	people	leave.behind.\textsc{pp}\\
    \glt `(The men's house) Saul Iwan Bopule was (still) there. It is just one (long) name. It isn't there (anymore). The people abandoned it.'\footnote{The relinquishment and destruction of the old men's house was a major point of change in people's experience and lifetime. There was a \textit{before} and an \textit{after} regarding this event. Thus the fate of the men's house could play the role of indicating time in one's life story. See also \textref{sc0204}.}
\z

\newpage
\ea\label{ex:nrexkoukip}
    1965 ko uki piyo. Ko Luppapyo nakap. 1966 ruri kopi sui roipi moniseso.\\
    \gll 1965	ko	uki	piyo	ko	Luppap-yo	nake-p	1966	ruri	ko-pi	sui	roipi	moniseso\\
    1965	I	husband	take.\textsc{pp}	I	Luppap-\textsc{loc}	live-\textsc{pc}	1966  child	1\textsc{sg}-\textsc{poss}	die	boy	very.small\\
    \glt `(In) 1965 I married and was then living in Luppap. (In) 1966 my child died, a baby boy.'
\z

\ea\label{ex:nrexburias}
    Ai Jeffreypi bûri aska. Ki kama solo. Ko luono lakiyo. Ai kopi ponamo nuni kopi ponamo luo namayo.\\
    \gll ai	Jeffrey-pi	bûri	aska	ki	kama	solo	ko	luo-no	lakiyo	ai	ko-pi	ponamo	nuni	ko-pi	ponamo	luo	namayo\\
    father	Jeffrey-\textsc{poss}	sister	none	\textsc{aph}	alone	only	I	money-\textsc{ins}	fetch.\textsc{pp}	father	1\textsc{sg}-\textsc{poss}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	maternal.uncle	1\textsc{sg}-\textsc{poss}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	money	give.\textsc{recp}.\textsc{pp}\\
    \glt `Jeffrey's father had no sister, he was alone. So he took me for money. He gave my father (some money) and gave my uncle (some money). They shared the money.'\footnote{Jeffrey Osi is the second son of Margaret's husband Lis Osi and his successor in the village. His older brother Paul had left the village and gone to Port Moresby.}
\z

\ea\label{ex:nrexlipkau}
    Memi kopi kau ponamo Lipi. Lipi pusayepo de luo ko ar powa.\\
    \gll memi	ko-pi	kau	ponamo	Lipi	Lipi	pusaye-po	de	luo	ko	ar	powa\\
    grandparent	1\textsc{sg}-\textsc{poss}	cow	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	Lipi	Lipi	reject-\textsc{lv}.\textsc{pp}	you	money	I	\textsc{neg}	give.1\textsc{sg}.\textsc{or}.\textsc{pp}\\
    \glt `He gave my grandmother Lipi a cow. Lipi first rejected it: ``You didn't give (any) money to me.'''
\z

\ea\label{ex:nrexkausil}
    Umul kep kauno sileno. Umul kep sneiwap. Kuru kuru. ``Kau de ko baponamko.''\\
    \gll umul	kep	kau-no	sile-no	umul	kep	snei-we-p	kuru	kuru	kau	de	ko	ba-poname-ko\\
    heart	3\textsc{sg}.\textsc{poss}	cow-\textsc{ins}	satisfy-3\textsc{sg}.\textsc{or}.\textsc{pp}	heart	3\textsc{sg}.\textsc{poss}	be.quiet-\textsc{ter}-\textsc{pc}	be.finished	be.finished	cow	you	I	\textsc{fac}-give.2\textsc{sg}.\textsc{or}-\textsc{fac}\\
    \glt `(Then) her heart was satisfied with the cow. Her mind was really quiet. (The argument) was finished. That's it. ``I have given a cow to you”, (said Jeffrey).'
\z

\ea\label{ex:nrexkuippu}
    Ruri kopi ipei klokni sui. Ruri ipei makiro. Yala ko makina wonake. kui kopi makiro ko makina wonake. Kui ppulae ko makina ar wonake. Kui kopi ko rapue ar powa, olo pi.\\
    \gll ruri	ko-pi	ipei	klokni	sui	ruri	ipei	maki=ro	yala	ko	maki-na	wo-nake	kui	ko-pi	maki=ro	ko	maki-na	wo-nake	kui	ppulae	ko	maki-na	ar	wo-nake	kui	ko-pi	ko	rapue	ar	powa	olo\_pi\\
    child	1\textsc{sg}-\textsc{poss}	firstborn	one	die	child	firstborn	good=\textsc{emph}	now	I	good-\textsc{adv}	\textsc{accom}-live	daughter.in.law	1\textsc{sg}-\textsc{poss}	good=\textsc{emph}	I	good-\textsc{adv}	\textsc{accom}-live	daughter.in.law	bad	I	good-\textsc{adv}	\textsc{neg}	\textsc{accom}-live	daughter.in.law	1\textsc{sg}-\textsc{poss}	I	food	\textsc{neg}	give.1\textsc{sg}.\textsc{or}.\textsc{pp}	be.greedy\\
    \glt `My firstborn child, one (child) died. (If) the firstborn child is good, then I will live fine in his company. (If) my daughter-in-law is good, I will live fine with her. (If) the daughter-in-law is bad, I won't live well with her. My daughter-in-law didn't (want to) give me food. She is greedy.'
\z

\ea\label{ex:nrexkuiime}
    Kui kopi Imelda bo kep as. Pon riso ar pi. Kui roke umul maki.\\
    \gll kui	ko-pi	Imelda	bo	kep	as	pon	ri-so	ar	pi	kui	ro-ke	umul	maki\\
    daughter.in.law	1\textsc{sg}-\textsc{poss}	Imelda	speech	3\textsc{sg}.\textsc{poss}	none	nose	wood-\textsc{sim}	\textsc{neg}	\textsc{lv}	daughter.in.law	\textsc{prox}.\textsc{emph}-\textsc{aph}	heart	good\\
    \glt `My daughter-in-law Imelda does not make words. Her face is not like wood. This daughter-in-law has a good heart.'
\z

\ea\label{ex:nrexruripp}
    Ko ruri ppusi nako David Yau Luppapyo nako. Ruri ba kini Charles Imop, ruri ba kini Rafael Ppisi. Kiniyo Luppapyo.\\
    \gll ko	ruri	ppusi	nako	David	Yau	Luppap-yo	nako	ruri	ba	kini	Charles	Imop	ruri	ba	kini	Rafael	Ppisi	kiniyo	Luppap-yo\\
    I	child	nextborn	gave.birth	David	Yau	Luppap-\textsc{loc}	gave.birth	child	other	one.\textsc{part}	Charles	Imop	child	other	one.\textsc{part}	Rafael	Ppisi	all	Luppap-\textsc{loc}\\
    \glt `I gave birth to the second child, David Yau. I gave birth (to him) in Luppap. (Then) to one of the other children, Charles Imop, still to another child, Rafael Ppisi. All (were born) in Luppap.'\footnote{Nowadays all children receive two names: a Western/Christian name and a traditional name relating the child to clan ancestors, usually the generation of the grandparents.}
\z

\ea\label{ex:nrexluppap}
    Yip ikoi Luppapyo polip. Yip moni duyo polip, Ouwinyo polip. Dete ol nem ponamo. Dete ol doriye pu lakiwepu, kiniyo lakiwepu. Pu Bilouppueki sowelaye. Pu lûpi maki.\\
    \gll yip	ikoi	Luppap-yo	poli-p	yip	moni	du-yo	poli-p	Ouwin-yo	poli-p	dete	ol	nem	ponamo	dete	ol	doriye	pu	laki-wepu	kiniyo	laki-wepu	pu	Bilouppueki	sowe\_laye	pu	lûpi	maki\\
    house	big	Luppap-\textsc{loc}	be.there-\textsc{pc}	house	small	bush-\textsc{loc}	be.there-\textsc{pc}	Ouwin-\textsc{loc}	be.there-\textsc{pc}	ancestor	mountain	name	give.3\textsc{sg}.\textsc{or}	ancestor	mountain	swamp	water	count-\textsc{quant}.\textsc{o}.\textsc{pp}	all	count-\textsc{quant}.\textsc{o}.\textsc{pp}	water	Bilouppueki	cover.well	water	clear	good\\
    \glt `The big house was in Luppap. A small house was in the bush. It stood (on the knoll) Ouwin. The ancestors gave the name to that knoll. They named hills, swamps, water bodies, the ancestors named everything. (Later) the pond Bilouppueki was filled up, (before) it had been good clear water.'\footnote{Luppap is the old Kilmeri name for Ossima Asples at the Puwani River.}
\z

\ea\label{ex:nrexbiyaip}
    Nuko yilauyo molap, yilauyo pulup. Ko bi yaip biopo biep biwi. Ko yaiwepu. Suo yasiyo, ral yasiyo, ul luan yasiyo, sawa yasiyo, puel yasiyo, due doriyeyo yasiyo.\\
    \gll nuko	yilau-yo	mole-p	yilau-yo	pulip	ko	bi	yai-p	biopo	biep	biwi	ko	yai-wepu	suo	yasiyo	ral	yasiyo	ul	luan	yasiyo	sawa	yasiyo	puel	yasiyo	due	doriye-yo	yasiyo\\
    we.\textsc{incl}	place-\textsc{loc}	go.\textsc{pl}-\textsc{pc}	place-\textsc{loc}	come.\textsc{pl}.\textsc{pp}	I	animal	take.care.of-\textsc{pc}	domesticated.pig	boar	sow	I	take.care.of-\textsc{quant}.\textsc{o}.\textsc{pp}	coconut	plant.\textsc{pp}	kind.of.tree	plant.\textsc{pp}	bamboo	breadfruit.tree	plant.\textsc{pp}	mango.tree	plant.\textsc{pp}	betel.nut	plant.\textsc{pp}	sago	swamp-\textsc{loc}	plant.\textsc{pp}	\\
    \glt `We (often) went to the place (in the bush). We came to the place, and I took care of the animals. I took care of the domesticated pigs, the boar, and the sows. I planted coconut palms, \textit{ral}-trees, bamboo, breadfruit trees, mango trees, and I planted betel nut palms. I (also) planted sago palms in the swamp.'
\z

\ea\label{ex:nrexpaulwa}
    Paul Waia bi kopi lu. Bipupi lu. Bipupi ruri nainpela nako. Ko smep paliyen, kiniyo yalaka molo ewo lipelip, ewo ilap.\\
    \gll Paul	Waia	bi	ko-pi	lu	Bipupi	lu	Bipupi	ruri	nainpela	nako	ko	smep	paliye-en	kiniyo	yalaka	molo	ewo	lipeli-p	ewo	ile-p\\
    Paul	Waia	pig	1\textsc{sg}-\textsc{poss}	shoot.\textsc{pp}	Bipupi	shoot.\textsc{pp}	Bipupi	child	nine(\textsc{tp})	gave.birth	I	door	open-\textsc{nsg}.\textsc{or}.\textsc{pp}	all	now	go.\textsc{pl}.\textsc{pp}	kind.of.worm	seek-\textsc{pc}	kind.of.worm	eat.\textsc{pl}.\textsc{a}-\textsc{pc}\\
    \glt `Paul Waia shot my pig, he shot Bipupi. Bipupi had nine piglets. I opened them the door (unintentionally), and so they all went away. They were looking for worms. They were feeding on worms.'
\z

\ea\label{ex:nrexkobiik}
    Ko bi ikap lipelip. Lipeliou ari. Ana luro? Ono ko ar reyo. Ko bo solo malo, bo pulo.\\
    \gll ko	bi	ikap	lipeli-p	lipeli-ou	ari	ana	lu=ro	ono	ko	ar	reyo	ko	bo	solo	malo	bo	pulo\\
    I	pig	1\textsc{sg}.\textsc{poss}.\textsc{emph}	seek-\textsc{pc}	seek-\textsc{frus}	no	who	shoot.\textsc{pp}=\textsc{emph}	person	I	\textsc{neg}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	I	speech	only	hear.\textsc{pp}	speech	come.\textsc{pp}\\
    \glt `I was searching for my pig, (but) I searched in vain. Nothing. Who shot it? I didn't see anybody. I only heard a rumour, the rumour came (to me).'
\z

\ea\label{ex:nrexmosupi}
    Uki kopi ruri duyo woko. Du mosupinap. Depi oki, depi oki. Du dob riyewap, mosupien. Du kiniyo mosupiwepien.\\
    \gll uki	ko-pi	ruri	du-yo	woko	du	mosupi-ne-p	de-pi	o-ki	de-pi	o-ki	du	dob	riye-we-p	mosupi-en	du	kiniyo	mosupi-wepi-en\\
    husband	1\textsc{sg}-\textsc{poss}	child	bush-\textsc{loc}	go.together.with.\textsc{pp}	bush	show-3\textsc{sg}.\textsc{or}-\textsc{pc}	2\textsc{sg}-\textsc{poss}	\textsc{prox}-\textsc{aph}	2\textsc{sg}-\textsc{poss}	\textsc{prox}-\textsc{aph}	bush	eye	see.\textsc{o}.\textsc{inanim}-\textsc{ter}-\textsc{imp}	show-\textsc{nsg}.\textsc{or}.\textsc{pp}	bush	all	show-\textsc{quant}.\textsc{o}-\textsc{nsg}.\textsc{or}.\textsc{pp}\\
    \glt `My husband went into the bush together with the sons. He was showing them the bush: here is yours, here is yours. Look carefully! He showed it to them. He showed to them all their bush lands.'\footnote{Grammatical comment: The first occurrence of the verb \textit{mosupi} with Recipient agreement shows singular agreement, while the second and third ones show plural agreement. The translation takes always the plural according to the intended meaning and reported situation.}
\z

\ea\label{ex:nrexkompan}
    Kompani kiniyo pulup oil lipelip. Yena kiniyo mapap. Riyopuno Jeffrey David Jerome Simon Tapi Joe Samou Sepik yilauyo kumune molo kompani roise. Wok pop Angoramyo krismas dupua. Riyopuno midoripulupip yilau kepyo.\\
    \gll kompani	kiniyo	pulup	oil	lipeli-p	yena	kiniyo	mape-p	riyopuno	Jeffrey	David	Jerome	Simon	Tapi	Joe	Samou	Sepik	yilau-yo	kumune	molo	kompani	roise	wok	pi-p	Angoram-yo	krismas	dupua	riyopuno	mi-doripulupi-p	yilau	kep-yo\\
    company	many	come.\textsc{pl}.\textsc{pp}	oil	seek-\textsc{pc}	people	many	sit-\textsc{pc}	then	Jeffrey	David	Jerome	Simon	Tapi	Joe	Samou	Sepik	location-\textsc{loc}	all.\textsc{coll}	go.\textsc{pl}.\textsc{pp}	company	together	work	do-\textsc{pc}	Angoram-\textsc{loc}	year	two	then	\textsc{iter}-come.back.\textsc{pl}-\textsc{pc}	village	3\textsc{sg}.\textsc{poss}-\textsc{loc}\\
    \glt `There came exploration companies. They searched for oil. Many people were there. Eventually Jeffrey, David, Jerome, Simon Tapi, and Joe Samou went to the Sepik region together with the company. Two years they worked near Angoram, then (everyone) returned to his village.'
\z

\ea\label{ex:nrexelpiam}
    Riyopuno ko el piamu. Ko karimpop Grace rumkari klokni solo. Ko wîs nainpela kanakapno ko ruri nako. Dob ko riye rumkari.\\
    \gll riyopuno	ko	el\_piamu	ko	karim-pi-p	Grace	rumkari	klokni	solo	ko	wîs	nainpela	k-nake-p-no	ko	ruri	nako	dob	ko	riye	rumkari\\
    then	I	get.pregnant.\textsc{pp}	I	be.pregnant(\textsc{tp})-\textsc{lv}-\textsc{pc}	Grace	girl	one	only	I	moon	nine(\textsc{tp})	\textsc{sub}-live-\textsc{pc}-\textsc{co}	I	child	gave.birth	eye	I	see.\textsc{o}.\textsc{inanim}	girl\\
    \glt `Then I got pregnant. I was carrying Grace, (my) only girl. After living for nine months, I gave birth to the child and I see: (She is) a girl.'
\z

\ea\label{ex:nrexbrataj}
    Yena ruri kopi woko. Brata Jim dob reyo. Pilot dob reyo. Dop kep ukeso! Yena mi koyo wepulupip.\\
    \gll yena	ruri	ko-pi	woko	brata	Jim	dob	reyo	pilot	dob	reyo	dop	kep	uke-so	yena	mi	ko-yo	wepulupi-p\\
    people	child	1\textsc{sg}-\textsc{poss}	go.together.with.\textsc{pp}	brother	Jim	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	pilot	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	skin	3\textsc{sg}.\textsc{poss}	we.\textsc{excl}-\textsc{sim}	people	again	1\textsc{sg}-\textsc{loc}	bring.\textsc{pl}.\textsc{a}-\textsc{pc}\\
    \glt `Some people went together with the child (in order to show her to other people). Brother Jim looked at her, and the pilot looked at her: ``Her skin is like ours.” The people brought her back to me.'
\z

\newpage
\ea\label{ex:nrex1}
    Josepin Bewa, Krisa yako, Bewa ruri nem ponamo. Grace neuli.\\
    \gll Josepin	Bewa	Krisa	yako	Bewa	ruri	nem	ponamo	Grace	ne-uli\\
    Josephine	Bewa	Krisa	woman	Bewa	child	name	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	Grace	go.thither-\textsc{prog}\\
    \glt `Josephine Bewa, a woman from Krisa, the daughter of Bewa, gave (the baby the) name. She named (the child) Grace.'\footnote{The literal meaning is `she goes forth (as) Grace'.}
\z

\ea\label{ex:nrexskulyo}
    Baikoipoko skulyo lo priskul. Helen skulim em tisa Amerikapi.\\
    \gll ba-ikoi-pi-ko	skul-yo	lo	priskul	Helen	skulim	em	tisa	Amerika-pi\\
    \textsc{fac}-big-\textsc{lv}-\textsc{fac}	school-\textsc{loc}	go.\textsc{pp}	primary.school	Helen	teach.\textsc{tr}	\textsc{aph}(\textsc{tp})	teacher	Amerika-\textsc{poss}\\
    \glt `She has become big and went to school, to primary school. Helen instructed her, a teacher from America.'
\z

\ea\label{ex:nrexlissui}
    Uki kopi basuik. Yar 1990no sui, Sarere September 16. Ko Imeldayo Charlesyo uke yip riyo mapap. Uki kopi puapno yena bo ppulaena womui womoliye. Uki kopi ar mariulipop. Upunaro makina nakap.\\
    \gll uki	ko-pi	ba-sui-ko	yar	1990-no	sui	Sarere	September	16	ko	Imelda-yo	Charles-yo	uke	yip	ri-yo	mape-p	uki	ko-pi	pue-p-no	yena	bo	ppulae-na	wo-mui	wo-moliye	uki	ko-pi	ar	mari-uli-pi-p	upuna=ro	maki-na	nake-p\\
    husband	1\textsc{sg}-\textsc{poss}	\textsc{fac}-die-\textsc{fac}	year	1990-\textsc{ins}	die	Saturday	September	16	I	Imelda-\textsc{loc}	Charles-\textsc{loc}	we.\textsc{excl}	house	\textsc{dist}-\textsc{loc}	sit.\textsc{pl}-\textsc{pc}	husband	1\textsc{sg}-\textsc{poss}	roam-\textsc{pc}-\textsc{co}	people	speech	bad-\textsc{adv}	\textsc{accom}-speak	\textsc{accom}-speak.\textsc{pl}	husband	1\textsc{sg}-\textsc{poss}	\textsc{neg}	be.sick-\textsc{prog}-\textsc{lv}-\textsc{pc}	alright=\textsc{emph}	good-\textsc{adv}	live-\textsc{pc}\\
    \glt `My husband has died. He died in the year 1990, on Saturday 16th of September. Me and Imelda and Charles, we were in that house (with him). While my husband was roaming (the bush), the people speak bad words on him. My husband was never sick. (Everything was) fine, he was in a good condition.'\footnote{The husband's health condition is spoken about in order to say that his death is caused by sorcery and is not a natural death.}
\z

\ea\label{ex:nrexyaeaul}
    Yena lotu yaeau nuro duruwa. Yaeau puni nuknoko. Brata Jim wo mop umul polenap. ``Ko yala ar nake. Ko yala lam. De nakapno ko nakap. Yalaka ko le.''\\
    \gll yena	lotu	yaeau	nu=ro	duruwa	yaeau	puni	nuknoko	brata	Jim	wo\_mop	umul\_pole-ne-p	ko	yala	ar	nake	ko	yala	le-m	de	nake-p-no	ko	nake-p	yalaka	ko	le\\
    people	church	traditional.celebration	do.intentionally.\textsc{pp}=\textsc{emph}	early.morning	traditional.celebration	night	having.spent.several.nights	brother	Jim	cry.\textsc{pp}	be.sad-3\textsc{sg}.\textsc{or}-\textsc{pc}	I	now	\textsc{neg}	stay	I	now	go-\textsc{pos}	you	live-\textsc{pc}-\textsc{co}	I	live-\textsc{pc}	now	I	go\\
    \glt `The people organised a traditional funeral service and songs until the early morning. They held a wake during night. Brother Jim was crying and sad over him: ``I won't stay (any longer), I will go. While you were alive, I was living (here). Now I'll leave.'''
\z

\ea\label{ex:nrexjeffre}
    Jeffrey bo ar mui. Ruri kep ari. Bo ar mui, ar mekiyo. Brata Jim Vanimoyo lo. Bisnis kep ppulaepowolo. Kimike upuna polip, kanakapno. Yalaka ari.\\
    \gll Jeffrey	bo	ar	mui	ruri	kep	ari	bo	ar	mui	ar	mekiyo	brata	Jim	Vanimo-yo	lo	bisnis	kep	ppulae-pi-wolo	kimike	upuna	poli-p	k-nake-p-no	yalaka	ari\\
    Jeffrey	speech	\textsc{neg}	speak	child	3\textsc{sg}.\textsc{poss}	no	speech	\textsc{neg}	speak	\textsc{neg}	help.\textsc{pp}	brother	Jim	Vanimo-\textsc{loc}	go.\textsc{pp}	business	3\textsc{sg}.\textsc{poss}	bad-\textsc{lv}-\textsc{cpl}.\textsc{pp}	before	alright	be.there-\textsc{pc}	\textsc{sub}-stay-\textsc{pc}-\textsc{co}	now	no\\
    \glt `Jeffrey is not articulate. Neither are his children. They aren't eloquent and didn't help (with the farm business). Brother Jim went to Vanimo. His farm business stagnated. Before everything was fine, when he was still there. (But) now it doesn't work.'
\z

\ea\label{ex:nrexepueka}
    Umul kep maki yilau makina polip. Epue ar polip. Yalaka epue kauna. Epue sowelaye. Yol kaupi epue sowelaye. Epue ikoina po. Ri luap poyana, ri lop poyana, ri bayana roise. Ri rur ri rupopin kauna. Kiniyo poyana.\\
    \gll umul	kep	maki	yilau	maki-na	poli-p	epue	ar	poli-p	yalaka	epue	kauna	epue	sowe\_laye	yol	kau-pi	epue	sowe\_laye	epue	ikoi-na	po	ri\_luap	poyana	ri\_lop	poyana	ri	bayana	roise	ri\_rur	ri\_rupopin	kauna	kiniyo	poyana\\
    heart	3\textsc{sg}.\textsc{poss}	happy	place	good-\textsc{adv}	be.there-\textsc{pc}	weeds	\textsc{neg}	be.there-\textsc{pc}	now	weeds	in.large.amounts	weeds	cover.thoroughly	fence	cow-\textsc{poss}	weeds	cover.thoroughly	weeds	great-\textsc{adv}	\textsc{lv}.\textsc{pp}	kind.of.tree	rise.thither.\textsc{pp}	kind.of.tree	rise.thither.\textsc{pp}	tree	different	together	kind.of.tree	Pandanus	numerous	all	rise.thither.\textsc{pp}\\
    \glt `(Back then Brother Jim) was happy. The place was in a good condition, and there were no weeds. Now (there are) weeds in large amounts. Weeds cover (everything) densely. The cow fence is covered throughout by weeds. The weeds are a plague. \textit{Luap}-trees rose, \textit{lop}-trees rose together with other trees. \textit{Rur}-trees and Pandanus are numerous. All (kinds of unwanted trees) grew big in any direction.'
\z

\ea\label{ex:nrexouwine}
    Ko yilauyo nake. Ko ruripiyo nake. Yip kopi Charles ako nake. Kau kopi Ouwin elno.\\
    \gll ko yilau-yo nake ko ruri-pi-yo nake yip ko-pi Charles ako nake kau ko-pi Ouwin elno\\
    I	village-\textsc{loc}	live	I	child-\textsc{poss}-\textsc{loc}	live	house	1\textsc{sg}-\textsc{poss}	Charles	wife	live	cow	1\textsc{sg}-\textsc{poss}	Ouwin	pregnant\\
    \glt `I live in the village. I live with my son. In my house, Charles's wife lives there. My cow Ouwin is pregnant.'\footnote{Ouwin is the name of the place in the bush, where the family used to have a small house (see sequence \ref{ex:nrexluppap} above). Domesticated animals used to and still bear a name of those bush places.}
\z
