\ea\label{ex:nrexakonon}
    Uki akono namueyo inakap, due soniip. Ruri dupua ba ikoi ba moni, moni wepulyo.\\
    \gll uki	ako-no	namue-yo	i-nake-p	due	soni-i-p	ruri	dupua	ba	ikoi	ba	moni	moni	wepul-yo\\
    husband	wife-\textsc{ins}	sago.swamp-\textsc{loc}	\textsc{du}.\textsc{s}-stay-\textsc{pc}	sago	pound.sago.pith-\textsc{du}.\textsc{a}-\textsc{pc}	child	two	other	big	other	small	small	baby.sling-\textsc{loc}\\
    \glt `Husband and wife were staying in the sago swamp and pounding sago pith. They  had two children, a big one and a small one. The small one was in the baby sling (for carrying).'
\z

\ea\label{ex:nrexduekso}
    Due kosoneipno wik klokni inakap. Bi kuluipno pulo yipyo.\\
    \gll due	k-soni-i-p-no	wik	klokni	i-nake-p	bi	k-lui-p-no	pulo	yip-yo\\
    sago	\textsc{sub}-pound.sago.pith-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	week	one	\textsc{du.s}-stay-\textsc{pc}	pig	\textsc{sub}-shoot-\textsc{pc}-\textsc{co}	come.\textsc{pp}	house-\textsc{loc}\\
    \glt `They pounded sago pith, and they were staying one week. The husband shot a pig and came to the (bush) hut.'
\z

\ea\label{ex:nrexkokepr}
    Ako kep ruri yipyo ukenakap.\\
    \gll ako	kep	ruri	yip-yo	uke-nake-p\\
    wife	\textsc{3sg}.\textsc{poss}	child	house-\textsc{loc}	jointly-stay-\textsc{pc}\\
    \glt `He joined his wife and the children in the house.'
\z

\ea\label{ex:nrexbipuen}
    Uke bi puenpoi, yipyo wepuloi, ipino sepoi. Kisipno napno pului, ureyo royepakoi.\\
    \gll uke	bi	puenpo-i	yip-yo	wepulo-i	ipi-no	sepo-i	k-si-p-no	nap-no	pulu-i	ure-yo	royepako-i\\
    jointly	pig	cut.meat.\textsc{pp}-\textsc{du}.\textsc{a}	house-\textsc{loc}	bring.\textsc{pp}-\textsc{du}.\textsc{a}	pot-\textsc{ins}	cook.\textsc{pl}.\textsc{pp}-\textsc{du}.\textsc{a}	\textsc{sub}-cook-\textsc{pc}-\textsc{co}	bamboo.tongs-\textsc{ins}	take.off.\textsc{pp}-\textsc{du}.\textsc{a}	smoking.container-\textsc{loc}	lay.down.\textsc{pp}-\textsc{du}.\textsc{a}\\
    \glt `They cut the meat together, brought it to the house, and cooked it in pots. After cooking  it they took it off with bamboo tongs and put it into smoking containers.'
\z

\ea\label{ex:nrexpeppuo}
    Sû mappeppuoi. Pupuol poppuo, bi slau po. Em so, yilauyo iloi.\\
    \gll sû	mappeppuo-i	pupuol	pi\_ppuo	bi	slau	po	em	so	yilau-yo	i-lo-i\\
    fire	light.up.\textsc{pp}-\textsc{du}.\textsc{a}	heat	\textsc{lv}\_go.up.\textsc{pp}	pig	dry	\textsc{lv}.\textsc{pp}	tomorrow	like	village-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}\\
    \glt `They lit the fire. The heat went up, and the meat dried. The following day the same. (Then) they went to the village.'
\z

\ea\label{ex:nrexnopuko}
    Uki akono pu koryo inekip. Uki rop akopi piyo, wel ruri ikoi roise.\\
    \gll uki	ako-no	pu	koryo	i-neki-p	uki	rop	ako-pi	piyo	wel	ruri	ikoi	roise\\
    husband	wife-\textsc{ins}	river	beside	\textsc{du}.\textsc{s}-stand-\textsc{pc}	husband	basket	wife-\textsc{poss}	take.\textsc{pp}	carry.\textsc{pp}	child	big	with\\
    \glt `Husband and wife were standing at the river. The husband took the wife's basket. He carried it, together with the big child.'
\z

\ea\label{ex:nrexiwilim}
    Ri wili mono sianai. Amaka rop ruri ikoi wor roise wapoko.\\
    \gll ri\_wili	mono	siana-i	amaka	rop	ruri	ikoi	wor	roise	wapo-ko\\
    log	path	cross.thither.\textsc{pp}-\textsc{du}.\textsc{a}	over.there	basket	child	big	dog	together		put.together.\textsc{pp}-\textsc{rts}\\
    \glt `They crossed (the river) over logs. Over there he placed the basket together with the big child and the dog.'
\z

\ea\label{ex:nrexunomid}
    Riyopuno mi dorisiana, ako kep lakiyap ruri moni roise.\\
    \gll riyopuno	mi	dori\_siana	ako	kep	lakiye-p	ruri	moni	roise\\
    then	again	turn.back\_cross.thither.\textsc{pp}	wife	\textsc{3sg}.\textsc{poss}	fetch-\textsc{pc}	child	small	with\\
    \glt `Then he crossed back again and took his wife together with the small child.'
\z

\ea\label{ex:nrexkesiam}
    Ukesiamu. Ako anno wiyo, ruri moni wepulyo. Ukesiamipno ol asayeya, urai puipulo.\\
    \gll uke-siamu	ako	an-no	wiyo	ruri	moni	wepul-yo	uke-siami-p-no	ol	a-saye=ya	urai	puipulo\\
    jointly-cross.hither.\textsc{pp}	wife	hand-\textsc{ins}	hold.\textsc{pp}	child	small	baby.sling-\textsc{loc}	jointly-cross.hither-\textsc{pc}-\textsc{co}	hill	\textsc{imp}3-touch=\textsc{emph}(\textsc{tp})	crocodile	come.to.emerge.\textsc{pp}\\
    \glt `They crossed the river jointly. He held the hand of his wife, and the small child was in the baby sling. When they had crossed and should have touched the river bank, a crocodile surfaced.'
\z

\ea\label{ex:nrexkikepw}
    Uki kep wiyo. Ako kep ruri moni roise pu wena.\\
    \gll uki	kep	wiyo	ako	kep	ruri	moni	roise	pu	wena\\
    husband	\textsc{3sg}.\textsc{poss}	hold.\textsc{pp}	wife	\textsc{3sg}.\textsc{poss}	child	small	with	river	carry.inside.\textsc{pp}\\
    \glt `(The crocodile) caught her husband, (while) the river carried away his wife and the small child.'
\z

\newpage
\ea\label{ex:nrexkepura}
    Uki kep urai no, koniyepako. El kepyo uraipiyo lilip.\\
    \gll uki	kep	urai	no	koniyepako	el	kep-yo	urai-pi-yo	lili-p\\
    husband	\textsc{3sg}.\textsc{poss}	crocodile	eat.\textsc{pp}	swallow.down.\textsc{pp}	belly	\textsc{3sg}.\textsc{poss}-\textsc{loc}	crocodile-\textsc{poss}-\textsc{loc}	be.there-\textsc{pc}\\
    \glt `The crocodile ate the man, it gobbled him up. He was in the crocodile's belly.'
\z

\ea\label{ex:nrexlepeyi}
    ``De lape yilauyo yena mueliinep! Ko ike ppulaero.''\\
    \gll de	le-p-e	yilau-yo	yena	mueli-ini-p	ko\_ike	ppulae=ro\\
    you	go-\textsc{imp}-\textsc{voc}	village-\textsc{loc}	people	talk.to-\textsc{nsg}.\textsc{or}-\textsc{imp}	I.myself	bad=\textsc{emph}\\
    \glt `(The mother said to the big child:) ``You go straight to the village and tell the people (what happened). I am just badly off.'''
\z

\ea\label{ex:nrexwelbur}
    ``Epe aino, epe kopi pu wel bûri kopi roise. Ai kopi urai wiyo.''\\
    \gll epe	ai-no	epe	ko-pi	pu	wel	bûri	ko-pi	roise	ai	ko-pi	urai	wiyo\\
    mother	father-\textsc{ins}	mother	\textsc{1sg}-\textsc{poss}	river	carry.\textsc{pp}	sister	\textsc{1sg}-\textsc{poss}	with	father	\textsc{1sg}-\textsc{poss}	crocodile	hold.\textsc{pp}\\
    \glt `{``}My parents! The river carried (away) my mother and my sister. My father was caught by a crocodile.'''
\z

\ea\label{ex:nrexahdedu}
    ``Ah de dupuapi!''\\
    \gll ah	de	dupuapi\\
    ah	you	lie\\
    \glt `{``}Ah, you are lying!'''
\z

\ea\label{ex:nrexodupua}
    ``Ko dupuapi ari dukiro! Ine awe, ine mipiyap!''\\
    \gll ko	dupuapi	ari	duki=ro	ine	awe	ine	mipi-yep\\
    I	lie	no	true=\textsc{emph}	you.\textsc{pl}	come.\textsc{imp}	you.\textsc{pl}	come.hither.\textsc{pl}-\textsc{imp}.\textsc{pl}\\
    \glt `{``}I am not lying, it's really true. You come, you come here!'''
\z

\ea\label{ex:nrexokeduk}
    ``Ko so de dupuapapro! Oke dukiro.''\\
    \gll ko	so	de	dupuapi-p=ro	o-ke	duki=ro\\
    I	believe	you	lie-\textsc{pc}=\textsc{emph}	\textsc{prox}-\textsc{aph}	true=\textsc{emph}\\
    \glt `{``}I thought you were lying, (but) this is true!'''
\z

\ea\label{ex:nrexkoardu}
    ``Ko ar dupuapap. Ko bo duki mulip.''\\
    \gll ko	ar	dupuapi-p	ko	bo	duki	muli-p\\
    I	\textsc{neg}	lie-\textsc{pc}	I	word	true	say-\textsc{pc}\\
    \glt `{``}I was not lying, I was saying the truth!'''
\z

\ea\label{ex:nrexriyeri}
    Yena kiniyo dob riye: ri wili wewolo, ri wili rupuewolo.\\
    \gll yena	kiniyo	dob	riye	ri\_wili	we-wolo	ri\_wili	rupue-wolo\\
    people	all	eye	see.\textsc{o}.\textsc{inanim}	log	break-\textsc{cpl}.\textsc{pp}	log	break-\textsc{cpl}.\textsc{pp}\\
    \glt `All the people see it, (the crocodile) broke the log, it broke the log completely.'
\z

\ea\label{ex:nrexwekepw}
    Ewe kep wo roise puap. Pu epi mono puapno urai puyo puipulo. Mek nieno, ko banok diri depi.\\
    \gll ewe	kep	wo\_roise	pue-p	pu	epi\_mono	pue-p-no	urai	pu-yo	puipulo	mek	nie-no	ko	ba-ni-ko	diri	de-pi\\
    older.brother	\textsc{3sg}.\textsc{poss}	crying	stroll-\textsc{pc}	river	side.path	stroll-\textsc{pc}-\textsc{co}	crocodile	river-\textsc{loc}	come.to.emerge.\textsc{pp}	mouth	show-\textsc{3sg}.\textsc{or}.\textsc{pp}	I	\textsc{fac}-eat-\textsc{fac}	younger.brother	\textsc{2sg}-\textsc{poss}\\
    \glt `The (husband's) older brother was walking around with tears. While walking along the river edge the crocodile surfaced and showed him its mouth: ``I have eaten up your younger brother.'''
\z

\ea\label{ex:nrexfignat}
    Yena pulupip. Father Ignatius muelno: ``Urai ono bawiyoko. Nem kep Iwei. Ewe kep Bou, wo roise puap.''\\
    \gll yena	pulupi-p	Father Ignatius	mueli-no	urai	ono	ba-wiye-ko	nem	kep	Iwei	ewe	kep	Bou	wo\_roise	pue-p\\
    people	come.\textsc{pl}-\textsc{pc}	Father Ignatius	talk.to-\textsc{3sg}.\textsc{or}.\textsc{pp}	crocodile	man	\textsc{fac}-catch-\textsc{fac}	name	\textsc{3sg}.\textsc{poss}	Iwei	older.brother	\textsc{3sg}.\textsc{poss}	Bou	crying	stroll-\textsc{pc}\\
    \glt `The people were coming and said to Father Ignatius: ``A crocodile has caught a man.  His name is Iwei. His older brother is Bou. He was walking around crying.'''
\z

\ea\label{ex:nrextiuspu}
    Father Ignatius pulo. Yena kiniyo ukepulupip pu koryo. Pu koryo nekip.\\
    \gll Father Ignatius	pulo	yena	kiniyo	uke-pulupi-p	pu	koryo	pu	koryo	neki-p\\
    F.Ignatius	come.\textsc{pp}	people	all	jointly-come.\textsc{pl}-\textsc{pc}	river	beside	river	beside	stand-\textsc{pc}\\
    \glt `Father Ignatius came. All the people were gathering beside the river. They were standing along the river.'
\z

\ea\label{ex:nrexraipuw}
    Urai pu wenap. Pu ûliyo mono nap.\\
    \gll urai	pu	wena-p	pu	ûliyo	mono	na-p\\
    crocodile	river	carry.inside-\textsc{pc}	river	inside	path	go.inside-\textsc{pc}\\
    \glt `The river current was carrying the crocodile. It was floating under water.'
\z

\ea\label{ex:nrexkouura}
    ``Ine dob powayep! Kou urai baluiko. Urai ere napi.''\\
    \gll ine	dob	pi-we-yep	Kou	urai	ba-lui-ko	urai	ere	napi\\
    you.\textsc{pl}	eye	\textsc{lv}-\textsc{ter}-\textsc{imp}.\textsc{pl}	Kou	crocodile	\textsc{fac}-shoot-\textsc{fac}	crocodile	here	go.inside.\textsc{pl}\\
    \glt `{``}Look carefully! Kou has shot at the crocodile. Here the crocodile is drifting!'''
\z

\ea\label{ex:nrexiweied}
    ``Iweie! De urai ko nieipep.''\\
    \gll Iwei-e	de	urai	ko	nie-ipi-p\\
    Iwei-\textsc{voc}	you	crocodile	I	show-\textsc{1sg}.\textsc{or}-\textsc{pc}\\
    \glt `(Bou sight:) ``Oh Iwei, the crocodile was showing you to me.'''
\z

\ea\label{ex:nrexmipuip}
    Mi puipulo urai. Father Ignatius muelien: ``Elyo kuluem! Puakuyo luiyep! Elyo ono lili.''\\
    \gll mi	puipulo	urai	Father Ignatius	mueli-en	el-yo	k-lui-m	puaku-yo	lui-yep	el-yo	ono	lili\\
    again	come.to.emerge.\textsc{pp}	crocodile	Father Ignatius	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	belly-\textsc{loc}	\textsc{proh}-shoot-\textsc{proh}	head-\textsc{loc}	shoot-\textsc{imp}.\textsc{pl}	belly-\textsc{loc}	man	be.there\\
    \glt `It surfaced again, the crocodile. Father Ignatius said to them: ``Don't shoot at the belly! Shoot at the head! In the belly there is the man.'''
\z

\ea\label{ex:nrexsuipal}
    Urai sui. Paliya. Yena kiniyo urai lapiyo, olyo layowe.\\
    \gll urai	sui	paliya	yena	kiniyo	urai	lapiyo	ol-yo	layo-we\\
    crocodile	die	be.dead	people	all	crocodile	pull.up.\textsc{pp}	hill-\textsc{loc}	lay.\textsc{pp}-\textsc{ter}\\
    \glt `The crocodile dies. It is dead. Many people pulled it (jointly) to the river bank  	and put it there.'
\z

\ea\label{ex:nrexaunapu}
    ``Ine auna puenpi. Ono elyo lili. El kemiyeyo nana riyo akûne.'' Uke kiniyo dob pop, Father Ignatius roise.\\
    \gll ine	auna	puenpi	ono	el-yo	lili	el	kemiye-yo	nana	ri-yo	a-kûne	uke	kiniyo	dob	pi-p	Father Ignatius	roise\\
    you.\textsc{pl}	slowly	cut.meat	man	belly-\textsc{loc}	be.there	belly	soft-\textsc{loc}	small.knife	\textsc{dist}-\textsc{loc}	\textsc{imp}3-go.down	jointly	all	eye	\textsc{lv}-\textsc{pc}	Father Ignatius	with\\
    \glt `{``}Cut slowly, the man is inside the belly! At the soft belly, there the knife should go down.'' All were watching, together with Father Ignatius.'
\z

\ea\label{ex:nrexnatius}
    Father Ignatius muelien: ``Ono puaku ereki.'' Kupuenpopno ono puaku anno piyo. Puaku Iweipi urai elyo lilip. Kipiyapno layowe, suo peleyo layowe.\\
    \gll Father Ignatius	mueli-en	ono	puaku	ere-ki	k-puenpi-p-no	ono	puaku	an-no	piyo	puaku	Iwei-pi	urai	el-yo	lili-p	k-piye-p-no	layo-we	suo	pele-yo	layo-we\\
    Father Ignatius	talk.to-\textsc{nsg}.\textsc{or}	man	head	\textsc{prox}-\textsc{aph}	\textsc{sub}-cut.meat-\textsc{pc}-\textsc{co}	man	head	hand-\textsc{ins}	take.\textsc{pp}	head	Iwei-\textsc{poss}	crocodile	belly-\textsc{loc}	be.there-\textsc{pc}	\textsc{sub}-take-\textsc{pc}-\textsc{co}	lay.\textsc{pp}-\textsc{ter}	coconut	leaf-\textsc{loc}	lay.\textsc{pp}-\textsc{ter}\\
    \glt `Father Ignatius said to them: ``Here is the head of the man!'' When they had cut, they took (out) the head of the man with their hands. Iwei’s head, it was in the crocodile’s belly. After taking it (out) they put it (there). They put it on coconut fronds.'
\z

\ea\label{ex:nrexpelipe}
    Yena lipelip. Eye bouno rapiyo, suo peleyo royowe, kili dupua roise.\\
    \gll yena	lipeli-p	eye	bou-no	rapiyo	suo	pele-yo	royo-we	kili	dupua	roise\\
    people	seek-\textsc{pc}	arm	leg-\textsc{ins}	fetch.\textsc{pp}	coconut.palm	leaf-\textsc{loc}	put.\textsc{pp}-\textsc{du}.\textsc{o}	bone	two	with\\
    \glt `The people were searching on. They took the limbs and put them on coconut palm leaves together with two bones.'
\z

\ea\label{ex:nrexbakase}
    An	baka sepolo, puasi sepolo, om sepolo u roise. Urai elyo ar lilip. ``Arka layero? Uke ar riye.''\\
    \gll an	baka	sepolo	puasi	sepolo	om	sepolo	u	roise	urai	el-yo	ar	lili-p	arka	laye=ro	uke	ar	riye\\
    arm	other	vanish.\textsc{pp}	bladder	vanish.\textsc{pp}	testicles	vanish.\textsc{pp}	penis	with	crocodile	belly-\textsc{loc}	\textsc{neg}	be.there-\textsc{pc}	where	lay=\textsc{emph}	we.\textsc{excl}	\textsc{neg}	see.\textsc{o}.\textsc{inanim}\\
    \glt `The other arm is lost, the bladder is lost, the testicles and the penis are lost, aren't in the crocodile's belly. ``Where are they, we don't see them?'''
\z

\ea\label{ex:nrexkismon}
    Bokis moniyo nis. Kiniseipno yena bî rar. Bîyo wel, bîyo pakûpo, bî sowo.\\
    \gll bokis	moni-yo	nis	k-nisei-p-no	yena	bî	rar	bî-yo	wel	bî-yo	pakûpo	bî	sowo\\
    box	small-\textsc{loc}	put.in.\textsc{pp}	\textsc{sub}-put.in-\textsc{pc}-\textsc{co}	people	hole	dig.\textsc{pp}	hole-\textsc{loc}	carry.\textsc{pp}	hole-\textsc{loc}	throw.down.\textsc{pl}.\textsc{a}.\textsc{pp}	hole	cover.\textsc{pp}\\
    \glt `They put (the remains) into a small box. Having put them in, the people dug a hole. (Then) they carried (the box) to the hole, let it down into the hole, and covered it.'
\z

\ea\label{ex:nrexraiile}
    Yena ba urai ile. Yena ba ar ile ilopi. Yena urai ilopi.\\
    \gll yena	ba	urai	ile	yena	ba	ar	ile	ilopi	yena	urai	ilopi\\
    people	other	crocodile	eat.\textsc{pl}.\textsc{a}	people	other	\textsc{neg}	eat.\textsc{pl}.\textsc{a}	taboo.food	people	crocodile	taboo.food\\
    \glt `Some people eat the crocodile's meat. Other people don't eat it, they taboo it. These people taboo the crocodile's flesh.'
\z

\ea\label{ex:nrexrarmon}
    Riyopuno yena bî rar, moni ikoina rar. Baka royepakûpo bîyo. Bî sowo. Yena kiniyo yipyo molo.\\
    \gll riyopuno	yena	bî	rar	moni	ikoi-na	rar	baka	roye\_pakûpo	bî-yo	bî	sowo	yena	kiniyo	yip-yo	molo\\
    then	people	hole	dig.\textsc{pp}	small	big-\textsc{adv}	dig.\textsc{pp}	other	put\_throw.down.\textsc{pl}.\textsc{a}.\textsc{pp}	hole-\textsc{loc}	hole	cover.\textsc{pp}	people	all	house-\textsc{loc}	go.\textsc{pl}.\textsc{pp}\\
    \glt `Then the people dug (one more) hole, a little bigger hole. They threw the remains  of the crocodile into there. (Then) they covered the hole. (Finally) all the people went back to their houses.'
\z
