\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Uki kopi bi dupua luwe. Bike luro, biep luro, dupua. Uke bike kupuenpopno koyo ropyo niskûnoi, ko Katlin. Bike ropyo kiniskûneipno uke meli puyo. Puppoyo uke meli, pu koryo meli. Koyo doriilo. Uki kopi biep riyo nakap. Koyo paeau. }}
    \ParallelRText{My husband shot two animals. He shot a cassowary and a boar, two animals. We cut up the cassowary and put the flesh into baskets, Katlin and me. After filling the baskets with the meat we carried them to the river. We carried them to the sandy river bank. Then we turned back to the place, where my husband stayed with the boar. We reached the place.}
\end{Parallel}
    \medskip
\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Ko bi eye bouno wono lolowe. Eye bouno ba lolowero. Ko ike bi kama piyeppuo. Ko wel puyo. Pu koryo uke molo. Uke pu epiyo mapap. Uke apla kopopno bi umaeauana aplayo. Ko sû mappeana bi apla. Ko sû kamappapno bi aplayo sû pupuol ppuo aplayo. Bi slau po. Bi kisileipno mi koyo ropyo nisuii. Bike ko appyo lauyayo. Biep ko appyo lauyayo. Koyo ropyo nisuii. Uki kopi dob solo pop. Uke molo Vani\-moyo maketyo. Uke maket po. Koyo maket kopoipno ko uki ikap isiye yilauyo ilo. Katlin uki kep roise nakap Vani\-moyo.}}
    \ParallelRText{With a \textit{kanda}-liana I tied the front limbs of the pig together. Then I tied together the back limbs. I myself lifted the pig up all alone. I carried it to the river. We walked along the river. Then we were staying beside the river. Having made the smoking planks, we put the meat high up on the planks. I lit a fire near the meat planks. After lighting it under the meat planks, the heat went up to the planks. So the meat slowly dried. When the meat had dried up enough, we put it in baskets. I had the cassowary cut up in two halves to share. I had also the boar cut up in two halfs to share. Now we filled the baskets with the meat. My husband was only watching. Having finished with the meat we left for Vani\-mo, to the market. We sold the meat on the market. Having sold the meat, my husband and I returned together to the village. Katlin remained in Vani\-mo together with her husband.}
\end{Parallel}
    \medskip