\ea\label{ex:nrexbieplu}
Uki kopi bi dupua luwe. Bike luro, biep luro, dupua.\\
\gll uki	ko-pi	bi	dupua	lu-we	bike	lu=ro	biep	lu=ro	dupua\\
     husband	1\textsc{sg}-\textsc{poss}	animal	two	shoot.\textsc{pp}-\textsc{du}.\textsc{o}	cassowary	shoot.\textsc{pp}=\textsc{emph}	boar	shoot.\textsc{pp}=\textsc{emph}	two\\
\glt `My husband shot two animals. He shot a cassowary and he shot a boar, two (animals).'
\z

\ea\label{ex:nrexbikekp}
Uke bike kupuenpopno koyo ropyo niskûnoi, ko Katlin.\\
\gll uke	bike	k-puenpi-p-no	koyo	rop-yo	niskûno-i	ko	Katlin\\
     we.\textsc{excl}	cassowary	\textsc{sub}-cut.meat-\textsc{pc}-\textsc{co}	we.\textsc{du}.\textsc{excl}	basket-\textsc{loc}	fill.into.\textsc{pp}-\textsc{du}.\textsc{a}	I	Katlin\\
\glt `Having cut up the cassowary, we filled the meat into baskets, Katlin and me.'
\z

\ea\label{ex:nrexkoryom}
Bike ropyo kiniskûneipno uke meli puyo. Puppoyo uke meli, pu koryo meli.\\
\gll bike	rop-yo	k-niskûne-i-p-no	uke	meli	pu-yo	puppo-yo	uke	meli	pu	koryo	meli\\
     cassowary	basket-\textsc{loc}	\textsc{sub}-fill.into-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	we.\textsc{excl}	carry.\textsc{pl}.\textsc{o}	river-\textsc{loc}	sand-\textsc{loc}	we.\textsc{excl}	carry.\textsc{pl}.\textsc{o}	river	along	carry.\textsc{pl}.\textsc{o}\\
\glt `Having put the cassowary meat into the baskets we carry it to the river. To a sand bar we carry it. We carry it along the river.'
\z

\newpage
\ea\label{ex:nrexkoyodo}
Koyo doriilo. Uki kopi biep riyo nakap. Koyo paeau.\\
\gll koyo	dori\_i-lo	uki	ko-pi	biep	ri-yo	nake-p	koyo	paeau\\
     we.\textsc{du}.\textsc{excl}	return\_\textsc{du}.\textsc{s}-go.\textsc{pp}	husband	1\textsc{sg}-\textsc{poss}	boar	\textsc{dist}-\textsc{loc}	stay-\textsc{pc}	we.\textsc{du}.\textsc{excl}	arrive\\
\glt `We went back. My husband and the boar were (still) staying there (where we had left them). We arrived there.'
\z

\ea\label{ex:nrexobieye}
Ko bi eye bouno wono lolowe. Eye bouno ba lolowero.\\
\gll ko	bi	eye	bou-no	wo-no	lolo-we	eye	bou-no	ba	lolo-we=ro\\
     I	pig	front.limbs	back.limbs-\textsc{ins}	liana-\textsc{ins}	tie.\textsc{pp}-\textsc{du}.\textsc{o}	front.limbs	back.limbs-\textsc{ins}	other	tie.\textsc{pp}-\textsc{du}.\textsc{o}=\textsc{emph}\\
\glt `With a (kanda)-liana I tied the front limbs of the pig together. (Then) I tied together the back limbs.'
\z

\ea\label{ex:nrexikebik}
Ko ike bi kama piyeppuo. Ko wel puyo. Pu koryo uke molo.\\
\gll ko\_ike	bi	kama	piyeppuo	ko	wel	pu-yo	pu	koryo	uke	molo\\
     I.myself	pig	alone	lift.up.\textsc{pp}	I	carry.\textsc{pp}	river-\textsc{loc}	river	along	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}\\
\glt `I myself lifted up the pig  all alone. I carried it to the river. We walked along the river.'
\z

\ea\label{ex:nrexaplakp}
Uke pu epiyo mapap. Uke apla kopopno bi umaeauana aplayo.\\
\gll uke	pu	epiyo	mape-p	uke	apla	k-pi-p-no	bi	umaeauana	apla-yo\\
     we.\textsc{excl}	river	beside	sit.\textsc{pl}-\textsc{pc}	we.\textsc{excl}	plank	\textsc{sub}-make-\textsc{pc}-\textsc{co}	meat	put.high.up.thither.\textsc{pp}	plank-\textsc{loc}\\
\glt `We were sitting beside the river. Having made the smoking planks, we put the meat high up on the planks.'
\z

\ea\label{ex:nrexkosuma}
Ko sû mappeana bi apla. Ko sû kamappapno bi aplayo. Sû pupuol ppuo aplayo. Bi slau po.\\
\gll ko	sû\_mappeana	bi	apla	ko	sû\_k-mappe-p-no	bi	apla-yo	sû	pupuol	ppuo	apla-yo	bi	slau	po\\
     I	light.a.fire.thither.\textsc{pp}	meat	plank	I	light.a.fire-\textsc{pc}-\textsc{co}	meat	plank-\textsc{loc}	fire	heat	go.up.\textsc{pp}	plank-\textsc{loc}	meat	dry	\textsc{lv}.\textsc{pp}\\
\glt `I lit a fire near the meat planks. After lighting it (under) the meat planks the heat of the fire went up to the planks. The meat dried.'
\z

\ea\label{ex:nrexbiksil}
Bi kisileipno mi koyo ropyo nisuii. Bike ko appyo lauyayo. Biep ko appyo lauyayo. Koyo ropyo nisuii.\\
\gll bi	k-silei-p-no	mi	koyo	rop-yo	nisui-i	bike	ko	appyo	lauyayo	biep	ko	appyo	lauyayo	koyo	rop-yo	nisui-i\\
     meat	\textsc{sub}-dry-\textsc{pc}-\textsc{co}	again	we.\textsc{du}.\textsc{excl}	basket-\textsc{loc}	put.in.\textsc{pp}-\textsc{du}.\textsc{a}	cassowary	I	half	carve.up.and.share.\textsc{pp}	boar	I	half	carve.up.and.share.\textsc{pp}	we.\textsc{du}.\textsc{excl}	basket-\textsc{loc}	put.in.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `When the meat had dried we put it again into baskets. I carved up the cassowary in halfs for sharing. I carved up the boar in halfs for sharing. (Then) we put (the meat) into the baskets.'
\z

\ea\label{ex:nrexyomake}
Uki kopi dob solo pop. Uke molo Vanimoyo maketyo. Uke maket po.\\
\gll uki	ko-pi	dob	solo	pi-p	uke	molo	Vanimo-yo	maket-yo	uke	maket	po\\
     husband	1\textsc{sg}-\textsc{poss}	eye	only	\textsc{lv}-\textsc{pc}	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}	Vanimo-\textsc{loc}	market-\textsc{loc}	we.\textsc{excl}	market	\textsc{lv}.\textsc{pp}\\
\glt `My husband was only watching. (Having finished with the meat) we went to Vanimo, to the market. We sold (the meat) on the market.'
\z

\ea\label{ex:nrexyomake1}
Koyo maket kopoipno ko uki ikap isiye yilauyo ilo.\\
\gll koyo	maket	k-pi-i-p-no	ko	uki	ikap	isiye	yilau-yo	i-lo\\
     we.\textsc{du}.\textsc{excl}	market	\textsc{sub}-\textsc{lv}-\textsc{du}.\textsc{a}-\textsc{pc}-\textsc{co}	I	husband	1\textsc{sg}.\textsc{poss}.\textsc{emph}	together	village-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}\\
\glt `Having sold the meat, my husband and I went (back) together to the village.'
\z

\ea\label{ex:nrexkatlin}
Katlin uki kep roise nakap Vanimoyo.\\
\gll Katlin	uki	kep	roise	nake-p	Vanimo-yo\\
     Katlin	husband	3\textsc{sg}.\textsc{poss}	together	stay-\textsc{pc}	Vanimo-\textsc{loc}\\
\glt `Katlin was staying in Vanimo together with her husband.'
\z
