\ea\label{ex:nrexheleny}
Helen yeloyo ye. Yipyo ye. Sementyo ye. Ipol pûkewolo, pikewolo. Ipol yul pikewolo.\\
\gll Helen	yelo-yo	ye	yip-yo	ye	sement-yo	ye	ipol	pûke-wolo	pike-wolo	ipol	yul	pike-wolo\\
	Helen	ground-\textsc{loc}	fall.over.\textsc{pp}	house-\textsc{loc}	fall.over.\textsc{pp}	cement-\textsc{loc}	fall.over.\textsc{pp}	hip.joint	lose-\textsc{cpl}.\textsc{pp}	tear-\textsc{cpl}.\textsc{pp}	hip.joint	joint	tear-\textsc{cpl}.\textsc{pp}\\
\glt `Helen fell over on the ground. She fell at the house. She fell on the cement (floor).  She hurt her hip. The joint tore. She dislocated her hip.'
\z

\ea\label{ex:nrexrafael}
Rafael Cindyyo lakiyepappoi. Rafael Cindyyo mekiyoi.\\
\gll Rafael	Cindy-yo	lakiyepappo-i	Rafael	Cindy-yo	mekiyo-i\\
     Rafael	Cindy-\textsc{loc}	lift.a.motionless.person.\textsc{pp}-\textsc{du}.\textsc{a}	Rafael	Cindy-\textsc{loc}	help.\textsc{pp}-\textsc{du}.\textsc{a}\\
\glt `Rafael and Cindy lifted her up. Rafael and Cindy helped her.'
\z

\ea\label{ex:nrexkoklam}
Helen kok lamo. Yeniyo nuiweyoi. Helen kok lamenakap. Dupuni duwani sipinakap. Ikoina sipip.\\
\gll Helen	kok\_lamo	yeni-yo	nuiweyo-i	Helen	kok\_lame-nake-p	dupuni	duwani	sipi-nake-p	ikoi-na	sipi-p\\
     Helen	scream.out.of.pain.\textsc{pp}	bed-\textsc{loc}	bed.\textsc{pp}-\textsc{du}.\textsc{a}	Helen	scream.out.of.pain-\textsc{dur}-\textsc{pc}	night	daylight	hurt-\textsc{dur}-\textsc{pc}	big-\textsc{adv}	hurt-\textsc{pc}\\
\glt `Helen cried from pain. They laid her on the bed. Helen continued to cry from pain. Day and night she whimpered. It was hurting severely.'
\z

\ea\label{ex:nrexkusopo}
Kuso polinakap dupuni duwani. Asa nuem.\\
\gll kuso	poli-nake-p	dupuni	duwani	asa	nui-m\\
     always	be.there-\textsc{dur}-\textsc{pc}	night	daylight	how	sleep-\textsc{pos}\\
\glt `She was (in this condition) for a long time, day and night. She couldn't sleep.'
\z

\ea\label{ex:nrexabaidj}
Abaidja rapue ponamap. Abaidja mekiyenakap. Dor asa rupiyam, kaepul dupua appue wipilal.\\
\gll Abaidja	rapue	poname-p	Abaidja	mekiye-nake-p	dor	asa	rupiye-m	kaepul	dupua	appue	wipilal\\
     Abaidja	food	give.3\textsc{sg}.\textsc{or}-\textsc{pc}	Abaidja	help-\textsc{dur}-\textsc{pc}	foot	how	stretch-\textsc{pos}	knee	two	sinew	cramp.\textsc{pp}\\
\glt `(Her granddaughter) Abaidja was giving her food. Abaidja helped her all the time. (Helen) couldn't stretch her legs. The sinews cramped (in) both knees.'
\z

\ea\label{ex:nrexplasku}
Plas kuso nuinakap. Plas kauna walpuap puso. Plas dûkû. Plas ili dûkûke.\\
\gll plas	kuso	nui-nake-p	plas	kauna	walpue-p	pu-so	plas	dûkû	plas	ili	dûkû-ke\\
     urine	often	do.intentionally-\textsc{dur}-\textsc{pc}	urine	in.large.amounts	spread.about-\textsc{pc}	water-\textsc{sim}	urine	stink	urine	smell	stink-\textsc{ingr}\\
\glt `She often urinated. Large amounts of urine were spreading about like water. The urine stinks, the urine starts to stink.'
\z

\largerpage
\ea\label{ex:nrexsuipan}
Abaidja ki kama mekiyap. Yena kiniyo suipanepo. Helen kama nakap Abaidja roise. Kiyo kama inuip.\\
\gll Abaidja	ki	kama	mekiye-p	yena	kiniyo	suipane-po	Helen	kama	nake-p	Abaidja	roise	kiyo	kama	i-nui-p\\
     Abaidja	\textsc{aph}	alone	help-\textsc{pc}	people	many	neglect-\textsc{lv}.\textsc{pp}	Helen	alone	sit-\textsc{pc}	Abaidja	with	\textsc{aph}.\textsc{du}	alone	\textsc{du}.\textsc{s}-sleep-\textsc{pc}\\
\glt `Abaidja alone was helping her. All the people neglected her. Helen was alone with Abaidja. They two were sleeping together (in one room).'
\z

\ea\label{ex:nrexpumoab}
Riyopuno Abaidja nomar. Mari ikoina po. Rumkari kep, Sabeth, epe kep ar mekiyo. Kipi ponamo.\\
\gll riyopuno	Abaidja	nomar	mari	ikoi-na	po	rumkari	kep	Sabeth	epe	kep	ar	mekiyo	kipi	ponamo\\
     then	Abaidja	be.very.sick.\textsc{pp}	be.sick	big-\textsc{adv}	\textsc{lv}.\textsc{pp}	girl	3\textsc{sg}.\textsc{poss}	Sabeth	mother	3\textsc{sg}.\textsc{poss}	\textsc{neg}	help.\textsc{pp}	back	give.3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Then Abaidja became very sick. She was very sick. (Helen's) daughter, Sabeth, didn't help her mother. She turned the back on her.'
\z

\ea\label{ex:nrexkohele}
Ko Helen mekiyo. Ko Theresiayo le kep pusiyeno. Koyo sopno pul mopnoi. Ikil kep ikoiele, iripories.\\
\gll ko	Helen	mekiyo	ko	Theresia-yo	le	kep	pusiye-no	koyo	sop-no	pul\_mopi-no-i	ikil	kep	ikoiele	iripories\\
     I	Helen	help.\textsc{pp}	I	Theresia-\textsc{loc}	things	3\textsc{sg}.\textsc{poss}	wash-3\textsc{sg}.\textsc{or}.\textsc{pp}	we.\textsc{du}.\textsc{excl}	soap-\textsc{ins}	bathe-3\textsc{sg}.\textsc{or}.\textsc{pp}-\textsc{du}.\textsc{a}	dirt	3\textsc{sg}.\textsc{poss}	very.big	be.pitiful\\
\glt `(Then) I helped Helen. Theresia and I washed her things, and we washed Helen with soap. She was very dirty. She was in a pitiful state.'
\z

\ea\label{ex:nrexdenisa}
Denis aepu loleno. Lûs kauna doryo. Lûs doryo po. Aepu po ekuyo po. Kok solo lamenakap duwani dupuni.\\
\gll Denis	aepu	lole-no	lûs	kauna	dor-yo	lûs	dor-yo	po	aepu	po	eku-yo	po	kok	solo	lame-nake-p	duwani	dupuni\\
     Denis	ulcer	wrap-3\textsc{sg}.\textsc{or}.\textsc{pp}	worms	in.great.numbers	foot-\textsc{loc}	worms	foot-\textsc{loc}	\textsc{lv}.\textsc{pp}	ulcer	\textsc{lv}.\textsc{pp}	behind-\textsc{loc}	\textsc{lv}.\textsc{pp}	out.of.pain	only	scream-\textsc{dur}-\textsc{pc}	daylight	night\\
\glt `Denis (the nurse) dressed her sores. So many worms on her legs. The worms covered the legs. There were ulcers on her behind. She was only whimpering from pain, day and night.'
\z

\largerpage[2]
\ea\label{ex:nrexyessic}
Yessica sû layo. Helen dorno kosiyepamu. Daunam sûro dor eku ipol roise. Sûro soloro.\\
\gll Yessica	sû	layo	Helen	dor-no	kosiyepamu	daunam	sû=ro	dor	eku	ipol	roise	sû=ro	solo=ro\\
     Yessica	lamp	put.\textsc{pp}	Helen	foot-\textsc{ins}	push.hither.\textsc{pp}	mosquito.net	fire=\textsc{emph}	foot	behind	hip.joint	with	fire=\textsc{emph}	only=\textsc{emph}\\
\glt `Yessica put a lamp (there). (But) Helen knocked it over with her foot towards herself. The mosquito net (caught) fire, her legs, her behind, and her hip. Only flames.'
\z

\ea\label{ex:nrexlenrap}
Helen rapue ilop. Ya ilop. Mek kep bo as. Mek kep kumau. Umul solo wiyap. Umul solo wiyenakap numuelna.\\
\gll Helen	rapue	ilop	ya	ilop	mek	kep	bo	as	mek	kep	kumau	umul	solo	wiye-p	umul	solo	wiye-nake-p	numuelna\\
     Helen	food	taboo.food.\textsc{pp}	sago	taboo.food.\textsc{pp}	mouth	3\textsc{sg}.\textsc{poss}	speech	none	mouth	3\textsc{sg}.\textsc{poss}	shut	heart	only	hold-\textsc{pc}	heart	only	hold-\textsc{dur}-\textsc{pc}	for.a.long.time\\
\glt `Helen refused food. She refused sago. Her mouth didn't make sounds (any more). Her mouth is shut. Only the heart was still beating. Only the heart still continued to beat for a long time.'  
\z

\ea\label{ex:nrexunisui}
Puni sui. Yena pulip dob reyo. Umul solo wiyap. Ko ai sano, ai kuso muelno: ``Mono ponamap, smep paliyenap, de lil elepno pusiyap! Puri lilno.''\\
\gll puni	sui	yena	pulip	dob	reyo	umul	solo	wiye-p	ko	ai	sa-no	ai	kuso	mueli-no	mono	poname-p	smep	paliye-ne-p	de	lil	elep-no	pusiye-p	puri	lil-no\\
     night	die	people	come.\textsc{pl}.\textsc{pp}	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	heart	only	hold-\textsc{pc}	I	father	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	father	always	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	path	give.2\textsc{sg}.\textsc{or}-\textsc{imp}	door	open-3\textsc{sg}.\textsc{or}-\textsc{imp}	you	blood	2\textsc{sg}.\textsc{poss}.\textsc{emph}-\textsc{ins}	wash-\textsc{imp}	clean	blood-\textsc{ins}\\
\glt `At night she dies. People came and looked at her. Only the heart was still beating.  I asked the (heavenly) Father, I prayed to God: ``Prepare the way for her, open the door for her, wash her with your blood! She (shall be) clean with your blood.'''
\z

\ea\label{ex:nrexkesuiw}
Riyopuno kesuiwepu. Sukei kep kelewepu. Balok. Yena wo mop, wo mopiyap. Yena Helen milino.\\
\gll riyopuno	ke-sui-wepu	sukei	kep	ke-le-wepu	ba-le-ko	yena	wo\_mop	wo\_mopiye-p	yena	Helen	mili-no\\
     then	\textsc{aph}-die-\textsc{quant}.\textsc{s}.\textsc{pp}	spirit	3\textsc{sg}.\textsc{poss}	\textsc{aph}-go-\textsc{quant}.\textsc{s}.\textsc{pp}	\textsc{fac}-go-\textsc{fac}	people	weep.\textsc{pp}	weep.\textsc{pl}-\textsc{pc}	people	Helen	surround.a.sick.person-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `Then she finally died. Her spirit went back to its home. It has gone. The people cried. They were crying. (Then) the people surrounded Helen (in their traditional way).'
\z

\ea\label{ex:nrexsarere}
Sarereno sui. Jeffrey Vanimoyo nakap. Pius Jeffrey muelno: ``Epe depi basuik. De lap!'' Jeffrey pulo yilauyo.\\
\gll Sarere-no	sui	Jeffrey	Vanimo-yo	nake-p	Pius	Jeffrey	mueli-no	epe	de-pi	ba-sui-ko	de	le-p	Jeffrey	pulo	yilau-yo\\
     Saturday-\textsc{ins}	die	Jeffrey	Vanimo-\textsc{loc}	sit-\textsc{pc}	Pius	Jeffrey	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	mother	2\textsc{sg}-\textsc{poss}	\textsc{fac}-die-\textsc{fac}	you	go-\textsc{imp}	Jeffrey	come.\textsc{pp}	village-\textsc{loc}\\
\glt `She died on a Saturday. Jeffrey (her second born son) was in Vanimo. Pius told Jeffrey (via cell-phone): ``Your mother has died. Come here!'' And Jeffrey came to the village.'
\z

\ea\label{ex:nrexdavidy}
David yeni wopiyeno. Yena yeniyo layo. Yena ba bî rarip. Bî kraripno bî kuru. Bî bawopiyeko.\\
\gll David	yeni	wopiye-no	yena	yeni-yo	layo	yena	ba	bî	rari-p	bî	k-rari-p-no	bî	kuru	bî	ba-wopiye-ko\\
     David	bed	get.ready-3\textsc{sg}.\textsc{or}.\textsc{pp}	people	bed-\textsc{loc}	lay.\textsc{pp}	people	other	hole	dig-\textsc{pc}	hole	\textsc{sub}-dig-\textsc{pc}-\textsc{co}	hole	be.finished	hole	\textsc{fac}-get.ready-\textsc{fac}\\
\glt `David fixed the coffin for her. The people put her in the coffin. Other people were  digging a hole. Having dug the hole (deep) enough, the grave is ready.'
\z

\ea\label{ex:nrexpakupu}
Riyopuno Helen piyekûpu yeniyo. Bîyo ule. Yena yeniyo roise wel, pakûpu bîyo. Sowo. Basowoko. Yena yipyo molo.\\
\gll riyopuno	Helen	piyekûpu	yeni-yo	bî-yo	ule	yena	yeni-yo	roise	wel	pakûpu	bî-yo	sowo	ba-sowe-ko	yena	yip-yo	molo\\
     then	Helen	take.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	bed-\textsc{loc}	hole-\textsc{loc}	put.inside.\textsc{pp}	people	bed-\textsc{loc}	together	carry.\textsc{pp}	throw.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	hole-\textsc{loc}	cover.\textsc{pp}	\textsc{fac}-cover-\textsc{fac}	people	house-\textsc{loc}	go.\textsc{pl}.\textsc{pp}\\
\glt `Then they took Helen down into the coffin and put it inside the hole. The people carried her together with the coffin and let it down into the grave. They covered it. It is covered. The people went home.'
\z

\ea\label{ex:nrexmarisu}
Jeffrey bo mu: ``Epe kopi marisui. Mari ba pari.'' Yena muelien: ``Ine umul kenekem! Ine bo komoliyemapam! Bo ppulaepo. Layepaneyep!''\\
\gll Jeffrey	bo	mu	epe	ko-pi	mari\_sui	mari	ba	pari	yena	mueli-en	ine	umul\_k-neki-m	ine	bo	k-moliye\_mape-m	bo	ppulae-po	laye\_pane-yep\\
     Jeffrey	speech	speak.\textsc{pp}	mother	1\textsc{sg}-\textsc{poss}	be.sick\_die	sickness	other	\textsc{neg}.\textsc{cop}	people	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{pl}	\textsc{proh}-think-\textsc{proh}	you.\textsc{pl}   speech	\textsc{proh}-speak.\textsc{pl}\_stay.\textsc{pl}-\textsc{proh}	speech	bad-\textsc{lv}.\textsc{pp}	leave.behind-\textsc{imp}.\textsc{pl}\\
\glt `Jeffrey said: ``My mother died of sickness, there was no other harm.''\footnote{The notion of ``other sickness'' refers to illness caused by sorcery. Many people still don't understand the death of a person as a natural consequence of age or illness, but attribute it to the impact of man-made evil.} He said to the people: ``Don't give it any thoughts, don't spread gossip all the time! It is bad gossip, stop it!'''
\z
