\ea\label{ex:nrexkoklep}
Ko klapno rop dupua nek klupuapopno ropyo niskûno.\\
\gll ko	k-le-p-no	rop	dupua	nek	k-lupuape-p-no	rop-yo	niskûno\\
     I	\textsc{sub}-go-\textsc{pc}-\textsc{co}	basket	two	sago.pith	\textsc{sub}-shovel.by.hand.a.lot-\textsc{pc}-\textsc{co}	basket-\textsc{loc}	fill.in.\textsc{pp}\\
\glt `Having gone and shoveled two baskets of sago pith, I filled (the pith) into the baskets (for transport).'
\z

\ea\label{ex:nrexkopiap}
Ko piapeppuo ba puakuyo, ba ko laliyo puakuyo.\\
\gll ko	piapeppuo	ba	puaku-yo	ba	ko	laliyo	puaku-yo\\
     I	lift.up.\textsc{pp}	other	head-\textsc{loc}	other	I	carry.by.hanging.\textsc{pp}	head-\textsc{loc}\\
\glt `One I lift up on (my) head, the other one I carried slung around (my) head.'
\z

\newpage
\ea\label{ex:nrexwemonr}
Ko wemonro ul ko ar riyap. Ul ikoi monoyo lilip.\\
\gll ko	wemon=ro	ul	ko	ar	riye-p	ul	ikoi	mono-yo	lili-p\\
     I	carry.hither.\textsc{pp}=\textsc{emph}	bamboo	I	\textsc{neg}	see.\textsc{o}.\textsc{inanim}-\textsc{pc}	bamboo	big	path-\textsc{loc}	be.there-\textsc{pc}\\
\glt `I carried (the baskets). I didn't see the bamboo. There was a big (piece of) bamboo on the path.'
\z

\ea\label{ex:nrexdorbak}
Dor ba ko kana wopiyeko. Eh ul poro! Srrrrrr. Eh lil monro!\\
\gll dor	ba	ko	kana	wopiye-ko	eh	ul	po=ro	srrr	eh	lil	mon=ro\\
     foot	other	I	quickly	stretch-\textsc{rts}	eh	bamboo	\textsc{lv}.\textsc{pp}=\textsc{emph}	srrr	eh	blood	come.hither.\textsc{pp}=\textsc{emph}\\
\glt `I had already stretched one of my feet. Eh a bamboo! Srrrrrr! Eh the blood ran!'
\z

\ea\label{ex:nrexridise}
Ruri diseipiro Pita pol pele piyamu. Sûno mariyap.\\
\gll ruri	disei-pi=ro	Pita	pol	pele	piyamu	sû-no	mariye-p\\
     child	brother-\textsc{poss}=\textsc{emph}	Pita	betel.pepper	leaf	take.hither.\textsc{pp}	fire-\textsc{ins}	heat-\textsc{pc}\\
\glt `My brother's son Pita collected leaves of betel pepper. (Then) he was heating them  on the fire.'\footnote{The hot leaves are said to staunch the blood.}
\z

\ea\label{ex:nrexokanap}
O kana peou. Ari. Lil ke puso.\\
\gll o	kana	pi-ou	ari	lil	ke	pu-so\\
     \textsc{prox}	quickly	do-\textsc{frus}	no	blood	\textsc{aph}	water-\textsc{sim}\\
\glt `He did this almost in vain. No. The blood, it was like water.'
\z

\ea\label{ex:nrexlilpus}
Lil puso. Dob kopi duwelo.\\
\gll lil	pu-so	dob	ko-pi	duwelo\\
     blood	water-\textsc{sim}	eye	1\textsc{sg}-\textsc{poss}	feel.dizzy.\textsc{pp}\\
\glt `The blood was like water. I felt dizzy.'
\z

\newpage
\ea\label{ex:nrexkopidu}
Dob kopi duwelo. Lil ikoina pi. Lil ke monro.\\
\gll dob	ko-pi	duwelo	lil	ikoina	pi	lil	ke	mon=ro\\
     eye	1\textsc{sg}-\textsc{poss}	feel.dizzy.\textsc{pp}	blood	much	\textsc{lv}	blood	\textsc{aph}	come.hither.\textsc{pp}=\textsc{emph}\\
\glt `I felt dizzy. So much blood. The blood streamed (only).'
\z

\ea\label{ex:nrexyakowo}
``Ya ko powaip!'' Yena ya ko powa.\\
\gll ya	ko	powai-p	yena	ya	ko	powa\\
     sago	I	give.1\textsc{sg}.\textsc{or}-\textsc{imp}	people	sago	I	give.1\textsc{sg}.\textsc{or}.\textsc{pp}\\
\glt `{``}Give me sago!'' (I said). (So) the people gave me sago.'
\z

\ea\label{ex:nrexererud}
Ya ko kunopno ereru dop kopi. Ko muelpup: ``De nakap!''\\
\gll ya	ko	k-ni-p-no	ereru	dop	ko-pi	ko	mueli-ipi-p	de	nake-p\\
     sago	I	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	strong	body	1\textsc{sg}-\textsc{poss}	I	talk.to-1\textsc{sg}.\textsc{or}-\textsc{pc}	you	sit-\textsc{imp}\\
\glt `Having eaten some sago, my body was strong(er). (But) they told me: ``Sit down!'''
\z

\ea\label{ex:nrexnekrop}
Nek rop roise dupua ko woliwoko. Ruri kopi Ayu boyo wemon.\\
\gll nek	rop	roise	dupua	ko	woli-we-ko	ruri	ko-pi	Ayu	boyo	wemon\\
     sago.pith	basket	together	two	I	set-\textsc{du}.\textsc{o}-\textsc{rts}	child	1\textsc{sg}-\textsc{poss}	Ayu	later	carry.hither.\textsc{pp}\\
\glt `I had (already) set (down) the two baskets of sago pith. My son Ayu carried them later (to the kneading trough).'
\z

\ea\label{ex:nrexonakap}
Ko nakapno lil kopi ke pinispo.\\
\gll ko	nake-p-no	lil	ko-pi	ke	pinis-po\\
     I	sit-\textsc{pc}-\textsc{co}	blood	1\textsc{sg}-\textsc{poss}	\textsc{aph}	finish(\textsc{tp})-\textsc{lv}.\textsc{pp}\\
\glt `While I was sitting, my blood stopped (running).'
\z

\ea\label{ex:nrexopdupu}
Rop dupua kopi boyo lulkûne.\\
\gll rop	dupua	ko-pi	boyo	lulkûne\\
     basket	two	1\textsc{sg}-\textsc{poss}	later	pour.down.for.washing\\
\glt `My two baskets (of sago pith) will be washed later.'
\z

\newpage
\ea\label{ex:nrexduiyel}
Kwer duiye. Lil kopi ar pinispo. Dor kopiro ul ko luro.\\
\gll kwer	duiye	lil	ko-pi	ar	pinis-po	dor	ko-pi=ro	ul	ko	lu=ro\\
     frog	croak	blood	1\textsc{sg}-\textsc{poss}	\textsc{neg}	finish(\textsc{tp})-\textsc{lv}.\textsc{pp}	foot	1\textsc{sg}-\textsc{poss}=\textsc{emph}	bamboo	I	shoot.\textsc{pp}=\textsc{emph}\\
\glt `The frogs were croaking. (But) my blood didn't stop. The bamboo cut my foot (severely).'
\z

\ea\label{ex:nrexmiukep}
Mi uke pulupiwap yipyo. Ko nui.\\
\gll mi	uke	pulupi-we-p	yip-yo	ko	nui\\
     then	we.\textsc{excl}	come.\textsc{pl}-\textsc{ter}-\textsc{pc}	house-\textsc{loc}	I	sleep\\
\glt `Then we arrived at the house. I (went to) sleep.'
\z

\ea\label{ex:nrextriden}
Trinde punipino ko pulo oyo. Iki boyo lulilo.\\
\gll Trinde	punipino	ko	pulo	o-yo	iki	boyo	luli\_lo\\
     Wednesday	morning	I	come.\textsc{pp}	\textsc{prox}-\textsc{loc}	\textsc{aph}.\textsc{pl}	later	wash.sago\_go.\textsc{pp}\\
\glt `Wednesday morning I came here (to Claudia's house). Later (Susan's family) went to wash sago.'
\z

\ea\label{ex:nrexclaudd}
Claudia dor kopi puno kupusiyapno, sopno marasin kuputimpopno, banisno lolo.\\
\gll Claudia	dor	ko-pi	pu-no	k-pusiye-p-no	sop-no	marasin	k-putim-pi-p-no	banis-no	lolo\\
     Claudia	foot	1\textsc{sg}-\textsc{poss}	water-\textsc{ins}	\textsc{sub}-wash-\textsc{pc}-\textsc{co}	soap-\textsc{ins}	medicine	\textsc{sub}-put(\textsc{tp})-\textsc{lv}-\textsc{pc}-\textsc{co}	bandage-\textsc{ins}	wrap.\textsc{pp}\\
\glt `Having washed my foot with water and soap and put on some medicine, Claudia dressed it with a bandage.'
\z

\ea\label{ex:nrexkomuel}
Ko muelpup: ``De klam! De yipyo solo nake. De due monasinem! Marasin ko putimporo strongpela.''\\
\gll ko	mueli-ipi-p	de	k-le-m	de	yip-yo	solo	nake	de	due	mona-sini-m	marasin	ko	putim-po=ro	strongpela\\
     I	talk.to-1\textsc{sg}.\textsc{or}-\textsc{pc}	you	\textsc{proh}-go-\textsc{proh}	you	house-\textsc{loc}	only	sit	you	sago	\textsc{irr}-pound.sago.pith-\textsc{irr}	medicine	I	put(\textsc{tp})-\textsc{lv}.\textsc{pp}=\textsc{emph}	strong(\textsc{tp})\\
\glt `She told me: ``Don't go (anywhere), stay in the house only! (Suppose) you would pound sago. The medicine I gave you is strong.'''
\z

\ea\label{ex:nrexarikoe}
Ari, ko epul maloro. Ko ar malo.\\
\gll ari	ko	epul	malo=ro	ko	ar	malo\\
     no	I	ear	hear.\textsc{pp}=\textsc{emph}	I	\textsc{neg}	hear.\textsc{pp}\\
\glt `If I had (only) listened. (But) I didn't listen.'
\z

\ea\label{ex:nrexmatias}
Due ko ar lul. Ko monoyo nakap. Matias ko kana lenou.\\
\gll due	ko	ar	lul	ko	mono-yo	nake-p	Matias	ko	kana	le-ne-ou\\
     sago	I	\textsc{neg}	wash.sago.\textsc{pp}	I	path-\textsc{loc}	sit-\textsc{pc}	Matias	I	quickly	wait.for-3\textsc{sg}.\textsc{or}-\textsc{frus}\\
\glt `I didn't wash sago. I was sitting at the path. I waited in vain for Matias.'
\z

\ea\label{ex:nrexkusopu}
Ko dorimaliyo. Ko kuso pulo.\\
\gll ko	dorimaliyo	ko	kuso	pulo\\
     I	turn.back.to.one's.house.\textsc{pp}	I	always	come.\textsc{pp}\\
\glt `I turned back home. I (had) came (as) always.'
\z

\ea\label{ex:nrexenakom}
Yena ko monoyo relana. Yena so pulupip. Krelanapno uke kumune pulupip yipyo.\\
\gll yena	ko	mono-yo	relana	yena	so	pulupi-p	k-relane-p-no	uke	kumune	pulupi-p	yip-yo\\
     people	I	path-\textsc{loc}	meet.\textsc{o}.\textsc{anim}.\textsc{pl}.\textsc{pp}	people	like	come.\textsc{pl}-\textsc{pc}	\textsc{sub}-meet.\textsc{o}.\textsc{anim}.\textsc{pl}-\textsc{pc}-\textsc{co}	we.\textsc{excl}	all.\textsc{coll}	come.\textsc{pl}-\textsc{pc}	house-\textsc{loc}\\
\glt `I met some people on the way. The people were coming by accident. Having met them we all walked to the house together.'
\z

\ea\label{ex:nrexemkako}
Emka ko ba lo. Ko nakapno dupuni.\\
\gll emka	ko	ba	lo	ko	nake-p-no	dupuni\\
     yesterday	I	\textsc{neg}.\textsc{emph}	go.\textsc{pp}	I	sit-\textsc{pc}-\textsc{co}	night\\
\glt `Yesterday I did not go (anywhere). While I was sitting (in the house), (it became) night.'
\z

\ea\label{ex:nrexraunpo}
Ko ba raunpo. Ko yîs nakap. Yipyo nakap.\\
\gll ko	ba	raun-po	ko	yîs	nake-p	yip-yo	nake-p\\
     I	\textsc{neg}.\textsc{emph}	stroll.around(\textsc{tp})-\textsc{lv}.\textsc{pp}	I	idle	sit-\textsc{pc}	house-\textsc{loc}	sit-\textsc{pc}\\
\glt `I did not stroll around. I was staying idly. I was in (my) house.'
\z

\ea\label{ex:nrexkookip}
Yala ko oki pulo.\\
\gll yala	ko	o-ki	pulo\\
     today	I	\textsc{prox}-\textsc{aph}	come.\textsc{pp}\\
\glt `Today I came here.'
\z

\ea\label{ex:nrexbokkur}
Bo kopi kuru.\\
\gll bo	ko-pi	kuru\\
     story	1\textsc{sg}-\textsc{poss}	be.finished\\
\glt `My story is finished.'
\z
