\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Ko due ropyo ule. Ko piyeppuo puakuyo laliyo. Ko monomno wepulo yipyo. Ko wolomno ppuo yip wapoyo. Ko due kasiyopako. Ali waliyo lolouli. Ko yeloyo seku due roise. Ali waliyo, loloulko. Ko kipino ye el appka. Kipi kopi ikoina sipi. Bui kili sipi, am duwewolap. Rita ko lakiyopappo. Ko dob wippu malippu. Yako dupua ko we\-ppuoi yipyo, yeniyo nuweyoi. Kipi sipino ko wo mop. Ikoina sipip. Wi\-ppip, wiakap. Rita marasin kipi sipino piyoko. Kowepulapno ko no. Dupuni duruka kipi bui kilino, kiniyo sipuwepup. Dupuni duwani kipi sipinakap wîs dupua. Haus sikyo ko lo Vanimoyo. Marasin ko piyo, no. Sipi as, upuna. Yip ikap kanakapno kipi midorisipi.}}
    \ParallelRText{I put the sago into the baskets, lifted them up, and carried them home on my back and slung around my forehead. I brought them to the house walking along the trail. I went up the ladder to the porch of my house. Up above I let the sago fall with an uncontrolled, jerky movement. So the sling of one of the baskets wound around my neck. I fell down from the porch along with the sago. The sling had just wound around my neck. I fell over on my back, belly upwards. My back hurt terribly. My coccyx hurt terribly. It almost broke. Rita lifted me up. My eyes were spinning, and I felt dizzy. Two women carried me up the ladder and into the house and bedded me on my bed. With my hurting back I just cried. It was hurting terribly. I turned my back up and turned it down. Rita got a medication for my hurting back. When she brought it, I took it. Day and night my back and coccyx bones hurt badly all over. Day and night the pain in my back continued, for two months. Finally I went to the hospital in Vanimo. There I got some medicine and took it. No pain anymore. The back felt alright. But as soon as I was back home in my house, my back started to hurt again.}
\end{Parallel}
    \medskip