\ea\label{ex:t4exkoduer}
    Ko due ropyo ule. Ko piyeppuo puakuyo laliyo. Ko monomno wepulo yipyo.\\
    \gll ko	due	rop-yo	ule	ko	piyeppuo	puaku-yo	laliyo	ko	monomno	wepulo	yip-yo\\
    I	sago	basket-\textsc{loc}	put.inside.\textsc{pp}	I	lift.up.\textsc{pp}	head-\textsc{loc}	carry.by.hanging.\textsc{pp}	I	along.the.path	bring.\textsc{pp}	house-\textsc{loc}\\
    \glt `I put the sago into the baskets, lifted them up, carried them home slung around my forehead and brought them to the house (walking) along the trail.'\footnote{One type of basket used by the Kilmeri has a long, about 10cm broad sling. The forehead supports the sling, and the basket hangs on the back. Carrying two baskets like this isn't easy, and one has to walk and move carefully.}
\z

\ea\label{ex:t4exkowolo}
    Ko wolomno ppuo yip wapoyo. Ko due kasiyopako. Ali waliyo lolouli.\\
    \gll ko	wolomno	ppuo	yip	wapo-yo	ko	due	kasiyepako	ali	wali-yo	lole-uli\\
    I	along.the.ladder	go.up.\textsc{pp}	house	porch-\textsc{loc}	I	sago	fall.jerkily.\textsc{pp}	sling	neck-\textsc{loc}	wrap-\textsc{prog}\\
    \glt `I went up the ladder to the porch of the house. Up above I let the sago fall with an uncontrolled, jerky movement. So the sling (of one basket) was winding around my neck.'
\z

\ea\label{ex:t4exkoyelo}
    Ko yeloyo seku due roise. Ali waliyo loloulko. Ko kipino ye el appka.\\
    \gll ko	yelo-yo	seku	due	roise	ali	wali-yo	lole-uli-ko	ko	kipi-no	ye	el	app-ka\\
    I	ground-\textsc{loc}	fall.\textsc{pp}	sago	together	sling	neck-\textsc{loc}	wrap-\textsc{prog}-\textsc{rts}	I	back-\textsc{ins}	fall.over	belly	sky-\textsc{path}\\
    \glt `I fell on the ground along with the sago. The sling had just wound around the neck. I fell over on the back, belly upwards.'
\z

\ea\label{ex:t4exkipiko}
    Kipi kopi ikoina sipi. Bui kili sipi, am duwewolap.\\
    \gll kipi	ko-pi	ikoina	sipi	bui	kili	sipi	am	d-we-wole-p\\
    back	1\textsc{sg}-\textsc{poss}	badly	hurt	coccyx	bone	hurt	almost	\textsc{lkh}-break-\textsc{cpl}-\textsc{pc}\\
    \glt `My back hurt terribly. My coccyx hurt terribly. It would almost break.'
\z

\ea\label{ex:t4exritako}
    Rita ko lakiyopappo. Ko dob wippu malippu. Yako dupua ko weppuoi yipyo, yeniyo nuweyoi.\\
    \gll Rita	ko	lakiyepappo	ko	dob	wippu	malippu	yako	dupua	ko	weppuo-i	yip-yo	yeni-yo	nuweyo-i\\
    Rita	I	lift.a.motionless.person.\textsc{pp}	I	eye	stir.up.\textsc{pp}	feel.dizzy.\textsc{pp}	women	two	I	carry.up.\textsc{pp}-\textsc{du}.\textsc{a}	house-\textsc{loc}	bed-\textsc{loc}	bed.\textsc{pp}-\textsc{du}.\textsc{a}\\
    \glt `Rita lifted me up. My eyes were spinning, and I felt dizzy. Two women carried me up  (the ladder) into the house and bedded me on the bed.'
\z

\ea\label{ex:t4exkipisi}
    Kipi sipino ko wo mop. Ikoina sipip. Wippip, wiakap.\\
    \gll kipi	sipi-no	ko	wo\_mop	ikoina	sipi-p	wippi-p	wiake-p\\
    back	hurt-\textsc{co}	I	weep.\textsc{pp}	badly	hurt-\textsc{pc}	turn.up-\textsc{pc}	turn.down-\textsc{pc}\\
    \glt `With my hurting back I cried. It was hurting terribly. I was turning (my back) up and was turning it down.'
\z

\ea\label{ex:t4exritama}
    Rita marasin kipi sipino piyoko. Kowepulapno ko no.\\
    \gll Rita	marasin	kipi	sipi-no	piye-ko	k-wepule-p-no	ko	no\\
    Rita	medicine	back	hurt-\textsc{co}	take-\textsc{rts}	\textsc{sub}-bring-\textsc{pc}-\textsc{co}	I	eat.\textsc{pp}\\
    \glt `Rita got a medication for my hurting back. When she  brought it, I took it.'
\z

\ea\label{ex:t4ex}
    Dupuni duruka kipi bui kilino, kiniyo sipuwepup. Dupuni duwani kipi sipinakap wîs dupua.\\
    \gll dupuni	duruka	kipi	bui	kili-no	kiniyo	sipi-wepi-p	dupuni	duwani	kipi	sipi-nake-p	wîs	dupua\\
    night	day	back	coccyx	bone-\textsc{ins}	all	hurt-\textsc{quant}.\textsc{s}-\textsc{pc}	night	day	back	hurt-\textsc{dur}-\textsc{pc}	moon	two\\
    \glt `Day and night the back and the coccyx bones were hurting badly all over. Day and night the pain in the back continued, for two months.'
\z

\ea\label{ex:t4exhaussi}
    Haus sikyo ko lo Vanimoyo. Marasin ko piyo, no. Sipi as, upuna.\\
    \gll haus\_sik-yo	ko	lo	Vanimo-yo	marasin	ko	piyo	no	sipi	as	upuna\\
    hospital-\textsc{loc}	I	go.\textsc{pp}	Vanimo-\textsc{loc}	medicine	I	take.\textsc{pp}	eat.\textsc{pp}	pain	none	alright\\
    \glt `I went to the hospital, to Vanimo. I got some medicine and took it. No pain (anymore). (It felt) alright.'
\z

\ea\label{ex:t4exyipika}
    Yip ikap kanakapno kipi midorisipi.\\
    \gll yip	ikap	k-nake-p-no	kipi	mi-dorisipi\\
    house	1\textsc{sg}.\textsc{poss}.\textsc{emph}	\textsc{sub}-sit-\textsc{pc}-\textsc{co}	back	\textsc{iter}-hurt.again\\
    \glt `(But) when I was staying (back home) in my house, the back hurt again.'
\z