\ea\label{ex:nrexnikoka}
    Puni ko kau yek. Ko kau isol kau rurina. Ko ba reyo. Ko dob po: Kau rurina pini, kike solo.\\
    \gll puni	ko	kau	yek	ko	kau	isol	kau	ruri-na	ko	ba	reyo	ko	dob	po	kau	ruri-na	pini	kike	solo\\
    night	I	cow	follow.one's.traces.\textsc{pp}	I	cow	chase.away.\textsc{pp}	cow	child-\textsc{aff}	I	\textsc{neg}.\textsc{emph}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	I	eye	\textsc{lv}.\textsc{pp}	cow	child-\textsc{aff}	come.up.hither	run	only\\
    \glt `At night I followed a cow. I chased a cow, a cow and its young. (But) I didn't see (the young). (Then) I noticed: A cow with its young is coming  towards me. They are just running.'
\z

\ea\label{ex:nrexdobsor}
    Ko dob soreyep: ``Numu de awe! Kau ruri roise kike solo. Pini. Nuko kikei.''\\
    \gll ko	dob	soreye-p	Numu	de	awe	kau	ruri	roise	kike	solo	pini	nuko	kike-i\\
    I	eye	look.intently.\textsc{o}.\textsc{inanim}-\textsc{pc}	Numu	you	come.\textsc{imp}	cow	child	together	run	only	come.up.hither	we.\textsc{incl}	run-\textsc{du}.\textsc{s}\\
    \glt `I was watching it like magnetised: ``Numu, come! A cow with its young is running straight towards us. It's coming up. We (need to) run (ourselves).”'
\z

\ea\label{ex:nrexkikoin}
    Koyo kikoi. Numu sowo, yol bîka lupuana. Ko kiko, kau kiko. Ko seliwolo. Simon wonino:\\
    \gll koyo	kiko-i	Numu	sowo	yol	bî-ka	lupuana	ko	kiko	kau	kiko	ko	seli-wolo	Simon	woni-no\\
    we.\textsc{du}.\textsc{excl}	run.\textsc{pp}-\textsc{du}.\textsc{s}	Numu	hide.\textsc{pp}	fence	hole-\textsc{path}	go.in.\textsc{pp}	I	run.\textsc{pp}	cow	run.\textsc{pp}	I	slip-\textsc{cpl}.\textsc{pp}	Simon	call-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
    \glt `We both ran. Numu hid away. She sneaked through a hole in the fence. I ran. The cow ran. I slipped and fell over. I called out for Simon.'
\z

\ea\label{ex:nrexawedek}
    ``De awe! De kau rino pakip! De ko mekiyap!''\\
    \gll de	awe	de	kau	ri-no	paki-p	de	ko	mekiye-p\\
    you	come.\textsc{imp}	you	cow	stick-\textsc{ins}	beat-\textsc{imp}	you	I	help-\textsc{imp}\\
    \glt `{``}Come! Beat the cow with a stick! Help me!”'
\z

\ea\label{ex:nrexkopiki}
    An kopi kipika ye. Ulelipana. Ko umul nekpamu. Ko an baka wiyo, popompiyo.\\
    \gll an	ko-pi	kipi-ka	ye	ulelipana	ko	umul\_nekpamu	ko	an\_baka	wiyo	popompiyo\\
    arm	1\textsc{sg}-\textsc{poss}	back-\textsc{path}	fall.over.\textsc{pp}	twist.\textsc{pp}	I	ponder.\textsc{pp}	I	other.hand	hold.\textsc{pp}	stretch.straight.\textsc{pp}\\
    \glt `My arm fell on the back. It twisted. I pondered (about what to do). (Then) I held it with my other hand and stretched it straight.'
\z

\ea\label{ex:nreximonba}
    Simon bayekiko. Balok. Ko Numu wonino: ``De awe! Nuko ilei yipyo.''\\
    \gll Simon	ba-yeki-ko	ba-le-ko	ko	Numu	woni-no	de	awe	nuko	i-le-i	yip-yo\\
    Simon	\textsc{fac}-pursue-\textsc{fac}	\textsc{fac}-go-\textsc{fac}	I	Numu	call-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	come.\textsc{imp}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go-\textsc{du}.\textsc{s}	house-\textsc{loc}\\
    \glt `Simon has routed (the cow). It has gone. I called Numu: ``Come, let's go into  the house.”'
\z

\ea\label{ex:nrexoyeniy}
    Ko yeniyo nu. Ko asa nuem. An kopi ikoina sipip. Ko dupuniro puana, Numu muelno:\\
    \gll ko	yeni-yo	nu	ko	asa	nui-m	an	ko-pi	ikoi-na	sipi-p	ko	dupuni=ro	puana	Numu	mueli-no\\
    I	bed-\textsc{loc}	do.intentionally.\textsc{pp}	I	how	sleep-\textsc{pos}	arm	1\textsc{sg}-\textsc{poss}	big-\textsc{adv}	hurt-\textsc{pc}	I	night=\textsc{emph}	rise.\textsc{pp}	Numu	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
    \glt `I lay down to sleep. (But) I could not sleep. My arm was hurting terribly. I rose in the (middle of the) night and said to Numu.'
\z

\ea\label{ex:nrexdepuan}
    ``De puanap! Ko yaup yowe. De sû mappap!'' Sû bamappeko. ``Pu ipino ipiyap!''\\
    \gll de	puane-p	ko	yaup	yowe	de	sû\_mappe-p	sû\_ba-mappe-ko	pu	ipi-no	ipiye-p\\
    you	stand.up-\textsc{imp}	I	hot.water	boil	you	 light.a.fire-\textsc{imp}	fire\_\textsc{fac}-light-\textsc{fac}	water	pot-\textsc{ins}	put.in.a.pot-\textsc{imp}\\
    \glt `{``}Get up! I boil water. You light a fire!” The fire is lit. ``(Now) fill water in a pot!”'
\z

\ea\label{ex:nrexyauppu}
    Yaup pupuol po, mol. Puliyo an kloknino.\\
    \gll yaup	pupuol	po	mol	puliyo	an	klokni-no\\
    hot.water	heat	\textsc{lv}.\textsc{pp}	boil.\textsc{pp}	take.off.\textsc{pp}	hand	one-\textsc{ins}\\
    \glt `The water became hot and boiled. I took it away (from the fire) with one hand.'
\z

\ea\label{ex:nrexpililu}
    ``Pili lupi piyekap!'' Ko an ikap sililipi. Lil upuna lam dopyo. Upuna walwolam.\\
    \gll pili	lupi	piye-ke-p	ko	an	ikap	sililipi	lil	upuna	le-m	dop-yo	upuna	walwole-m\\
    cloth	end	take-\textsc{ingr}-\textsc{imp}	I	arm	1\textsc{sg}.\textsc{poss}.\textsc{emph}	press.a.hot.wet.cloth.onto.skin	blood	alright	go-\textsc{pos}	body-\textsc{loc}	alright	diffuse-\textsc{pos}\\
    \glt `{``}Go get a piece of cloth”, I said. I press a hot, wet cloth onto my arm. (So) the blood will flow well through the body. It will diffuse well.'
\z

\ea\label{ex:nrexiyopun}
    Riyopuno ko yeniyo lo. Koyo ilo. Ko due nu duruwa.\\
    \gll riyopuno	ko	yeni-yo	lo	koyo	i-lo	ko	due\_nu	duruwa\\
    then	I	bed-\textsc{loc}	go.\textsc{pp}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-go.\textsc{pp}	I	sleep.\textsc{pp}	dawn\\
    \glt `Then I went to bed. We two went (to bed). I slept until dawn.'
\z

\newpage
\ea\label{ex:nrexdemisu}
    ``De mi sû mappap! Ko yaup yowe. Mi ko yaupno sililipi.''\\
    \gll de	mi	sû\_mappe-p	ko	yaup	yowe	mi	ko	yaup-no	sililipi\\
    you	again	light.a.fire-\textsc{imp}	I	hot.water	boil	again	I	hot.water-\textsc{ins}	press.a.hot.wet.cloth.onto.skin\\
    \glt `{``}You light a fire again! I'll boil water. I'll press again a hot, wet cloth (onto my arm).”'
\z

\ea\label{ex:nrexmikopi}
    Mi ko pili lupi yaupyo îlako. Mi ko pakulyo sililipo.\\
    \gll mi	ko	pili	lupi	yaup-yo	îlako	mi	ko	pakul-yo	sililipo\\
    again	I	cloth	end	hot.water-\textsc{loc}	dip.into.water.\textsc{pp}	again	I	shoulder-\textsc{loc}	press.a.hot.wet.cloth.onto.skin.\textsc{pp}\\
    \glt `I dipped again the piece of cloth into the hot water. I pressed the hot, wet cloth onto my shoulder.'
\z

\ea\label{ex:nrexdopika}
    Ko dop ikap suko. Lil puso solo. Lil ba amaleipe. Lil bapok. Lil aska.\\
    \gll ko	dop	ikap	suko	lil	pu-so	solo	lil	ba	am-a-le-ipe	lil	ba-pi-ko	lil	aska\\
    I	skin	1\textsc{sg}.\textsc{poss}.\textsc{emph}	cut.\textsc{pp}	blood	water-\textsc{sim}	only	blood	other	\textsc{grad}-\textsc{imp}3-go-\textsc{ant}	blood	\textsc{fac}-do-\textsc{fac}	blood	none\\
    \glt `I cut my skin. The blood is only (transparent) like water. Other blood, (red blood) should flow first. The blood has flown out. No blood (anymore).'
\z

\ea\label{ex:nrexkoeyep}
    Ko eye piyeppuou. Ikoina sipi, yilipi.\\
    \gll ko	eye	piyeppue-ou	ikoi-na	sipi	yili-pi\\
    I	arm	lift.up-\textsc{frus}	big-\textsc{adv}	hurt	tense-\textsc{lv}\\
    \glt `I try to lift my arm in vain. It hurts terribly. It is tense.'
\z

\ea\label{ex:nrexkoasap}
    Pili ko asa poniyam. Mi ko auna poniye, auna kure. Ikoina sipinakap.\\
    \gll pili	ko	asa	poniye-m	mi	ko	auna	poniye	auna	kure	ikoi-na	sipi-nake-p\\
    cloth	I	how	wrap-\textsc{pos}	again	I	slowly	wrap	slowly	put.on	big-\textsc{adv}	hurt-\textsc{dur}-\textsc{pc}\\
    \glt `I can't put on clothes. I have to wrap slowly (a laplap). I put it on slowly. It continued to hurt terribly.'\footnote{A laplap is a nicely designed, but informal piece of cloth about 1,60m in length and 1m in width, which women sling around their body.}
\z

\ea\label{ex:nrexkolomo}
    Mi ko lo monomno. Ko Matyu iloi. Opo kompanipi kumuneyo lakiyowe.\\
    \gll mi	ko	lo	monomno	ko	Matyu	i-lo-i	opo	kompani-pi	kumune-yo	lakiyo-we\\
    then	I	go.\textsc{pp}	along.the.path	I	Matyu	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	car	company-\textsc{poss}	all.\textsc{coll}-\textsc{loc}	fetch.people.\textsc{pp}-\textsc{du}.\textsc{o}\\
    \glt `Then I went along the road (to Vanimo). Me and Matyu, we walked. (Later) a car of the Company took both of us along.'
\z

\ea\label{ex:nrexnoroke}
    Ono roke nem kep Joe Erima. Iloro Base Campyo.\\
    \gll ono	ro-ke	nem	kep	Joe	Erima	i-lo=ro	Base	Camp-yo\\
    man	\textsc{prox}.\textsc{emph}-\textsc{aph}	name	3\textsc{sg}.\textsc{poss}	Joe	Erima	\textsc{du}.\textsc{s}-go.\textsc{pp}=\textsc{emph}	Base	Camp-\textsc{loc}\\
    \glt `This man, (the driver) named Joe Erima, (picked us up.) We drove (with him) to the Base Camp.'
\z

\ea\label{ex:nrexkoyoop}
    Mi koyo opo baya îpîpui. Koyo iloi, mi ineki, koyo Waisan Campyo ikûnoi. Koyo inakap.\\
    \gll mi	koyo	opo	baya	îpîpu-i	koyo	i-lo-i	mi	i-neki	koyo	Waisan\_Camp-yo	i-kûno-i	koyo	i-nake-p\\
    then	we.\textsc{du}.\textsc{excl}	car	other	climb.on.\textsc{pp}-\textsc{du}.\textsc{s}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	again	\textsc{du}.\textsc{s}-stand	we.\textsc{du}.\textsc{excl}	Waisan.Camp-\textsc{loc}	\textsc{du}.\textsc{s}-go.down.\textsc{pp}-\textsc{du}.\textsc{s}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-sit-\textsc{pc}\\
    \glt `Then we jumped on another car. We drove, stood again (on the truck bed), and jumped off at Waisan Camp. There we were staying.'
\z

\ea\label{ex:nrexnipino}
    Punipino lo haus sikyo, ko kama. Ko marasin piyo, bayana. Ko puno no.\\
    \gll punipino	lo	haus\_sik-yo	ko	kama	ko	marasin	piyo	bayana	ko	pu-no	no\\
    morning	go.\textsc{pp}	hospital-\textsc{loc}	I	alone	I	medicine(\textsc{tp})	take.\textsc{pp}	different	I	water-\textsc{ins}	eat.\textsc{pp}\\
    \glt `In the morning I went to the hospital, I alone. I took a medication, a different one.  I took it with water.'
\z

\ea\label{ex:nrexkomara}
    Mi ko marasin kiniyo piyo. Ko yipyo wel. Rokloknino nep.\\
    \gll mi	ko	marasin	kiniyo	piyo	ko	yip-yo	wel	ro=kloknino	ni-p\\
    again	I	medicine	many	take.\textsc{pp}	I	house-\textsc{loc}	carry.\textsc{pp}	\textsc{emph}=one.time	eat-\textsc{pc}\\
    \glt `Then I took many capsules (of the medication). I took them home. I swallowed them daily.'
\z

\ea\label{ex:nrexmoniup}
    An kopiro sipi as. Moni upuna.\\
    \gll an	ko-pi=ro	sipi	as	moni	upuna\\
    arm	1\textsc{sg}-\textsc{poss}=\textsc{emph}	pain	none	little	alright\\
    \glt `My arm doesn't hurt any more. It feels a bit better.'
\z

\ea\label{ex:nrexlekoas}
    Sele ko asa pem. Yili ko asa wilem. Due ko asa sonem, lulem. Ko yala nake solo pi.\\
    \gll sele	ko	asa	pi-m	yili	ko	asa	wili-m	due	ko	asa	soni-m	luli-m	ko	yala	nake	solo	pi\\
    garden	I	how	make-\textsc{pos}	weight	I	how	carry-\textsc{pos}	raw.sago	I	how	pound.sago.pith-\textsc{pos}	wash.sago-\textsc{pos}	I	now	sit	only	do\\
    \glt `I can't work in the garden. I can't carry any weight. I can't pound and wash sago. Now all I do is sit (idly).'
\z

\ea\label{ex:nrexkowaye}
    Ko wayeou ko sele pi. Pakul waliyo sipi. Dop kiniyo, dop sipiwepi. Ko nake solo pi, sipi as, upuna.\\
    \gll ko	waye-ou	ko	sele	pi	pakul	wali-yo	sipi	dop	kiniyo	dop	sipi-wepi	ko	nake	solo	pi	sipi	as	upuna\\
    I	try-\textsc{frus}	I	garden	do	shoulder	neck-\textsc{loc}	hurt	body	many	body	hurt-\textsc{quant}.\textsc{s}	I	sit	only	do	pain	none	alright\\
    \glt `I tried in vain to work in the garden. (My) shoulder and neck hurt. The whole body (hurts). My body hurts all over. (If) I rest only, (there is) no pain, it is alright.'
\z
