\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Dika ko Margaretyo duyo iloi. Koyo yipp su lipeli. Koyo yûr pol riyepoip. Yûr pol ko bariyeko. Margaret yûr pol meip. Ko eku nakap. Ari, ko nekip, kles siyip. Riyopuno Margaret yipp su dupua paliyowe. Ba ura no. Dupua solo paliyowe. Margaret yipp su dupua rupueno lolowe. Ba kini ko powa, ba kini kep. Ki klokni ko klokni. Koyo yipyo iloi. Margaret weiau suelip. Ko nekip. Riyopuno mi ko emur suel. Emur ko ino pu siamu. Margaret an solo siana. Ko	ri wili laine. Ko emurno pulo. Emurno puloro, yipyo paeau. Emur mi ko yipyo layowe. Emur ulili. Ko ar piyeko.}}
    \ParallelRText{The day before yesterday Margaret and I went to the bush. We were looking for eggs of a wild fowl. We saw a bird's nest with eggs. Now I finally saw a wildfowl's nest with eggs! With her hands, Margaret digged for the eggs in the nest. I was sitting aside. No, I was standing and expelling the mosquitos! Then Margaret uncovered two eggs. A lizard had eaten another one. So she only dug out two. Margaret wrapped each of the two eggs of the wild fowl in a leaf. One she gave to me, one is hers. One for her, one for me. We went back home. On the way Margaret cut a \textit{kanda}-liana. I stood aside and looked. Then she cut a walking stick for me. I crossed the water ditch with the walking stick. Margaret crossed it without a stick. I balanced on a log. I walked with a stick. With the walking stick I came to the house and reached it. The walking stick I put into the house. It is here. I didn't throw it away.}
\end{Parallel}