\ea\label{ex:nrexdikako}
    Dika ko Margaretyo duyo iloi. Koyo yipp su lipelip.\\
    \gll dika	ko	Margaret-yo	du-yo	i-lo-i	koyo	yipp	su	lipeli-p\\
    day.before.yesterday	I	Margaret-\textsc{loc}	bush-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	we.\textsc{du}.\textsc{excl}	wild.fowl	egg	seek-\textsc{pc}\\
    \glt `The day before yesterday Margaret and I went to the bush. We were looking for eggs  	of a wild fowl.'
\z

\ea\label{ex:nrexkoyoyu}
    Koyo yûr pol riyepoip. Yûr pol ko bariyeko.\\
    \gll koyo	yûr	pol	riye-pi-i-p	yûr	pol	ko	ba-riye-ko\\
    we.\textsc{du}.\textsc{excl}	bird	nest.with.eggs	see.\textsc{o}.\textsc{inanim}-\textsc{lv}-\textsc{du}.\textsc{a}-\textsc{pc}	bird	nest.with.eggs	I	\textsc{fac}-see.\textsc{o}.\textsc{inanim}-\textsc{fac}\\
    \glt `We saw a bird's nest with eggs. (Now) I have seen a (wild fowl's) nest with eggs.''\footnote{Wild fowls breed on the ground. Deep in the forest they build huge nests from sticks and leaves.}
\z

\ea\label{ex:nrexpolmei}
    Margaret yûr pol meip. Ko eku nakap. Ari, ko nekip, kles siyip.\\
    \gll Margaret	yûr	pol	mei-p	ko	eku	nake-p	ari	ko	neki-p	kles	siyi-p\\
    Margaret	bird	nest.with.eggs	dig.with.one's.hands-\textsc{pc}	I	behind	sit-\textsc{pc}	no	I	stand-\textsc{pc}	mosquito	throw-\textsc{pc}\\
    \glt `Margaret was digging for the eggs in the nest with her hands. I was sitting (aside). No, I was standing and expelling the mosquitos.'
\z

\ea\label{ex:nrexpunoma}
    Riyopuno Margaret yipp su dupua paliyowe. Ba ura no. Dupua solo paliyowe.\\
    \gll riyopuno	Margaret	yipp	su	dupua	paliyo-we	ba	ura	no	dupua	solo	paliyo-we\\
    then	Margaret	wild.fowl	egg	two	open.\textsc{pp}-\textsc{du}.\textsc{o}	other	lizard	eat.\textsc{pp}	two	only	open.\textsc{pp}-\textsc{du}.\textsc{o}\\
    \glt `Then Margaret uncovered two eggs. One (more) a lizard (had) eaten. She dug out only two.'
\z

\ea\label{ex:nrexyippdu}
    Margaret yipp su dupua rupueno lolowe. Ba kini ko powa, ba kini kep. Ki klokni ko klokni.\\
    \gll Margaret	yipp	su	dupua	rupue-no	lolo-we	ba	kini	ko	powa	ba	kini	kep	ki	klokni	ko	klokni\\
    Margaret	wild.fowl	egg	two	leaf-\textsc{ins}	wrap.\textsc{pp}-\textsc{du}.\textsc{o}	other	one.\textsc{part}	I	give.1\textsc{sg}.\textsc{or}.\textsc{pp}	other	one.\textsc{part}	3\textsc{sg}.\textsc{poss}	\textsc{aph}	one	I	one\\
    \glt `Margaret wrapped the two eggs of the wild fowl in a leaf (each). One she gave to me,  one is her's. One for her, one for me.'
\z

\ea\label{ex:nrexyoyipy}
    Koyo yipyo iloi. Margaret weiau suelip. Ko nekip.\\
    \gll koyo	yip-yo	i-lo-i	Margaret	weiau	sueli-p	ko	neki-p\\
    we.\textsc{du}.\textsc{excl}	house-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}-\textsc{du}.\textsc{s}	Margaret	kanda	cut-\textsc{pc}	I	stand-\textsc{pc}\\
    \glt `We went (back) home. (On the way) Margaret was cutting to size a kanda(-liana). I was standing (aside).'
\z

\ea\label{ex:nrexmikoem}
    Riyopuno mi ko emur suel. Emur ko ino pu siamu. Margaret an solo siana.\\
    \gll riyopuno	mi	ko	emur	suel	emur	ko	i-no	pu	siamu	Margaret	an	solo	siana\\
    then	again	I	walking.stick	cut.\textsc{pp}	walking.stick	I	\textsc{dist}-\textsc{ins}	water	cross.hither.\textsc{pp}	Margaret	hand	only	cross.thither.\textsc{pp}\\
    \glt `Then she cut a walking stick (for me). I crossed the water (ditch) with the walking stick. Margaret crossed it without (a stick).'
\z

\ea\label{ex:nrexkoriwi}
    Ko ri wili laine. Ko emurno pulo. Emurno puloro, yipyo paeau.\\
    \gll ko	ri\_wili	laine	ko	emur-no	pulo	emur-no	pulo=ro	yip-yo	paeau\\
    I	log	seek.balance	I	stick-\textsc{ins}	come.\textsc{pp}	stick-\textsc{ins}	come.\textsc{pp}=\textsc{emph}	house-\textsc{loc}	arrive\\
    \glt `I balance on a log. I came with a stick. With a walking stick I came to the house and reached it.'
\z

\newpage
\ea\label{ex:nrexemurmi}
    Emur mi ko yipyo layowe. Emur ulili. Ko ar piyeko.\\
    \gll emur	mi	ko	yip-yo	layo-we	emur	u-lili	ko	ar	piyi-ko\\
    stick	then	I	house-\textsc{loc}	lay.\textsc{pp}-\textsc{ter}	stick	\textsc{dfac}-be.there	I	\textsc{neg}	throw.away-\textsc{rts}\\
    \glt `Then I put the walking stick into the house. The stick is here. I didn't throw it away.'
\z
