\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Ko Trindeno lo, 8 a clock. Ko loro. Ko loro Isa yilauyo, paeau. Mi ko eppi noyap. Ko Simon Joe, uke tripela. Isa ako aeu po apono. Nem kep Ipp. Aeu popno bese siro, bi roise yûr roise. Baroko. Wîlyo rupopo. Uke konupno uke molo. Uke kûno puyo Pumon. Yelo pur ol ari. Uke moloro Pusokyo. Uke ppuoppuo olyo. Nem kep Awolu. Olro uke pueliyekepuo. Mi uke ol baya ppuo ol Oni. Yilauyo paeau. Uke eppi noyap, yip Yawapi ako kep roise.}}
    \ParallelRText{I went on Wednesday, at 8 o'clock. I walked. I went to Isa's hamlet. I arrived. There I was resting. Me, Simon, Joe, the three of us were resting. Isa's wife made sago pancakes on a baking plate. Her name is Ipp. While making the sago pancakes she cooked \textit{tulip}-vegetable with pig and chicken. It was done. She distributed it on the plates. Having eaten we went on. We went down to Pumon creek. The country is flat, there are no hills. We went on to Pusok creek. Then we climbed up Mount Awolu. We left the mountain behind us and climbed one more mountain, Mount Oni. We arrived at a hamlet. There we were resting, in the house of Yawa and his wife.}
\end{Parallel}
    \medskip
\begin{Parallel}{0.47\textwidth}{0.47\textwidth}
    \ParallelLText{\textit{Uke moloro Pupualyo. Ko pulmop, Simon pulmop. Ko uro puakuyo ppuo. Koyo ilo. Pu po. Koyo pu isiye ilo. Uke ol Ayauiyo ippuo. Ayauiyo ippuo epi mono. Pu ikoiele. Koyo pu isiye iloro. Koyo pu monoyo piyei. Boyopuno pu poyana. Koyo iloro Uleiyo paeau. Koyo pueliyekoi. Koyo ono bayana Makoa reyanai. Koyo lewoinep monoyo. Boyo uke tripela moloro Pasiyo paeau. Riyopuno uke basno molo Waisan Campyo. Uke mapap.}}
    \ParallelRText{We went down to Pupual creek. I waded through, Simon waded through. Then I climbed up the riverbank with the netbag around the head. We went on. It started to rain. We walked through the rain. We climbed Mount Ayaui, but we took the trail along the flank. The rain became heavy. We walked on through the rain. We became very wet on the way. Later on the rain stopped. We went to Ulei and arrived there. Soon we left it behind and then met a man, Makoa. He was waiting for us at the road. Now the three of us continued to the village of Pasi and arrived there. Then we went on by bus to Waisan Camp. There we stayed.}
\end{Parallel}
    \medskip