\ea\label{ex:nrexkotrin}
    Ko Trindeno lo, 8 a clock. Ko loro. Ko loro Isa yilauyo, paeau. Mi ko eppi noyap. Ko Simon Joe, uke tripela. Isa ako aeu po apono. Nem kep Ipp.\\
    \gll ko	Trinde-no	lo	8aclock	ko	lo=ro	ko	lo=ro	Isa	yilau-yo	paeau	mi	ko	eppi\_noye-p	ko	Simon	Joe	uke	tripela	Isa	ako	aeu	po	apo-no	nem	kep	Ipp\\
    I	Wednesday-\textsc{ins}	go.\textsc{pp}	8\_o'clock	I	go.\textsc{pp}=\textsc{emph}	I	go.\textsc{pp}=\textsc{emph}	Isa	village-\textsc{loc}	arrive	then	I	rest-\textsc{pc}	I	Simon	Joe	we.\textsc{excl}	three(TP)	Isa	wife	sago.pancake	make.\textsc{pp}	pan-\textsc{ins}	name	3\textsc{sg}.\textsc{poss}	Ipp\\
    \glt `I went on Wednesday, at 8 o'clock. I walked. I went to Isa's hamlet and arrived. (There) I was resting. Me, Simon, Joe, the three of us (were resting). Isa's wife made sago pancakes on a baking plate. Her name is Ipp.'
\z

\largerpage
\ea\label{ex:nrexaeupip}
    Aeu popno bese siro, bi roise yûr roise. Baroko. Wîlyo rupopo.\\
    \gll aeu	pi-p-no	bese	si=ro	bi	roise	yûr	roise	ba-re-ko	wîl-yo	rupopo\\
    sago.pancake	make-\textsc{pc}-\textsc{co}	tulip.vegetable	cook.\textsc{pp}=\textsc{emph}	pig	with	chicken	with	\textsc{fac}-be.done-\textsc{fac}	plate-\textsc{loc}	distribute.\textsc{pl}.\textsc{o}.\textsc{pp}\\
    \glt `While making the sago pancakes, she cooked \textit{tulip}-vegetable with pig and chicken. It is done. She distributed it on the plates.'
\z

\ea\label{ex:nrexukekni}
    Uke konupno uke molo. Uke kûno puyo Pumon. Yelo pur ol ari. Uke moloro Pusokyo. Uke ppuoppuo olyo. Nem kep Awolu.\\
    \gll uke	k-ni-p-no	uke	molo	uke	kûno	pu-yo	Pumon	yelo	pur	ol	ari	uke	molo=ro	Pusok-yo	uke	ppuo-ppuo	ol-yo	nem	kep	Awolu\\
    we.\textsc{excl}	\textsc{sub}-eat-\textsc{pc}-\textsc{co}	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}	we.\textsc{excl}	go.down.\textsc{pp}	river-\textsc{loc}	Pumon	ground	flat	mountain	no	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	Pusok-\textsc{loc}	we.\textsc{excl}	go.up.\textsc{pp}-go.up.\textsc{pp}	mountain-\textsc{loc}	name	3\textsc{sg}.\textsc{poss}	Awolu\\
    \glt `Having eaten we went (on). We went down to the Pumon (creek). The country is flat, no hills. We went on to the Pusok (creek). (Then) we climbed up a mountain. Its name is Awolu.'
\z

\ea\label{ex:nrexolrouk}
    Olro uke pueliyekepuo. Mi uke ol baya ppuo ol Oni. Yilauyo paeau. Uke eppi noyap, yip Yawapi ako kep roise.\\
    \gll ol=ro	uke	pueliyekepuo	mi	uke	ol	baya	ppuo	ol	Oni	yilau-yo	paeau	uke	eppi\_noye-p	yip	Yawa-pi	ako	kep	roise\\
    mountain=\textsc{emph}	we.\textsc{excl}	leave.behind.\textsc{pp}	then	we.\textsc{excl}	mountain	other	go.up.\textsc{pp}	mountain	Oni	village-\textsc{loc}	arrive	we.\textsc{excl}	rest-\textsc{pc}	house	Yawa-\textsc{poss}	wife	3\textsc{sg}.\textsc{poss}	with\\
    \glt `We left the mountain behind (us) and climbed one more mountain, the mountain Oni. We arrived at a hamlet. (There) we were resting, (in) the house of Yawa and his wife.'
\z

\ea\label{ex:nrexkemolo}
    Uke moloro Pupualyo. Ko pulmop, Simon pulmop. Ko uro puakuyo ppuo. Koyo ilo. Pu po. Koyo pu isiye ilo.\\
    \gll uke	molo=ro	Pupual-yo	ko	pul\_mop	Simon	pul\_mop	ko	uro	puaku-yo	ppuo	koyo	i-lo	pu	po	koyo	pu	isiye	i-lo\\
    we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	Pupual-\textsc{loc}	I	wade.a.creek.\textsc{pp}	Simon	wade.a.creek.\textsc{pp}	I	netbag	head-\textsc{loc}	go.up.\textsc{pp}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-go.\textsc{pp}	rain	\textsc{lv}.\textsc{pp}	we.\textsc{du}.\textsc{excl}	rain	together	\textsc{du}.\textsc{s}-go.\textsc{pp}\\
    \glt `We went (down) to the Pupual (creek). I waded through, Simon waded through. (Then) I climbed up (the riverbank) with the netbag around the head. We went on. It rained. We went through the rain.'
\z

\newpage
\ea\label{ex:nrexolayau}
    Uke ol Ayauiyo ippuo. Ayauiyo ippuo epi mono. Pu ikoiele. Koyo pu isiye iloro.\\
    \gll uke	ol	Ayaui-yo	i-ppuo	Ayaui-yo	i-ppuo	epi\_mono	pu	ikoiele	koyo	pu	isiye	i-lo=ro\\
    we.\textsc{excl}	mountain	Ayaui-\textsc{loc}	\textsc{du}.\textsc{s}-go.up.\textsc{pp}	Ayaui-\textsc{loc}	\textsc{du}.\textsc{s}-go.up.\textsc{pp}	sidepath	rain	very.big	we.\textsc{du}.\textsc{excl}	rain	together	\textsc{du}.\textsc{s}-go.\textsc{pp}=\textsc{emph}\\
    \glt `We climbed the mountain Ayaui. At Mount Ayaui, we climbed the flank path. The rain was heavy. We walked through the rain.'
\z

\ea\label{ex:nrexyopumo}
    Koyo pu monoyo piyei. Boyopuno pu poyana.\\
    \gll koyo	pu	mono-yo	piye-i	boyopuno	pu	poyana\\
    we.\textsc{du}.\textsc{excl}	rain	path-\textsc{loc}	take-\textsc{du}.\textsc{a}	later	rain	stop.\textsc{pp}\\
    \glt `We became very wet on the way. Later the rain stopped.'
\z

\ea\label{ex:nrexilorou}
    Koyo iloro Uleiyo paeau. Koyo pueliyekoi. Koyo ono bayana Makoa reyanai. Koyo lewoinep monoyo.\\
    \gll koyo	i-lo=ro	Ulei-yo	paeau	koyo	pueliye-ko-i	koyo	ono	bayana	Makoa	reyana-i	koyo	lewo-ini-p	mono-yo\\
    we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-go.\textsc{pp}=\textsc{emph}	Ulei-\textsc{loc}	arrive	we.\textsc{du}.\textsc{excl}	leave.behind-\textsc{rts}-\textsc{du}.\textsc{a}	we.\textsc{du}.\textsc{excl}	man	different	Makoa	meet.\textsc{sg}.\textsc{o}.\textsc{pp}-\textsc{du}.\textsc{a}	we.\textsc{du}.\textsc{excl}	wait.for-\textsc{nsg}.\textsc{or}-\textsc{pc}	road-\textsc{loc}\\
    \glt `We went to Ulei and arrived there. (Soon) we left it behind and met another man, Makoa. He was waiting for us at the road.'
\z

\ea\label{ex:nrexripela}
    Boyo uke tripela moloro Pasiyo paeau. Riyopuno uke basno molo Waisan Campyo. Uke mapap.\\
    \gll boyo	uke	tripela	molo=ro	Pasi-yo	paeau	riyopuno	uke	bas-no	molo	Waisan\_Camp-yo	uke	mape-p\\
    later	we.\textsc{excl}	three(TP)	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	Pasi-\textsc{loc}	arrive	then	we.\textsc{excl}	bus-\textsc{ins}	go.\textsc{pl}.\textsc{pp}	Waisan.Camp-\textsc{loc}	we.\textsc{excl}	stay.\textsc{pl}-\textsc{pc}\\
    \glt `Later the three of us continued to Pasi and arrived there. Then we went on by bus to Waisan Camp. (There) we stayed.'
\z
