\ea\label{ex:nrexnodupu}
    Ono du puap. Puni lo, bi lipelip. Loro, du puapno.\\
    \gll ono	du	pue-p	puni	lo	bi	lipeli-p	lo=ro	du	pue-p-no\\
    man	bush	roam-\textsc{pc}	night	go.\textsc{pp}	pig	seek-\textsc{pc}	go.\textsc{pp}=\textsc{emph}	bush	roam-\textsc{pc}-\textsc{co}\\
    \glt `A man was roaming the bush. He went at night. He was looking for pigs. He set off. He was roaming the bush.'
\z

\ea\label{ex:nrexkosupi}
    Umul nekpamu, ``ko sû piye.'' An seku uroyo, sû piyo. Sû bayana. Tos duki ari.\\
    \gll umul\_nekpamu	ko	sû	piye	an	seku	uro-yo	sû	piyo	sû	bayana	tos	duki	ari\\
    ponder.\textsc{pp}	I	light	take	hand	fall.\textsc{pp}	netbag-\textsc{loc}	light	take.\textsc{pp}	light	different	torch(\textsc{tp})	true	no\\
    \glt `He thought: ``I'll take a light.'' His hand grabbed into his netbag, and he got a light. A different light, not a true torch light.'
\z

\ea\label{ex:nrexsusela}
    Sû selayap. Sû pîpîlip. Bayana. Pîli pîli pîli. Pîpîlip.\\
    \gll sû	selaye-p	sû	pîpîli-p	bayana	pîli	pîli	pîli	pîpîli-p\\
    light	lay.spreadingly-\textsc{pc}	light	flicker-\textsc{pc}	different	flicker	flicker	flicker	flicker-\textsc{pc}\\
    \glt `The light was beaming. The light was flickering. A different (type of light). Flicker, flicker, flicker, it was flickering.'
\z

\ea\label{ex:nrexpialdo}
    Pial dobro riye. Tststs tststs. Bo ppulaena.\\
    \gll pial	dob=ro	riye	tststs	tststs	bo	ppulae-na\\
    snake	eye=\textsc{emph}	see.\textsc{o}.\textsc{inanim}	tststs	tststs	sound	bad-\textsc{adv}\\
    \glt `A snake is focusing (the light): ``Tststs tststs.'' A bad sound.'
\z

\ea\label{ex:nrexkoanap}
    Ko ana poipi? Riyopuno tos duki piyo. Selayap. ``Eh pial, oke pial! Pial bekulu. Pial yala ko dupuesi.''\\
    \gll ko	ana	pi-ipi	riyopuno	tos	duki	piyo	selaye-p	eh	pial	o-ke	pial	pial	bekulu	pial	yala	ko	d-puesi\\
    I	who	do-1\textsc{sg}.\textsc{or}	eventually	torch	true	take.\textsc{pp}	lay.spreadingly-\textsc{pc}	eh	snake	\textsc{prox}-\textsc{aph}	snake	snake	huge	snake	\textsc{mod}	I	\textsc{lkh}-bite\\
    \glt `{``}Who does (this) to me?'' (thinks the man). Then he took a true torch. The light was spreading: ``Eh, a snake, this is a snake. A huge snake. The snake will certainly bite me.'''
\z

\ea\label{ex:nrexonodki}
    Ono dikike. Epemna lo. Pe pakono neppi royepana. Uro royepanouko. An solo kike. Tos solo wel, yilauyo puni.\\
    \gll ono	d-kike	epemna	lo	pe	pako-no	neppi	royepana	uro	royepane-ou-ko	an	solo	kike	tos	solo	wel	yilau-yo	puni\\
    man	\textsc{lkh}-run	fast	go.\textsc{pp}	arrow	bow-\textsc{ins}	bush.knife	put.thither.\textsc{pp}	netbag	put.thither-\textsc{frus}-\textsc{rts}	hand	only	run	torch(\textsc{tp})	only	carry.\textsc{pp}	village-\textsc{loc}	night\\
    \glt `The man must (have) run. He went fast. Bow and arrows and the bush knife he put there. The netbag he (also) left there, frustrated. He runs empty-handed. He took only the torch light, to the village, at night.'
\z

\ea\label{ex:nrexenamos}
    Yena mosaupoen: ``Pial ko puesiou. Koro pulo. Ko mepu po. Em le. Ko pe pakono uro neppi rapiyeke.''\\
    \gll yena	mosaupo-en	pial	ko	puesi-ou	ko=ro	pulo	ko	mepu\_po	em	le	ko	pe	pako-no	uro	neppi	rapiye-ke\\
    people	teach-\textsc{nsg}.\textsc{or}.\textsc{pp}	snake	I	bite-\textsc{frus}	I=\textsc{emph}	come.\textsc{pp}	I	be.afraid.\textsc{pp}	tomorrow	go	I	arrow	bow-\textsc{ins}	netbag	bush.knife	fetch-\textsc{ingr}\\
    \glt `He told the people: ``The snake didn't manage to bite me. I came (back here). I was afraid. Tomorrow I'll go. I go to get bow and arrows, the bush knife and the netbag.'''
\z

\ea\label{ex:nrexunipin}
    Punipino lo. Rapiyeko, yilauyo pulo. Barapiyeko.\\
    \gll punipino	lo	rapiye-ko	yilau-yo	pulo	ba-rapiye-ko\\
    morning	go.\textsc{pp}	fetch-\textsc{rts}	village-\textsc{loc}	come.\textsc{pp}	\textsc{fac}-fetch-\textsc{fac}\\
    \glt `In the morning he left (for the bush). He fetched (the things) and came (back) to the village. He has gotten (all the equipment).'
\z

\ea\label{ex:nrexlauyok}
    Yilauyo klapno nakap. Bo muelien: ``Sû bayana kipiyeyem! Ppulae! Sû roke imiyu pial pi. Pial yala ki riyeno sû mi dop kepso pîpîli.''\\
    \gll yilau-yo	k-le-p-no	nake-p	bo	mueli-en	sû	bayana	k-piye-yem	ppulae	sû	ro-ke	imiyu	pial	pi	pial	yala	ki	riye-no	sû	mi	dop	kep-so	pîpîli\\
    village-\textsc{loc}	\textsc{sub}-go-\textsc{pc}-\textsc{co}	stay-\textsc{pc}	word	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	light	different	\textsc{proh}-take-\textsc{proh}.\textsc{pl}	bad	light	\textsc{prox}.\textsc{emph}-\textsc{aph}	sorcerer	snake	make	snake	now	\textsc{aph}	see.\textsc{o}.\textsc{inanim}-\textsc{co}	light	again	skin	3\textsc{sg}.\textsc{poss}-\textsc{sim}	flicker\\
    \glt `Having come to the village, he was staying there. He said to the people: ``Don't take the different light! It's bad! This light makes a snake hostile. While the snake looks at it, the light flickers like a snake's skin.'''\footnote{Perhaps the fancy torch light flickers like a rainbow snake's skin.}
\z

\ea\label{ex:nrexalmiya}
    ``Pial mi yala nuko nowepi. Sû ereke layepanepiyap! Puni nuko rino mipi. Tos makiro oso poli.''\\
    \gll pial	mi	yala	nuko	ni-wepi	sû	ere-ke	layepane-pi-yep	puni	nuko	ri-no	mipi	tos	maki=ro	oso	poli\\
     snake	again	\textsc{mod}	we.\textsc{incl}	eat-\textsc{quant}.\textsc{o}	light	\textsc{prox}-\textsc{aph}	abandon-\textsc{lv}-\textsc{pl}.\textsc{imp}	night	we.\textsc{incl}	wood-\textsc{ins}	come.hither.\textsc{pl}	torch(\textsc{tp})	good=\textsc{emph}	more	be.there\\
    \glt `{``}Snakes will bite us often. Abandon this light! At night we come with a wooden torch, (which) a good torch. It is (much) better.'''\footnote{The ``wooden torch'' is a glowing stick whose light isn't glaring.}
\z