\ea\label{ex:nrexkoevam}
    Ko Eva muelno: ``De inalap! Margaret muelkenap: Ko de same. Epe kopi de muli. Awe!''\\
    \gll ko	Eva	mueli-no	de	inale-p	Margaret	mueli-ke-ne-p	ko	de	sa-me	epe	ko-pi	de	muli	awe\\
    I	Eva	talk.to-3\textsc{sg}.\textsc{or}	you	hurry-\textsc{imp}	Margaret	talk.to-\textsc{ingr}-3\textsc{sg}.\textsc{or}-\textsc{imp}	I	you	ask-2\textsc{sg}.\textsc{or}	mother	1\textsc{sg}-\textsc{poss}	you	want	come\\
    \glt `I said to Eva: ``Hurry up, go and tell Margaret: I am asking you, my mother wants you, come!”'
\z

\ea\label{ex:nrexbubude}
    ``Bubu de puanap! De inalap! Claudia de muli.''\\
    \gll bubu	de	puane-p	de	inale-p	Claudia	de	muli\\
    grandmother	you	rise-\textsc{imp}	you	hurry-\textsc{imp}	Claudia	you	want\\
    \glt `{``}Grandmother, get up! Hurry up! Claudia wants you!”'
\z

\ea\label{ex:nrexiyopun1}
    Riyopuno ko pulo. Ko de reyo.\\
    \gll riyopuno	ko	pulo	ko	de	reyo\\
    then	I	come.\textsc{pp}	I	you	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}\\
    \glt `Then I came and saw you.' (\textit{I} refers to Margaret, and now the story is told from her perspective.) 
\z

\newpage
\ea\label{ex:nrexkosial}
    ``Margaret ko sialipo. Ko le haus sikyo Vanimoyo. Nuko ile. De mini, de muli?''\\
    \gll Margaret	ko	siali-po	ko	le	haus\_sik-yo	Vanimo-yo	nuko	i-le	de	mini	de	muli\\
    Margaret	I	rash-\textsc{lv}.\textsc{pp}	I	go	hospital-\textsc{loc}	Vanimo-\textsc{loc}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	you	come.hither	you	want\\
    \glt `{``}Margaret, I got a rash. I'll go to the hospital, to Vanimo. Let's go. Will you come, do you want to?”'
\z

\ea\label{ex:nrexnekkou}
    Ko umul nek. Ko umul nekipe. ``Ko kaikai bapiyeko. Yena due kopi sini. Luo kiniyo ko bapakok. Kaikai kopi ko David ponamo. Upunaro. Apoliwe. Nuko ile, ko mini.''\\
    \gll ko	umul\_nek	ko	umul\_neki-ipe	ko	kaikai	ba-piye-ko	yena	due	ko-pi	sini	luo	kiniyo	ko	ba-pake-ko	kaikai	ko-pi	ko	David	ponamo	upuna=ro	a-poli-we	nuko	i-le	ko	mini\\
    I	think.\textsc{pp}	I	think-\textsc{ant}	I	food(\textsc{tp})	\textsc{fac}-take-\textsc{fac}	people	sago	1\textsc{sg}-\textsc{poss}	pound.sago.pith	money	all	I	\textsc{fac}-throw-\textsc{fac}	food(\textsc{tp})	1\textsc{sg}-\textsc{poss}	I	David	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	alright=\textsc{emph}	\textsc{imp}3-be.there-\textsc{ter}	we.\textsc{incl}	\textsc{du}.\textsc{s}-go	I	come.hither	\\
    \glt `I thought (about it). First I (wanted to) think (about it): ``I have bought food. The people will pound sago pith for me. All the money I have thrown away. My food I gave to David. (But) alright, be it (as it is). Let's go. I come (with you).”'
\z

\ea\label{ex:nrexkoamsa}
    ``Ko am ba saupi. Ko am ba malo. Nuko auno mole, nuko opono mole?'' Riyopuno nuko au lewone. Au amapuleipe.\\
    \gll ko	am	ba	saupi	ko	am	ba	malo	nuko	au-no	mole	nuko	opo-no	mole	riyopuno	nuko	au	lewo-ne	au	am-a-pule-ipe\\
    I	yet	\textsc{neg}.\textsc{emph}	know	I	yet	\textsc{neg}.\textsc{emph}	hear.\textsc{pp}	we.\textsc{incl}	plane-\textsc{ins}	go.\textsc{pl}	we.\textsc{incl}	car-\textsc{ins}	go.\textsc{pl}	then	we.\textsc{incl}	plane	wait.for-3\textsc{sg}.\textsc{or}	plane	\textsc{grad}-\textsc{imp}3-come-\textsc{ant}\\
    \glt `(Claudia said:) ``I don't know yet, I didn't hear yet. Will we go by plane, will we  go by car?'' Eventually we wait for the plane, it has yet to come.'
\z

\newpage
\ea\label{ex:nrexpuleau}
    Au pule. Au seku ples balusyo. Au liki kepyo seku. Le royepanepo balus bîyo.\\
    \gll au	pule	au	seku	ples	balus-yo	au	liki	kep-yo	seku	le	royepane-po	balus	bîyo\\
    plane	come	plane	fall.\textsc{pp}	place(\textsc{tp})	plane(\textsc{tp})-\textsc{loc}	plane	made.ready.place	3\textsc{sg}.\textsc{poss}-\textsc{loc}	fall.\textsc{pp}	belongings	put.thither-\textsc{lv}.\textsc{pp}	plane	inside\\
    \glt `The plane is coming. The plane landed on the airstrip. The plane landed on its landing place. We put (our) luggage inside the plane.'
\z

\ea\label{ex:nrexkevina}
    Riyopuno de Kevin sano: ``De wepulo sia fopela?'' Kevin muelno: ``Ou, ko wepulo, ou. Uke ronpua ronpua.'' Nuko lupuanepo, bili musi. Ensin kep po. Au puana, wariye, au lo nuko roise. Vanimoyo seku.\\
    \gll riyopuno	de	Kevin	sa-no	de	wepulo	sia	fopela	Kevin	mueli-no	ou	ko	wepulo	ou	uke	ronpua\_ronpua	nuko	lupuane-po	bili	musi	ensin	kep	po	au	puana	wariye	au	lo	nuko	roise	Vanimo-yo	seku\\
    then	you	Kevin	ask-3\textsc{sg}.\textsc{or}.\textsc{pp}	you	bring.\textsc{pp}	seat	four(\textsc{tp})	Kevin	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	yes	I	bring.\textsc{pp}	yes	we.\textsc{excl}	four	we.\textsc{incl}	enter-\textsc{lv}.\textsc{pp}	door	close	engine	3\textsc{sg}.\textsc{poss}	do.\textsc{pp}	plane	rise.\textsc{pp}	fly	plane	go.\textsc{pp}	we.\textsc{incl}	with	Vanimo-\textsc{loc}	fall.\textsc{pp}\\
    \glt `Then you asked Kevin: ``Did you bring four seats?” Kevin replied: ``Yes, I brought them, yes. We are four.” We entered (the plane), and the door closes. Its engine started, the plane rose, it is flying. The plane went with us and landed in Vanimo.'
\z

\ea\label{ex:nrexksekip}
    Kesekipno nuko haus sikyo molo. De dokta reyo. Dop depi riye. De muelme: ``De eol ikoina pi. Lil depi piye.''\\
    \gll k-seki-p-no	nuko	haus\_sik-yo	molo	de	dokta	reyo	dop	de-pi	riye	de	mueli-me	de	eol	ikoi-na	pi	lil	de-pi	piye\\
    \textsc{sub}-fall-\textsc{pc}-\textsc{co}	we.\textsc{incl}	hospital-\textsc{loc}	go.\textsc{pl}.\textsc{pp}	you	doctor	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	skin	2\textsc{sg}-\textsc{poss}	see.\textsc{o}.\textsc{inanim}	you	talk.to-2\textsc{sg}.\textsc{or}	you	sweat	big-\textsc{adv}	\textsc{lv}	blood	2\textsc{sg}-\textsc{poss}	take\\
    \glt `After landing we went to the hospital. You saw a doctor. He examines your skin and says to you: ``You sweat badly. I will make a blood test.”'
\z

\ea\label{ex:nrexmarpon}
    Riyopuno marasin de poname. Due ronpua ronpua. Dokta de muelme: ``Marasin so piye. Mi fraideno pule. Ko de reye. Aepu depi riye baslaupoko ari? Pul ar pi, slaupi. Roke upuna. Sapos pul pi, ar slaupi. Roke de marasin bayana piye.''\\
    \gll riyopuno	marasin	de	poname	due	ronpua\_ronpua	dokta	de	mueli-me	marasin	so	piye	mi	fraide-no	pule	ko	de	reye	aepu	de-pi	riye	ba-slau-pi-ko	ari	pul	ar	pi	slau-pi	ro-ke	upuna	sapos	pul	pi	ar	slau-pi	ro-ke	de	marasin	bayana	piye\\
    then	medicine	you	give.2\textsc{sg}.\textsc{or}	night	four	doctor	you	talk.to-2\textsc{sg}.\textsc{or}	medicine	so	take	again	Friday-\textsc{ins}	come	I	you	see.\textsc{o}.\textsc{anim}.\textsc{sg}	sore	2\textsc{sg}-\textsc{poss}	see.\textsc{o}.\textsc{inanim}	\textsc{fac}-dry-\textsc{lv}-\textsc{fac}	no	liquid	\textsc{neg}	\textsc{lv}	dry-\textsc{lv}	\textsc{prox}.\textsc{emph}-\textsc{aph}	alright	if(\textsc{tp})	liquid	\textsc{lv}	\textsc{neg}	dry-\textsc{lv}	\textsc{prox}.\textsc{emph}-\textsc{aph}	you	medicine	different	take\\
    \glt `Then he gives you some medicine. (For) four days. The doctor says to you: ``You take the medicine like this. Friday you come again. I (need to) see you. (I need) to see your sores. Are they dry or not? (If) there is no liquid and they are dry, (in) that (case) alright. If there is liquid and they are not dry, then you take a different medicine.”'
\z

\ea\label{ex:nrexyenkos}
    Yena ko saipop: ``Depe arka pulo? De mono pulo?'' ``Ko Claudiayo ipulo auno, Claudia uki kep roise ruri kep. Haus sikyo ipulo dokta reye. An kep aepu pi, bou kep aepu pi.''\\
    \gll yena	ko	sa-ipi-p	de=pe	arka	pulo	de	mono	pulo	ko	Claudia-yo	i-pulo	au-no	Claudia	uki	kep	roise	ruri	kep	haus\_sik-yo	i-pulo	dokta	reye	an	kep	aepu	pi	bou	kep	aepu	pi\\
    people	I	ask-1\textsc{sg}.\textsc{or}-\textsc{pc}	you=\textsc{q}	where	come.\textsc{pp}	you	when	come.\textsc{pp}	I	Claudia-\textsc{loc}	\textsc{du}.\textsc{s}-come.\textsc{pp}	plane-\textsc{ins}	Claudia	husband	3\textsc{sg}.\textsc{poss}	with	child	3\textsc{sg}.\textsc{poss}	hospital-\textsc{loc}	\textsc{du}.\textsc{s}-come.\textsc{pp}	doctor	see.\textsc{o}.\textsc{anim}.\textsc{sg}	arm	3\textsc{sg}.\textsc{poss}	sore	\textsc{lv}	leg	3\textsc{sg}.\textsc{poss}	sore	\textsc{lv}\\
    \glt `The people were asking me: ``Where did you come from? When did you come?” ``Claudia and me came by plane, Claudia together with her husband and child. We came to the hospital to see a doctor. There are sores on her arms and legs.”'
\z

\ea\label{ex:nrexromape}
    Mi uke romape, uke tripela. Eva Loteyo nake. Uke tripela solo. Ko bo komiye uke haus sikyo mape.\\
    \gll mi	uke	ro=mape	uke	tripela	Eva	Lote-yo	nake	uke	tripela	solo	ko	bo	komiye	uke	haus\_sik-yo	mape\\
    then	we.\textsc{excl}	\textsc{emph}=stay.\textsc{pl}	we.\textsc{excl}	three(\textsc{tp})	Eva	Lote-\textsc{loc}	stay	we.\textsc{excl}	three(\textsc{tp})	only	I	word	hide	we.\textsc{excl}	hospital-\textsc{loc}	stay.\textsc{pl}\\
    \glt `Well, we stay (on), the three of us. Eva is staying in Lote. Only we three (are in town). I hide the news (that) we are in the hospital (for outpatient treatment).'
\z

\ea\label{ex:nrexkosoar}
    So ko ar mueli uke yip riyo mape. Ari ari, ko bo komiyemaeu. Yip nukopi ari. Yip bisoppi. Yip nuko elep, yena upuna pulupi nuko rili.\\
    \gll so	ko	ar	mueli	uke	yip	ri-yo	mape	ari	ari	ko	bo	komiye\_maeu	yip	nuko-pi	ari	yip	bisop-pi	yip	nuko	elep	yena	upuna	pulupi	nuko	rili\\
    so	I	\textsc{neg}	talk.to	we.\textsc{excl}	house	\textsc{dist}-\textsc{loc}	stay.\textsc{pl}	no	no	I	word	hide\_belong.to	house	we.\textsc{incl}-\textsc{poss}	no	house	bishop-\textsc{poss}	house	we.\textsc{incl}	2\textsc{sg}.\textsc{poss}.\textsc{emph}	people	alright	come.\textsc{pl}	we.\textsc{incl}	see.\textsc{o}.\textsc{anim}.\textsc{pl}\\
    \glt `I don't talk (about the fact) that we are staying in that house. No, no, I hide the news as mine. The house doesn't belong to us. It is the bishop's house. Were it our house or yours, (then) alright, the people may come to visit us.'
\z

\ea\label{ex:nrexinemon}
    Yena ko saipop: ``Ine mono mole?'' ``Ko ar saupi uke sarereno molam? Dokta am ar mueli. Am bo malomaye. Aepu kep am ar sui. Yala asa muelnam bo. Yala uke opono molam yala uke auno molam.''\\
    \gll yena	ko	sa-ipi-p	ine	mono	mole	ko	ar	saupi	uke	sarere-no	mole-m	dokta	am	ar	mueli	am	bo	malomaye	aepu	kep	am	ar	sui	yala	asa	mueli-ne-m	bo	yala	uke	opo-no	mole-m	yala	uke	au-no	mole-m\\
    people	I	ask-1\textsc{sg}.\textsc{or}-\textsc{pc}	you.\textsc{pl}	when	go.\textsc{pl}	I	\textsc{neg}	know	we.\textsc{excl}	Saturday-\textsc{ins}	go.\textsc{pl}-\textsc{pos}	doctor	yet	\textsc{neg}	talk.to	yet	word	listen.carefully	sore	3\textsc{sg}.\textsc{poss}	yet	\textsc{neg}	die	\textsc{mod}	how	talk.to-3\textsc{sg}.\textsc{or}-\textsc{pos}	word	\textsc{mod}	we.\textsc{excl}	car-\textsc{ins}	go.\textsc{pl}-\textsc{pos}	\textsc{mod}	we.\textsc{excl}	plane-\textsc{ins}	go.\textsc{pl}-\textsc{pos}\\
    \glt `The people were asking me: ``When do you go (back)?” (I replied:) ``I don't know, maybe we'll leave on Saturday. The doctor doesn't say it yet. We (have) still to listen to his words. Her sores aren't healed yet. He cannot say anything to her. We might go by car, we might go by plane.'''
\z

\ea\label{ex:nrexdeeska}
    Yena mi ko saipop: ``De eska pulo, de eska pulo?''\\
    \gll yena	mi	ko	sa-ipi-p	de	eska	pulo	de	eska	pulo\\
    people	again	I	ask-1\textsc{sg}.\textsc{or}-\textsc{pc}	you	when	come.\textsc{pp}	you	when	come.\textsc{pp}\\
    \glt `The people kept asking me: ``When did you come? When did you come?”'
\z
