\ea\label{ex:t3exaimili}
    Ai Milipi yilauyo pulo. Opono pulo, lakiyoko, woko Vanimoyo.\\
    \gll ai	Mili-pi	yilau-yo	pulo	opo-no	pulo	lakiye-ko	woko	Vanimo-yo\\
    father	Mili-\textsc{poss}	village-\textsc{loc}	come.\textsc{pp}	car-\textsc{ins}	come.\textsc{pp}	fetch-\textsc{rts}	accompany.\textsc{pp}	Vanimo-\textsc{loc}\\
    \glt `Mili's father came to the village. He came by car, picked (Mili) up, and brought her to Vanimo.'
\z

\ea\label{ex:t3exkomuel}
    Ko muelpup: ``Epee, de ruri kopi wokap haus sikyo! Aepu ikoina pi. Aepu kana po.''\\
    \gll ko	mueli-ipi-p	epe-e	de	ruri	ko-pi	woke-p	haus\_sik-yo	aepu	ikoi-na	pi	aepu	kana	po\\
    I	talk.to-1\textsc{sg}.\textsc{or}-\textsc{pc}	mother-\textsc{voc}	you	child	1\textsc{sg}-\textsc{poss}	accompany-\textsc{imp}	hospital-\textsc{loc}	ulcer	big-\textsc{adv}	do	ulcer	fast	do.\textsc{pp}\\
    \glt `He said to me: ``Mother, accompany my child to the hospital! The ulcer is big, it grew very fast.'''
\z

\ea\label{ex:t3exdoktam}
    Dokta muelien: ``Deyo yala haus sikyo wonake. Aepu bueyo pusiyekap!''\\
    \gll dokta	mueli-en	deyo	yala	haus\_sik-yo	wo-nake	aepu	bue-yo	pusiye-ke-p\\
    doctor	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you.\textsc{du}	now	hospital-\textsc{loc}	\textsc{accom}-sit	ulcer	sea-\textsc{loc}	wash-\textsc{ingr}-\textsc{imp}\\
    \glt `The doctor said to us: ``The two of you stay together in the hospital now. You (two) go and wash the ulcer in the sea.'''\footnote{Salt water has disinfectant effects on wounds.}
\z

\ea\label{ex:t3exmiliwe}
    Ko Mili wepulo. Marasin pono aepuyo, lolono, dokta muelo: ``Fonde operesenyo le, de ni kepem! De yala lam.''\\
    \gll ko	Mili	wepulo	marasin	pi-no	aepu-yo	lole-no	dokta	mueli-no	Fonde	operesen-yo	le	de	ni	k-pi-m	de	yala	le-m\\
    I	Mili	bring.\textsc{pp}	medicine	do-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer-\textsc{loc}	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	doctor	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	Thursday	surgery-\textsc{loc}	go	you	eat	\textsc{proh}-do-\textsc{proh}	you	\textsc{mod}	go-\textsc{pos}\\
    \glt `I brought Mili (back). She got medicine on the ulcer and it was dressed. The doctor said to her: ``Thursday you undergo the surgery. You must not eat (anything), you just come.'''
\z

\ea\label{ex:t3exfondep}
    Fonde punipino tenklokno lo. Aepu dûkû ikoiele. Dokta kiniyo riyemayo. Ppulae, sut ponamo.\\
    \gll Fonde	punipino	ten\_klok-no	lo	aepu	dûkû	ikoiele	dokta	kiniyo	riyemayo	ppulae	sut	ponamo\\
    Thursday	morning	ten.a.clock-\textsc{ins}	go.\textsc{pp}	ulcer	smell	very.big	doctor	all	look.carefully.\textsc{o}.\textsc{inanim}.\textsc{pp}	bad	injection	give.3\textsc{sg}.\textsc{or}.\textsc{pp}\\
    \glt `Thursday morning at 10 o'clock she went (to the treatment room). The ulcer smelled strongly. The doctor looked carefully at everything. (It looked) bad. He gave her an injection.'
\z

\ea\label{ex:t3exvenepu}
    Yeniyo wenepu. Nanayo puenpo, sesiyepiyo. Aepu ppulae puppuli polip. Marasin pono, aepuyo lolono. Mili wemipu yeniyo, yeni koyopi.\\
    \gll yeni-yo	wenepu	nana-no	puenpo	sesiyepiyo	aepu	ppulae	puppuli	poli-p	marasin	pi-no	aepu-yo	lole-no	Mili	wemipu	yeni-yo	yeni	koyo-pi\\
    bed-\textsc{loc}	carry.thither.\textsc{pl}.\textsc{a}.\textsc{pp}	small.knife-\textsc{ins}	cut.meat.\textsc{pp}	take.away.\textsc{pp}	ulcer	bad	fat	be.there-\textsc{pc}	medicine	do-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer-\textsc{loc}	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	Mili	bring.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	bed-\textsc{loc}	bed	we.\textsc{du}.\textsc{excl}-\textsc{poss}\\
    \glt `(The nurses) carried her to the operation table. (The doctor) cut with a lancet. He cut  (the ulcer) out. The ulcer is bad. There were (depositions of) fat. He put medicine on the ulcer and dressed it. They brought Mili to her bed, to our bed.'\footnote{Margaret and Mili share a bed in the hospital.}
\z


\ea\label{ex:t3exnuveze}
    Yeniyo nuweikûpu. Yeniyo nuip aua klokni. Riyopuno puana, ko muelpup: ``Bubu, ko ni muli.''  Nes: ``Kenem, umul maki amapoipe!''\\
    \gll yeni-yo	nuweyekûpu	yeni-yo	nui-p	aua	klokni	riyopuno	puana	ko	mueli-ipi-p	bubu	ko	ni	muli	nes	k-ni-m	umul	maki	am-a-pi-ipe\\
    bed-\textsc{loc}	lay.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	bed-\textsc{loc}	sleep-\textsc{pc}	hour	one	then	wake.up.\textsc{pp}	I	talk.to-1\textsc{sg}.\textsc{or}-\textsc{pc}	granny(\textsc{tp})	I	eat	want	nurse	\textsc{proh}-eat-\textsc{proh}	mind	good	\textsc{grad}-\textsc{imp}3-\textsc{lv}-\textsc{ant}\\
    \glt `They laid her down on the bed. She was sleeping on the bed for one hour. Then she woke up and said to me: ``Granny, I want to eat.'' The nurse (said): ``She must not eat! First (her) mind should be clear.'''
\z

\ea\label{ex:t3exbubuqo}
    ``Bubu, ko el sui, ni muli!''  Ko ni ponamo, ya yûr dû yûr su roise. No. Ya kesiyo.\\
    \gll bubu	ko	el\_sui	ni	muli	ko	ni	ponamo	ya	yûr	dû	yûr	su	roise	no	ya	kesiyo\\
    granny(\textsc{tp})	I	be.hungry	eat	want	I	eat	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	sago	chicken	meat	chicken	egg	with	eat.\textsc{pp}	sago	use.up.\textsc{pp}\\
    \glt `{``}Granny, I am hungry and want to eat.'' I gave her to eat, sago together with chicken meat and eggs. She ate. She finished the sago.'
\z

\ea\label{ex:t3exaepuik}
    Aepu ikil am poli, puppuli polip. Tunde seken operesen lo. Nanano puenpo, sesiyepiyo, aepu suelpiyo. Lolono, wemipu yeniyo.\\
    \gll aepu	ikil	am	poli	puppuli	poli-p	Tunde	seken	operesen	lo	nana-no	puenpo	sesiyepiyo	aepu	suelipiyo	lole-no	wemipu	yeni-yo\\
    ulcer	dirt	still	be.there	fat	be.there-\textsc{pc}	Tuesday	second(\textsc{tp})	surgery	go.\textsc{pp}	small.knife-\textsc{ins}	cut.meat.\textsc{pp}	take.away.\textsc{pp}	ulcer	cut.away.\textsc{pp}	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	bring.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	bed-\textsc{loc}\\
    \glt `The ulcer is still dirty. There were (still depositions of) fat. On Tuesday she underwent the second surgery. (The doctor) cut with a small knife, cut along the ulcer, and cut it out. He dressed her wound. (Then) they brought her here to her bed.'
\z

\ea\label{ex:t3exwemipu}
    Wemipu yeniyo nuweikûpu. Aua klokni nuip, bo mulanapop: ``Epo ko ni, pili yasiyap! Ekuyo laye.''\\
    \gll wemipu	yeni-yo	nuweyekûpu	aua	klokni	nui-p	bo	mulane-pi-p	epo	ko	ni	pili	yasiye-p	eku-yo	laye\\
    bring.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	bed-\textsc{loc}	lay.down.hither.\textsc{pl}.\textsc{a}.\textsc{pp}	hour	one	sleep-\textsc{pc}	speech	babble-\textsc{lv}-\textsc{pc}	faeces	I	eat	cloth	place-\textsc{imp}	behind-\textsc{loc}	lay\\
    \glt `They brought her here and laid her down on the bed. She was sleeping for one hour and was talking confused things: ``I eat faeces, spread a cloth! Put it on (my) behind.'''
\z

\ea\label{ex:t3expuanak}
    Puana: ``Ko aiyo le.''  Mi due nu, due nuro. Mipuana: ``Bubu, ko el sui. Ko ni pi.''  Bo mulanapi layepana, kuru. Puaku maki pi.\\
    \gll puana	ko	ai-yo	le	mi	due\_nu	due\_nu=ro	mi-puana	bubu	ko	el\_sui	ko	ni	pi	bo	mulane-pi	layepana	kuru	puaku	maki	pi\\
    wake.up.\textsc{pp}	I	father-\textsc{loc}	go	again	sleep.\textsc{pp}	sleep.\textsc{pp}=\textsc{emph}	\textsc{iter}-wake.up.\textsc{pp}	granny	I	be.hungry	I	eat	do	speech	babble-\textsc{lv}	leave.behind.\textsc{pp}	be.finished	head	good	\textsc{lv}\\
    \glt `She woke up: ``I go to my father.'' She slept again and was sound asleep. (Then) she woke up again: ``Granny, I am hungry, I do eat.'' She quit babbling nonsense. It came to an end. Her head is fine.'
\z

\ea\label{ex:t3exnikowo}
    ``Ni ko powaip!''  Ko wal dû yûr su roise yano ponamo. Banok, bakesiyok.\\
    \gll ni	ko	powai-p	ko	wal	dû	yûr	su	roise	ya-no	ponamo	ba-ni-ko	ba-kesiye-ko\\
    eat	I	give.1\textsc{sg}.\textsc{or}-\textsc{imp}	I	fish	meat	chicken	egg	with	sago-\textsc{ins}	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	\textsc{fac}-eat-\textsc{fac}	\textsc{fac}-use.up-\textsc{fac}\\
    \glt `{``}Give me to eat!'' I gave her fish and eggs with sago. She has eaten it up. She has finished it.'
\z

\ea\label{ex:t3exukenak}
    Ukenakap. Yena kiniyo mapap wot operesenpiyo. Aepu solo loloulipop wîs dupua. Aepu milolap, milolap, loloulipop.\\
    \gll uke-nake-p	yena	kiniyo	mape-p	wot	operesen-pi-yo	aepu	solo	lole-uli-pi-p	wîs	dupua	aepu	mi-lole-p	mi-lole-p	lole-uli-pi-p\\
    jointly-sit-\textsc{pc}	people	many	sit.\textsc{pl}-\textsc{pc}	ward	surgery-\textsc{poss}-\textsc{loc}	ulcer	only	tie-\textsc{prog}-\textsc{lv}-\textsc{pc}	moon	two	ulcer	\textsc{iter}-tie-\textsc{pc}	\textsc{iter}-tie-\textsc{pc}	tie-\textsc{prog}-\textsc{lv}-\textsc{pc}\\
    \glt `We were staying together. Many people were staying in the surgery ward. For two months, (the nurses) just kept dressing the ulcer. They redressed the ulcer again and again. They kept dressing it.'
\z

\ea\label{ex:t3exdoktaa}
    Dokta aepu riyeno: ``Aepu depi maki. Pepualso bapok. Aeppu pon suli.''\\
    \gll dokta	aepu	riye-no	aepu	de-pi	maki	pepual-so	ba-pi-ko	aeppu	pon	suli\\
    doctor	ulcer	see.\textsc{o}.\textsc{inanim}-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer	2\textsc{sg}-\textsc{poss}	good	sound.flesh-\textsc{sim}	\textsc{fac}-\textsc{lv}-\textsc{fac}	red	nose	close\\
    \glt `The doctor looked closely at her ulcer: ``Your ulcer is good. It has become like sound flesh. It is red, (with) straight closing edges.'''
\z

\ea\label{ex:t3exdoktax}
    Dokta muelo: ``Ko lil depi riyeipe.'' Lil aska.\\
    \gll dokta	mueli-no	ko	lil	de-pi	riye-ipe	lil	aska\\
    doctor	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	I	blood	2\textsc{sg}-\textsc{poss}	see.\textsc{o}.\textsc{inanim}-\textsc{ant}	blood	none\\
    \glt `The doctor said to her: ``Before (you go home) I'll check your blood.'' There is no blood.'\footnote{That means, there aren't enough red blood cells, and Mili needs a transfusion.}
\z

\ea\label{ex:t3exdoktaw}
    Dokta Wi ko musiyo: ``De ko upuna wulimonpi lil riye. Milipiso upuna, riye solo pi. Dedukoyo ilei.''\\
    \gll dokta	Wi	ko	musiyo	de	ko	upuna	wulimini-ipi	lil	riye	Mili-pi-so	upuna	riye	solo	pi	dedukoyo	i-le\\
    doctor	Wi	I	send.\textsc{pp}	you	I	kindly	follow.hither-1\textsc{sg}.\textsc{or}	blood	see.\textsc{o}.\textsc{inanim}	Mili-\textsc{poss}-\textsc{sim}	alright	see.\textsc{o}.\textsc{inanim}	only	\textsc{lv}	we.\textsc{du}.\textsc{incl}	\textsc{du}.\textsc{s}-go\\
    \glt `Doctor Wi sent me (for a blood sample): ``Please follow me kindly to check your blood. (In case) it is like Mili's, alright. Only to check it, let's go.'''
\z

\ea\label{ex:t3exlilriy}
    Dokta lil riyeno. Lil poro. Lil Charlespi riye: ``Lil depi upuna, Milipiso.'' Lil wepulo.\\
    \gll dokta	lil	riye-no	lil	pi=ro	lil	Charles-pi	riye	lil	de-pi	upuna	Mili-pi-so	lil	wepulo\\
    doctor	blood	see.\textsc{o}.\textsc{inanim}-3\textsc{sg}.\textsc{or}.\textsc{pp}	blood	do=\textsc{emph}	blood	Charles-\textsc{poss}	see.\textsc{o}.\textsc{inanim}	blood	2\textsc{sg}-\textsc{poss}	alright	Mili-\textsc{poss}-\textsc{sim}	blood	bring.\textsc{pp}\\
    \glt `The doctor checked the blood for (Mili's) benefit. He takes some blood. (Then) he checks (her father) Charles's blood: ``Your blood is alright, it is like Mili's.'' (Later) they brought (Charles') blood.'
\z

\ea\label{ex:t3exmilize}
    Mili yeniyo nuip, lil sutyo po. Lil sut mono na, ppae mono na.\\
    \gll Mili	yeni-yo	nui-p	lil	sut-yo	po	lil	sut	mono	na	ppae	mono	na\\
    Mili	bed-\textsc{loc}	sleep-\textsc{pc}	blood	injection-\textsc{loc}	do.\textsc{pp}	blood	injection	road	go.inside	blood.vessel	road	go.inside\\
    \glt `Mili was lying on the bed. They prepared the blood transfusion. The blood entered through the needle and flowed through the blood vessel.'
\z

\ea\label{ex:t3exainriy}
    Ain riyo koliyo nes lil riyo koliyo. Lil dop kepyo lo. So solo nakap, lil leip. Amakesiyowole. Bakesiyowoloko, aska.\\
    \gll ain	ri-yo	koliyo	nes	lil	ri-yo	koliyo	lil	dop	kep-yo	lo	so	solo	nake-p	lil	le-p	am-a-kesiye\_wole	ba-kesiye-wole-ko	aska\\
    iron	\textsc{dist}-\textsc{loc}	hang.\textsc{pp}	nurse	blood	\textsc{dist}-\textsc{loc}	hang.\textsc{pp}	blood	body	3\textsc{sg}.\textsc{poss}-\textsc{loc}	go.\textsc{pp}	like	only	sit-\textsc{pc}	blood	go-\textsc{pc}	\textsc{grad}-\textsc{imp}3-use.up\_move.further	\textsc{fac}-use.up-\textsc{cpl}-\textsc{fac}	none\\
    \glt `(There was) a metal stand. She hung it there. The nurse hung the blood bag there. The blood went into (Mili's) body. (Mili) was staying like this; the blood was flowing. It still has to be used up. It is used up (now), none (is left).'
\z

\ea\label{ex:t3exsutpuk}
    Sut pûke, an plastano penei. Lil mi yala keminem.\\
    \gll sut	pûke	an	plasta-no	penei	lil	mi	yala	k-mini-m\\
    injection	take.away	hand	bandage-\textsc{ins}	press	blood	again	now	\textsc{proh}-come.hither-\textsc{proh}\\
    \glt `(The nurse) takes away the needle and presses a bandage on the (crook of her) arm. Now the blood won't come out again.'
\z

\ea\label{ex:t3exkoyoin}
    Koyo inakap, Dokta muelien: ``De operesenyo le. Ko dop pili depi srene. Pili monina srene aepuyo pi. Aepuyo panapne.''\\
    \gll koyo	i-nake-p	dokta	mueli-en	de	operesen-yo	le	ko	dop	pili	depi	sre-ne	pili	moni-na	sre-ne	aepu-yo	pi	aepu-yo	pane-pi-ne\\
    we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-sit-\textsc{pc}	doctor	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	you	surgery-\textsc{loc}	go	I	body	skin	2\textsc{sg}-\textsc{poss}	scrape.off-3\textsc{sg}.\textsc{or}	skin	small-\textsc{adv}	scrape.off-3\textsc{sg}.\textsc{or}	ulcer-\textsc{loc}	do	ulcer-\textsc{loc}	put.thither-\textsc{lv}-3\textsc{sg}.\textsc{or}\\
    \glt `The two of us were staying on, and the doctor said to us: ``You (need to) undergo a (further) surgery. I will scrape off a piece of skin from you. Scrape off a small piece and do it on the ulcer. (I'll) put it on the ulcer for you (to heal).'''
\z

\ea\label{ex:t3exkobamu}
    ``Ko ba muli!''  ``De lap! De epemna upuna lam!''  Umul kep nek, ``ko le operesenyo.''\\
    \gll ko	ba	muli	de	le-p	de	epemna	upuna	le-m	umul	kep	nek	ko	le	operesen-yo\\
    I	\textsc{neg}.\textsc{emph}	want	you	go-\textsc{imp}	you	quickly	kindly	go-\textsc{pos}	heart	3\textsc{sg}.\textsc{poss}	erect.\textsc{pp}	I	go	surgery-\textsc{loc}\\
    \glt `(Mili replied:) ``I don't want to.'' (Doctor Wi repeated:) ``You go, please you should go quickly.'' She thought about it: ``I will go for the surgery.'''
\z

\ea\label{ex:t3exyipbiy}
    Yip bîyo lupuana, sut ponamo. Due nuip. Nes dupua, nes kini dokta kini wonino. Bo ar muel. Aepu nepeino, pusiyono. Aepu maki. Bou pili srono, aepuyo pono.\\
    \gll yip	bîyo	lupuana	sut	ponamo	due\_nui-p	nes	dupua	nes	kini	dokta	kini	woni-no	bo	ar	muel	aepu	nepei-no	pusiye-no	aepu	maki	bou	pili	sre-no	aepu-yo	pi-no\\
    house	inside	enter.thither.\textsc{pp}	injection	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	sleep-\textsc{pc}	nurse	two	nurse	one.\textsc{part}	doctor	one.\textsc{part}	call-3\textsc{sg}.\textsc{or}.\textsc{pp}	speech	\textsc{neg}	talk.\textsc{pp}	ulcer	undress-3\textsc{sg}.\textsc{or}.\textsc{pp}	wash-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer	good	thigh	skin	scrape.off-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer-\textsc{loc}	do-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
    \glt `She went into the treatment pavilion. They gave her an injection. She was sleeping. Two nurses, (no), a nurse and a doctor each called her. She did not talk. They undressed the ulcer and washed it. The ulcer is good. He scraped off a piece of her skin from the thigh and put it on her wound.'
\z

\ea\label{ex:t3exriyopu}
    Riyopuno aepu lolono, bou lolono. Nes wemon yeniyo. Nes klokni nuweiko numuelna nuro. Puana, ``Ko el sui. Ni muli.''  Ko ya namo yûr su wal roise. Kesiyei.\\
    \gll riyopuno	aepu	lole-no	bou	lole-no	nes	wemon	yeni-yo	nes	klokni	nuweye-ko	numuelna	nu=ro	puana	ko	el\_sui	ni	muli	ko	ya	namo	yûr	su	wal	roise	kesiye\\
    then	ulcer	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	thigh	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	nurse	bring.hither.\textsc{pp}	bed-\textsc{loc}	nurse	one	lay.down.for.resting-\textsc{rts}	for.a.long.time	sleep.\textsc{pp}=\textsc{emph}	wake.up.\textsc{pp}	I	be.hungry	eat	want	I	sago	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	chicken	egg	fish	with	use.up\\
    \glt `Then they dressed her ulcer and dressed her thigh. The nurse brought her to the bed. One nurse laid her down before she slept for a long time. She awoke: ``I am hungry and want to eat.'' I gave her sago with eggs and fish. She finishes (the food).'
\z

\ea\label{ex:t3exkonesm}
    Ko nes muelno: ``Aepu sipine, ikoina lolono. Ikoina kikipiyo. Minepeinap.''\\
    \gll ko	nes	mueli-no	aepu	sipi-ne	ikoi-na	lole-no	ikoi-na	kikipiyo	mi-nepei-ne-p\\
    I	nurse	talk.to-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer	hurt-3\textsc{sg}.\textsc{or}	big-\textsc{adv}	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	big-\textsc{adv}	hold.strong.\textsc{pp}	\textsc{iter}-undress-3\textsc{sg}.\textsc{or}-\textsc{imp}\\
    \glt `I said to the nurse: ``Her ulcer hurts, you dressed it tightly. It held too tightly, take off her bandage!'''
\z

\ea\label{ex:t3exneskob}
    Nes ko boi malo. Auna lolono. Upuna ar sipi. Aepu sipi kep as. Nes muelien: ``Due an baka dupua!''  Koyo due an baka dupua inakap.\\
    \gll nes	ko	bo	malo	auna	lole-no	upuna	ar	sipi	aepu	sipi	kep	as	nes	mueli-en	due	an\_baka	dupua	koyo	due	an\_baka	dupua	i-nake-p\\
    nurse	I	word	hear.\textsc{pp}	carefully	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}	alright	\textsc{neg}	hurt	ulcer	hurt	3\textsc{sg}.\textsc{poss}	none	nurse	talk.to-\textsc{nsg}.\textsc{or}.\textsc{pp}	night	five	two	we.\textsc{du}.\textsc{excl}	night	five	two	\textsc{du}.\textsc{s}-sit-\textsc{pc}\\
    \glt `The nurse listened to me. She (re)dressed her (wound) carefully. Alright, it doesn't hurt (anymore). No hurting ulcer. (Soon) the nurse said to us: ``Seven days!'' (So) we were staying on for seven days.'
\z

\ea\label{ex:t3exmilide}
    ``Mili de awe! Uke aepu depi nepei.'' Kenepeipno dob aepu riyeno: ``Aepu depi maki.''  Aepu epi pusiyeno, marasin peneino, aepuyo lolono.\\
    \gll Mili	de	awe	uke	aepu	de-pi	nepei	k-nepei-p-no	dob	aepu	riye-no	aepu	de-pi	maki	aepu	epi	pusiye-no	marasin	penei-no	aepu-yo	lole-no\\
    Mili	you	come.\textsc{imp}	we.\textsc{excl}	ulcer	2\textsc{sg}-\textsc{poss}	undress	\textsc{sub}-undress-\textsc{pc}-\textsc{co}	eye	ulcer	see.\textsc{o}.\textsc{inanim}-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer	2\textsc{sg}-\textsc{poss}	good	ulcer	side	wash-3\textsc{sg}.\textsc{or}.\textsc{pp}	medicine	press-3\textsc{sg}.\textsc{or}.\textsc{pp}	ulcer-\textsc{loc}	tie-3\textsc{sg}.\textsc{or}.\textsc{pp}\\
    \glt `{``}Mili, come, we undress your ulcer!'' Having undressed it, they looked thoroughly at her ulcer: ``Your ulcer is good.'' They washed the edges of her ulcer, pressed medicine upon the wound, and redressed it for her.'
\z

\ea\label{ex:t3exkoyoye}
    Koyo yeniyo ilo. Koyo inakap, dokta Wi pulo: ``Deyo yala Fraideno ilei.'' Pulo, muel: ``De lap, kuru. De yilauyo lap! Deyo yilauyo ilap!''\\
    \gll koyo	yeni-yo	i-lo	koyo	i-nake-p	dokta	Wi	pulo	deyo	yala	Fraide-no	i-le	pulo	muel	de	le-p	kuru	de	yilau-yo	le-p	deyo	yilau-yo	i-le-p\\
    we.\textsc{du}.\textsc{excl}	bed-\textsc{loc}	\textsc{du}.\textsc{s}-go.\textsc{pp}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-sit-\textsc{pc}	doctor	Wi	come.\textsc{pp}	you.\textsc{du}	\textsc{mod}	Friday-\textsc{ins}	\textsc{du}.\textsc{s}-go	come.\textsc{pp}	talk.to.\textsc{pp}	you	go-\textsc{imp}	be.finished	you	village-\textsc{loc}	go-\textsc{imp}	you.\textsc{du}	village-\textsc{loc}	\textsc{du}.\textsc{s}-go-\textsc{imp}\\
    \glt `We went (back) to the bed and were staying on. (Then) Doctor Wi came: ``The two of you leave on Friday.'' (Some days later) he came and said: ``You leave now, it is finished. Go to (your) village! You two return to the village!'''
\z

\largerpage[2]
\ea\label{ex:t3exkoyowa}
    Koyo Waisan Camp ilo. Koyo due dupua inakap. Ai kep ka hairimpo. Mili mike lo yilauyo.\\
    \gll koyo	Waisan\_Camp	i-lo	koyo	due	dupua	i-nake-p	ai	kep	ka	hairim-po	Mili	mike	lo	yilau-yo\\
    we.\textsc{du}.\textsc{excl}	Waisan.Camp	\textsc{du}.\textsc{s}-go.\textsc{pp}	we.\textsc{du}.\textsc{excl}	night	two	\textsc{du}.\textsc{s}-sit-\textsc{pc}	father	3\textsc{sg}.\textsc{poss}	car(\textsc{tp})	hire(\textsc{tp})-\textsc{lv}.\textsc{pp}	Mili	first	go.\textsc{pp}	village-\textsc{loc}\\
    \glt `We went to Waisan Camp and were staying there for two days. Her father hired a car, and Mili went (back) to the village first.'
\z

\ea\label{ex:t3exkoboyo}
    Ko boyo lo due bano. Ko Theresiayo opo ikoino iloi. Simon koyo ukel. Junksenyo wilikûno, koyo dorno iloi dupuni.\\
    \gll ko	boyo	lo	due\_ba-no	ko	Theresia-yo	opo	ikoi-no	i-lo	Simon	koyo	ukel	junksen-yo	wilikûno	koyo	dor-no	i-lo	dupuni\\
    I	later	go.\textsc{pp}	three.days.later-\textsc{ins}	I	Theresia-\textsc{loc}	car	big-\textsc{ins}	\textsc{du}.\textsc{s}-go.\textsc{pp}	Simon	we.\textsc{du}.\textsc{excl}	take.with.oneself.\textsc{pp}	junction-\textsc{loc}	drop.off.\textsc{pp}	we.\textsc{du}.\textsc{excl}	foot-\textsc{ins}	\textsc{du}.\textsc{s}-go.\textsc{pp}	night\\
    \glt `I went later, three days later. Theresia and I went on a big truck. Simon picked us up. At the junction he let us off, and we went by foot until night.'
\z

\ea\label{ex:t3exjerryd}
    Jerry disei kopi wonino. Yilau kepyo nakap, koyo riyo inu yip kepyo. Ako kep yûr si pewo si. Ya aska. Yûr pewono koyo inoi.\\
    \gll Jerry	disei	ko-pi	woni-no	yilau	kep-yo	nake-p	koyo	ri-yo	i-nu	yip	kep-yo	ako	kep	yûr	si	pewo	si	ya	aska	yûr	pewo-no	koyo	i-no\\
    Jerry	brother	1\textsc{sg}-\textsc{poss}	call-3\textsc{sg}.\textsc{or}.\textsc{pp}	village	3\textsc{sg}.\textsc{poss}-\textsc{loc}	sit-\textsc{pc}	we.\textsc{du}.\textsc{excl}	\textsc{dist}-\textsc{loc}	\textsc{du}.\textsc{s}-sleep.\textsc{pp}	house	3\textsc{sg}.\textsc{poss}-\textsc{loc}	wife	3\textsc{sg}.\textsc{poss}	chicken	cook.\textsc{pp}	banana	cook.\textsc{pp}	sago	none	chicken	banana-\textsc{ins}	we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{a}-eat.\textsc{pp}\\
    \glt `We called out for Jerry, my brother. We were staying in his village and slept there, in his house. His wife cooked chicken, and she cooked bananas. (There was) no sago. We ate the chicken with bananas.'
\z

\ea\label{ex:t3exukekum}
    Uke kumune due sap. Yûr bo mo. Yaep bo mo ``kukukuku'', yipp bo mo ``nananana'', yopp bo mo ``kopokopo lolololo''. Duruwa. Uke dueyo puana.\\
    \gll uke	kumune	due\_sap	yûr	bo	mu	yaep	bo	mu	kukukuku	yipp	bo	mu	nananana	yopp	bo	mu	kopokopo	lolololo	duruwa	uke	due-yo	puana\\
    we.\textsc{excl}	all.\textsc{coll}	sleep.\textsc{pl}.\textsc{pp}	rooster	sound	say.\textsc{pp}	kind.of.bird	sound	say.\textsc{pp}	kukukuku	wild.fowl	sound	say.\textsc{pp}	nananana	kind.of.bird	sound	say.\textsc{pp}	kopokopo	lolololo	dawn	we.\textsc{excl}	sleep-\textsc{loc}	rise.\textsc{pp}\\
    \glt `We all slept. (Then) the rooster crowed. Birds called ``kukukuku'', the wild fowl called ``nananana'', birds called ``kopokopo lolololo''. (It is) dawn. We rose from sleep.'
\z

\ea\label{ex:t3exsumapp}
    Sû mappo, pewo si. Ipino waeupp si biper si. Puliyo ipino layewo. Pupuol polip, pupuol nisi. Wîlyo wapo.\\
    \gll sû\_mappo	pewo	si	ipi-no	waeupp	si	biper	si	puliyo	ipi-no	layo-we	pupuol	poli-p	pupuol	nisi	wîl-yo	wapo\\
    light.fire.\textsc{pp}	banana	cook.\textsc{pp}	pot-\textsc{ins}	eel	cook.\textsc{pp}	possum	cook.\textsc{pp}	take.out.\textsc{pp}	pot-\textsc{ins}	lay.\textsc{pp}-\textsc{ter}	heat	be.there-\textsc{pc}	heat	become.cool	dish-\textsc{loc}	put.for.serving.\textsc{pp}\\
    \glt `(Jerry's wife) lit a fire and cooked bananas. In a pot she cooked eel and possum. She took it off (the fire), and she put (the cooked food) aside with the pot. The heat lasted, the heat ceases. She served it in a dish.'
\z

\ea\label{ex:t3expowath}
    Ko powa Theresia namo. Biper waeupp roise, koyo pewono ino, Theresia ruri kep roise Joanna.\\
    \gll ko	powa	Theresia	namo	biper	waeupp	roise	koyo	pewo-no	i-no	Theresia	ruri	kep	roise	Joanna\\
    I	give.1\textsc{sg}.\textsc{or}.\textsc{pp}	Theresia	give.3\textsc{sg}.\textsc{or}.\textsc{pp}	possum	eel	with	we.\textsc{du}.\textsc{excl}	banana-\textsc{ins}	\textsc{du}.\textsc{a}-eat.\textsc{pp}	Theresia	child	3\textsc{sg}.\textsc{poss}	together	Joanna\\
    \glt `She gave (some to) me and gave (some to) Theresia. We ate possum and eel with bananas, me and Theresia together with her child Joanna.'
\z

\ea\label{ex:t3exkukuno}
    Kukunopno uke molo dorno monomno. Uke moloro Pupualyo, uke pulmopip. Uke pul komopipno uke mapap pewo ilo. Molo yilauyo moloro yilauyo paeau.\\
    \gll k-konupi-p-no	uke	molo	dor-no	monomno	uke	molo=ro	pu	Pual-yo	uke	pul\_mopi-p	uke	pul\_k-mopi-p-no	uke	mape-p	pewo	ilo	molo	yilau-yo	molo=ro	yilau-yo	paeau\\
    \textsc{sub}-eat.\textsc{pl}.\textsc{o}-\textsc{pc}-\textsc{co}	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}	foot-\textsc{ins}	along.the.path	we.\textsc{excl}	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	river	Pual-\textsc{loc}	we.\textsc{excl}	wade.a.creek-\textsc{pc}	we.\textsc{excl}	liquid\_\textsc{sub}-wade.a.creek-\textsc{pc}-\textsc{co}	we.\textsc{excl}	sit.\textsc{pl}-\textsc{pc}	banana	eat.\textsc{pl}.\textsc{a}.\textsc{pp}	go.\textsc{pl}.\textsc{pp}	village-\textsc{loc}	go.\textsc{pl}.\textsc{pp}=\textsc{emph}	village-\textsc{loc}	arrive\\
    \glt `Having eaten we walked along the trail. We went to the Pual river and forded it. When we had crossed we were sitting and ate bananas. (Then) we walked on to the village. We walked to (our) village and arrived there.'
\z

\newpage
\ea\label{ex:t3exbokuku}
    Bo kuru.\\
    \gll bo	kuru\\
    story	be.finished\\
    \glt `The story is finished.'
\z
