\largerpage
\ea\label{ex:nrexrpepsu}
Ko seleyo lo. Ko sele pop. Ko ri sukap ripep. Ripep ini sukap.\\
\gll ko	sele-yo	lo	ko	sele	pi-p	ko	ri	suke-p	ri\_pep	ri\_pep	ini	suke-p\\
     I	garden-\textsc{loc}	go.\textsc{pp}	I	garden	do-\textsc{pc}	I	tree	cut-\textsc{pc}	ton(\textsc{tp})-tree	ton(\textsc{tp})-tree	branch	cut-\textsc{pc}\\
\glt `I went to the garden. I was working in the garden. I was cutting a tree, a \textit{ton}-tree. I was cutting a branch of a \textit{ton}-tree.'
\z

\ea\label{ex:nrexkodiwi}
Ko neppino suelip. Baka ko dorno piyelayowe. Baka piowemon. Ri puaku kopi papiyo. Ko duwi.\\
\gll ko	neppi-no	sueli-p	baka	ko	dor-no	piyelayo-we	baka	piowe\_mon	ri	puaku	ko-pi	papiyo	ko	d-wi\\
     I	bush.knife-\textsc{ins}	cut-\textsc{pc}	half	I	foot-\textsc{ins}	trample.\textsc{pp}-\textsc{ter}	half	jump\_come.hither.\textsc{pp}	tree	head	1\textsc{sg}-\textsc{poss}	injure.by.hitting.\textsc{pp}	I	\textsc{lkh}-turn\\
\glt `I was cutting it with the bush knife. One end (of the branch) I hold firmly with my foot. (But) the other end flipped against me. The branch hit and injured my head. I must have  turned away.'
\z

\ea\label{ex:nrexkikkop}
Riyopuno kiki kopi nowo. Soru ri ponyo poro, roke ko basuiko.\\
\gll riyopuno	kiki	ko-pi	nowo	soru	ri	pon-yo	po=ro	roke	ko	ba-sui-ko\\
     then	forehead	1\textsc{sg}-\textsc{poss}	swell.\textsc{pp}	if	tree	nose-\textsc{loc}	do.\textsc{pp}=\textsc{emph}	then	I	\textsc{fac}-die-\textsc{fac}\\
\glt `Then my forehead swelled up. If the branch (had) hit my face, then I would have died.'
\z

\ea\label{ex:nrexdeboyp}
De boyo pulero yala de David sane: ``Margaret ari?''\\
\gll de	boyo	pule=ro	yala	de	David	sa-ne	Margaret	ari\\
     you	later	come=\textsc{emph}	soon	you	David	ask-3\textsc{sg}.\textsc{or}	Margaret	no\\
\glt `Later on you (would) come and (would) ask David: ``Margaret isn't there?'''
\z

\ea\label{ex:nrexargare}
``Margaret basuik. Uke bîyo bararko.''\\
\gll Margaret	ba-sui-ko	uke	bî-yo	ba-rari-ko\\
     Margaret	\textsc{fac}-die-\textsc{fac}	we.\textsc{excl}	hole-\textsc{loc}	\textsc{fac}-bury-\textsc{fac}\\
\glt `{``}Margaret has died. We have buried her in a grave.'''
\z
