\ea\label{ex:nrexaibikk}
Ko yip epiyo yesi yasiye. Ko yesi auna yasiye. Ko yelo ikoina rarpi.\\
\gll ko	yip	epiyo	yesi	yasiye	ko	yesi	auna	yasiye	ko	yelo	ikoi-na	rari-pi\\
     I	house	next.to	aibika	plant	I	aibika	slowly	plant	I	ground	big-\textsc{adv}	dig-\textsc{lv}\\
\glt `I plant \textit{aibika}-shrubs next to the house. I plant the \textit{aibika} slowly, and I dig the ground thoroughly.'
\z

\ea\label{ex:nrexminmar}
Ko minimari. Dop ikoina sipiwepi. Pakul sipi, paui sipi, kipi sipi. Ko mimaripi. Dop kiniyo sipiwepi.\\
\gll ko	mini\_mari	dop	ikoi-na	sipi-wepi	pakul	sipi	paui	sipi	kipi	sipi	ko	mi-mari-pi	dop	kiniyo	sipi-wepi\\
     I	come.hither\_be.sick	body	big-\textsc{adv}	hurt-\textsc{quant}.\textsc{s}	shoulder	hurt	collar.bone	hurt	back	hurt	I	\textsc{iter}-be.sick-\textsc{lv}	body	all	hurt-\textsc{quant}.\textsc{s}\\
\glt `I come to be sick. My body hurts badly. My shoulder hurts, my collar bones hurt, and 	the back hurts. I am sick again. The whole body hurts all over.'
\z

\ea\label{ex:nrexsusups}
Ko so susup ikoina poro. Mi ko ikoina mari. Sele ikoina ko pino.\\
\gll ko	so	susup	ikoi-na	pi=ro	mi	ko	ikoi-na	mari	sele	ikoi-na	ko	pi-no\\
     I	believe	grass	big-\textsc{adv}	\textsc{lv}=\textsc{emph}	again	I	big-\textsc{adv}	be.sick	garden	big-\textsc{adv}	I	\textsc{lv}-\textsc{co}\\
\glt `I believe the grass grows high. Now I am again really sick (after) working the garden so hard.'
\z

\ea\label{ex:nrexanawai}
Ruri ko ar mekiye. Ko ike an ikapno pi. Ko ana powaem? Luo ko ana powaem?\\
\gll ruri	ko	ar	mekiye	ko\_ike	an	ikap-no	pi	ko	ana	powai-m	luo	ko	ana	powai-m\\
     child	I	\textsc{neg}	help	I.myself	hand	1\textsc{sg}.\textsc{poss}.\textsc{emph}-\textsc{ins}	make	I	who	give.1\textsc{sg}.\textsc{or}-\textsc{pos}	money	I	who	give.1\textsc{sg}.\textsc{or}-\textsc{pos}\\
\glt `The (adult) children don't help me. I myself do (everything) with my own hands. Who would give me (anything)? Who would give me money?'
\z