\ea\label{ex:nrexdeyoya}
Ko deyo reyanawe yilau depiyo. Pu ikoiele, luo ikoi ulap.\\
\gll ko	deyo	reyana-we	yilau	de-pi-yo	pu	ikoiele	luo	ikoi	ule-p\\
     I	you.\textsc{du}	meet.\textsc{pp}-\textsc{du}.\textsc{o}	place	2\textsc{sg}-\textsc{poss}-\textsc{loc}	river	very.big	rock	big	be.there.\textsc{pl}-\textsc{pc}\\
\glt `I met you two of at your place. (There was) a very big river, and there were big rocks.'
\z

\ea\label{ex:nrexevaruw}
Eva ruwaesi ba roise pu epiyo ikûn deyo. Dob pokûnep olyo pokûnep. Ol inepi.\\
\gll Eva	ruwaesi	ba	roise	pu	epiyo	i-kûn	de-yo	dob	pokûni-p	ol-yo	pokûni-p	ol	i-nepi\\
     Eva	small.child	other	together	river	next.to	\textsc{du}.\textsc{s}-go.down.\textsc{pp}	you.\textsc{sg}-\textsc{loc} eye	move.down.hither-\textsc{pc}	riverbank-\textsc{loc}	move.down.hither-\textsc{pc}	riverbank	\textsc{du}.\textsc{s}-go.thither\\
\glt `Eva and another small child went down along the river edge towards you.  They were looking down, looking down here from the riverbank. They walk downriver along the bank.'
\z

\ea\label{ex:nrexluoysu}
Ko luoyo nakap puyo. Sû pulip. Ko riyo nakap.\\
\gll ko	luo-yo	nake-p	pu-yo	sû	puli-p	ko	ri-yo	nake-p\\
     I	stone-\textsc{loc}	sit-\textsc{pc}	river-\textsc{loc}	light	shine-\textsc{pc}	I	\textsc{dist}-\textsc{loc}	sit-\textsc{pc}\\
\glt `I was sitting on a rock in the river. A light was shining. I was sitting there.'
\z

\ea\label{ex:nrexanapww}
Eva ruri bayana koyo wepulo. Ko dob reyo: ``Ruri eroke de anapi wepulo?''\\
\gll Eva	ruri	bayana	ko-yo	wepulo	ko	dob	reyo	ruri	ero-ke	de	ana-pi	wepulo\\
     Eva	child	different	1\textsc{sg}-\textsc{loc}	bring.\textsc{pp}	I	eye	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	child	\textsc{prox}-\textsc{aph}	you	who-\textsc{poss}	bring.\textsc{pp}\\
\glt `Eva brought a different child to me. I looked at her: ``This child (which) you brought, whose (child) is it?”'
\z

\ea\label{ex:nrexrumkbs}
``Ko nunipi wepulo.'' Ruri baisui. Rumkari baisui.\\
\gll ko	nuni-pi	wepulo	ruri	baisui	rumkari	baisui\\
     I	maternal.uncle-\textsc{poss}	bring.\textsc{pp}	child	beautiful	girl	beautiful\\
\glt `(Eva replied:) ``I brought my uncle's (child).” A beautiful child. A beautiful girl. '
\z

\ea\label{ex:nrexbaisui}
Mi yeniyo nakap. Sû beri kleina pulip. Ko meniyekoniyap. Baisui rumkari, seke kep maki.\\
\gll mi	yeni-yo	nake-p	sû	beri	klei-na	puli-p	ko	meniye\_koniye-p	baisui	rumkari	seke	kep	maki\\
     then	plank-\textsc{loc}	sit-\textsc{pc}	light	burn	clear-\textsc{adv}	shine-\textsc{pc}	I	be.jealous-\textsc{pc}	pretty	girl	hair	3\textsc{sg}.\textsc{poss}	nice\\
\glt `I was still sitting on a plank. A light burns. It was shining brightly. I was jealous. What a pretty girl, her hair is nice.'
\z

\ea\label{ex:nrexluikar}
Ko pu lilika riye, luika riye. Luo ikoi walpuap ol ikoi roise. Mi ko puana. Kuru.\\
\gll ko	pu	lilika	riye	luika	riye	luo	ikoi	walpue-p	ol	ikoi	roise	mi	ko	puana	kuru\\
     I	river	up.river	see.\textsc{o}.\textsc{inanim}	down.river	see.\textsc{o}.\textsc{inanim}	rock	big	spread.about-\textsc{pc}	mountain	big	together	then	I	wake.up.\textsc{pp}	be.finished\\
\glt `I am looking upriver, I am looking downriver. Big rocks were spreading about and big mountains. Then I woke up. Over.'
\z