\ea\label{ex:nrexpaekma}
Ko ri wopiyako pu paekyo. Riyo malapo puyo. Malapo riyo.\\
\gll ko	ri	wopiyako	pu\_paek-yo	ri-yo	malapo	pu-yo	malapo	ri-yo\\
     1\textsc{sg}	stick	stretch.down.\textsc{pp}	water.hole-\textsc{loc}	stick-\textsc{loc} measure.\textsc{pp}	water-\textsc{loc}	measure.\textsc{pp}	stick-\textsc{loc}\\
\glt `I put a stick into a waterhole. I measured with the stick (put upright) in the water. I measured (the mark) on the stick.'
\z

\newpage
\ea\label{ex:nrexpewewe}
Ko ri peweko puyo.\\
\gll ko	ri	pewe-ko	pu-yo\\
     1\textsc{sg}	stick	erect-\textsc{rts}	water-\textsc{loc}\\
\glt `(Some days ago) I had put a stick upright into the water.'
\z

\ea\label{ex:nrexpubaik}
Pu ere îake. Pu baîko.\\
\gll pu	ere	îake	pu	ba-î-ko\\
     water	now	recede.down	water	\textsc{fac}-recede-\textsc{fac}\\
\glt `The water is going down now. The water (level) has gone down.'
\z

\ea\label{ex:nrexpubaik1}
Ko ul peweko puyo.\\
\gll ko	ul	pewe-ko	pu-yo\\
     water	bamboo	erect-\textsc{rts}	water-\textsc{loc}\\
\glt `I had put a bamboo rod in the water.'\footnote{Alternatively, one can use a bamboo rod instead of a stick for measuring the depth of the water.}
\z

\ea\label{ex:nrexpinkop}
Pu pini. Ri ko peworo. Pu ri sowelayo.\\
\gll pu	pini	ri	ko	pewo=ro	pu	ri	sowelayo\\
     water	go.up.hither	stick	1\textsc{sg}	erect.\textsc{pp}=\textsc{emph}	water	stick	cover.entirely.\textsc{pp}\\
\glt `The water is going up. I took a stick. The water covered the stick entirely.'
\z

\ea\label{ex:nrexyalyop}
Yala ikoina yopi. Pu bayopko.\\
\gll yala	ikoi-na	yopi	pu	ba-yopi-ko\\
     now	big-\textsc{adv}	rise.of.water	water	\textsc{fac}-rise.of.water-\textsc{fac}\\
\glt `Now (the water) rises high. The water has risen.'
\z
