\ea\label{ex:nrexrowekk}
Mi ko le piyowe. Mi ko rowekûne yeloka. Wolmo bayo koliye, yeloyo polino.\\
\gll mi	ko	le	piye-we	mi	ko	ro=wekûne	yeloka	wolmo	ba-yo	koliye	yelo-yo	poli-no\\
     again	I	things	take-\textsc{ter}	again	I	\textsc{emph}=carry.down	outside	clothes.line	other-\textsc{loc}	hang	ground-\textsc{loc}	be.there-\textsc{co}\\
\glt `I take the (washed) things again, carry them down outside again, and hang them on the other clothes line which is outside.'
\z

\ea\label{ex:nrexyyipip}
Mi ko yipyo îpîpî. Yipyo nake.\\
\gll mi	ko	yip-yo	îpîpî	yip-yo	nake\\
     again	I	house-\textsc{loc}	climb.up	house-\textsc{loc}	sit\\
\glt `(Then) I climb up (the stairs) into the house and sit again in the house.'
\z

\ea\label{ex:nrexpimmik}
Yala pu pem. Mi ko îpake. Mi ko le piyeke. Mi ko wepini yipyo. Koliyewe. Mi ko nake.\\
\gll yala	pu	pi-m	mi	ko	îpake	mi	ko	le	piye-ke	mi	ko	wepini	yip-yo	koliye-we	mi	ko	nake\\
     now	rain	\textsc{lv}-\textsc{pos}	again	I	climb.down	again	I	things	take-\textsc{ingr}	again	I	carry.up.hither	house-\textsc{loc}	hang-\textsc{ter}	again	I	sit\\
\glt `(Oh) it may rain now. Again I climb down, go take the laundry, carry  them up here into the house, hang them, and sit (there) again.'
\z

\ea\label{ex:nrexkoekua}
Nini yala pulem. Mi ko rowekûne, koliyewe. Ko mipini yipyo. Ko eku nake.\\
\gll nini	yala	puli-m	mi	ko	ro=wekûne	koliye-we	ko	mi-pini	yip-yo	ko	eku\_nake\\
     sun	soon	shine-\textsc{pos}	again	I	\textsc{emph}=carry.down	hang-\textsc{ter}	I	\textsc{iter}-come.up.hither	house-\textsc{loc}	I	sit.down\\
\glt `Suddenly it seems to get sunny. Again I carry the things down and hang them. Again I come  up here into the house and sit down.'
\z

\ea\label{ex:nrexarpull}
Oh! Yala pu poro! Nini ar puli. Mi ko piyeke, wepini. Pu pi. Ko due nui.\\
\gll oh	yala	pu	pi=ro	nini	ar	puli	mi	ko	piye-ke	wepini	pu	pi	ko	due\_nui\\
     oh	now	rain	\textsc{lv}=\textsc{emph}	sun	\textsc{neg}	shine	again	I	take-\textsc{ingr}	carry.up.hither	rain	\textsc{lv}	I	sleep\\
\glt `Oh, now it is raining! The sun doesn't shine (anymore). I go again to take (the things) and carry them up here. It rains. I will sleep.'
\z