\ea\label{ex:nrexyemriy}
Yem riyo wariyeppue. Ri iniyo nakeko, re kep supuaspi. Yem ili kep puakuyo poli.\\
\gll yem	ri-yo	wariye\_ppue	ri	ini-yo	nake-ko	re	kep	supuas\_pi	yem	ili	kep	puaku-yo	poli\\
     crowned.pigeon	tree-\textsc{loc}	fly.up	tree	branch-\textsc{loc}	sit-\textsc{rts}	feather	3\textsc{sg}.\textsc{poss}	wobble	crowned.pigeon	bunch	3\textsc{sg}.\textsc{poss}	head-\textsc{loc}	be.there\\
\glt `The crowned pigeon flies up into a tree. After alighting on a branch the feathers (of its crown) wobble. The crowned pigeon has a crest on its head.'
\z

\ea\label{ex:nrexyemlep}
Yem re lepapouli. Yem re boli sikero poli. Boli kep ar lipiyo, ar lepapo. Epi kep solo lepapouli.\\
\gll yem	re	lepapi-uli	yem	re	boli	sike=ro	poli	boli	kep	ar	lipiyo	ar	lepapo	epi	kep	solo	lepapi-uli\\
     crowned.pigeon	feather	have.a.pattern-\textsc{prog}	crowned.pigeon	feather	origin	hard=\textsc{emph}	be.there	origin	3\textsc{sg}.\textsc{poss}	\textsc{neg}	paint.\textsc{pp}	\textsc{neg}	have.a.pattern.\textsc{pp}	side	3\textsc{sg}.\textsc{poss}	only	have.a.pattern-\textsc{prog}\\
\glt `The feathers of the crowned pigeon have a pattern. The quills of the crowned pigeon are strong; they have no colour and no pattern. Only its side has a pattern.'
\z

\ea\label{ex:nrexyemppi}
Puaku re ba sumi ba ilei. Yem re ri ppipeso.\\
\gll puaku	re	ba	sumi	ba	ilei	yem	re	ri	ppipe-so\\
     head	feather	other	short	other	long	crowned.pigeon	feather	tree	flower-\textsc{sim}\\
\glt `Some of the head feathers are short, some are long. (The crest of) the crowned pigeon  (looks) like a flower of a tree.'
\z

\ea\label{ex:nrexyemsek}
Yem re puaku pili roise sesiyepiye. Ko yem re sekeyo papi.\\
\gll yem	re	puaku	pili	roise	sesiye\_piye	ko	yem	re	seke-yo	papi\\
     crowned.pigeon	feather	head	skin	together	take.away	I	crowned.pigeon	feather	hair-\textsc{loc}	prepare.\textsc{pl}.\textsc{o}\\
\glt `One takes away the crowned pigeon's (crest) feathers together with some skin  of the head. I prepare the (crest) feathers of the crowned pigeon for my hair.'
\z

\ea\label{ex:nrexyemkoi}
Yem ikoiele. Dop kep dû roise kakarukso pari.\\
\gll yem	ikoiele	dop	kep	dû	roise	kakaruk-so	pari\\
     crowned.pigeon	very.big	body	3\textsc{sg}.\textsc{poss}	meat	together	chicken-\textsc{sim}	\textsc{neg}.\textsc{cop}\\
\glt `The crowned pigeon is very big. Its body and meat aren't like a chicken, (but substantially more).'
\z