\ea\label{ex:nrexmaruma}
Margaret Uma due kana soni. Due uki kep wel. Due melro, rapno wel.\\
\gll Margaret	Uma	due	kana	soni	due	uki	kep	wel	due	mel=ro	rap-no	wel\\
     Margaret	Uma	sago	quickly	pound.sago.pith	sago	husband	3\textsc{sg}.\textsc{poss}	carry.\textsc{pp}	sago	carry.\textsc{pl}.\textsc{o}.\textsc{pp}=\textsc{emph}	raft-\textsc{ins}	carry.\textsc{pp}\\
\glt `Margaret Uma pounded sago pith quite quickly. Her husband took the sago. He carried a lot of sago and transported it by raft.'
\z

\ea\label{ex:nrexrapwel}
Uki kep rapno welro. Puyo masakaikûno.\\
\gll uki	kep	rap-no	wel=ro	pu-yo	masakaikûno\\
     husband	3\textsc{sg}.\textsc{poss}	raft-\textsc{ins}	carry.\textsc{pp}=\textsc{emph}	river-\textsc{loc}	fall.down.in.plenty.\textsc{pp}\\
\glt `Her husband transported it by raft. (Unfortunately all the sago bags) tipped over into the river.'
\z

\ea\label{ex:nrexlorapn}
Uki kep an solo lo. Rapno due aska. Due pu meli. Due pu wena.\\
\gll uki	kep	an	solo	lo	rap-no	due	aska	due	pu	meli	due	pu	wena\\
     husband	3\textsc{sg}.\textsc{poss}	hand	only	go.\textsc{pp}	raft-\textsc{ins}	sago	none	sago	river	carry.\textsc{pl}.\textsc{o}	sago	river	carry.inside\\
\glt `Her husband went empty-handed. There is no sago on the raft. The river carries all the sago. The river carries the sago with the current.'
\z

\ea\label{ex:nrexjoerap}
Joe rap makina ar wel. Riyo kûpiyo. Luoyo kûpiyo. Riyopuno due masakaikûno.\\
\gll Joe	rap	maki-na	ar	wel	ri-yo	kûpiyo	luo-yo	kûpiyo	riyopuno	due	masakaikûno\\
     Joe	raft	good-\textsc{adv}	\textsc{neg}	carry.\textsc{pp}	tree-\textsc{loc}	bump.against.\textsc{pp}	rock-\textsc{loc}	bump.against.\textsc{pp}	then	sago	fall.down.in.plenty.\textsc{pp}\\
\glt `Joe didn't carry the raft well. It bumped against a tree (or) it bumped against a rock. Then all the sago (bags) fell down.'
\z