\ea\label{ex:nrexurabis}
Koyo ilo duyo. Uki kopi bike dupua luwe, biep klokni, ura, bisem dupua, yem klokni, iwan klokni.\\
\gll koyo	i-lo	du-yo	uki	ko-pi	bike	dupua	lu-we	biep	klokni	ura	bisem	dupua	yem	klokni	iwan	klokni\\
     we.\textsc{du}.\textsc{excl}	\textsc{du}.\textsc{s}-go.\textsc{pp}	bush-\textsc{loc}	husband	1\textsc{sg}-\textsc{poss}	cassowary	two	shoot.\textsc{pp}-\textsc{du}.\textsc{o}	boar	one	lizard	kangaroo	two	crowned.pigeon	one	hornbill	one\\
\glt `We went into the (deep) bush. (There) my husband shot two cassowarries, one boar, lizards, two kangaroos, one crowned pigeon, and one hornbill.'
\z

\ea\label{ex:nrexsuslau}
Mi ko apla po bino bikeno. Ko sûyo royo. Ko kiniyo royepepo. Ko sû mappeana. Sû beri poro. Ko sû piana pianpana. Ko aplayo royo, bi slau.\\
\gll mi	ko	apla	po	bi-no	bike-no	ko	sû-yo	royo	ko	kiniyo	royepepo	ko	sû\_mappeana	sû	beri	po=ro	ko	sû	piana	pianpana	ko	apla-yo	royo	bi	slau\\
     then	I	plank	do.\textsc{pp}	pig-\textsc{ins}	cassowary-\textsc{ins}	I	fire-\textsc{loc}	put.\textsc{pp}	I	all	lay.on.top.\textsc{pp}	I	light.a.fire.thither.\textsc{pp}	fire	flame	\textsc{lv}.\textsc{pp}=\textsc{emph}	I	fire	blow.on.\textsc{pp}	blaze.up.\textsc{pp}	I	plank-\textsc{loc}	put.\textsc{pp}	meat	dry\\
\glt `I made the planks (and filled them) with pig and cassowary meat. I put them above the fire. I put everything on top (of the planks). I lit a fire. The fire did flames. I blew the fire, and it blazed up. I put meat on the planks, and the meat dries.'\footnote{The narrator describes repeated acts of filling the wooden planks with meat in order to smoke and dry all the meat.}
\z

\ea\label{ex:nrexurayil}
Punipino uki kopi muel: ``Nuko mole yilauyo.'' Uke yilauyo pulip. Paeau.\\
\gll punipino	uki	ko-pi	muel	nuko	mole	yilau-yo	uke	yilau-yo	pulip	paeau\\
     morning	husband	1\textsc{sg}-\textsc{poss}	talk.to.\textsc{pp}	we.\textsc{incl}	go.\textsc{pl}	village-\textsc{loc}	we.\textsc{excl}	village-\textsc{loc}	come.\textsc{pl}.\textsc{pp}	arrive\\
\glt `One morning my husband said: ``We'll go (back) to the village.” We came to the village 	and arrived (there).'
\z