\ea\label{ex:nrexumekot}
Bi seleyo mini. Uke wok ar pi, bi wip ni, arme ni, umeko ni.\\
\gll bi	sele-yo	mini	uke	wok	ar	pi	bi	wip	ni	arme	ni	umeko	ni\\
     pig	garden-\textsc{loc}	come.hither	we.\textsc{excl}	work	\textsc{neg}	\textsc{lv}	pig	taro	eat	kind.of.taro	eat	kind.of.taro	eat\\
\glt `The pigs come to the gardens. (When) we don't work (there), the pigs eat the taro. They eat \textit{arme}-taro and they eat \textit{umeko}-taro.'
\z

\ea\label{ex:nrexdormal}
Uke sû mappe seleyo, bi ar mini. Dop ili onopi nikiyei. Bi dorimaliye.\\
\gll uke	sû\_mappe	sele-yo	bi	ar	mini	dop	ili	ono-pi	nikiyei	bi	dorimaliye\\
     we.\textsc{excl}	light.a.fire	garden-\textsc{loc}	pig	\textsc{neg}	come.hither	body	smell	man-\textsc{poss}	smell	pig	turn.back\\
\glt `(When) we light fires in the gardens, the pigs don't come. They smell the body odour of people. (Then) the pigs turn back (to their hidden places).'
\z

\ea\label{ex:nrexsulink}
Yala sele kuso ponake. Sele klayepanam. Sû sûli nikiyei. Bi dop ili sû sûli roise nikiyei.\\
\gll yala	sele	kuso	pi-nake	sele	k-layepane-m	sû	sûli	nikiyei	bi	dop	ili	sû	sûli	roise	nikiyei\\
     now	garden	always	\textsc{lv}-\textsc{dur}	garden	\textsc{proh}-neglect-\textsc{proh}	fire	smoke	smell	pig	body	smell	fire	smoke	together	smell\\
\glt `(Better) one always does the gardens. One must not neglect the gardens. (The pigs) smell  the smoke of fire. They smell the body odour and the smoke.'
\z

\ea\label{ex:nreximiyne}
Dop ili le. Ono neki. Imiyu neki. Iner dop ili, ono iner ili le.\\
\gll dop	ili	le	ono	neki	imiyu	neki	iner	dop	ili	ono	iner	ili	le\\
     body	smell	go	person	stand	sorcerer	stand	armpit	body	smell	man	armpit	smell	go\\
\glt `The body odour goes. Somebody is (nearby). A sorcerer is (nearby). The body odour of the armpit, the armpit odour of a man hangs in the air.'
\z