\ea\label{ex:nrexsaproo}
Bariyeko. Basuiko. Bia barirko. Uke kumune sapro.\\
\gll ba-riye-ko	ba-sui-ko	bia	ba-rire-ko	uke	kumune	sap=ro\\
     \textsc{fac}-see.\textsc{o}.\textsc{inanim}-\textsc{fac}	\textsc{fac}-die-\textsc{fac}	corpse	\textsc{fac}-be.stiff-\textsc{fac}	we.\textsc{excl}	all.\textsc{coll}	sleep.\textsc{pl}.\textsc{pp}=\textsc{emph}\\
\glt `We have seen it. He has died. The corpse is stiff. We all were fast asleep.'
\z

\ea\label{ex:nrexuenuir}
Puni uke ar reyo yeni kepyo. Uke so due nuiro.\\
\gll puni	uke	ar	reyo	yeni	kep-yo	uke	so	due\_nui=ro\\
     night	we.\textsc{excl}	\textsc{neg}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	bed	3\textsc{sg}.\textsc{poss}-\textsc{loc}	we.\textsc{excl}	believe	sleep=\textsc{emph}\\
\glt `During the night we didn't see him on his bed. We thought he was fast asleep.'
\z

\ea\label{ex:nrexsapawe}
Yena kiniyo wonien: ``Uke ar reyo. Uke due sap. Ono basuik. Ine awe.''\\
\gll yena	kiniyo	woni-en	uke	ar	reyo	uke	due\_sap	ono	ba-sui-ko	ine	awe\\
     people	all	call-\textsc{nsg}.\textsc{or}.\textsc{pp}	we.\textsc{excl}	\textsc{neg}	see.\textsc{o}.\textsc{anim}.\textsc{sg}.\textsc{pp}	we.\textsc{excl}	sleep.\textsc{pl}.\textsc{pc}	man	\textsc{fac}-die-\textsc{fac}	you.\textsc{pl}	come.\textsc{imp}\\
\glt `We called all the people: ``We didn't look for him. We were sleeping. The man has died. You all come!'''
\z

\ea\label{ex:nrexppusik}
Yena pulupi, eku mape. Wo mopine, womopinake. Nuni kep, ewe kep, ai kep, ai ppusi kep, buka ruri kep, memi kep.\\
\gll yena	pulupi	eku\_mape	wo\_mopi-ne	wo\_mopi-nake	nuni	kep	ewe	kep	ai	kep	ai\_ppusi	kep	buka	ruri	kep	memi	kep\\
     people	come.\textsc{pl}	sit.down.\textsc{pl}	cry-3\textsc{sg}.\textsc{or}	cry-\textsc{dur}	maternal.uncle	3\textsc{sg}.\textsc{poss}	older.brother	3\textsc{sg}.\textsc{poss}	father	3\textsc{sg}.\textsc{poss}	paternal.uncle	3\textsc{sg}.\textsc{poss}	nephew	child	3\textsc{sg}.\textsc{poss}	grandparent	3\textsc{sg}.\textsc{poss}\\
\glt `The people come and sit down. They cry for him, they stay crying. His maternal uncle, his older brother, his father, his paternal uncle, his nephews, his grandparents.'
\z

\ea\label{ex:nrexwokmol}
Wo kuru. Ine molewe yilauyo.\\
\gll wo	kuru	ine	mole-we	yilau-yo\\
     crying	be.finished	you.\textsc{pl}	go.\textsc{pl}-\textsc{ter}	place-\textsc{loc}\\
\glt `The crying is over. You go (back) to (your) places.'
\z